### 2022-12-12 README.Rmd
initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# mb11CalculusPilot

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/inst/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.pdf`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/inst/data): Data used in the analysis.
  - [:file\_folder: figures](/inst/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/inst/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2022-12-12 inst/paper/_discussion.qmd
initial commit

<!-- Big summary/important statement for the science -->
In this study we were able to extract and identify multiple alkaloids from the
dental calculus of individuals from Middenbeemster, a 19th century Dutch
archaeological site. We applied ultra-high-performance liquid chromatography-tandem
mass spectrometry (UHPLC-MS/MS), using a method that was validated the presence
of drugs and metabolites in dental calculus with the equivalent compounds in blood
[@sorensenDrugsCalculus2021].
Here we have shown that the method can also be successfully applied to archaeological
dental calculus.
We initially detected cannabinoids, cocaine, nicotine, cotinine, caffeine,
theopyhlline, and salicylic acid in multiple individuals; however, the detection
of cannabinoids and cocaine could not be replicated in a second run on different
samples from the same individuals.

<!-- discuss any correlations, spurious or otherwise
e.g. more caffeine, salicyl -> less calculus (acidic environ?)
tobacco -> sinusitis -->

<!-- compounds present and absent with potential sources-->

<!-- nicotine and cotinine: tobacco, obviously...-->
Nicotine and cotinine were strongly positively correlated, both in concentration
and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
The detection of nicotine and cotinine is not surprising, as pipe-smoking in
the Beemsterpolder is well-documented in the literature
[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
remains [@lemmersMiddenbeemster2013].
There is also documented medicinal use of nicotine in the Beemsterpolder,
where a tobacco
smoke enema was used for headaches, respiratory problems, colds, and drowziness,
among others, from around 1780 to 1830 [@aten400JaarBeemster2012].
<!-- Scientific findings as extension of previous work -->
The detection of nicotine in dental calculus has previously been presented by
Eerkens and colleagues [-@eerkensDentalCalculus2018] in two individuals from
pre-contact California. They also targeted caffeine, cotinine, and theophylline
in their samples, but were unable to detect any of them. It remains to be seen
whether this is due to differences in methods used, or due to our samples being
more recent.
Other alkaloids have been detected in archaeological samples, though these are more
often associated with other materials, including pipes
[@raffertyCurrentResearch2012] and ceramics [@smithDetectionOpium2018].
Direct analysis on human remains
has also included hair [@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013].

<!-- theophylline and caffeine: coffee and tea? -->
Theophylline and caffeine concentrations were positively correlated,
suggesting that they may have originated from the same source.
Caffeine is present in coffee, tea, and cocoa beans, with concentrations slightly higher
in coffee [@stavricVariabilityCaffeine1988; @chinCaffeineContent2008; @bispoSimultaneousDetermination2002].
Theophylline is present in both coffee beans and tea leaves, but in negligible
quantities [@stavricVariabilityCaffeine1988]. It is also metabolised from
caffeine in the liver, and given the low quantity relative to caffeine, it likely
originated from either tea or coffee, which contain higher concentrations of
the compounds [@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
Caffeine and theophylline are
structurally very similar, and we expect they would experience similar rates of
degradation, allowing us to make interpretations on the ratio between the compounds.
Tea consumption had become widespread in the Netherlands by 1820, reaching all aspects of
the population (some argue that this happened already by early 18th century)
[@nierstraszTeaTrade2015, p.91].
Historically, we also know that both tea and coffee were consumed in the Beemsterpolder
during the 19th century. 'Theegasten' (teatime) was a special
occasion occurring from 15.00-20.00 hours, where tea was served along with the evening
bread [@schuijtemakerTeTheegasten2011]. Many households also owned at least one coffee pot and tea pot [@boumanBegravenis2017].
Both the presence of theophylline and caffeine could also be explained by chocolate,
although given the large quantities of caffeine compared to theophylline, it is more
likely that the source is tea and/or coffee
[@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
Distinguishing between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is present in
higher quantities in chocolate compared to caffeine and theophylline
[@stavricVariabilityCaffeine1988; @bispoSimultaneousDetermination2002; @alanonAssessmentFlavanol2016].
Although there is some indication that theobromine does not preserve well
[@velskoDentalCalculus2017], and frequent consumption of all three
items would be difficult to parse.

<!-- salicylic acid: diet, medicine, or environmental??? -->
The ubiquitous presence of salicylic acid poses a different problem. It has
been shown that salicilyc acid is a very mobile organic acid and the ubuquitous
presence may be due to environmental contamination
[@badriRegulationFunction2009; @velskoDentalCalculus2017].
The bark of a willow tree, *Salix alba*, has long been known for its
pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119],
and is also present in many plant-based foods
[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017]. This includes
potatoes, which were a mainstay in the Beemsterpolder diet [@aten400JaarBeemster2012].
It will be necessary to explore the extent to which salicylic acid can leech
into the dental calculus from the soil, and what the rate of degradation is for
salicylic acid when trapped in dental calculus.

<!-- cannabinoids: medicinal? wait, where did they go..? -->
Cannabinoids---specifically THC, THCA-A, THCVA, CBD, CBN---were found in the
first batch, but none were replicated in the second batch. Medicinal use of
cannabinoids has been well-established in Europe since Medieval-times, and it
was also grown in the Netherlands [@bruinsmaBijdragenTot1872].
Administration was most common in the form of concoctions containing various portions
of the cannabis plant for ingestion; not until the late 19th century did it become
recommended to smoke it for more immediate effects [@clarkeCannabisEvolution2013].
John Parkinson described a Dutch medicinal use of hemp. An emulsion was prepared
from the seeds of the plants to treat pain and various stomach ailments. Another
preparation involving the roots of the plants was used for inflammation, gout, and joint
pains [@clarkeCannabisEvolution2013].
The ability to detect cannabinoids in calculus may be limited by a reduced ability
to diffuse from serum to salivary glands due to an affinity for protein-binding,
[@sorensenEffectAntioxidants2018; @sorensenDrugsCalculus2021],
meaning detection would rely on oral consumption of the cannabinoids. Even then,
the overall instability of some cannabinoids could also affect detection
[@sorensenEffectAntioxidants2018; @lindholstLongTerm2010].

<!-- opioids: medicinal. Damn... that would've been cool-->
Despite many of the individuals living during the height of the opium era in the
Netherlands, none of the targeted opioids
(morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
were detected. 
Sylvius de la Boe (1614–1672), a German-born physician who studied anatomy at
Leiden University, the Netherlands, relied heavily on opium [@machtHistoryOpium1915].
The absence of opioids could be a result of the people ascribing more to the
"traditional" rather than "scientific" medicine, although laudanum and
another opium containing concoction was part of the "traditional" medicine in
the Netherlands [@leuwProhibitionLegalizationDutch1994], including Middenbeemster
[@aten400JaarBeemster2012]. It was also generally
considered a drug of the upper class [@scheltemaOpiumTrade1907], and may be
more common in an urban center.
The absence could also be attributed to postmortem degradation.
It has been shown that, while abundant in opium, morphine
degrades rapidly. The more stable opioids such as thebaine and papaverine
[@chovanecOpiumMasses2012] were also absent, however.
<!--Similar to cannabinoids, many opioids are preferentially bound to (plasma) proteins
[@sorensenDrugsCalculus2021].-->

We explored the relationship between detected compounds and various skeletal
indicators, such as pathological and dental lesions, preservation, and pipe notches.
We found no evidence to suggest that preservation of the skeleton influences
the recovery of compounds from the dental calculus. This could be due to
the good overall preservation of both teeth and dental calculus, which often
preserves better than the rest of the skeleton.
We found no correlation between the number of pipe notches
and the concentration of nicotine and cotinine, suggesting that our ability
to detect tobacco consumption in dental calculus does not necessarily rely
targeting frequent smokers; here, we consider individuals with multiple pipe notches
as likely to have been heavy smokers.
There does seem to be a positive correlation between the weight of the calculus
sample and recovered quantities of the compounds, so weight of the sample may affect
the ability to detect compounds; although, we were able to detect
compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].

[[ Correlation between caffeine, nicotine, and sinusitis - Maia ]]

Authentication

<!-- authentication -->
The list of targeted compounds included both naturally occurring compounds known
to have been used in the past, as well as synthetic modern drugs did not exist
at the time (e.g. Fentanyl, MDMA, Amphetamine). These were part of the toxicology
screening for the original method [@sorensenDrugsCalculus2021], originally
developed on cadavers. In our study they serve as an authentication step, as their
presence in archaeological samples could only be the result of contamination.
<!-- the importance of replication - brief discussion of cocaine -->
The only modern compound (at least in a European context) detected in the sample
was cocaine, which was detected in the first batch of sampels. The population
analysed is from the early-mid 18th century, and cocaine was isolated in 1860 by
Albert Niemann, and entered popular medical
practice in 1884. Coca arrived in Europe as early as 1771, but as botanical
specimens rather than for consumption, and there were also issues importing enough
viable specimens of coca for cocaine extraction
[@mortimerHistoryCoca1901 p. 179; @abucaCocaTrade2019, p.108].
We considered it possible that it would be present in a sample with
most individuals originating from the early- to mid-19th century. If corroborated,
this would have been the first case of coca-consumption in Europe. Given that cocaine
had yet to be isolated, the only way it could have gotten into the calculus was
through consumption of the coca leaf. In our replication batch, we included all
of the individuals who had been cocaine-positive in the first batch. We were
unable to replicate any of the cocaine results, and we were unable to detect the
principal metabolite, benzoylecgonine, in either batch; although, the original
study only detected trace amounts of benzoylecgonine, and only in some of the
individuals that were cocaine-positive. We hope this will serve as
a reminder of the importance of replication. We suspect that the original detection
of cocaine was a result lab contamination during the UHPLC-MS/MS analysis.

Accuracy

We used the presence of a pipe notch and concurrent detection of nicotine
and/or cotinine was as a crude estimate of the accuracy of the method. Nicotine
has been widely studied in the archaeological record due to its apparent stability
and ability to survive over long periods of time
[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].
The accuracy was found to be around
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
We acknowledge that this is
a very rough estimate, as the presence of a pipe notch is likely not a perfect
indicator of whether or not someone consumed tobacco, and it may not be
representative of other compounds with different rates of degradation.

<!-- Findings' scientific implications -->

<!-- secondary findings-->
The choice of tooth for sampling may impact the detection of certain compounds,
as the incorporation in dental calculus may depend on the mode of consumption.
Tobacco smokers may have more
nicotine present in calculus on incisors, whereas tobacco chewers may have more
on molars [@eerkensDentalCalculus2018]. This has yet to be systematically
tested in archaeological dental calculus. Direct incorporation by ingesting
items, food or otherwise, containing the compounds
is not the only potential pathway to incorporation in the dental calculus. 
The presence of cotinine, the primary
metabolite of nicotine, suggests that saliva is also a source of deposition, and
that deposition of alkaloids in dental calculus can occur both on the way into the body,
i.e. during consumption, and on the way out, i.e. disposal of waste products via
saliva secretion into the mouth. Especially mucin-rich saliva from the sublingual
and submandibular glands preferentially binds toxins [@doddsHealthBenefits2005],
and since these glands are located closest to the lower incisors, they may be the
most effective target for these studies. Due to homogenisation of samples in our
study, we were unable to test this effect.

<!-- How this engineering will facilitate future research -->
<!-- both calculus and teeth preserve well-->

<!-- Forward-looking statements about the field as a whole -->

<!-- a summary/importance of the engineering aspect of the work -->

<!-- How it facilitates scientific research -->

<!-- limitations of the platform show which questions would be unanswered -->
Limitations

Since we targeted individuals with moderate-to-large calculus deposits,
it is likely a biased sample. The presence of calculus may increase risk of premature death
[@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
be associated with dental calculus build-up) is a risk-factor for
respiratory diseases, if periodontal and respiratory pathogens
enter the blood stream
[@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
In our sample, the percentage of CMS
(`r scales::percent(mean(explore_select$CMS, na.rm = T), accuracy = 0.1)`)
is lower than in another (more representative) male sample (44.1%)
[@casnaUrbanizationRespiratory2021], and the caries percentage is similarly lower
in our sample (`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`) than
a more representative sample (22.9%) [@lemmersMiddenbeemster2013].
The bias may also affect the detection of compounds, as caffeinated drinks are often
acidic, and may cause a lower calculus formation<!-- citation?-->.
We only found a weak negative correlation (`r signif(corr_tib[15,2], 3)`)
between caffeine concentration and dental calculus index, but since we targeted individuals with
calculus, this may inadvertently have been controlled for during sample collection.

Quantitation of the detected compounds may have limited value in archaeological
samples, given the effects of taphonomy added on top of individual variability
and relative stochasticity of dental calculus formation, and the metabolic
degradation of compounds during life [@sorensenDrugsCalculus2021; @velskoAncientDental2022a]. <!-- the results of comparisons with concentration
had limited outcomes, and the correlations between concentration and other variables
were similar to the correlations with presence/absence, suggesting that the actual
concentration was somewhat random.-->

Perhaps the main limitation, which is also true for many other analyses on
archaeological materials, is that detecting the absence of a compound is not possible.
The absence of a compound is
not evidence of its absence, which complicates the interpretation of our results.
We have attempted to minimise errors occurring due to this limitation by including
a relatively large sample of individuals. Although given the relatively low
detection rate seen in tobacco, this remains a major limitation, and will likely
been compounded by increasing antiquity of the samples.
The replication of samples increases the evidence of absence if the compound
was absent in both batches.

<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->

While a major limitation is the uncertainty surrounding whether or not a compound
is actually absent, the power of the method is in detecting food and other items
that are consumed in a manner that does not allow macroscopic detection. Taking
tobacco consumption as an example; while pipe
notches are a useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, which includes chewing, drinking, cigars, snuff,
[@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
was mostly common in males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018].
Our method does not, at least to our knowledge, discriminate between modes of
consumption,
and may be more suitable for exploring tobacco consumption in the entire population,
not just males. In our sample, we detected nicotine and cotinine (replicated)
in an old adult, probable female individual. Because the individual was an old
adult, it is unlikely that the compounds entered the dental calculus through
pipe-smoking, as there would be pipe notches. More likely that the tobacco entered
through alternative consumption or secondhand smoke.

Future studies should explore the effect of tooth position on the detection of
compounds.
Also on the properties increasing incorporation potential of different compounds,
and which pathways facilitate maximum incorporation (direct or indirect pathways?),
which likely differs between alkaloids. How is the preservation once trapped
inside the calculus, for example using artificial ageing on real or artificial
dental calculus?
In this study, we only targeted drugs that were included in the forensic toxicological
screenings, and therefore only covered a limited number of the potential compounds
that could be of interest for exploring past diets and medicinal treatments.

<!--final remarks-->

This preliminary study outlines the benefits of using calculus to target a
variety of compounds that could have been consumed as medicine or diet. 
This method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in historic
documentation, such as lower status, children and women.
We also
show that there are many limitations that will need to be addressed going forward
with this type of analysis, and stress the need for more systematic research on
the consumption of alkaloid-containing items and their subsequent concentration
and preservation in dental calculus,
in addition to how mode of consumption may affect concentrations on different parts
of the dentition. <!-- also individual variation -->
Another limitation of dental calculus as a medium is the variability its formation
and the many factors that can influence incorporation and retention of molecules
and particles; however, in the absence
of hair and serum (quite uncommon in archaeology), dental calculus represents
an impressive reservoir of information regarding the consumption of various
alkaloids, whether dietary, medicinal, recreational, or otherwise.

### 2022-12-12 inst/paper/_intro.qmd
initial commit

<!-- intro to calculus and residues -->
Dental calculus has proven to be excellent source of a wide variety of information
about our past.

the increased availability of methods such as aDNA, paleoproteomics, and
chromatography, has expanded our ability to identify specific pathogens, dietary,
and possible medicinal compounds<!--cite-->.

<!--how are compounds incorporated in dental calculus-->
Increased interest in using oral fluid<!--more accurate terminology than saliva?--> as a means of detecting alkaloids in living
individuals, due to the non-invasive nature of the testing compared to blood
and urine sampling<!--citations-->.

Alkaloids enter the dental calculus via two routes: (1) direct
incorporation through oral consumption of alkaloid-containing plants, whether
deliberate or accidental;
and (2) passive diffusion through the
saliva as toxins are transferred from serum to saliva [@coneInterpretationOral2007],
and then into the oral cavity through the salivary glands. The relation to serum
is why there is often a close relation between oral fluid and blood concentrations
of drugs [@coneInterpretationOral2007; @sorensenDrugsCalculus2021]
The second route allows the
detection of compounds that not only are consumed orally, but also any other pathway.
We also expect to detect metabolites through the second pathway.

Many of the components involved in the formation and growth of dental calculus
originate from saliva. Therefore the transfer of salts and other
compounds between saliva and oral biofilms on the tooth surface is likely to also
allow various alkaloids of dietary and medicinal origin to become entrapped in
dental plaque, eventually mineralising and becoming dental calculus.

Previous studies have been conducted using various chromatographic methods on
archaeological dental calculus, employing both targeted
[@eerkensDentalCalculus2018] and untargeted approaches
[@buckleyDentalCalculus2014; @gismondiMultidisciplinaryApproach2020].

<!--brief description of method validation-->
In the method validation study, all compounds detected in the blood were also
detected in the dental calculus,
with additional compounds present in dental calculus that were not present
in blood, suggesting that dental calculus represents a comprehensive history of
consumption over a long time <!-- rephrase-->[@sorensenDrugsCalculus2021].

<!--summary of results-->
We were able to detect both parent compounds, such as caffeine and nicotine, as
well as their metabolites theophylline and cotinine. These indicate the consumption
of tea and coffee and smoking of tobacco, which is also confirmed by historic
documentation and identification of pipe notches in the dentition.

### 2022-12-12 inst/paper/_materials.qmd
initial commit

<!-- Demographics and description of MB11 sample, selection criteria. -->
The sample consists of `r nrow(demography)` individuals from Middenbeemster,
a 19th century rural Dutch site. 
The village of Middenbeemster and the surrounding Beemsterpolder was established
in the beginning of the 17th century, when the Beemster lake was drained to
create more farm land, mainly for the cultivation of cole seeds (de Vries 1978).
In 1615, a decision was made to dedicate a church to Hendrick de Keyser, and
construction started in 1618 (Hakvoort 2013). The excavated cemetery is associated
with the Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
[@lemmersMiddenbeemster2013]. Archival documents are available for those buried
between AD 1829 and 1866, when the majority of individuals were interred
[@palmerActivityReconstruction2016]. The main occupation of the inhabitants
was dairy farming, consisting largely of manual labour prior to the industrial
revolution [@aten400JaarBeemster2012; @palmerActivityReconstruction2016].

To reduce the number of factors to account for
in the analysis, we preferentially selected males from the middle adult age category
(35-49). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
`r nrow(filter(demography, sex == "pm"))` probable males,
`r nrow(filter(demography, sex == "pf"))` probable female, and
`r nrow(filter(demography, sex == "f"))` female.
We selected males due to a higher
occurrence of dental calculus deposits than females. Dental calculus was
preferentially selected from incisors.

```{r}
# create tables for demographics if there are tables to spare in the article
demography %>%
  count(sex) %>%
  knitr::kable()

demography %>%
  count(age) %>% 
  knitr::kable()
```

### 2022-12-12 inst/paper/_methods.qmd
initial commit

New dental calculus samples were taken under sterile conditions in a positive
pressure laminar flow hood in a dedicated dental calculus lab at Leiden University.
The surface of the tooth was lightly brushed with a sterile, disposable toothbrush
to get rid of surface contaminants. A sterile dental curette was then used to
scrape calculus from the tooth onto weighing paper, which was transferred to 1.5 ml
Eppendorf tubes.
Where possible, previously sampled material was used to prevent repeated sampling
of individuals. The sampling methods can be found in the published material
[@ziesemer16SChallenges2015; @ziesemerGenomeCalculus2018]. 
The samples were sent to the Department of Forensic Medicine at Aarhus University
for ultra-high-performance liquid chromatography tandem mass spectrometry (UHPLC-MS/MS).

The UHPLC-MS/MS method was validated in a separate study on cadavers received
for forensic autopsy and toxicological analysis. Results from dental calculus were
validated against compounds detected in whole blood samples from the same
individuals.

Briefly, dental calculus was treated with citric acid and the dissolution extracts
were cleaned using weak and strong polymeric cation-exchange sorbents.

[[ UHPLC method description needs to be elaborated ]]
<!--rephrase and expand-->

<!-- the samples were washed in 0.5 mL MeOH for 10 seconds, and centrifuged
at 10,000 $\times$ g for 1 minute. This process was repeated. Supernatants were
mixed with 10 mL dimethyl sulphoxide (DMSO) and 50 mL SIL-IS solution in a conical
glass centrifuge tube. The solvent was evaporated under a stream of nitrogen at
40&deg;C until ca. 10 mL residue remained, which was vortexed with 50 ml 30% (v/v)
MeOH. The calculus residue was air-dried in the dark for 24 h and weighed. The
washing/drying procedure was repeated twice using 1 mL MeOH, resulting in three
wash extracts for analysis. -->
In the original method, samples were washed three times to remove residual
substances from the surface of the calculus that originated from oral fluids,
and only extract substances from the calculus. In our samples the washes served
to remove potential contaminants from the burial environment and post-excavation
handling.
<!-- 
The remaining calculus was dissolved using lysing tube beads in 800 $\mu$L of 0.5
$\small{M}$ citric acid (CA) and 50 $\mu$L SIL-IS solution. During dissolution the
suspension was kept at ambient temperature for 1 h under continuous gentle shaking,
and subsequently mixed with 800 $\mu$L MeOH and centrifuged at 10,000 $\times$ g for
5 mins.
-->
<!--
2.4.3. Clean-up by WCX
The dissolution extract was mixed with 6300 m L of water,
100 m L of a 25% ammonia solution to obtain a pH of 6–7 and 50 m L
of the SIL-IS working solution. The solution was passed through a
WCX SPE cartridge with a maximum flow rate of 0.5 mL/min
(mainly using gravitational force (GF)). The cartridge was pre-
conditioned with 1 mL of MeOH followed by 1 mL of water. The
passed extract was collected in a glass tube for clean-up by SCX.
The cartridge was washed with 1 mL of water, then full vacuum
( 70 kPa) was applied for 5 s to remove the bulk volume of water
with a minimum of air flow passing through the sorbent. The wash
water was collected in the same tube. The substances were eluted
from the column using a 2-mL volume of freshly prepared 2% FA in
MeOH. A 10- m L volume of DMSO was added to the eluate, and then
the solvent was evaporated to a residual volume of approximately
10 m L under a stream of nitrogen at 40  C. The residue was vortexed
with a 50- m L volume of 30% MeOH, and the solution was
transferred to an autosampler vial. This solution was used for
the analysis of most substances.

2.4.4. Clean-up by SCX
The sample fraction saved from WCX clean-up was mixed with
100 m L of 12 M HCl and passed through an SCX SPE cartridge that
was pre-conditioned with 1 mL of MeOH followed by 1 mL of
water. The cartridge was washed with 1 mL of water, then full
vacuum was applied for 5 s before it was eluted with 2 mL of 5%
ammonia in MeOH. The eluate was mixed with 10 m L of DMSO
and evaporated to a residual volume of approximately 10 m L
under a stream of nitrogen at 40  C. The residue was vortexed
with a 50- m L volume of 30% MeOH, and the solution was
transferred to an autosampler vial. This solution was used for the
analysis of morphine-3-glu, morphine-6-glu, pregabalin, gaba-
pentin and salicylic acid.

2.4.5. UHPLC
A 5- m L volume was injected onto the column running 99%
mobile phase A. The column flow rate was 400 m L/min, and the
column temperature was maintained at 40  2  C. The mobile
phase was changed through a linear gradient to 0% A and 100% B
over 5 min. Nine minutes after the initial injection, the gradient
was returned to 99% A over 0.1 min, and the column was
equilibrated for 2.9 min before the next injection, resulting in a
total run time of 12 min. The same column and gradient
programme were used for the wash, WCX and SCX extracts. The
eluent was diverted to waste during the time intervals of 0–1.5 and
8–12 min after injection using a post-column switch.

2.4.6. MS-MS
The probe temperature (TEM) was set to 400  C. The curtain
gas (CUR), ion source gas 1 (GS1), ion source gas 2 (GS2) and
collision gas (CAD) were set at 20, 80, 80 and 8 psi, respectively.
The mass spectrometer was operated in both positive ion mode
and negative ion mode at unit mass resolution using polarity
switching to obtain the maximum sensitivity for both analytes
that favour positive ionisation and analytes that favour negative
ionisation. The ion spray voltages were set at 4.5 kV. Scheduled
multiple reaction monitoring (MRM) with windows of 30 s was
applied, and at least 12 data points were obtained across each
peak. The applied MRM conditions are listed in Supplementary
Table S2. Nitrogen was used as the CAD gas. Data acquisition and
processing were performed using Analyst 1.7 and MultiQuant
3.0.3 (Sciex), respectively. -->

For more details about the method and validation, see the original study by
Sørensen and colleagues [-@sorensenDrugsCalculus2021].

Of the `r nrow(demography)` individuals that were analysed,
`r nrow(filter(metadata, replicated == TRUE))` individuals were subsequently
replicated using leftover calculus from the original DNA study.
in <!-- the same lab? different lab? -->.

### Skeletal analysis

<!-- methods for age, sex, and stature estimation -->
Demographic and pathological analyses were conducted in the Laboratory for Human
Osteoarchaeology at Leiden University. Sex was estimated using cranial
and pelvic morphological traits [@Standards1994]. Age was estimated using dental wear,
auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985].
<!-- differential diagnoses -->

#### Paleopathology

Various pathological conditions were included in the analysis and were
dichotomised to presence/absence to allow statistical analysis.
Osteoarthritis (OA) was considered present in cases with eburnation.
No distinction was made between active or <!-- ? --> lesions in cribra femora
and cribra orbitalia.
Data on chronic maxillary sinusitis were obtained from another study
[@casnaUrbanizationRespiratory2021].

[[ Description of method for sinusitis - Maia ]]
<!-- Description of lesions included in path conditions - Supp Material?

- OA = osteoarthritis
- IVDD = intervertebral disc disease
- TB = tuberculosis
- DISH = diffuse idiopathic skeletal hyperostosis
- VOP = vertebral osteophytosis
- SN = schmorl's nodes
- DDD = degenerative disc disease
- PNBF = periosteal new bone formation
- OD = osteochondritis dissecans
- CF = cribra femora
- CO = cribra orbitalia

Sinusitis as well
-->

#### Dental pathology

Caries ratios were calculated by dividing the number of lesions by the number
of teeth scored, resulting in a single caries ratio per individual. If the surface
where the lesion originated is not visible, i.e. if the lesion covered multiple
surfaces, this was scored as "crown".

Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
Calculus was scored with a four-stage scoring system (0-3) to score absent, slight,
moderate, and heavy calculus deposits [@brothwellDiggingBones1981]
on the lingual, buccal (labial), and interproximal surfaces of each tooth. Only
one score was used for the combined interproximal surfaces, resulting in three
scores per tooth (when surfaces are intact), and four calculus indices per individual;
upper anterior (UA), upper posterior (UP), lower anterior (LA), lower posterior (LP).
Each index was calculated by dividing
the sum of calculus scores for each surface by the total number of surfaces scored
in each quadrant. If a tooth could not be scored on all three surfaces, the tooth
was not included [@greeneQuantifyingCalculus2005].

### Statistical analysis

All compounds and pathological conditions/lesions were converted to a presence/absence
score. Pearson product-moment correlation was applied to the
dichotomised pathological lesions (point-biserial correlation),
compound concentrations, calculus indices, and caries ratios to explore relationships
paired continuous-continuous variables and paired continuous-binary variables.
Compound concentrations were then dichotomised to presence/absence, and
the caries ratio and calculus index for each individual were converted to an
ordinal score from 0 to 4 by using quartiles. Polychoric correlation was applied
to the paired dichotomous variables and dichotomous-ordinal variables.
<!--Median periodontal scores were calculated for each individual. -->

All statistical analysis was conducted in `r version$version.string`,
`r version$nickname`, [@Rbase]. Data wrangling was conducted with the **tidyverse**
[@tidyverse2019] and visualisations were created using **ggplot2** [@ggplot2].
Polychoric correlations were calculated with the **psych** package [@Rpsych].

### 2022-12-12 inst/paper/_results.qmd
initial commit

<!-- detected compounds -->
Multiple compounds were detected in the dental calculus samples. Compounds detected
at a lower concentration than the LLOQ were considered not present. Not all the
compounds detected in the first batch could be replicated in the second batch
(@tbl-compound-detect). For a full list of targeted compounds, see Supplementary
Material.
<!-- maybe move full list to supp. mat and only include compounds here that were detected in one of the batches. -->

```{r}
#| label: detection-tbl-setup

uhplc_calculus_detect <- uhplc_calculus_long %>%
#uhplc_calculus_detect <- uhplc_calculus_demography %>%
  #remove_missing() %>%
  ungroup() %>% 
  filter(quant > 0) %>%
  distinct(compound, batch) %>%
  pivot_wider(names_from = batch, values_from = batch) %>% 
  mutate(across(!compound, ~ if_else(!is.na(.x), TRUE, FALSE)))
```

<!-- May need to limit the list to compounds that were detected in either batch and keep the full list in supplementary materials -->
```{r}
#| label: tbl-compound-detect
#| tbl-cap: "Target compound including whether it was detected (TRUE) or not (FALSE) in each batch, as well as the lower limit of quantitation (LLOQ) in ng."
uhplc_calculus_detect %>%
  left_join(compound_name_abbrev, by = c("compound" = "abbrev")) %>% 
  full_join(lloq, by = c("name" = "compound")) %>%
  mutate(across(batch1:batch2, ~ replace_na(.x, FALSE))) %>%
  filter(!(batch1 == FALSE & batch2 == FALSE)) %>% 
  select(name, batch1, batch2, lloq) %>% 
  arrange(name) %>% 
  knitr::kable(col.names = c("Compound", "Batch 1", "Batch 2", "LLOQ"))
```

Most plots show a large increase in extracted mass of a compound between the
calculus wash extracts (wash 1-3) and the dissolved calculus (calc). Most samples
containing theophylline and caffeine had the largest quantity of the compound
extracted from the first wash, then decreasing in washes 2 and 3. There is
an increase between wash 3 and the dissolved calculus in all samples.
The patterns are consistent across batches 1 and 2.
Nicotine and cotinine have the same relative quantities in the samples, i.e., the
sample with the highest extracted quantity of nicotine also had the highest extracted
quantity of cotinine [@fig-auth-plot-batch2].

```{r}
#| label: fig-auth-plot-batch2
#| fig-cap: "Quantity (ng) of each compound extracted from each sample across washes 1-3 and final calculus extraction (calc). Colours represent different calculus samples. Only results from batch 2 are shown."

uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    #sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    #facet_wrap(~ compound, scales = "free_y", labeller = compound_name_repair) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names)) +
    theme_bw() +
    theme(
      legend.position = "none"
    ) +
    labs(y = "Quantity (ng)", x = "") +
    scale_colour_viridis_d()
```

<!-- detection vs. preservation -->
To see if preservation of the skeletal remains had any effect on the detection of
compounds, concentrations of compounds were compared to the various levels of
preservation. No apparent patterns were present between preservation and detection
of compounds in the samples (@fig-detection-preservation).

```{r}
#| label: fig-detection-preservation
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names)) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()
```

<!-- detection of tobacco as accuracy -->

```{r}
#| label: tobacco-accuracy-setup
#tobacco <- uhplc_calculus_demography %>%
tobacco <- uhplc_calculus_long %>%
  filter(compound %in% c("nicotine", "cotinine")) %>% 
  mutate(
    detection = if_else(quant == 0, 0, 1) # 0 = not detected; NA = not included in batch 2.

    ) # 0 = not detected; NA = not included in batch 2.

tobacco_accuracy <- tobacco %>%
  remove_missing(vars = "quant") %>%
  group_by(sample, id, .drop = F) %>% 
  summarise(
    detection = sum(detection),
    #compound = compound
    ) %>% 
  left_join(select(demography, id, pipe_notch, preservation, age), by = "id") %>% 
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(
      detection == 0 & pipe_notch == "N" ~ 1,
      detection > 0 & pipe_notch == "Y" ~ 1,
      detection > 0 & pipe_notch == "N" ~ NaN, # no way of knowing macroscopically if the person smoked without pipe
      TRUE ~ 0
    )
  )

accuracy_age <- tobacco_accuracy %>% # move to supplementary material
  group_by(age) %>% 
  summarise(mean = mean(correct, na.rm = T))

# accuracy in replicated samples
replicated_accuracy <- tobacco %>%
  group_by(sample,id) %>%
  summarise(detection = sum(detection)) %>%
  remove_missing() %>% 
  left_join(select(demography, id, pipe_notch, preservation, age), by = "id") %>% 
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(
      detection == 0 & pipe_notch == "N" ~ 1,
      detection > 0 & pipe_notch == "Y" ~ 1,
      detection > 0 & pipe_notch == "N" ~ NaN, # no way of knowing macroscopically if the person smoked without pipe
      TRUE ~ 0
    )
  )
```

The presence of pipe notch(es) in an individual and concurrent detection of nicotine
and/or cotinine is used as a crude indicator of the accuracy of the method. When
combining the results of both batches, the method was able to detect some form
of tobacco in
`r nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0))`
of `r nrow(filter(tobacco_accuracy, pipe_notch == "Y"))`
individuals with a pipe notch
(`r scales::percent(nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(tobacco_accuracy, pipe_notch == "Y")), accuracy = 0.1)`).
When also considering correct the absence of a tobacco alkaloid together with the absence
of a pipe notch, the accuracy of the method is
`r scales::percent(mean(tobacco_accuracy$correct, na.rm = T), accuracy = 0.1)`.
Accuracy in the old adult age category is
`r scales::percent(filter(accuracy_age, age == "old")$mean, accuracy = 0.1)`.

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0))`
out of
`r nrow(filter(replicated_accuracy, pipe_notch == "Y"))`
pipe smokers
(`r scales::percent(nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(replicated_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.

One individual---an old adult, probable female---was positive
for both nicotine and cotinine, and had no signs of a pipe notch.

<!-- missing data -->

Additional results from dental and skeletal lesions can be found in the supplementary
materials.

### Correlations between detected alkaloids and diseases

For further statistical analyses, only the UHPLC-MS/MS results from batch 2 were
used, as batch 1 had results that could not be replicated and may have contained
contamination.

```{r}
#| label: tbl-pearson
corr_tib <- conc_cor %>% 
  as_tibble(rownames = "var") %>% 
  filter(var != "age",
         var != "periodont_status") %>% 
  column_to_rownames("var") %>%
  mutate(across(everything(), signif, 3))
corr_tib %>% 
  knitr::kable(caption = "Pearson correlation on dichotomous skeletal lesions and compound concentrations from batch 2. Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation.")
```

Point-biserial (Pearson) correlation on caries and calculus versus
presence/absence variables. We found only weak correlations between the calculus
index and caries ratio and other variables.

Polychoric correlation was conducted on the dichotomised compounds and pathological
conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample, and is
likely to cause spurious correlations.
Strong correlations were found between `r cor_statement(polycorr_tib, strength = "strong")`.
Moderate correlations were found between `r cor_statement(polycorr_tib, strength = "moderate")`. Remaining
correlations were weak or absent (@fig-polycorr).
Correlations with age will be depressed because age was largely controlled for
in the sample selection.

```{r}
#| label: fig-polycorr
#| fig-cap: "Plot of the polychoric correlations. Larger circles and stronger calculus indicate stronger correlation coefficient."
polycorr$rho %>%
  ggcorrplot::ggcorrplot(
    method = "circle", type = "lower", 
    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::plasma(1, begin = 1))
    ) +
  scale_x_discrete(labels = compound_names) +
  scale_y_discrete(position = "right") +
  theme(
    legend.position = "left",
  )
```

### 2022-12-12 inst/paper/paper.qmd
initial commit

---
title: "Title Goes Here"
author:
  - Bjørn Peare Bartholdy:
      correspondence: "yes"
      email: seriousname@email.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
  - Jørgen B. Hasselstrøm
  - Lambert K. Sørensen
  - Maia Casna
  - Historisch Genootschap Beemster
  - Amanda G. Henry
institute:
  - fosg:
      name: ZZ9 Plural Z Alpha
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
#date: now
#date-format: long
format: 
  html: 
    theme: cosmo
execute: 
  freeze: true
knitr: 
  opts_chunk:
    echo: false
    warning: false
    message: false
    comment: "#>"
    fig-path: "../figures/"
    fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  #- ../templates/pagebreak.lua
bibliography: references.bib
csl: "apa.csl" # Insert path for the bib-style
abstract: |
  This is an abstract.
keywords: |
  stuff; and; things
highlights: |
  These are the highlights. 
---

```{r}
#| label: setup
#| include: false
#| cache: false

library(here)
library(readr)

# upload data
metadata <- read_tsv(here("inst/data/raw_data/metadata.tsv"))
demography <- read_csv(here("inst/data/raw_data/demography.csv"))
lloq <- read_tsv(here("inst/data/raw_data/lloq.tsv"))
uhplc_data_comb <- read_csv(here("inst/data/derived_data/uhplc-data_combined.csv"))
#uhplc_calculus <- read_csv(here("inst/data/derived_data/uhplc-calculus_cleaned.csv"))
dental_inv <- read_csv(here("inst/data/raw_data/dental-inv.csv"))
caries <- read_csv(here("inst/data/raw_data/caries.csv"))
periodont <- read_csv(here("inst/data/raw_data/periodontitis.csv"))
periap <- read_csv(here("inst/data/raw_data/periapical.csv"))
calculus <- read_csv(here("inst/data/raw_data/calculus.csv"))
calculus_full <- read_csv(here("inst/data/raw_data/calculus_full.csv"))
sinusitis_clean <- read_csv(here("inst/data/derived_data/sinusitis_cleaned.csv"))
path_cond_clean <- read_csv(here("inst/data/derived_data/path-conditions_cleaned.csv"))

source(here("inst/scripts/setup-qmd.R"))
#source(here("inst/supplementary-materials/supp-mat.qmd"), local = knitr::knit_global())
# set viridis as default pallette

options(ggplot.discrete.colour = function() scale_fill_viridis_d())
```

Keywords: {{< meta keywords >}}

<!-- Highlights: {{< meta highlights >}} -->

<!-- The actual document text starts here: -->

## Introduction

{{< include _intro.qmd >}}

## Materials

{{< include _materials.qmd >}}

## Methods

{{< include _methods.qmd >}}

## Results

{{< include _results.qmd >}}

## Discussion

{{< include _discussion.qmd >}}

## Conclusion

## Acknowledgements {-}

Cats and dogs

## References {-}

::: {#refs}
:::

### 2022-12-12 inst/supplementary-materials/supp-mat.qmd
initial commit

---
title: "Supplementary Materials"
knitr: 
  opts_chunk: 
    warning: false
    message: false
format: 
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false
library(here)
library(readr)
devtools::load_all()

# upload data
metadata <- read_tsv(here("inst/data/raw_data/metadata.tsv"))
demography <- read_csv(here("inst/data/raw_data/demography.csv"))
lloq <- read_tsv(here("inst/data/raw_data/lloq.tsv"))
uhplc_data_comb <- read_csv(here("inst/data/derived_data/uhplc-data_combined.csv"))
#uhplc_calculus <- read_csv(here("inst/data/derived_data/uhplc-calculus_cleaned.csv"))
dental_inv <- read_csv(here("inst/data/raw_data/dental-inv.csv"))
caries <- read_csv(here("inst/data/raw_data/caries.csv"))
periodont <- read_csv(here("inst/data/raw_data/periodontitis.csv"))
periap <- read_csv(here("inst/data/raw_data/periapical.csv"))
calculus <- read_csv(here("inst/data/raw_data/calculus.csv"))
calculus_full <- read_csv(here("inst/data/raw_data/calculus_full.csv"))
sinusitis_clean <- read_csv(here("inst/data/derived_data/sinusitis_cleaned.csv"))
path_cond_clean <- read_csv(here("inst/data/derived_data/path-conditions_cleaned.csv"))

source(here("inst/scripts/setup-qmd.R"), local = knitr::knit_global())
```

## Session information

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
print(sessionInfo(), locale = F)
```

## Data dictionary

### metadata

| variable | description |
|---|---|
| id | unique identifier for the individual |
| sample | sample number for the UHPLC-MS/MS analysis |

### dental-inv

Dental inventory

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and status |

Dental inventory key:

- p = present
- m = missing (for unknown reason)
- aml = ante-mortem loss
- dna = removed for DNA sampling

### caries

Caries lesions location

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and location of caries |

Caries location key:

- none = no caries present
- mes = mesial
- dis = distal
- lin = lingual (and labial)
- buc = buccal
- occ = occlusal
- crown = crown (caries-related loss of much of/entire tooth crown)
- root = root
- blank = tooth not present

### periodontitis

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and periodontitis score (0-3) |

Periodontitis score:

- 0 = none
- 1 = slight
- 2 = moderate
- 3 = severe

- blank = tooth not present

### periapical

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and periapical lesion |

- none = no lesion
- bucc = buccal
- lin = lingual
- perf = perforated alveolar bone
- blank = not scoreable

### calculus

| variable | description |
|---|---|
| feat | Identifier: Feature number of individual |
| find | Identifier: Find number of individual |
| t11..t48 | Tooth number (FDI system) and deposit size (0-3) |

- 0 = no calculus
- 1 = slight calculus
- 2 = moderate calculus
- 3 = heavy calculus

Scored according to Standards.

### path-conditions

Pathological conditions

- OA = osteoarthritis
- IVDD = intervertebral disc disease
- TB = tuberculosis
- DISH = diffuse idiopathic skeletal hyperostosis
- VOP = vertebral osteophytosis
- SN = schmorl's nodes
- DDD = degenerative disc disease
- PNBF = periosteal new bone formation
- OD = osteochondritis dissecans
- CF = cribra femora
- CO = cribra orbitalia

## Sample demographics

```{r}
#| label: male-sample
males <- demography %>%
  filter(sex == "m" | sex == "pm")
```

The sample consists of `r nrow(demography)` individuals, most of which are
middle adult male individuals (@fig-age-distribution).

```{r}
#| label: fig-age-distribution
demography %>%
  ggplot(aes(x = age, fill = sex)) +
    geom_bar() +
    theme_minimal()
```

## Missing data

Missing data per dataset

```{r}
#| label: fig-missing-hex
caries %>%
  mutate(across(!id, ~ is.na(.x))) %>%
  select(!id) %>%
  colSums()

caries %>%
  mutate(across(!id, ~ is.na(.x))) %>% 
  pivot_longer(-id) %>% 
  ggplot(aes(x = name, y = id, fill = value)) +
    geom_tile()

```

Missing data across all dental datasets

```{r}
#| label: fig-missing-dental
dental_long %>%
  mutate(across(c(caries, periodont, periap, calculus), is.na)) %>%
  select(c(caries, periodont, periap, calculus, id, tooth)) %>%
  #group_by(id, tooth) %>% 
  rowwise() %>% 
  mutate(missing = sum(caries, periodont, periap, calculus)) %>% 
  ggplot(aes(x = tooth, y = id, fill = missing)) +
    geom_tile()
```

## UHPLC analysis

```{r}
#| label: tbl-lloq
#| tbl-cap: "Target compounds and lower limits of quantitation."
lloq %>%
  arrange(compound)
```

```{r}
#| label: tbl-uhplc-batch-1
#| tbl-cap: "Results from the UHPLC analysis first batch. Quantity of compound in the dissolved calculus, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch1") %>%
  #select(sample, contains("batch1")) %>%
  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
  pivot_wider(names_from = "compound", values_from = "quant") %>%
  select(!c(extraction, batch))
```

```{r}
#| label: tbl-uhplc-batch-2
#| tbl-cap: "Results from the UHPLC analysis second batch. Quantity of compound in calculus after third wash, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch2") %>%
  #select(sample, contains("batch1")) %>%
  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
  pivot_wider(names_from = "compound", values_from = "quant") %>%
  select(!c(extraction, batch))
```

### Authentication

No modern synthetic drugs were detected in any of the samples.

Samples were replicated to verify results from the initial analysis. Of the
`r nrow(demography)` samples initially analysed,
`r nrow(filter(metadata, replicated == TRUE))` samples were replicated.

Only caffeine, theophylline, nicotine, cotinine, and salicylic acid were found
in the replicated samples.

```{r}
#| label: tbl-uhplc-replicated
uhplc_calculus_replicated %>%
  filter(sum(presence) != 0) %>% # remove compounds absent in all individuals in sample
  pivot_wider(names_from = "compound", values_from = "presence")
```

<!-- line plot with concentration across washes and to the final calculus sample -->

Most plots show a large increase in extracted mass of a compound between the
calculus wash extracts (wash 1-3) and the dissolved calculus (calc). Most samples
containing theophylline and caffeine had the largest quantity of the compound
extracted from the first wash, then decreasing in washes 2 and 3. There is
an increase between wash 3 and the dissolved calculus in all samples.
The patterns are consistent across batches 1 and 2. The pattern we expect to see
in a sample is a reduction in the quantity from wash 1 to wash 3, and then another
spike in the final extraction from the dissolved calculus, which means the compound
is actually 'ancient' or authentic. The compounds that are completely absent in
all three washes and present in high quantities in the final extraction may also
be suggestive of lab contamination.
This has not been thoroughly tested and is
only based on what we expect to see. Therefore, the interpretation of these graphs
is itself up for interpretation.

```{r auth-plot-batch1}
#| label: fig-auth-plot-batch1

uhplc_data_long %>%
  filter(
    batch == "batch1",
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
```

```{r}
#| label: fig-auth-plot-batch2
uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
```

### Quantity vs. sample weight

There is no clear relationship between the sample weight and the amount of compound
detected, except for salicylic acid, where the amount of extracted compound increases
with increasing sample weight. In batch 2 there is also a slight positive trend
for caffeine, nicotine, and cotinine.
Nicotine and cotinine display the same relative relationship between samples. Where
the nicotine quantity is high compared to other samples, the cotinine quantity
will be similarly high (@fig-quant-weight-1 and @fig-quant-weight-1).

```{r}
#| label: fig-quant-weight-1
#| #| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 1."
uhplc_data_comb %>%

  select(sample, contains("batch1")) %>% 
  select(sample, batch1_weight, contains("calc")) %>% 
  pivot_longer(
    -c(sample, batch1_weight),
    names_to = c("compound", "batch"),
    names_pattern = "(.*)_(.*)",
    values_to = c("conc")
  ) %>% 
  filter(conc > 0) %>% 
  mutate(compound = str_remove(compound, "_calc")) %>%
  ggplot(aes(x = batch1_weight, y = conc, col = as.factor(sample))) +
    geom_point() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = "", y = "Quantity (ng)")
```

```{r}
#| label: fig-quant-weight-2
#| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 2."
uhplc_data_comb %>%
  select(sample, contains("batch2")) %>% 
  select(sample, batch2_weight, contains("calc")) %>% 
  pivot_longer(
    -c(sample, batch2_weight),
    names_to = c("compound", "batch"),
    names_pattern = "(.*)_(.*)",
    values_to = c("conc")
  ) %>% 
  mutate(compound = str_remove(compound, "_calc")) %>%
  #remove_missing() %>% 
  filter(conc > 0) %>% 
  ggplot(aes(x = batch2_weight, y = conc, col = as.factor(sample))) +
    geom_point() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw() +
    labs(y = "Quantity (ng)", y = "", col = "Sample number")
```

### Distribution of compounds detected in the samples

<!-- absolute counts in each batch -->

```{r}
#| label: fig-compounds-detect
#| fig-cap: "Number of individuals in which each compound was detected between batch 1 and 2."

# awkward wrangling required because the information on which individuals were included in replication was lost when combining batches 1 and 2
#n_batch1 <- length(metadata$replicated)
#n_batch2 <- sum(metadata$replicated) # sum boolean vector to get length of TRUE values
uhplc_calculus_long %>%
  #remove_missing() %>%
  filter(quant > 0) %>% 
  #mutate(detect = replace_na(conc, 0)) %>% 
  #mutate(detect = if_else(detect > 0, 1, 0)) %>%
  #group_by(compound, batch) %>% 
  # summarise(
  #   n_detect = sum(detect)
  #           ) %>% 
  # mutate(n = case_when(batch == 1 ~ n_batch1,
  #                      batch == 2 ~ n_batch2)) %>% 
  # mutate(prop = n_detect / n) %>%

ggplot(aes(x = compound, fill = compound)) +
    #geom_jitter(width = 0.2) +
    #geom_boxplot() +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "none"
      )
```

The replication showed that caffeine, theophylline, cotinine, nicotine, and
salicylic acid could be consistently detected in the samples, although theophylline
detection decreased between batches 1 and 2. CBD, CBN, cocaine, and THCA-A was not
detected at all in the second batch.

<!-- absolute counts in the replicated individuals -->

```{r}
#| label: fig-compounds-detect2
#| fig-cap: "Number of individuals in which each compound was detected between batch 1 and 2. Only showing replicated individuals."

uhplc_calculus_long %>%
  filter(id %in% filter(metadata, replicated == T)$id,
         quant > 0) %>%
  #remove_missing() %>%
  ggplot(aes(x = compound, fill = compound)) +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "none"
      )
```

### Detection and preservation

To see if preservation of the skeletal remains had any effect on the detection of
compounds, absolute quantities of compounds were compared to the various levels of
preservation.

```{r}
#| label: fig-detection-preservation
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected
    ) %>% 
  ggplot(aes(x = preservation, y = quant, fill = preservation)) +
    geom_violin() +
    geom_jitter(width = 0.2) +
    #geom_boxplot(alpha = 0.6, width = 0.2) +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
# 
# uhplc_calculus_demography %>%
#   filter(
#     !is.na(preservation),
#     !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected
#     ) %>% 
#   ggplot(aes(x = preservation, y = quant, fill = preservation)) +
#     geom_violin() +
#     geom_jitter(width = 0.2) +
#     #geom_boxplot(alpha = 0.6, width = 0.2) +
#     facet_wrap(~ compound, scales = "free_y") +
#     theme_bw()
```

```{r}
#| label: fig-detection-preservation2
#uhplc_calculus_demography %>%
uhplc_calculus_long %>% 
  filter(!is.na(preservation)) %>% 
  ggplot(aes(x = preservation, y = quant, fill = preservation)) +
    geom_boxplot() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
```

### Detection of tobacco

Given that pipe notches are present in the majority of individuals, the presence
of pipe notch(es) in an individual and concurrent detection of nicotine and/or
cotinine is used as a rough indicator of the accuracy of the method.

```{r}
#| label: fig-corr-notches
# correlation between nicotine and cotinine conc, and pipe notches
uhplc_calculus_long %>%
  ggplot(aes(x = pipe_notch, y = conc)) +
    geom_point() +
    facet_wrap(~ compound + batch, scales = "free_y")
```

```{r}
tobacco <- uhplc_calculus_long %>%
  filter(compound %in% c("nicotine", "cotinine")) %>% 
  mutate(detection = if_else(is.na(conc), FALSE, TRUE))

tobacco_comb <- tobacco %>% 
  group_by(batch, sample, id, .add = TRUE) %>%
  summarise(
    detection = sum(detection), 
    #sex, age, completeness, preservation, pipe_notch, 
    .groups = "keep" # combine compounds: 0 = none detected; 1 = 1 detected; 2 = both detected
  ) %>%
  ungroup() %>%
  left_join(select(demography, id, pipe_notch), by = "id") %>%
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(
      detection > 0 & pipe_notch == "Y" ~ TRUE,
      detection == 0 & pipe_notch == "N" ~ TRUE,
      detection > 0 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
      TRUE ~ FALSE
      )
    )

nicotine <- tobacco %>%
  filter(compound == "nicotine") %>% 
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(
      detection > 0 & pipe_notch == "Y" ~ TRUE,
      detection == 0 & pipe_notch == "N" ~ TRUE,
      detection > 0 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
      TRUE ~ FALSE
    )
  )

cotinine <- tobacco %>%
  filter(compound == "cotinine") %>% 
  mutate(
  correct = case_when(
    detection > 0 & pipe_notch == "Y" ~ TRUE,
    detection == 0 & pipe_notch == "N" ~ TRUE,
    detection > 0 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
    TRUE ~ FALSE
    )
  )

# accuracy in replicated samples
uhplc_accuracy <- uhplc_calculus_replicated %>%
  left_join(select(demography, id, pipe_notch), by = "id") %>% 
  filter(compound == "nicotine" | compound == "cotinine") %>%
  group_by(id) %>% 
  arrange(desc(presence)) %>%
  slice(which(compound == "nicotine" || compound == "cotinine")) %>%
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(presence == 1 & pipe_notch == "Y" ~ TRUE,
                             presence == 0 & pipe_notch == "N" ~ TRUE, # can we be sure if this is correct?
                             presence == 1 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
                             TRUE ~ FALSE))

# ratio of nicotine to cotinine
tobacco_ratio <- tobacco %>%
  select(id, batch, compound, conc) %>% 
  pivot_wider(names_from = compound, values_from = conc) %>% 
  group_by(id, batch) %>% 
  summarise(ratio = cotinine / nicotine) %>%
  remove_missing() %>% 
  left_join(select(demography, id, age, sex, preservation, pipe_notch)) %>% 
  filter(batch == "batch2",
         ratio != Inf) %>% 
  mutate(
    age = case_when(
      age == "eya" ~ 0,
      age == "lya" ~ 1,
      age == "ma" ~ 2,
      age == "old" ~ 3
    ),
    preservation = case_when(
      preservation == "fair" ~ 0,
      preservation == "good" ~ 1,
      preservation == "excellent" ~ 2
    )
  )

tobacco_ratio %>%
  ggplot(aes(y = ratio, x = pipe_notch)) +
    #geom_boxplot()
    geom_point()

  #lm(ratio ~ age + sex + preservation + pipe_notch, tobacco_ratio)
```

No apparent correlation between the number of pipe notches
and the concentration of nicotine or cotinine [@fig-corr-notches].

Nicotine and cotinine detection successfully identified
`r nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y", detection > 0))`
smokers out of
`r nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y"))`
(`r scales::percent(nrow(filter(tobacco_comb, pipe_notch == "Y", batch == 1, detection > 0)) / nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`)
individuals with pipe notches in the first analysis.

Nicotine detection successfully identified
`r nrow(filter(nicotine, batch == 1, pipe_notch == "Y", detection > 0))`
smokers out of
`r nrow(filter(nicotine, batch == 1, pipe_notch == "Y"))`
individuals with pipe notches
(`r scales::percent(nrow(filter(nicotine, batch == 1, pipe_notch == "Y", detection > 0)) / nrow(filter(nicotine, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`).

Cotinine detection successfully identified
`r nrow(filter(cotinine, batch == 1, pipe_notch == "Y", detection > 0))`
smokers out of
`r nrow(filter(cotinine, batch == 1, pipe_notch == "Y"))`
individuals with pipe notches
(`r scales::percent(nrow(filter(cotinine, batch == 1, pipe_notch == "Y", detection > 0)) / nrow(filter(cotinine, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`).

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0))`
out of
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y"))`
smokers
(`r scales::percent(mean(filter(uhplc_accuracy, pipe_notch == "Y")$correct, na.rm = T))`).
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(uhplc_accuracy$correct, na.rm = T), accuracy = 0.1)`.

One individual---an old adult, probable female---was positive
for both nicotine and cotinine, and had no signs of a pipe notch.

## Dental analysis

```{r}
#| label: fig-dental-plots
#| layout-ncol: 2
#| fig-cap: 
#|   - "Antemortem tooth loss in sample."
#|   - "Periodontal scores in the sample."
dental_long %>%
  dental_plot(fill = status)

dental_long %>%
  filter(!is.na(periodont)) %>% 
  dental_plot(fill = periodont)
```

### AMTL

### Caries

Caries were scored based on location in each individual. Multiple locations on
a single tooth is possible. The size of caries was also
recorded, but not used in further analysis. Large caries that cover multiple
surfaces were recorded as 'crown'. Caries counts were extracted based on the
number of locations recorded.

| code | surface |
|---|---|
| mes | mesial surface |
| dis | distal surface |
| lin | lingual surface |
| buc | buccal surface (including labial surface) |
| occ | occlusal surface (including incisal surface) |
| crown | caries covers 2+ surfaces |
| none | No caries visible on surface |
| NA | Not observable/tooth missing |

```{r}
surface <- c("mes", "dis", "occ", "buc", "lin", "root", "crown") # here occ = incisal

# caries_count <- dental_long %>% 
#   select(!c(periodont, periap, calculus)) %>% 
#   #filter(status == "p") %>%
#   separate_rows(caries, sep = ";") %>% # one lesion per row
#   mutate(caries_count = if_else(
#     caries == "none", 0L, 1L
#     )
#   ) %>%
#   group_by(id, tooth) %>%
#   summarise(
#     caries_count = sum(caries_count, na.rm = F)
#     ) %>%
#   tooth_position()
# 
# site_wide_caries <- caries_count %>%
#   caries_ratio(.caries = caries_count)
```

In the `r caries_ratio_site$n_teeth` that were examined, `r caries_ratio_site$count`
teeth had caries
(`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`).
This frequency has very little meaning, and was further broken down into tooth
class and individuals.

```{r}
caries_count %>%
  group_by(class) %>% 
  caries_ratio(.caries = count)
```

As expected, the molars have a higher frequency of caries than the other teeth.

```{r}
caries_count %>% 
  group_by(id, class) %>% 
  caries_ratio(.caries = count)
```

```{r}
caries_count %>%
  caries_ratio(.caries = count, id, class)
```

```{r}
caries_count %>%
  caries_ratio(.caries = count, class, id) %>%
  ggplot(aes(x = class, y = ratio)) +
    geom_violin(aes(fill = class)) +
    geom_boxplot(width = 0.1)
```

```{r}
upper_order <- c(paste0("t", 18:11), paste0("t", 21:28))
lower_order <- c(paste0("t", 48:41), paste0("t", 31:38))

maxilla <- caries_count %>%
  filter(region == "maxilla") %>% 
  mutate(tooth = factor(tooth, levels = c(upper_order, lower_order))) %>% 
  group_by(tooth) %>% 
  summarise(
    n_teeth = n(),
    count = sum(count, na.rm = T),
    rate = count / n_teeth
    ) %>% 
  ggplot(aes(x = tooth, y = rate)) +
    geom_col(fill = "white") +
    theme_dark() +
    scale_y_reverse(limits = c(0.33,0), sec.axis = sec_axis(~.)) +
    scale_x_discrete(position = "top") +
    theme(
      axis.title.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line = element_line(colour = "red", size = 1),
      axis.line.x.top = element_line(colour = "red", size = 4)
      ) +
  labs(y = "")

mandible <- caries_count %>%
  filter(region == "mandible") %>% 
  mutate(tooth = factor(tooth, levels = c(upper_order, lower_order))) %>% 
  group_by(tooth) %>% 
  summarise(
    n_teeth = n(),
    count = sum(count, na.rm = T),
    rate = count / n_teeth
    ) %>% 
  ggplot(aes(x = tooth, y = rate)) +
    geom_col(fill = "white") +
    scale_y_continuous(sec.axis = sec_axis(~.)) +
    theme_dark() +
    theme(axis.line = element_line(colour = "red", size = 1),
          axis.line.y.right = element_line(colour = "red", size = 1),
          axis.line.x.bottom = element_line(colour = "red", size = 6),
          axis.ticks.x = element_blank()) +
    labs(y = "Caries rate")

maxilla / mandible + plot_layout(guides = "collect")
```

### Periodontitis

```{r}
#| label: fig-periodont
dental_long %>%
  filter(!is.na(periodont)) %>% 
  dental_plot(fill = periodont)
```

### Calculus

```{r}
calc_index <- calculus_full %>%
  dental_longer(-id) %>% 
    #   group_by(id, quadrant) %>%
    # summarise(
    #   n_surf = sum(!is.na(score)),
    #   score_sum = sum(score, na.rm = T),
    #   calc_index = score_sum / n_surf
    # )

  calculus_index()
```

Distribution of calculus indices within the sample.
<!-- it is not calculating per surface (I think) -->

```{r}
#| label: fig-calculus-index
calc_index %>%
  ggplot(aes(x = quadrant, y = calc_index)) +
    geom_violin(aes(fill = quadrant), alpha = 0.6) +
    geom_boxplot(width = 0.1) +
    theme_minimal() +
    theme(panel.grid.major.x = element_blank())
```

```{r}
calc_index %>%
  left_join(uhplc_calculus_long, by = "id") %>%
  filter(
    batch == "batch1",
    compound != "thc",
    compound != "cbd",
    compound != "thcva",
    quadrant == "LA"
    ) %>% 
  ggplot(aes(x = as.factor(presence), y = calc_index)) +
    geom_violin(aes(fill = as.factor(presence)), alpha = 0.6) +
    geom_boxplot(width = 0.1) +
    facet_wrap(~ compound)
```

No apparent influence of lower anterior calculus index on the presence/absence
of a compound (or vice versa).

## Pathological conditions

## Statistical analysis

### Point-biserial correlation

Point-biserial (Pearson) correlation was conducted on compound concentrations,
calculus index, caries ratio, and binary variables.
This is done to see if any correlations exist prior to discretisation of continuous
variables.

```{r}
#| label: fig-pearson-corr
conc_cor %>%
  as_tibble(rownames = "var") %>% 
  mutate(across(!var, ~ if_else(.x == 1, NaN, .x))) %>%
  column_to_rownames("var") %>% 
ggcorrplot::ggcorrplot()
```

### Polychoric correlation

The calculus index and caries ratio for each individual was converted to an
ordinal variable by using quartiles, providing a score from 0--4.

```{r caries-calculus-quartiles}
caries_discrete
calculus_discrete
```

```{r polycorr}
polycorr$rho
```

```{r fig-polycorr}
ggcorrplot::ggcorrplot(polycorr$rho, type = "lower", show.diag = F)
```

### 2022-12-12 R/calculus_index.R

@@ -1,0 +1,62 @@
+#' Calculate calculus index
+#'
+#' @details Calculate the calculus index for each quadrant of an individual.
+#' Based on Greene et al. 2005.
+#' @param .data Data frame in long format, with one tooth surface per row, and one calculus score per surface.
+#' @return Returns a data frame with four rows per individual containing the calculated calculus
+#' index for each quadrant (CI-UP, CI-LP, CI-UA, CI-LA).
+#' @export
+
+calculus_index <- function(.data, simple = FALSE){
+
+  if(simple == FALSE){
+    group_data <- .data %>%
+      group_by(id, quadrant)
+  }
+  else{
+    group_data <- .data %>%
+      group_by(id)
+  }
+
+  group_data %>%
+  # minimally, each surface must be scorable on at least one side of the dentition
+    dentition_checker() %>%
+    mutate(
+      index = case_when(
+        region == "maxilla" & position == "anterior" ~ "CI-UA",
+        region == "maxilla" & position == "posterior" ~ "CI-UP",
+        region == "mandible" & position == "anterior" ~ "CI-LA",
+        region == "mandible" & position == "posterior" ~ "CI-LP",
+      )
+    ) %>%
+    #group_by(id) %>%
+    summarise(
+      n_surf = sum(!is.na(score)),
+      n_teeth = n_surf / 3,
+      score_sum = sum(score, na.rm = T),
+      calc_index = score_sum / n_surf
+    ) %>%
+    ungroup()
+}
+
+#' Calculate caries ratio
+#'
+#' @details Calculate the caries ratio for each tooth type. Calculated as the number
+#' of caries divided by the number of visible surfaces.
+#'
+#' @param .data Data frame in long format, with one tooth surface per row, and one calculus score per surface.
+#' @param .caries Caries count variable.
+#' @param ... Can be used to add grouping variable(s). Group by tooth, tooth type, individual, etc.
+#' @return Returns a data frame with caries ratio per group.
+#' @export
+
+caries_ratio <- function(.data, .caries, ...){
+  .data %>%
+    group_by(..., .add = T) %>%
+    summarise(
+      #n_teeth = n(),
+      n_teeth = sum(!is.na( {{ .caries }} ), na.rm = T),
+      count = sum( {{ .caries }} , na.rm = T),
+      ratio = count / n_teeth
+    )
+}

### 2022-12-12 R/dental-plot.R

@@ -1,0 +1,41 @@
+#' Plotting dental features
+#'
+#' @param .data Data frame containing...
+#' @param fill What variable is used for the 'fill' argument. Passed to ggplot().
+#' @param notation Currently not in use. May be used to support different notation in the future.
+#' @export dental_plot
+
+dental_plot <- function(.data, fill, notation){
+
+  upper_order <- c(paste0("t", 18:11), paste0("t", 21:28))
+  lower_order <- c(paste0("t", 48:41), paste0("t", 31:38))
+
+  tooth_names <- c("M3", "M2", "M1", "PM2", "PM1", "C1", "I2", "I1")
+  #upper_order <- c(tooth_names, rev(tooth_names))
+  #lower_order <- upper_order
+  tooth_data <- .data %>%
+    mutate(tooth = factor(tooth, levels = c(upper_order, lower_order)),
+           group = as.factor( {{ fill }} ))
+
+  maxilla <- tooth_data %>%
+    filter(region == "maxilla") %>%
+    ggplot(aes(x = tooth, fill = group)) +
+    geom_bar(position = "fill") +
+    scale_x_discrete(position = "top") +
+    theme_minimal() +
+    theme(axis.title.x = element_blank(),
+          panel.grid = element_blank()) +
+    scale_y_reverse()
+
+  mandible <- tooth_data %>%
+    filter(region == "mandible") %>%
+    ggplot(aes(x = tooth, fill = group)) +
+    geom_bar(position = "fill") +
+    scale_x_discrete(position = "bottom") +
+    theme_minimal() +
+    theme(axis.title.x = element_blank(),
+          panel.grid = element_blank())
+
+  require(patchwork)
+  maxilla / mandible + plot_layout(guides = "collect")
+}

### 2022-12-12 R/generate-bib.R

@@ -1,0 +1,10 @@
+library(rbbt)
+
+cite_keys <- bbt_detect_citations(list.files("./inst/paper/", "*.qmd", full.names = T))
+ignore_keys <- grep("fig-|tbl-", cite_keys, value = T)
+bbt_write_bib(
+  here("inst/paper/references.bib"),
+  keys = cite_keys,
+  ignore = ignore_keys,
+  overwrite = T
+)

### 2022-12-12 R/utils.R

@@ -1,0 +1,253 @@
+#' Pipe operator
+#'
+#' See \code{magrittr::\link[magrittr:pipe]{\%>\%}} for details.
+#'
+#' @name %>%
+#' @rdname pipe
+#' @keywords internal
+#' @export
+#' @importFrom magrittr %>%
+#' @usage lhs \%>\% rhs
+#' @param lhs A value or the magrittr placeholder.
+#' @param rhs A function call using the magrittr semantics.
+#' @return The result of calling `rhs(lhs)`.
+NULL
+
+#' Produce detailed tooth region information from tooth number (FDI notation)
+#'
+#' @param .data data frame containing tooth number (t + FDI notation)
+#' @return Returns a data frame with tooth number, region (maxilla, mandible),
+#' position (anterior, posterior), side, tooth_type (incisor, canine, etc.)
+#' @export tooth_position
+
+tooth_position <- function(.data){
+  # vectors to categorise teeth by position and type (FDI notation)
+  maxilla <- c(paste0("t", 11:18), paste0("t", 21:28))
+  mandible <- c(paste0("t", 31:38), paste0("t", 41:48))
+  all_teeth <- c(maxilla, mandible)
+  left <- all_teeth[str_detect(all_teeth, "2.|3.")]
+  posterior <- all_teeth[str_detect(all_teeth, ".[4-8]")]
+  molar <- all_teeth[str_detect(all_teeth, "\\d([6-8])")]
+  premolar <- all_teeth[str_detect(all_teeth, "\\d([4-5])")]
+  canine <- all_teeth[str_detect(all_teeth, "\\d(3)")]
+  incisor <- all_teeth[str_detect(all_teeth, "\\d([1-2])")]
+
+  central_incisor <- all_teeth[str_detect(all_teeth, "\\d(1)")]
+  lateral_incisor <- all_teeth[str_detect(all_teeth, "\\d(2)")]
+  premolar_1 <- all_teeth[str_detect(all_teeth, "\\d(4)")]
+  premolar_2 <- all_teeth[str_detect(all_teeth, "\\d(5)")]
+  molar_1 <- all_teeth[str_detect(all_teeth, "\\d(6)")]
+  molar_2 <- all_teeth[str_detect(all_teeth, "\\d(7)")]
+  molar_3 <- all_teeth[str_detect(all_teeth, "\\d(8)")]
+
+  .data %>%
+    mutate(
+      region = if_else(tooth %in% maxilla, "maxilla", "mandible"),
[…]

### 2022-12-12 inst/scripts/01-data_prep.R

@@ -1,0 +1,194 @@
+# Preparing data for analysis
+
+library(readr)
+library(dplyr)
+library(tidyr)
+library(stringr)
+library(tidyselect)
+library(purrr)
+
+# Upload data -------------------------------------------------------------
+
+# See README for details
+metadata <- read_tsv("inst/data/raw_data/metadata.tsv") # ids need to be changed to match key
+lloq <- read_tsv("inst/data/raw_data/lloq.tsv")
+uhplc_data_raw <- read_csv("inst/data/raw_data/uhplc-results.csv")
+uhplc_data_raw2 <- read_csv("inst/data/raw_data/uhplc-results_batch2.csv")
+dental_inv <- read_csv("inst/data/raw_data/dental-inv.csv")
+sinusitis <- read_csv("inst/data/raw_data/sinusitis.csv")
+path_cond <- read_csv("inst/data/raw_data/path-conditions.csv")
+caries <- read_csv("inst/data/raw_data/caries.csv")
+periodont <- read_csv("inst/data/raw_data/periodontitis.csv")
+periap <- read_csv("inst/data/raw_data/periapical.csv")
+calculus <- read_csv("inst/data/raw_data/calculus.csv")
+calculus_full <- read_csv("inst/data/raw_data/calculus_full.csv")
+demography <- read_csv("inst/data/raw_data/demography.csv")
+
+
+# CMS = chronic maxillary sinusitis
+# IPR = periosteal reaction on visceral surface of ribs
+# present only if active (pulmonary infection)
+
+# Data cleaning -----------------------------------------------------------
+
+# UHPLC-MS/MS results
+
+# removal internal validation samples
+uhplc_data <- uhplc_data_raw %>%
+  filter(sample != "16b") %>% # sample 16b removed (test sample that differs from 16 in sampling location)
+  group_by(sample) %>%
+  mutate(across(weight:cocaine_calculus, sum)) %>% # combine 12.1 and 12.2 (split samples from same sampling location - for internal validation)
+  filter(sample != "12.2") %>% # remove duplicate
+  select(!contains("weight")) %>% # remove unnecessary weights of calculus samples
+  rename(theophyl_calc = thyephyl_calc, # fix typos
+         theophyl_wash1 = thephyl_wash1,
+         theophyl_wash2 = thephyl_wash2,
+         theophyl_wash3 = thephyl_wash3)
+
[…]

### 2022-12-12 inst/scripts/02_data-explore.R

@@ -1,0 +1,69 @@
+# Soft-deprecated
+# should this all be done in the report?
+library(readr)
+library(dplyr)
+library(tidyr)
+library(purrr)
+
+# Upload data -------------------------------------------------------------
+
+metadata <- read_tsv("inst/data/raw_data/metadata.tsv")
+lloq <- read_tsv("inst/data/raw_data/lloq.tsv")
+uhplc_calculus <- read_tsv("inst/data/derived_data/uhplc-data_cleaned.csv")
+dental_inv <- read_csv("inst/data/raw_data/dental-inv.csv")
+caries <- read_csv("inst/data/raw_data/caries.csv")
+periodont <- read_csv("inst/data/raw_data/periodontitis.csv")
+periap <- read_csv("inst/data/raw_data/periapical.csv")
+calculus <- read_csv("inst/data/raw_data/calculus.csv")
+demography <- read_csv("inst/data/raw_data/demography.csv")
+sinusitis <- read_csv("inst/data/derived_data/sinusitis_cleaned.csv")
+
+
+# Convert to long ---------------------------------------------------------
+
+dental_inv_long <- dental_inv %>% # Inventory
+  right_join(demography) %>%
+  select(!c(occupation, own_grave, tax, pipe_notch)) %>%
+  pivot_longer(t11:t48, names_to = "tooth",
+               values_to = "status") %>%
+  mutate(status = case_when(status == "dna" ~ "m", # teeth missing due to DNA sampling
+                            TRUE ~ status))
+
+teeth_list <- list(caries, periodont, periap, calculus)
+
+test<- lapply(
+  teeth_list,
+  pivot_longer,
+  cols = t11:t48,
+  names_to = "tooth",
+  values_to = "score"
+)
+
+caries_long <- caries %>% # Caries
+  pivot_longer(t11:t48, names_to = "tooth",
+               values_to = "caries_score")
+
+periodont_long <- periodont %>% # Periodontitis
+  pivot_longer(cols = t11:t48,
+               names_to = "tooth",
+               values_to = "periodont_score")
+
+periap_long <- periap %>% # Periapical lesions
+  pivot_longer(cols = t11:t48,
+               names_to = "tooth",
+               values_to = "periap_score")
+
+calculus_long <- calculus %>% # Calculus
+  pivot_longer(cols = t11:t48,
+               names_to = "tooth",
+               values_to = "calculus_score")
+
+# prepare to merge data frames
[…]

### 2022-12-12 inst/scripts/03_data-analyse.R

@@ -1,0 +1,104 @@
+# data analysis
+library(psych)
+library(dplyr)
+library(readr)
+library(corrplot)
+
+# Upload data -------------------------------------------------------------
+
+#efa_data <- read_csv("inst/data/derived_data/efa-analysis_data.csv")
+
+# combine nicotine and cotinine to tobacco
+efa_comb <- efa_data %>%
+  mutate(tobacco = case_when(nicotine == 1 | cotinine == 1 ~ 1,
+                             TRUE ~ 0)) %>%
+  select(!c(nicotine, cotinine))
+
+# Exploratory factor analysis (psych package?)
+
+  # tetrachoric correlation for dichotomous variables
+
+
+# Missing values
+
+
+
+# selection of variables (remove very low and high frequencies)
+
+# dichotomise caries and calculus ratios for polychoric correlation
+  # caries ratios falling below the 25th percentile are considered absent
+  # discretisation of caries and calculus
+caries_discrete <- quantile(caries_rate$caries_rate)
+calculus_discrete <- quantile(calc_index$calc_index)
+
+caries_bin <- caries_rate %>%
+  mutate(caries_rate = case_when(caries_rate > quantile(caries_rate)[2] ~ 1,
+                                 TRUE ~ 0
+  ))
+
+
+efa_filter <- efa_comb %>%
+  select(!c(OD)) %>% # vertebral osteophytosis highly correlated with OA
+  summarise(across(where(is.numeric), function(x) sum(x) / length(x))) %>%
+  select(which(colSums(.) > 0.1))
+
+efa_select <- efa_comb %>%
+  select(names(efa_filter)) %>%
+  select(!c(periap_num)) %>%
+  filter(complete.cases(.)) #%>%
+  as.matrix()
+
+efa_discrete <- efa_select %>%
+  mutate(
+    calc_index = case_when(
+      calc_index <= calculus_discrete[1] ~ 0,
+      calc_index >= calculus_discrete[1] & calc_index < calculus_discrete[2] ~ 1,
+      calc_index >= calculus_discrete[2] & calc_index < calculus_discrete[3] ~ 2,
+      calc_index >= calculus_discrete[3] & calc_index < calculus_discrete[4] ~ 3,
+      calc_index >= calculus_discrete[4] & calc_index <= calculus_discrete[5] ~ 4,
+      ),
+    caries_rate = case_when(
+      caries_rate <= caries_discrete[1] ~ 0,
[…]

### 2022-12-12 inst/scripts/new-explore.R

@@ -1,0 +1,118 @@
+library(tidyverse)
+library(here)
+source(here("scripts/dental-plot.R"))
+
+dental_long <- readr::read_csv(here("data/derived_data/dental-full_long.csv"))
+
+
+
+# Data wrangling ----------------------------------------------------------
+
+# Caries rate
+
+# convert caries data from location to number of caries and caries rate
+# number of lesions per tooth observed
+# separate anterior and posterior teeth
+surface <- c("mes", "dis", "occ", "buc", "lin", "root", "crown") # here occ = incisal
+
+caries_rate <- dental_long %>%
+  filter(status == "p") %>%
+  separate_rows(caries_score, sep = ";") %>% # one lesion per row
+  mutate(caries_count = if_else(
+    caries_score == "none", 0L, 1L
+    )
+  ) %>%
+  group_by(id, tooth_type, position) %>%
+  summarise(
+    n_teeth = n(),
+         count = sum(caries_count, na.rm = T),
+         rate = count / n_teeth
+            )
+
+
+  # mutate(caries_count = if_else(caries_score == "none", 0L, 
+  #                               stringi::stri_count_words(caries_score)
+  #                               )
+  #       )
+  
+#caries_rate <- 
+# caries_long %>%
+#   group_by(id, position) %>%
+#   filter(status == "p") %>%
+#   summarise(n_teeth = n(),
+#             count = sum(caries_count, na.rm = T),
+#             rate = count / n_teeth
+#             ) %>%
+#   ungroup()
+#filter(status == "p") %>%
+#  summarise(caries_count = sum(caries_count, na.rm = T)) %>% #view
+#  group_by(id) %>%
+#  summarise(n_teeth = n(),
+#            caries_rate = sum(caries_count, na.rm = T) / n_teeth)
+
+# caries_rate_type <- dental_long %>%
+#   group_by(tooth_type) %>%
+#   summarise(n_teeth = sum(!is.na(caries_count)),
+#             n_caries = sum(caries_count, na.rm = T),
+#             caries_rate = n_caries / n_teeth) # shouldn't be any NA??
+
+# CARIES RATE CALCULATION
+
+# Caries longform data with row for each surface of each tooth?
+
+# convert periodontitis and calculus to mean/median score?  
+
+
+
+#full_data_long
+
+# compare replicated samples across batch 1 and 2
+
+caffeine_rep <- uhplc_data_comb %>%
+  select(sample, contains("caffeine")) %>%
[…]

### 2022-12-12 inst/scripts/new-prep.R

@@ -1,0 +1,185 @@
+library(tidyverse)
+library(here)
+
+# Upload data -------------------------------------------------------------
+
+# See README for details
+lloq <- readr::read_tsv(here("data/lloq.tsv"))
+uhplc_data_raw <- readr::read_csv(here("data/raw/anon/uhplc-results.csv"))
+uhplc_data_raw2 <- readr::read_csv(here("data/raw/anon/uhplc-results_batch2.csv"))
+dental_inv <- readr::read_csv(here("data/raw/anon/dental-inv.csv"))
+caries <- readr::read_csv(here("data/raw/anon/caries.csv"))
+periodont <- readr::read_csv(here("data/raw/anon/periodontitis.csv"))
+periap <- readr::read_csv(here("data/raw/anon/periapical.csv"))
+calculus <- readr::read_csv(here("data/raw/anon/calculus.csv"))
+demography <- readr::read_csv(here("data/raw/anon/demography.csv"))
+sinusitis <- readr::read_csv(here("data/raw/anon/sinusitis.csv"), na = "UNOBS")
+# CMS = chronic maxillary sinusitis
+# IPR = periosteal reaction on visceral surface of ribs 
+  # present only if active (pulmonary infection)
+
+# Helper objects ----------------------------------------------------------
+
+# vectors to categorise teeth by position and type (FDI notation)
+maxilla <- c(paste0("t", 11:18), paste0("t", 21:28))
+mandible <- c(paste0("t", 31:38), paste0("t", 41:48)) 
+left <- c(paste0("t", 21:28), paste0("t", 31:38))
+posterior <- c(paste0("t", 14:18), paste0("t", 24:28), 
+               paste0("t", 34:38), paste0("t", 44:48))
+molar <- c(paste0("t", 16:18), paste0("t", 26:28), 
+           paste0("t", 36:38), paste0("t", 46:48))
+incisor <- c(paste0("t", 11:12), paste0("t", 21:22), 
+             paste0("t", 31:32), paste0("t", 41:42))
+premolar <- c(paste0("t", 14:15), paste0("t", 24:25), 
+              paste0("t", 34:35), paste0("t", 44:45))
+canine <- c(paste0("t", 13), paste0("t", 23), 
+            paste0("t", 33), paste0("t", 43))
+
+# Data cleaning -----------------------------------------------------------
+
+# UHPLC-MS/MS results
+
+uhplc_data <- uhplc_data_raw %>%
[…]

### 2022-12-12 inst/scripts/setup-qmd.R

@@ -1,0 +1,378 @@
+library(here)
+library(readr)
+library(dplyr)
+library(tidyr)
+library(tibble)
+library(ggplot2)
+library(ggcorrplot)
+library(stringr)
+library(glue)
+library(purrr)
+library(psych)
+devtools::load_all()
+
+# UHPLC analysis ----------------------------------------------------------
+
+weight <- uhplc_data_comb %>%
+  select(sample, batch1_weight, batch2_weight) %>%
+  pivot_longer(
+    cols = -sample,
+    names_to = c("batch"),
+    #names_pattern = "(.*)_(.*)",
+    values_to = "weight"
+  ) %>%
+  mutate(batch = str_remove(batch, "_weight"),
+         sample = as.character(sample))
+
+uhplc_data_long <- uhplc_data_comb %>%
+  #select(!contains("id_batch")) %>%
+  mutate(sample = as.character(sample)) %>%
+  select(!c(batch1_weight, batch2_weight)) %>%
+  left_join(select(metadata, id, sample), by = "sample") %>%
+  pivot_longer(
+    -c(sample, id),
+    names_to = c("compound", "extraction", "batch"),
+    names_pattern = "(.*)_(.*)_(.*)",
+    #names_to = c("compound", "batch"),
+    #names_pattern = "(.*)_(.*)",
+    values_to = "quant"
+  )
+
+# filter to remove all compounds that weren't detected within each sample (they clog up the lower part of the plots and are uninformative)
+quant_filter <- uhplc_data_long %>%
+  group_by(sample, compound) %>%
+  summarise(
+    sum = sum(quant, na.rm = T)
+  ) %>%
+  filter(sum > 0) %>%
+  select(!sum)
+
+# uhplc_calculus_id <- uhplc_calculus %>%
+#   mutate(sample = as.character(sample)) %>%
+#   left_join(select(metadata, id, sample), by = "sample") # add individual IDs to data frame
+
+uhplc_calculus_long <- uhplc_data_long %>%
+  # pivot_longer(
+  #   -c(sample, id),
+  #   names_to = c("compound", "batch"),
+  #   names_pattern = "(.*)_(.*)",
+  #   values_to = c("quant")
+  # ) %>%
+  filter(extraction == "calc") %>%
+  #mutate(compound = str_remove(compound, "_calc")) %>%
+  #remove_missing() %>%
+  left_join(weight) %>%  # add weight of calculus samples
[…]

### 2023-01-04 inst/paper/_conclusions.qmd
add: revisions based on co-authors feedback

--- 2022-12-12 inst/paper/_discussion.qmd
@@
-<!-- Big summary/important statement for the science -->
-In this study we were able to extract and identify multiple alkaloids from the
-dental calculus of individuals from Middenbeemster, a 19th century Dutch
-archaeological site. We applied ultra-high-performance liquid chromatography-tandem
-mass spectrometry (UHPLC-MS/MS), using a method that was validated the presence
-of drugs and metabolites in dental calculus with the equivalent compounds in blood
-[@sorensenDrugsCalculus2021].
-Here we have shown that the method can also be successfully applied to archaeological
-dental calculus.
-We initially detected cannabinoids, cocaine, nicotine, cotinine, caffeine,
-theopyhlline, and salicylic acid in multiple individuals; however, the detection
-of cannabinoids and cocaine could not be replicated in a second run on different
-samples from the same individuals.
-
-<!-- discuss any correlations, spurious or otherwise
-e.g. more caffeine, salicyl -> less calculus (acidic environ?)
-tobacco -> sinusitis -->
-
-<!-- compounds present and absent with potential sources-->
-
-<!-- nicotine and cotinine: tobacco, obviously...-->
-Nicotine and cotinine were strongly positively correlated, both in concentration
-and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
-The detection of nicotine and cotinine is not surprising, as pipe-smoking in
-the Beemsterpolder is well-documented in the literature
-[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
-remains [@lemmersMiddenbeemster2013].
-There is also documented medicinal use of nicotine in the Beemsterpolder,
-where a tobacco
-smoke enema was used for headaches, respiratory problems, colds, and drowziness,
-among others, from around 1780 to 1830 [@aten400JaarBeemster2012].
-<!-- Scientific findings as extension of previous work -->
-The detection of nicotine in dental calculus has previously been presented by
-Eerkens and colleagues [-@eerkensDentalCalculus2018] in two individuals from
-pre-contact California. They also targeted caffeine, cotinine, and theophylline
-in their samples, but were unable to detect any of them. It remains to be seen
-whether this is due to differences in methods used, or due to our samples being
-more recent.
-Other alkaloids have been detected in archaeological samples, though these are more
-often associated with other materials, including pipes
-[@raffertyCurrentResearch2012] and ceramics [@smithDetectionOpium2018].
-Direct analysis on human remains
-has also included hair [@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013].
-
-<!-- theophylline and caffeine: coffee and tea? -->
-Theophylline and caffeine concentrations were positively correlated,
-suggesting that they may have originated from the same source.
-Caffeine is present in coffee, tea, and cocoa beans, with concentrations slightly higher
-in coffee [@stavricVariabilityCaffeine1988; @chinCaffeineContent2008; @bispoSimultaneousDetermination2002].
-Theophylline is present in both coffee beans and tea leaves, but in negligible
-quantities [@stavricVariabilityCaffeine1988]. It is also metabolised from
-caffeine in the liver, and given the low quantity relative to caffeine, it likely
-originated from either tea or coffee, which contain higher concentrations of
-the compounds [@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
-Caffeine and theophylline are
-structurally very similar, and we expect they would experience similar rates of
-degradation, allowing us to make interpretations on the ratio between the compounds.
-Tea consumption had become widespread in the Netherlands by 1820, reaching all aspects of
-the population (some argue that this happened already by early 18th century)
-[@nierstraszTeaTrade2015, p.91].
-Historically, we also know that both tea and coffee were consumed in the Beemsterpolder
-during the 19th century. 'Theegasten' (teatime) was a special
-occasion occurring from 15.00-20.00 hours, where tea was served along with the evening
-bread [@schuijtemakerTeTheegasten2011]. Many households also owned at least one coffee pot and tea pot [@boumanBegravenis2017].
-Both the presence of theophylline and caffeine could also be explained by chocolate,
-although given the large quantities of caffeine compared to theophylline, it is more
-likely that the source is tea and/or coffee
-[@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
-Distinguishing between tea, coffee, and chocolate may be possible by also including
-theobromine and comparing ratios of the compounds, as theobromine is present in
-higher quantities in chocolate compared to caffeine and theophylline
-[@stavricVariabilityCaffeine1988; @bispoSimultaneousDetermination2002; @alanonAssessmentFlavanol2016].
-Although there is some indication that theobromine does not preserve well
-[@velskoDentalCalculus2017], and frequent consumption of all three
-items would be difficult to parse.
-
-<!-- salicylic acid: diet, medicine, or environmental??? -->
-The ubiquitous presence of salicylic acid poses a different problem. It has
-been shown that salicilyc acid is a very mobile organic acid and the ubuquitous
-presence may be due to environmental contamination
-[@badriRegulationFunction2009; @velskoDentalCalculus2017].
-The bark of a willow tree, *Salix alba*, has long been known for its
-pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119],
-and is also present in many plant-based foods
-[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017]. This includes
-potatoes, which were a mainstay in the Beemsterpolder diet [@aten400JaarBeemster2012].
-It will be necessary to explore the extent to which salicylic acid can leech
-into the dental calculus from the soil, and what the rate of degradation is for
-salicylic acid when trapped in dental calculus.
-
-<!-- cannabinoids: medicinal? wait, where did they go..? -->
-Cannabinoids---specifically THC, THCA-A, THCVA, CBD, CBN---were found in the
-first batch, but none were replicated in the second batch. Medicinal use of
-cannabinoids has been well-established in Europe since Medieval-times, and it
-was also grown in the Netherlands [@bruinsmaBijdragenTot1872].
-Administration was most common in the form of concoctions containing various portions
-of the cannabis plant for ingestion; not until the late 19th century did it become
-recommended to smoke it for more immediate effects [@clarkeCannabisEvolution2013].
-John Parkinson described a Dutch medicinal use of hemp. An emulsion was prepared
-from the seeds of the plants to treat pain and various stomach ailments. Another
-preparation involving the roots of the plants was used for inflammation, gout, and joint
-pains [@clarkeCannabisEvolution2013].
-The ability to detect cannabinoids in calculus may be limited by a reduced ability
-to diffuse from serum to salivary glands due to an affinity for protein-binding,
-[@sorensenEffectAntioxidants2018; @sorensenDrugsCalculus2021],
-meaning detection would rely on oral consumption of the cannabinoids. Even then,
-the overall instability of some cannabinoids could also affect detection
-[@sorensenEffectAntioxidants2018; @lindholstLongTerm2010].
-
-<!-- opioids: medicinal. Damn... that would've been cool-->
-Despite many of the individuals living during the height of the opium era in the
-Netherlands, none of the targeted opioids
-(morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
-were detected. 
-Sylvius de la Boe (1614–1672), a German-born physician who studied anatomy at
-Leiden University, the Netherlands, relied heavily on opium [@machtHistoryOpium1915].
-The absence of opioids could be a result of the people ascribing more to the
-"traditional" rather than "scientific" medicine, although laudanum and
-another opium containing concoction was part of the "traditional" medicine in
-the Netherlands [@leuwProhibitionLegalizationDutch1994], including Middenbeemster
-[@aten400JaarBeemster2012]. It was also generally
-considered a drug of the upper class [@scheltemaOpiumTrade1907], and may be
-more common in an urban center.
-The absence could also be attributed to postmortem degradation.
-It has been shown that, while abundant in opium, morphine
-degrades rapidly. The more stable opioids such as thebaine and papaverine
-[@chovanecOpiumMasses2012] were also absent, however.
-<!--Similar to cannabinoids, many opioids are preferentially bound to (plasma) proteins
-[@sorensenDrugsCalculus2021].-->
-
-We explored the relationship between detected compounds and various skeletal
-indicators, such as pathological and dental lesions, preservation, and pipe notches.
-We found no evidence to suggest that preservation of the skeleton influences
-the recovery of compounds from the dental calculus. This could be due to
-the good overall preservation of both teeth and dental calculus, which often
-preserves better than the rest of the skeleton.
-We found no correlation between the number of pipe notches
-and the concentration of nicotine and cotinine, suggesting that our ability
-to detect tobacco consumption in dental calculus does not necessarily rely
-targeting frequent smokers; here, we consider individuals with multiple pipe notches
-as likely to have been heavy smokers.
-There does seem to be a positive correlation between the weight of the calculus
-sample and recovered quantities of the compounds, so weight of the sample may affect
-the ability to detect compounds; although, we were able to detect
-compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].
-
-[[ Correlation between caffeine, nicotine, and sinusitis - Maia ]]
-
-Authentication
-
-<!-- authentication -->
-The list of targeted compounds included both naturally occurring compounds known
-to have been used in the past, as well as synthetic modern drugs did not exist
-at the time (e.g. Fentanyl, MDMA, Amphetamine). These were part of the toxicology
-screening for the original method [@sorensenDrugsCalculus2021], originally
-developed on cadavers. In our study they serve as an authentication step, as their
-presence in archaeological samples could only be the result of contamination.
-<!-- the importance of replication - brief discussion of cocaine -->
-The only modern compound (at least in a European context) detected in the sample
-was cocaine, which was detected in the first batch of sampels. The population
-analysed is from the early-mid 18th century, and cocaine was isolated in 1860 by
-Albert Niemann, and entered popular medical
-practice in 1884. Coca arrived in Europe as early as 1771, but as botanical
-specimens rather than for consumption, and there were also issues importing enough
-viable specimens of coca for cocaine extraction
-[@mortimerHistoryCoca1901 p. 179; @abucaCocaTrade2019, p.108].
-We considered it possible that it would be present in a sample with
-most individuals originating from the early- to mid-19th century. If corroborated,
-this would have been the first case of coca-consumption in Europe. Given that cocaine
-had yet to be isolated, the only way it could have gotten into the calculus was
-through consumption of the coca leaf. In our replication batch, we included all
-of the individuals who had been cocaine-positive in the first batch. We were
-unable to replicate any of the cocaine results, and we were unable to detect the
-principal metabolite, benzoylecgonine, in either batch; although, the original
-study only detected trace amounts of benzoylecgonine, and only in some of the
-individuals that were cocaine-positive. We hope this will serve as
-a reminder of the importance of replication. We suspect that the original detection
-of cocaine was a result lab contamination during the UHPLC-MS/MS analysis.
-
-Accuracy
-
-We used the presence of a pipe notch and concurrent detection of nicotine
-and/or cotinine was as a crude estimate of the accuracy of the method. Nicotine
-has been widely studied in the archaeological record due to its apparent stability
-and ability to survive over long periods of time
-[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].
-The accuracy was found to be around
-`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
-We acknowledge that this is
-a very rough estimate, as the presence of a pipe notch is likely not a perfect
-indicator of whether or not someone consumed tobacco, and it may not be
-representative of other compounds with different rates of degradation.
-
-<!-- Findings' scientific implications -->
-
-<!-- secondary findings-->
-The choice of tooth for sampling may impact the detection of certain compounds,
-as the incorporation in dental calculus may depend on the mode of consumption.
-Tobacco smokers may have more
-nicotine present in calculus on incisors, whereas tobacco chewers may have more
-on molars [@eerkensDentalCalculus2018]. This has yet to be systematically
-tested in archaeological dental calculus. Direct incorporation by ingesting
-items, food or otherwise, containing the compounds
-is not the only potential pathway to incorporation in the dental calculus. 
-The presence of cotinine, the primary
-metabolite of nicotine, suggests that saliva is also a source of deposition, and
-that deposition of alkaloids in dental calculus can occur both on the way into the body,
-i.e. during consumption, and on the way out, i.e. disposal of waste products via
-saliva secretion into the mouth. Especially mucin-rich saliva from the sublingual
-and submandibular glands preferentially binds toxins [@doddsHealthBenefits2005],
-and since these glands are located closest to the lower incisors, they may be the
-most effective target for these studies. Due to homogenisation of samples in our
-study, we were unable to test this effect.
-
-<!-- How this engineering will facilitate future research -->
-<!-- both calculus and teeth preserve well-->
-
-<!-- Forward-looking statements about the field as a whole -->
-
-<!-- a summary/importance of the engineering aspect of the work -->
-
-<!-- How it facilitates scientific research -->
-
-<!-- limitations of the platform show which questions would be unanswered -->
-Limitations
-
-Since we targeted individuals with moderate-to-large calculus deposits,
-it is likely a biased sample. The presence of calculus may increase risk of premature death
-[@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
-be associated with dental calculus build-up) is a risk-factor for
-respiratory diseases, if periodontal and respiratory pathogens
-enter the blood stream
-[@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
-In our sample, the percentage of CMS
-(`r scales::percent(mean(explore_select$CMS, na.rm = T), accuracy = 0.1)`)
-is lower than in another (more representative) male sample (44.1%)
-[@casnaUrbanizationRespiratory2021], and the caries percentage is similarly lower
-in our sample (`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`) than
-a more representative sample (22.9%) [@lemmersMiddenbeemster2013].
-The bias may also affect the detection of compounds, as caffeinated drinks are often
-acidic, and may cause a lower calculus formation<!-- citation?-->.
-We only found a weak negative correlation (`r signif(corr_tib[15,2], 3)`)
-between caffeine concentration and dental calculus index, but since we targeted individuals with
-calculus, this may inadvertently have been controlled for during sample collection.
-
-Quantitation of the detected compounds may have limited value in archaeological
-samples, given the effects of taphonomy added on top of individual variability
-and relative stochasticity of dental calculus formation, and the metabolic
-degradation of compounds during life [@sorensenDrugsCalculus2021; @velskoAncientDental2022a]. <!-- the results of comparisons with concentration
-had limited outcomes, and the correlations between concentration and other variables
-were similar to the correlations with presence/absence, suggesting that the actual
-concentration was somewhat random.-->
-
-Perhaps the main limitation, which is also true for many other analyses on
-archaeological materials, is that detecting the absence of a compound is not possible.
-The absence of a compound is
-not evidence of its absence, which complicates the interpretation of our results.
-We have attempted to minimise errors occurring due to this limitation by including
-a relatively large sample of individuals. Although given the relatively low
-detection rate seen in tobacco, this remains a major limitation, and will likely
-been compounded by increasing antiquity of the samples.
-The replication of samples increases the evidence of absence if the compound
-was absent in both batches.
-
-<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
-
-While a major limitation is the uncertainty surrounding whether or not a compound
-is actually absent, the power of the method is in detecting food and other items
-that are consumed in a manner that does not allow macroscopic detection. Taking
-tobacco consumption as an example; while pipe
-notches are a useful way to identify tobacco consumption, pipe smoking was not the
-only mode of tobacco consumption, which includes chewing, drinking, cigars, snuff,
-[@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
-was mostly common in males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018].
-Our method does not, at least to our knowledge, discriminate between modes of
-consumption,
-and may be more suitable for exploring tobacco consumption in the entire population,
-not just males. In our sample, we detected nicotine and cotinine (replicated)
-in an old adult, probable female individual. Because the individual was an old
-adult, it is unlikely that the compounds entered the dental calculus through
-pipe-smoking, as there would be pipe notches. More likely that the tobacco entered
-through alternative consumption or secondhand smoke.
-
-Future studies should explore the effect of tooth position on the detection of
-compounds.
-Also on the properties increasing incorporation potential of different compounds,
-and which pathways facilitate maximum incorporation (direct or indirect pathways?),
-which likely differs between alkaloids. How is the preservation once trapped
-inside the calculus, for example using artificial ageing on real or artificial
-dental calculus?
-In this study, we only targeted drugs that were included in the forensic toxicological
-screenings, and therefore only covered a limited number of the potential compounds
-that could be of interest for exploring past diets and medicinal treatments.
-
-<!--final remarks-->
-
 This preliminary study outlines the benefits of using calculus to target a
@@
 in addition to how mode of consumption may affect concentrations on different parts
-of the dentition. <!-- also individual variation -->
-Another limitation of dental calculus as a medium is the variability its formation
-and the many factors that can influence incorporation and retention of molecules
-and particles; however, in the absence
+of the dentition.
+Another limitation of dental calculus as a medium is the inter- and intra-individual
+variability of its formation and the many factors that can influence incorporation
+and retention of molecules and particles; however, in the absence
 of hair and serum (quite uncommon in archaeology), dental calculus represents
-an impressive reservoir of information regarding the consumption of various
+an impressive long-term reservoir of information regarding the consumption of various
 alkaloids, whether dietary, medicinal, recreational, or otherwise.

### 2023-01-04 inst/paper/_discussion.qmd
add: revisions based on co-authors feedback

@@
 archaeological site. We applied ultra-high-performance liquid chromatography-tandem
-mass spectrometry (UHPLC-MS/MS), using a method that was validated the presence
-of drugs and metabolites in dental calculus with the equivalent compounds in blood
-[@sorensenDrugsCalculus2021].
+mass spectrometry (UHPLC-MS/MS), using a method that was validated by co-occurrence
+of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
 Here we have shown that the method can also be successfully applied to archaeological
 dental calculus.
-We initially detected cannabinoids, cocaine, nicotine, cotinine, caffeine,
-theopyhlline, and salicylic acid in multiple individuals; however, the detection
-of cannabinoids and cocaine could not be replicated in a second run on different
-samples from the same individuals.
+We extend findings from previous studies on alkaloids in archaeological samples
+by extracting multiple different alkaloids from dental calculus, including
+cannabinoids, cocaine, nicotine, cotinine, caffeine, theophylline, and salicylic
+acid in multiple individuals. Of these compounds, the cannabinoids and cocaine
+could not be replicated in a second run on different samples from the same individuals.
+We discuss the implications of these findings in light of historical and
+archaeological evidence for the consumption of these drugs.
 
-<!-- discuss any correlations, spurious or otherwise
-e.g. more caffeine, salicyl -> less calculus (acidic environ?)
-tobacco -> sinusitis -->
-
-<!-- compounds present and absent with potential sources-->
+<!-- compounds detected with potential sources, and those that were surprisingly missing -->
 
 <!-- nicotine and cotinine: tobacco, obviously...-->
-Nicotine and cotinine were strongly positively correlated, both in concentration
-and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
-The detection of nicotine and cotinine is not surprising, as pipe-smoking in
-the Beemsterpolder is well-documented in the literature
+Nicotine and its principal metabolite, cotinine, were strongly positively correlated,
+both in concentration and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
+The detection of nicotine and cotinine is not surprising, as tobacco-use and
+pipe-smoking in the Beemsterpolder is well-documented in the literature
 [@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
-remains [@lemmersMiddenbeemster2013].
+remains as pipe notches [@lemmersMiddenbeemster2013].
 There is also documented medicinal use of nicotine in the Beemsterpolder,
-where a tobacco
-smoke enema was used for headaches, respiratory problems, colds, and drowziness,
-among others, from around 1780 to 1830 [@aten400JaarBeemster2012].
+where a tobacco-smoke enema was used for headaches, respiratory problems, colds,
+and drowsiness from around 1780 to 1830 [@aten400JaarBeemster2012].
+In our sample, we also detected nicotine and cotinine (replicated)
+in an old adult, probable female individual.
+It is unlikely that the compounds entered the dental calculus through
+pipe-smoking, as the individual had no visible pipe notches.
+More likely that the tobacco entered through an alternate mode of consumption,
+secondhand smoke, or the aforementioned tobacco-smoke enema.
+
 <!-- Scientific findings as extension of previous work -->
-The detection of nicotine in dental calculus has previously been presented by
-Eerkens and colleagues [-@eerkensDentalCalculus2018] in two individuals from
-pre-contact California. They also targeted caffeine, cotinine, and theophylline
-in their samples, but were unable to detect any of them. It remains to be seen
-whether this is due to differences in methods used, or due to our samples being
-more recent.
-Other alkaloids have been detected in archaeological samples, though these are more
-often associated with other materials, including pipes
-[@raffertyCurrentResearch2012] and ceramics [@smithDetectionOpium2018].
-Direct analysis on human remains
-has also included hair [@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013].
+
+<!-- strong correlation between nicotine and cotinine concentration is important?
+      Could suggest that the incorporation pathway is excretion, not consumption
+        - longer time in body than mouth (consumption)
+-->
 
 <!-- theophylline and caffeine: coffee and tea? -->
-Theophylline and caffeine concentrations were positively correlated,
-suggesting that they may have originated from the same source.
-Caffeine is present in coffee, tea, and cocoa beans, with concentrations slightly higher
-in coffee [@stavricVariabilityCaffeine1988; @chinCaffeineContent2008; @bispoSimultaneousDetermination2002].
+Theophylline and caffeine were positively correlated in our samples, though to a
+lesser extent than nicotine and cotinine, so we are unable to determine if they
+originated from the same source (@tbl-pearson and @fig-polycorr).
+Caffeine and theophylline have very similar chemical structures, so we expect
+they would experience similar rates of incorporation and degradation
+degradation, allowing us to make interpretations the ratio and correlations
+between the compounds.
+Caffeine is present in coffee, tea, and cocoa beans, with concentrations slightly
+higher in coffee
+[@stavricVariabilityCaffeine1988; @chinCaffeineContent2008; @bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
 Theophylline is present in both coffee beans and tea leaves, but in negligible
-quantities [@stavricVariabilityCaffeine1988]. It is also metabolised from
-caffeine in the liver, and given the low quantity relative to caffeine, it likely
-originated from either tea or coffee, which contain higher concentrations of
-the compounds [@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
-Caffeine and theophylline are
-structurally very similar, and we expect they would experience similar rates of
-degradation, allowing us to make interpretations on the ratio between the compounds.
-Tea consumption had become widespread in the Netherlands by 1820, reaching all aspects of
-the population (some argue that this happened already by early 18th century)
-[@nierstraszTeaTrade2015, p.91].
-Historically, we also know that both tea and coffee were consumed in the Beemsterpolder
-during the 19th century. 'Theegasten' (teatime) was a special
+quantities [@stavricVariabilityCaffeine1988]. It is also a primary metabolite of
+caffeine produced by the liver. Given the low correlation, there are likely
+multiple sources of caffeine and theophylline in the population, with tea and
+coffee being the most obvious.  
+Tea consumption had become widespread in the Netherlands by 1820, reaching all
+parts of society [@nierstraszTeaTrade2015, p.91].
+Historically, we also know that both tea and coffee were consumed in the
+Beemsterpolder during the 19th century. 'Theegasten' (teatime) was a special
 occasion occurring from 15.00-20.00 hours, where tea was served along with the evening
-bread [@schuijtemakerTeTheegasten2011]. Many households also owned at least one coffee pot and tea pot [@boumanBegravenis2017].
-Both the presence of theophylline and caffeine could also be explained by chocolate,
-although given the large quantities of caffeine compared to theophylline, it is more
-likely that the source is tea and/or coffee
-[@bispoSimultaneousDetermination2002; @srdjenovicSimultaneousHPLC2008].
+bread [@schuijtemakerTeTheegasten2011]. Many households also owned at least one
+coffee pot and tea pot [@boumanBegravenis2017].
 Distinguishing between tea, coffee, and chocolate may be possible by also including
@@
 [@stavricVariabilityCaffeine1988; @bispoSimultaneousDetermination2002; @alanonAssessmentFlavanol2016].
-Although there is some indication that theobromine does not preserve well
-[@velskoDentalCalculus2017], and frequent consumption of all three
-items would be difficult to parse.
+However, In addition to oral factors
+affecting alkaloid uptake in dental calculus, there is some indication that
+theobromine does not preserve well in the archaeological record
+[@velskoDentalCalculus2017], and frequent consumption of all three items would
+be difficult to parse.
 
 <!-- salicylic acid: diet, medicine, or environmental??? -->
-The ubiquitous presence of salicylic acid poses a different problem. It has
-been shown that salicilyc acid is a very mobile organic acid and the ubuquitous
-presence may be due to environmental contamination
-[@badriRegulationFunction2009; @velskoDentalCalculus2017].
-The bark of a willow tree, *Salix alba*, has long been known for its
-pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119],
-and is also present in many plant-based foods
-[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017]. This includes
-potatoes, which were a mainstay in the Beemsterpolder diet [@aten400JaarBeemster2012].
-It will be necessary to explore the extent to which salicylic acid can leech
+Salicylic acid was found in all but one individual in our sample.
+It can be extracted from the bark of willow trees, *Salix alba*, and
+has long been used for its
+pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119]. It is also present
+in many plant-based foods
+[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017], including
+potatoes, which were a staple of the Beemsterpolder diet [@aten400JaarBeemster2012].
+The extracted quantity from our samples decreased over the three washes, followed
+by a sharp increase in the final calculus extraction, which is what we would expect
+to see if the salicylic acid was incorporated during life [@fig-auth-plot-batch2].
+However, it has been shown that salicilyc acid is a very mobile organic acid and
+the ubiquitous presence may be due to environmental contamination, which would
+also explain the high quantity in the washes
+[@chenCa2Dependent2001; @badriRegulationFunction2009].
+Given the multiple plausible sources of this residue, it will be necessary to
+explore the extent to which salicylic acid can leach
 into the dental calculus from the soil, and what the rate of degradation is for
 salicylic acid when trapped in dental calculus.
+<!-- perhaps using an oral biofilm model ;) -->
 
-<!-- cannabinoids: medicinal? wait, where did they go..? -->
+<!-- cannabinoids: oh cool! medicinal? wait, where did they go..? -->
 Cannabinoids---specifically THC, THCA-A, THCVA, CBD, CBN---were found in the
@@
 recommended to smoke it for more immediate effects [@clarkeCannabisEvolution2013].
-John Parkinson described a Dutch medicinal use of hemp. An emulsion was prepared
-from the seeds of the plants to treat pain and various stomach ailments. Another
-preparation involving the roots of the plants was used for inflammation, gout, and joint
-pains [@clarkeCannabisEvolution2013].
+A Dutch medicinal use of hemp involved an emulsion prepared from the seeds of
+the plants to treat pain and various stomach ailments. Another
+preparation involving the roots of the plants was used for inflammation, gout,
+and joint pains [@clarkeCannabisEvolution2013].
 The ability to detect cannabinoids in calculus may be limited by a reduced ability
@@
 [@sorensenEffectAntioxidants2018; @lindholstLongTerm2010].
+However, given the lack of replication, we cannot with security confirm that
+cannabis was used by the Beemster population.
 
-<!-- opioids: medicinal. Damn... that would've been cool-->
-Despite many of the individuals living during the height of the opium era in the
-Netherlands, none of the targeted opioids
+<!-- opioids: Damn... that would've been awesome -->
+Despite many of our sampled individuals living during the height of the opium era
+in the Netherlands, none of the targeted opioids
 (morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
@@
 It has been shown that, while abundant in opium, morphine
-degrades rapidly. The more stable opioids such as thebaine and papaverine
-[@chovanecOpiumMasses2012] were also absent, however.
-<!--Similar to cannabinoids, many opioids are preferentially bound to (plasma) proteins
-[@sorensenDrugsCalculus2021].-->
+degrades rapidly, while thebaine and papaverine are more resistant to various
+ageing processes [@chovanecOpiumMasses2012]. The latter were also absent from
+our samples.
 
+<!-- authentication -->
+<!-- the importance of replication - brief discussion of cocaine -->
+The only strictly modern compound (at least in a European context) detected in the sample
+was cocaine, which was detected in the first batch of samples. Our sample is
+derived from an early--mid 19th century population, and cocaine was isolated in 1860
+by Albert Niemann, and entered popular medical
+practice in 1884. Coca arrived in Europe as early as 1771, but as botanical
+specimens rather than for consumption, and there were also issues importing enough
+viable specimens of coca for cocaine extraction
+[@mortimerHistoryCoca1901 p. 179; @abucaCocaTrade2019, p.108].
+We considered it possible that it would be present in a sample with
+most individuals originating from the early- to mid-19th century. If corroborated,
+this would have been the first case of coca-leaf-consumption in Europe.
+In our replication batch, we included all
+of the individuals who had been cocaine-positive in the first batch. We were
+unable to replicate any of the cocaine results, and we were unable to detect the
+principal metabolite, benzoylecgonine, in either batch.
+We suspect that the original detection of cocaine was a result of lab contamination
+during analysis.<!--and hope this will serve as a reminder of the importance of replication.-->
+
+<!-- correlations with skeletal indicators -->
 We explored the relationship between detected compounds and various skeletal
@@
 and the concentration of nicotine and cotinine, suggesting that our ability
-to detect tobacco consumption in dental calculus does not necessarily rely
+to detect tobacco consumption in dental calculus does not necessarily rely on
 targeting frequent smokers; here, we consider individuals with multiple pipe notches
 as likely to have been heavy smokers.  
-There does seem to be a positive correlation between the weight of the calculus
-sample and recovered quantities of the compounds, so weight of the sample may affect
-the ability to detect compounds; although, we were able to detect
-compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].
-
-[[ Correlation between caffeine, nicotine, and sinusitis - Maia ]]
-
-Authentication
-
-<!-- authentication -->
-The list of targeted compounds included both naturally occurring compounds known
-to have been used in the past, as well as synthetic modern drugs did not exist
-at the time (e.g. Fentanyl, MDMA, Amphetamine). These were part of the toxicology
-screening for the original method [@sorensenDrugsCalculus2021], originally
-developed on cadavers. In our study they serve as an authentication step, as their
-presence in archaeological samples could only be the result of contamination.
-<!-- the importance of replication - brief discussion of cocaine -->
-The only modern compound (at least in a European context) detected in the sample
-was cocaine, which was detected in the first batch of sampels. The population
-analysed is from the early-mid 18th century, and cocaine was isolated in 1860 by
-Albert Niemann, and entered popular medical
-practice in 1884. Coca arrived in Europe as early as 1771, but as botanical
-specimens rather than for consumption, and there were also issues importing enough
-viable specimens of coca for cocaine extraction
-[@mortimerHistoryCoca1901 p. 179; @abucaCocaTrade2019, p.108].
-We considered it possible that it would be present in a sample with
-most individuals originating from the early- to mid-19th century. If corroborated,
-this would have been the first case of coca-consumption in Europe. Given that cocaine
-had yet to be isolated, the only way it could have gotten into the calculus was
-through consumption of the coca leaf. In our replication batch, we included all
-of the individuals who had been cocaine-positive in the first batch. We were
-unable to replicate any of the cocaine results, and we were unable to detect the
-principal metabolite, benzoylecgonine, in either batch; although, the original
-study only detected trace amounts of benzoylecgonine, and only in some of the
-individuals that were cocaine-positive. We hope this will serve as
-a reminder of the importance of replication. We suspect that the original detection
-of cocaine was a result lab contamination during the UHPLC-MS/MS analysis.
-
-Accuracy
-
-We used the presence of a pipe notch and concurrent detection of nicotine
-and/or cotinine was as a crude estimate of the accuracy of the method. Nicotine
-has been widely studied in the archaeological record due to its apparent stability
-and ability to survive over long periods of time
-[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].
-The accuracy was found to be around
-`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
-We acknowledge that this is
-a very rough estimate, as the presence of a pipe notch is likely not a perfect
-indicator of whether or not someone consumed tobacco, and it may not be
-representative of other compounds with different rates of degradation.
+<!-- inter-individual variability in calculus formation and degradation depending
+on microbiome, diet, etc. -->
+We found a positive correlation between CMS and nicotine, which may be
+indicative of the impact tobacco smoking had on the respiratory health of the
+Beemster inhabitants.
+Tobacco smoke may play a significant role in diseases
+of the upper respiratory tract, including CMS [@rehImpactTobacco2012]. Although
+the mechanisms by which smoking increases the risk of infections is not fully
+understood, solid evidence has been presented linking tobacco smoke to
+increased mucosal permeability and impairment of mucociliary clearance
+[@arcaviCigaretteSmoking2004]. Such changes, together with an altered immunologic
+response, are thought to predispose to the development of CMS
+[@slavinDiagnosisManagement2005].  
+We also observed a moderate positive correlation between CMS and caffeine which
+contradicts previous research linking chronic coffee consumption with a positive
+effect on the respiratory system, suggesting a preventive association between
+caffeine intake and pneumonia
+[e.g. @alfaroChronicCoffee2018; @kondoAssociationCoffee2021].
+However, while the
+upper respiratory tract seems to benefit from chronic coffee consumption, it is
+possible that elevated caffeine intake impacts mucosal moisture due to its
+dehydrating effect [@maughanCaffeineIngestion2003], thereby exposing individuals
+to greater risk of sinus infection.
 
@@
 <!-- secondary findings-->
-The choice of tooth for sampling may impact the detection of certain compounds,
+
+The detection of nicotine in dental calculus has previously been presented by
+Eerkens and colleagues [-@eerkensDentalCalculus2018] in two individuals from
+pre-contact California. They also targeted caffeine, cotinine, and theophylline
+in their samples, but were unable to detect any of them. It remains to be seen
+whether this is due to differences in methods used, or due to our samples being
+more recent.
+They also suggest that the choice of tooth for sampling may impact the detection
+of certain compounds,
 as the incorporation in dental calculus may depend on the mode of consumption.
@@
 nicotine present in calculus on incisors, whereas tobacco chewers may have more
-on molars [@eerkensDentalCalculus2018]. This has yet to be systematically
-tested in archaeological dental calculus. Direct incorporation by ingesting
-items, food or otherwise, containing the compounds
-is not the only potential pathway to incorporation in the dental calculus. 
-The presence of cotinine, the primary
-metabolite of nicotine, suggests that saliva is also a source of deposition, and
-that deposition of alkaloids in dental calculus can occur both on the way into the body,
-i.e. during consumption, and on the way out, i.e. disposal of waste products via
-saliva secretion into the mouth. Especially mucin-rich saliva from the sublingual
+on molars [@eerkensDentalCalculus2018].
+However, sampling may not be limited to mode of consumption.
+The presence of cotinine suggests that the excretion of a compound after being
+metabolised in the body is also a source of deposition, and
+that deposition of alkaloids in dental calculus can occur both on the way into
+the body, i.e. during consumption, and on the way out, i.e. disposal of waste
+products via saliva secretion into the mouth.
+Especially mucin-rich saliva from the sublingual
 and submandibular glands preferentially binds toxins [@doddsHealthBenefits2005],
 and since these glands are located closest to the lower incisors, they may be the
-most effective target for these studies. Due to homogenisation of samples in our
-study, we were unable to test this effect.
+most effective target for these studies.
+This has yet to be systematically
+tested in archaeological dental calculus.
+Because we homogenised samples from multiple teeth of an indiviual, we were unable
+to test the effect of oral biogeography.
+It is also possible that resident microflora within
+biofilms contribute to alkaloid breakdown and that the presence of caffeine and
+nicotine metabolites following direct ingestion can be explained by this pathway.
+However, the literature on
+biofilm biodegradation of alkaloids is limited, and *in vitro* studies have only
+found minimal contributions by certain oral bacteria in isolation
+[@sunMetabolomicsEvaluation2016; @cogoVitroEvaluation2008]; it is possible that
+a larger role is played by oral bacteria within larger, more metabolically active
+communities, e.g. biofilms [@takahashiOralMicrobiome2015].
 
 <!-- How this engineering will facilitate future research -->
-<!-- both calculus and teeth preserve well-->
 
@@
 <!-- limitations of the platform show which questions would be unanswered -->
-Limitations
-
 Since we targeted individuals with moderate-to-large calculus deposits,
@@
 a more representative sample (22.9%) [@lemmersMiddenbeemster2013].  
-The bias may also affect the detection of compounds, as caffeinated drinks are often
-acidic, and may cause a lower calculus formation<!-- citation?-->.
-We only found a weak negative correlation (`r signif(corr_tib[15,2], 3)`)
-between caffeine concentration and dental calculus index, but since we targeted individuals with
-calculus, this may inadvertently have been controlled for during sample collection.
-
+We used the presence/absence of a pipe notch and concurrent detection of tobacco
+as a crude estimate of the accuracy of the method, which we found to be around
+`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
+This is a very rough estimate, as the presence of a pipe notch is likely not a
+perfect indicator of whether or not someone consumed tobacco.
+Dental calculus is also
+more transient than for example bone, as it can be mechanically removed,
+intentionally or unintentionally, during life<!--cite-->. This would remove all
+trace of the alkaloids consumed prior to its removal, and provide a blank slate
+for newly formed calculus.  
 Quantitation of the detected compounds may have limited value in archaeological
-samples, given the effects of taphonomy added on top of individual variability
-and relative stochasticity of dental calculus formation, and the metabolic
-degradation of compounds during life [@sorensenDrugsCalculus2021; @velskoAncientDental2022a]. <!-- the results of comparisons with concentration
-had limited outcomes, and the correlations between concentration and other variables
-were similar to the correlations with presence/absence, suggesting that the actual
-concentration was somewhat random.-->
-
+samples, and will greatly affect correlations related to concentration.
+The detected quantity of a compound will depend on the quantity in dental
+calculus during life, which is largely dependent on quantity of consumption,
+how often the
+calculus was disrupted/removed, and inter-individual factors such as calculus
+formation, and metabolic breakdown of the compound
+[@mackiePreservationMetaproteome2017; @velskoAncientDental2022a].
+Following burial, compound stability over
+time will play a large role, as will microbial degradation of compounds by bacteria
+and funghi in soil [@liuNicotinedegradingMicroorganisms2015], as well as
+the soil environment, such as temperature, pH, and oxygen availability
+[@lindholstLongTerm2010].  
 Perhaps the main limitation, which is also true for many other analyses on
@@
 
+Future studies should explore how sampling from various types of teeth
+and their position in the mouth effects the probability of a compound becoming
+entrapped in dental calculus.
+This may also be related to properties within the oral cavity, as well as chemical
+properties of the compounds, which facilitate or reduce the incorporation-potential,
+and which incorporation pathways are more likely for a given compound.  
+We only targeted drugs that were included in the forensic toxicological
+screenings, and therefore only covered a limited number of the potential compounds
+that could be of interest for exploring past diets and medicinal treatments. The
+list of targeted compounds can be expanded as we discover more potential targets
+based on which specific compounds/metabolites are more likely to be incorporated
+and preserved in dental calculus.  
+We found a strong correlation between nicotine and cotinine, but
+a lesser correlation between caffeine and theophylline. This is likely because
+caffeine and theophylline have multiple plausible sources in our population, but
+more research is needed on the relationship between parent compounds and their
+metabolites in dental calculus.  
+There is an increasing interest in using oral fluid as a means of detecting alkaloids
+in living
+individuals due to the non-invasive nature of the testing compared to blood
+and urine sampling [@coneSalivaTesting1993; @valenDetermination212017].
+These *in vivo* studies are a valuable
+source of method validation and can help determine the feasibility of detecting
+certain alkaloids in oral fluid and, subsequently, dental calculus. Archaeologists,
+though, will likely be responsible for exploring dental calculus specific
+incorporation and retention of alkaloids, as well as their long-term preservation
+in the burial environment. Here, *in vitro* studies
+may provide valuable insights<!--shameless self-citation?-->.
+
 <!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
-
 While a major limitation is the uncertainty surrounding whether or not a compound
@@
 notches are a useful way to identify tobacco consumption, pipe smoking was not the
-only mode of tobacco consumption, which includes chewing, drinking, cigars, snuff,
-[@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
-was mostly common in males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018].
-Our method does not, at least to our knowledge, discriminate between modes of
-consumption,
-and may be more suitable for exploring tobacco consumption in the entire population,
-not just males. In our sample, we detected nicotine and cotinine (replicated)
-in an old adult, probable female individual. Because the individual was an old
-adult, it is unlikely that the compounds entered the dental calculus through
-pipe-smoking, as there would be pipe notches. More likely that the tobacco entered
-through alternative consumption or secondhand smoke.
-
-Future studies should explore the effect of tooth position on the detection of
-compounds.
-Also on the properties increasing incorporation potential of different compounds,
-and which pathways facilitate maximum incorporation (direct or indirect pathways?),
-which likely differs between alkaloids. How is the preservation once trapped
-inside the calculus, for example using artificial ageing on real or artificial
-dental calculus?
-In this study, we only targeted drugs that were included in the forensic toxicological
-screenings, and therefore only covered a limited number of the potential compounds
-that could be of interest for exploring past diets and medicinal treatments.
-
-<!--final remarks-->
-
-This preliminary study outlines the benefits of using calculus to target a
-variety of compounds that could have been consumed as medicine or diet. 
-This method allows us to directly address specific individuals, which can be
-especially useful in individuals that are not always well-documented in historic
-documentation, such as lower status, children and women.
-We also
-show that there are many limitations that will need to be addressed going forward
-with this type of analysis, and stress the need for more systematic research on
-the consumption of alkaloid-containing items and their subsequent concentration
-and preservation in dental calculus,
-in addition to how mode of consumption may affect concentrations on different parts
-of the dentition. <!-- also individual variation -->
-Another limitation of dental calculus as a medium is the variability its formation
-and the many factors that can influence incorporation and retention of molecules
-and particles; however, in the absence
-of hair and serum (quite uncommon in archaeology), dental calculus represents
-an impressive reservoir of information regarding the consumption of various
-alkaloids, whether dietary, medicinal, recreational, or otherwise.
+only mode of tobacco consumption, with others including chewing, drinking, cigars,
+and snuff [@goodmanTobaccoHistory1994, p.67]. Pipe-smoking
+was mainly practised by males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018],
+so chromatographic methods, such as the one presented here,
+are suitable for exploring tobacco consumption in an entire population, rather
+than a trivial subset of past populations.<!--admittedly, this study did just that...-->

### 2023-01-04 inst/paper/_intro.qmd
add: revisions based on co-authors feedback

@@
 <!-- intro to calculus and residues -->
-Dental calculus has proven to be excellent source of a wide variety of information
-about our past.
+Dental calculus has proven to be an excellent source of a wide variety of information
+about our past. The increased accessibility and advancement of methods such as
+aDNA, paleoproteomics, and chromatography, has expanded our ability to identify
+biomarkers of diet and disease on an increasingly large scale
+[@warinnerEvidenceMilk2014; @velskoDentalCalculus2017; @gismondiMultidisciplinaryApproach2020].
 
-the increased availability of methods such as aDNA, paleoproteomics, and
-chromatography, has expanded our ability to identify specific pathogens, dietary,
-and possible medicinal compounds<!--cite-->.
+<!-- define alkaloids -->
+One such biomarker is alkaloids, a plant-derived group of compounds. Many alkaloids
+have important medicinal and psychoactive effects in humans, and their direct
+detection, or detection of their metabolites, is of great interest to archaeologists.
+Previous studies have successfully recovered alkaloids archaeological
+contexts including ceramics [@smithDetectionOpium2018], pipes
+[@raffertyCurrentResearch2012], human hair
+[@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013], and even
+dental calculus employing both targeted [@eerkensDentalCalculus2018] and
+untargeted approaches
+[@buckleyDentalCalculus2014; @gismondiMultidisciplinaryApproach2020].
+Especially nicotine, the principal alkaloid in tobacco leaves,
+has been widely studied in the archaeological record due to its apparent stability
+and ability to survive over long periods of time
+[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].
 
 <!--how are compounds incorporated in dental calculus-->
-Increased interest in using oral fluid<!--more accurate terminology than saliva?--> as a means of detecting alkaloids in living
-individuals, due to the non-invasive nature of the testing compared to blood
-and urine sampling<!--citations-->.
-
-Alkaloids enter the dental calculus via two routes: (1) direct
+Alkaloids may enter the oral cavity via two pathways: (1) direct
 incorporation through oral consumption of alkaloid-containing plants, whether
 deliberate or accidental;
-and (2) passive diffusion through the
-saliva as toxins are transferred from serum to saliva [@coneInterpretationOral2007],
+and (2) passive diffusion as toxins are transferred from serum to saliva
+[@coneInterpretationOral2007],
 and then into the oral cavity through the salivary glands. The relation to serum
-is why there is often a close relation between oral fluid and blood concentrations
-of drugs [@coneInterpretationOral2007; @sorensenDrugsCalculus2021]
-The second route allows the
-detection of compounds that not only are consumed orally, but also any other pathway.
-We also expect to detect metabolites through the second pathway.
+is why there is often a close correlation between presence (not concentration)
+of drugs in oral fluid and blood/serum
+[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
+The second pathway allows the
+detection of compounds that not only are consumed orally, but also any other mode
+of consumption or administration<!--cite e.g. injection or topical-->.
+<!-- Additionally, alkaloids will spend shorter time in mouth during direct consumption than
+during excretion through saliva, so we expect the second pathway to be more
+feasible for detection. -->
 
 Many of the components involved in the formation and growth of dental calculus
-originate from saliva. Therefore the transfer of salts and other
-compounds between saliva and oral biofilms on the tooth surface is likely to also
-allow various alkaloids of dietary and medicinal origin to become entrapped in
-dental plaque, eventually mineralising and becoming dental calculus.
-
-Previous studies have been conducted using various chromatographic methods on
-archaeological dental calculus, employing both targeted
-[@eerkensDentalCalculus2018] and untargeted approaches
-[@buckleyDentalCalculus2014; @gismondiMultidisciplinaryApproach2020].
+originate from oral fluid. Proteins, bacteria, salts and other compounds are
+transferred from saliva to biofilms on the tooth surface
+[@whiteDentalCalculus1997; @jinSupragingivalCalculus2002].
+This may also allow various alkaloids of dietary and medicinal origin to become
+incorporated in dental plaque. Dental plaque undergoes frequent mineralisation
+events, ultimately causing the entrapped alkaloids and their metabolites to become
+preserved within the dental calculus. Barring intentional or accidental removal of
+the calculus during life,
+burial, excavation, and post-excavation cleaning, the alkaloids can then be detected
+by various methods to show a record of consumption during life.
 
 <!--brief description of method validation-->
-In the method validation study, all compounds detected in the blood were also
-detected in the dental calculus,
+The method we use in this study was developed in a previous study on dental
+calculus from cadavers and validated by comparing the results to compounds detected
+in the blood of the same individuals.
+All compounds that were detected in the blood were also detected in dental calculus,
 with additional compounds present in dental calculus that were not present
@@
 consumption over a long time <!-- rephrase-->[@sorensenDrugsCalculus2021].
-
 <!--summary of results-->
-We were able to detect both parent compounds, such as caffeine and nicotine, as
-well as their metabolites theophylline and cotinine. These indicate the consumption
-of tea and coffee and smoking of tobacco, which is also confirmed by historic
-documentation and identification of pipe notches in the dentition.
+We were able to detect both parent compounds and metabolites, including caffeine,
+nicotine, theophylline, and cotinine, in the dental calculus of individuals from
+a 19th century Dutch population from Middenbeemster. By detecting these compounds
+we are able to show the consumption
+of tea and coffee and smoking of tobacco on an individual scale, which is also
+confirmed by historic documentation and identification of pipe notches in the
+dentition.

### 2023-01-04 inst/paper/_materials.qmd
add: revisions based on co-authors feedback

@@
 We selected males due to a higher
-occurrence of dental calculus deposits than females. Dental calculus was
-preferentially selected from incisors.
+occurrence of dental calculus deposits than females (unpublished observation).
+Dental calculus was preferentially selected from incisors.
 
+<!-- use plot to show original and replicated sample demographics? -->
 ```{r}
+#| label: fig-sample-demography
+#| fig-cap: "Overview of sample demography.
+Left plot is the first batch and right plot is replication batch.
+eya = early young adult (18-24); lya = late young adult (25-34); ma = middle adult (35-49); old = old adult (50+)."
+
 # create tables for demographics if there are tables to spare in the article
-demography %>%
-  count(sex) %>%
-  knitr::kable()
+first_batch <- demography %>%
+  #left_join(select(metadata, id, replicated)) %>%
+  mutate(sex = factor(sex, levels = c("f", "pf", "pm", "m"))) %>% 
+  ggplot(aes(x = age, fill = sex)) +
+    geom_bar() +
+    scale_fill_viridis_d(drop = FALSE, labels = c("Female", "Female?", "Male?", "Male")) +
+    theme_minimal() +
+    labs(y = "Count", fill = "Sex") +
+    theme(
+      panel.grid.major.x = element_blank(),
+      axis.title.x = element_blank()
+    )
 
-demography %>%
-  count(age) %>% 
-  knitr::kable()
+second_batch <- demography %>%
+  left_join(select(metadata, id, replicated)) %>%
+  filter(replicated == TRUE) %>% 
+  mutate(sex = factor(sex, levels = c("f", "pf", "pm", "m"))) %>% 
+  ggplot(aes(x = age, fill = sex)) +
+    geom_bar() +
+    scale_fill_viridis_d(drop = FALSE, labels = c("Female", "Female?", "Male?", "Male")) +
+    scale_y_continuous(limits = c(0,25)) +
+    theme_minimal() +
+    labs(fill = "Sex") +
+    theme(
+      axis.title = element_blank(),
+      axis.text.y = element_blank(),
+      axis.ticks.y = element_blank(),
+      panel.grid.major.x = element_blank()
+    )
+
+first_batch + second_batch + plot_layout(guides = "collect")
+
+grid::grid.draw(grid::textGrob("Age-at-death", x =.45, y=0.01))
+
+# demography %>% 
+#   count(sex) %>%
+#   knitr::kable()
+# 
+# demography %>%
+#   count(age) %>% 
+#   knitr::kable()
 ```

### 2023-01-07 inst/paper/_conclusions.qmd
incorporate feedback from coauthors

@@
 especially useful in individuals that are not always well-documented in historic
-documentation, such as lower status, children and women.
+documentation, such as rural communities, children and women.
 We also

### 2023-01-07 inst/paper/_discussion.qmd
incorporate feedback from coauthors

@@
 <!-- Big summary/important statement for the science -->
-In this study we were able to extract and identify multiple alkaloids from the
+In this study we were able to extract and identify multiple alkaloids and salicylic
+acid from the
 dental calculus of individuals from Middenbeemster, a 19th century Dutch
 archaeological site. We applied ultra-high-performance liquid chromatography-tandem
-mass spectrometry (UHPLC-MS/MS), using a method that was validated by co-occurrence
+mass spectrometry method using pneumatically assisted electrospray ionisation
+(UHPLC-ESI-MS/MS), a method that was validated by co-occurrence
 of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
@@
 in an old adult, probable female individual.
-It is unlikely that the compounds entered the dental calculus through
-pipe-smoking, as the individual had no visible pipe notches.
-More likely that the tobacco entered through an alternate mode of consumption,
+In this particular case it is unlikely that the compounds entered the dental
+calculus through pipe-smoking, as the individual had no visible pipe notches;
+more likely the tobacco entered through an alternate mode of consumption,
 secondhand smoke, or the aforementioned tobacco-smoke enema.
@@
 so chromatographic methods, such as the one presented here,
-are suitable for exploring tobacco consumption in an entire population, rather
+are suitable for exploring tobacco consumption in an entire society, rather
 than a trivial subset of past populations.<!--admittedly, this study did just that...-->

### 2023-01-07 inst/paper/_intro.qmd
incorporate feedback from coauthors

@@
 deliberate or accidental;
-and (2) passive diffusion as toxins are transferred from serum to saliva
-[@coneInterpretationOral2007],
-and then into the oral cavity through the salivary glands. The relation to serum
+and (2) passive diffusion as alkaloids and other compounds are transferred from
+serum to saliva, and then into the oral cavity through the salivary glands
+in the hours-to-days following consumption [@coneInterpretationOral2007]. The
+relation to serum
 is why there is often a close correlation between presence (not concentration)
@@
 detection of compounds that not only are consumed orally, but also any other mode
-of consumption or administration<!--cite e.g. injection or topical-->.
-<!-- Additionally, alkaloids will spend shorter time in mouth during direct consumption than
-during excretion through saliva, so we expect the second pathway to be more
-feasible for detection. -->
+of consumption or administration, as long as it is excreted through saliva<!--cite e.g. injection or topical The second pathway may also be more
+detectable, as the excretion of compounds through saliva may take hours to days
+whereas compounds spend less time in the oral cavity during direct consumption-->.
 
@@
 <!--brief description of method validation-->
-The method we use in this study was developed in a previous study on dental
+In this study we use a ultra-high-performance liquid chromatography-tandem mass
+spectrometry method using pneumatically assisted electrospray ionisation
+(UHPLC-ESI-MS/MS) that was developed in a previous study on dental
 calculus from cadavers and validated by comparing the results to compounds detected
@@
 in blood, suggesting that dental calculus represents a comprehensive history of
-consumption over a long time <!-- rephrase-->[@sorensenDrugsCalculus2021].
+consumption over a long period of time [@sorensenDrugsCalculus2021].
 <!--summary of results-->

### 2023-01-07 inst/paper/_materials.qmd
incorporate feedback from coauthors

@@
 `r nrow(filter(demography, sex == "pm"))` probable males,
-`r nrow(filter(demography, sex == "pf"))` probable female, and
-`r nrow(filter(demography, sex == "f"))` female.
+`r nrow(filter(demography, sex == "pf"))` probable females, and
+`r nrow(filter(demography, sex == "f"))` female [@fig-sample-demography].
 We selected males due to a higher
 occurrence of dental calculus deposits than females (unpublished observation).
-Dental calculus was preferentially selected from incisors.
 
@@
       panel.grid.major.x = element_blank(),
-      axis.title.x = element_blank()
+      axis.title.x = element_blank(),
+      plot.margin = unit(c(0,0,0.5,0), "cm")
     )
@@
 
-grid::grid.draw(grid::textGrob("Age-at-death", x =.45, y=0.01))
-
-# demography %>% 
-#   count(sex) %>%
-#   knitr::kable()
-# 
-# demography %>%
-#   count(age) %>% 
-#   knitr::kable()
+grid::grid.draw(grid::textGrob("Age-at-death", x =.45, y=0.045))
 ```

### 2023-01-07 inst/paper/_methods.qmd
incorporate feedback from coauthors

@@
 The samples were sent to the Department of Forensic Medicine at Aarhus University
-for ultra-high-performance liquid chromatography tandem mass spectrometry (UHPLC-MS/MS).
+for ultra-high-performance liquid chromatography-tandem mass spectrometry method
+using pneumatically assisted electrospray ionisation (UHPLC-ESI-MS/MS).
 
-The UHPLC-MS/MS method was validated in a separate study on cadavers received
+The UHPLC-ESI-MS/MS method was validated in a separate study on cadavers received
 for forensic autopsy and toxicological analysis. Results from dental calculus were
@@
 individuals.  
-
-Briefly, dental calculus was treated with citric acid and the dissolution extracts
-were cleaned using weak and strong polymeric cation-exchange sorbents.
-
-[[ UHPLC method description needs to be elaborated ]]
-<!--rephrase and expand-->
-
-<!-- the samples were washed in 0.5 mL MeOH for 10 seconds, and centrifuged
-at 10,000 $\times$ g for 1 minute. This process was repeated. Supernatants were
-mixed with 10 mL dimethyl sulphoxide (DMSO) and 50 mL SIL-IS solution in a conical
-glass centrifuge tube. The solvent was evaporated under a stream of nitrogen at
-40&deg;C until ca. 10 mL residue remained, which was vortexed with 50 ml 30% (v/v)
-MeOH. The calculus residue was air-dried in the dark for 24 h and weighed. The
-washing/drying procedure was repeated twice using 1 mL MeOH, resulting in three
-wash extracts for analysis. -->
 In the original method, samples were washed three times to remove residual
@@
 handling.  
-<!-- 
+The list of targeted compounds included both naturally occurring compounds known
+to have been used in the past, as well as synthetic modern drugs that did not exist
+at the time (e.g. Fentanyl, MDMA, Amphetamine). These were part of the toxicology
+screening for the original method [@sorensenDrugsCalculus2021],
+developed on cadavers. In our study they serve as an authentication step, as their
+presence in archaeological samples could only be the result of contamination.
+
+Briefly, dental calculus was treated with citric acid and the dissolution extracts
+were cleaned using weak and strong polymeric cation-exchange sorbents. Samples
+were washed with 0.5 mL MeOH for 10 seconds. Samples were weighed before and after
+each wash. Samples were air-dried for 24 hours at room temperature after each wash.
+Extracts from each wash were analysed by injecting 5 mL volume into the column
+of an Exion UHPLC system that consisted of two Exion AD pumps, an Exion AD
+multiplate autosampler set at 10 $\pm$ 2 &deg;C and an Exion AC column oven set
+at 40 $\pm$ 2 &deg;C (Sciex, Ontario, Canada).
+Separation was performed using a Raptor Biphenyl UHPLC column (2.7 mm, 2.1 mm I.D.
+$\times$100 mm) (Restek, Bellefonte, PA). The mass spectrometer was a Sciex QTRAP
+6500+ with a TurboIonSpray probe for electrospray ionisation.  
 The remaining calculus was dissolved using lysing tube beads in 800 $\mu$L of 0.5
-$\small{M}$ citric acid (CA) and 50 $\mu$L SIL-IS solution. During dissolution the
-suspension was kept at ambient temperature for 1 h under continuous gentle shaking,
-and subsequently mixed with 800 $\mu$L MeOH and centrifuged at 10,000 $\times$ g for
-5 mins.
--->
-<!--
-2.4.3. Clean-up by WCX
-The dissolution extract was mixed with 6300 m L of water,
-100 m L of a 25% ammonia solution to obtain a pH of 6–7 and 50 m L
-of the SIL-IS working solution. The solution was passed through a
-WCX SPE cartridge with a maximum flow rate of 0.5 mL/min
-(mainly using gravitational force (GF)). The cartridge was pre-
-conditioned with 1 mL of MeOH followed by 1 mL of water. The
-passed extract was collected in a glass tube for clean-up by SCX.
-The cartridge was washed with 1 mL of water, then full vacuum
-( 70 kPa) was applied for 5 s to remove the bulk volume of water
-with a minimum of air flow passing through the sorbent. The wash
-water was collected in the same tube. The substances were eluted
-from the column using a 2-mL volume of freshly prepared 2% FA in
-MeOH. A 10- m L volume of DMSO was added to the eluate, and then
-the solvent was evaporated to a residual volume of approximately
-10 m L under a stream of nitrogen at 40  C. The residue was vortexed
-with a 50- m L volume of 30% MeOH, and the solution was
-transferred to an autosampler vial. This solution was used for
-the analysis of most substances.
-
-2.4.4. Clean-up by SCX
-The sample fraction saved from WCX clean-up was mixed with
-100 m L of 12 M HCl and passed through an SCX SPE cartridge that
-was pre-conditioned with 1 mL of MeOH followed by 1 mL of
-water. The cartridge was washed with 1 mL of water, then full
-vacuum was applied for 5 s before it was eluted with 2 mL of 5%
-ammonia in MeOH. The eluate was mixed with 10 m L of DMSO
-and evaporated to a residual volume of approximately 10 m L
-under a stream of nitrogen at 40  C. The residue was vortexed
-with a 50- m L volume of 30% MeOH, and the solution was
-transferred to an autosampler vial. This solution was used for the
-analysis of morphine-3-glu, morphine-6-glu, pregabalin, gaba-
-pentin and salicylic acid.
-
-2.4.5. UHPLC
-A 5- m L volume was injected onto the column running 99%
-mobile phase A. The column flow rate was 400 m L/min, and the
-column temperature was maintained at 40  2  C. The mobile
-phase was changed through a linear gradient to 0% A and 100% B
-over 5 min. Nine minutes after the initial injection, the gradient
-was returned to 99% A over 0.1 min, and the column was
-equilibrated for 2.9 min before the next injection, resulting in a
-total run time of 12 min. The same column and gradient
-programme were used for the wash, WCX and SCX extracts. The
-eluent was diverted to waste during the time intervals of 0–1.5 and
-8–12 min after injection using a post-column switch.
-
-2.4.6. MS-MS
-The probe temperature (TEM) was set to 400  C. The curtain
-gas (CUR), ion source gas 1 (GS1), ion source gas 2 (GS2) and
-collision gas (CAD) were set at 20, 80, 80 and 8 psi, respectively.
-The mass spectrometer was operated in both positive ion mode
-and negative ion mode at unit mass resolution using polarity
-switching to obtain the maximum sensitivity for both analytes
-that favour positive ionisation and analytes that favour negative
-ionisation. The ion spray voltages were set at 4.5 kV. Scheduled
-multiple reaction monitoring (MRM) with windows of 30 s was
-applied, and at least 12 data points were obtained across each
-peak. The applied MRM conditions are listed in Supplementary
-Table S2. Nitrogen was used as the CAD gas. Data acquisition and
-processing were performed using Analyst 1.7 and MultiQuant
-3.0.3 (Sciex), respectively. -->
-
+$\small{M}$ citric acid (CA) and 50 $\mu$L stable isotope-labelled analogue used
+as internal standards (SIL-IS) solution for 1 h at ambient
+temperature with gentle shaking. The suspension was then mixed with 800 $\mu$L
+MeOH and centrifuged at 10,000 $\times$ g for 5 mins, and analysed by the same
+method as the wash extracts.
+Data analysis was performed using Analyst 1.7 and MultiQuant 3.0.3 (Sciex).
+Raw quantities of compounds are presented in ng and concentrations as ng / mg.  
 For more details about the method and validation, see the original study by
@@
 replicated using leftover calculus from the original DNA study.
-in <!-- the same lab? different lab? -->.
+<!--in the same lab? different lab? -->
 
@@
 Osteoarchaeology at Leiden University. Sex was estimated using cranial
-and pelvic morphological traits [@Standards1994]. Age was estimated using dental wear,
+and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
+dental wear,
 auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
-[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985].
-<!-- differential diagnoses -->
+[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985], and divided into the following categories:
+early young adult (EYA, 18-24), late young adult (LYA, 25-34), middle adult (MA, 
+35-49), old adult (Old, 50+).
 
@@
 
-Various pathological conditions were included in the analysis and were
+Pathological conditions and lesions that occur frequently in the population were
+included in the analysis. Data were
 dichotomised to presence/absence to allow statistical analysis.
-Osteoarthritis (OA) was considered present in cases with eburnation.
-No distinction was made between active or <!-- ? --> lesions in cribra femora
-and cribra orbitalia.
-Data on chronic maxillary sinusitis were obtained from another study
-[@casnaUrbanizationRespiratory2021].
-
-[[ Description of method for sinusitis - Maia ]]
-<!-- Description of lesions included in path conditions - Supp Material?
-
-- OA = osteoarthritis
-- IVDD = intervertebral disc disease
-- TB = tuberculosis
-- DISH = diffuse idiopathic skeletal hyperostosis
-- VOP = vertebral osteophytosis
-- SN = schmorl's nodes
-- DDD = degenerative disc disease
-- PNBF = periosteal new bone formation
-- OD = osteochondritis dissecans
-- CF = cribra femora
-- CO = cribra orbitalia
-
-Sinusitis as well
--->
+Osteoarthritis (OA) was considered present in cases where eburnation was visible
+on one or more joint surface.
+Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
+formation on the margin of the superior and inferior surfaces of the vertebral
+body.
+Cribra orbitalia (CO) was diagnosed based on presence of pitting on the superior
+surface of the orbit. No distinction was made between active or <!-- ? --> lesions.
+Degenerative disc disease (DDD), or spondylosis, is identified as a large diffuse
+depression of the
+superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
+Schmorl's nodes (SN) are identified as any cortical depressions on the surface of
+the vertebral body.
+Data on chronic maxillary sinusitis from @casnaUrbanizationRespiratory2021 were
+included in this study to assess the relationship between upper respiratory
+diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
+Lesions associated with chronic maxillary sinusitis (CMS) as defined by
+@boocockMaxillarySinusitis1995 were
+recorded for each individual and classified as "pitting", "spicule-type bone
+formation", "remodeled spicules", or "white pitted bone". CMS was scored as absent
+when the sinus presented smooth surfaces with little or no associated pitting.
 
@@
 surfaces, this was scored as "crown".
-
 Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
@@
 was not included [@greeneQuantifyingCalculus2005].
+Periodontitis was scored on a visual four-stage (0-3) scoring system according to
+distance from CEJ of each tooth to alveolar bone [@maatManualPhysical2005].
 

### 2023-01-07 inst/paper/_results.qmd
incorporate feedback from coauthors

@@
 Material.
-<!-- maybe move full list to supp. mat and only include compounds here that were detected in one of the batches. -->
 
@@
 #| label: tbl-compound-detect
-#| tbl-cap: "Target compound including whether it was detected (TRUE) or not (FALSE) in each batch, as well as the lower limit of quantitation (LLOQ) in ng."
+#| tbl-cap: "Target compound including whether it was detected (TRUE) or not (FALSE) in each batch, as well as the lower limit of quantitation (LLOQ) in ng.
+CBD = cannabidiol; CBN = cannabinol; THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid; THCVA = tetrahydrocannabivarin acid."
 uhplc_calculus_detect %>%
@@
 #| label: fig-auth-plot-batch2
-#| fig-cap: "Quantity (ng) of each compound extracted from each sample across washes 1-3 and final calculus extraction (calc).
+#| fig-cap: "(A) Number of samples in which the each compound was detected in the first and second batch.
+(B) Quantity (ng) of each compound extracted from each sample across the three washes and final calculus extraction (calc).
 Colours represent different calculus samples.
 Only results from batch 2 are shown."
+#| fig-width: 8
+#| fig-height: 7
 
-uhplc_data_long %>%
+auth_plot <- uhplc_data_long %>%
   filter(
@@
     #facet_wrap(~ compound, scales = "free_y", labeller = compound_name_repair) +
-    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names)) +
+    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
     theme_bw() +
+    scale_colour_viridis_d() +
+    scale_y_continuous(position = "right") +
+    labs(y = "Quantity (ng)") +
     theme(
-      legend.position = "none"
+      legend.position = "none",
+      axis.title.x = element_blank(),
+      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
+    )
+
+compound_detect_plot <- uhplc_calculus_long %>%
+  #remove_missing() %>%
+  filter(quant > 0) %>% 
+  ggplot(aes(x = compound, fill = batch)) +
+    #geom_jitter(width = 0.2) +
+    #geom_boxplot() +
+    geom_bar(position = "dodge") +
+    #facet_wrap(~ batch) +
+    theme_bw() +
+    scale_fill_viridis_d() +
+    scale_x_discrete(labels = compound_names) +
+    theme(
+      axis.title.x = element_blank(),
+      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
+      legend.position = "none",
+      panel.grid.major.x = element_blank()
       ) +
-    labs(y = "Quantity (ng)", x = "") +
-    scale_colour_viridis_d()
+    labs(y = "Count")
+
+compound_detect_plot + auth_plot + plot_layout(widths = c(1,1.8)) + plot_annotation(tag_levels = "A")
+
 ```
@@
 #| label: fig-detection-preservation
-uhplc_calculus_long %>%
+#| fig-cap: "(A) Distribution of extracted quantities of each compound, separated by state of preservation of the skeleton.
+(B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2."
+#| fig-height: 6
+#| fig-width: 7
+
+preservation_plot <- uhplc_calculus_long %>%
   filter(
@@
     geom_boxplot(width = 0.2) +
-    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names)) +
+    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
     theme_bw() +
-    theme(legend.position = "none") +
+    theme(
+      legend.position = "none",
+      panel.grid.major.x = element_blank(),
+      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
+      ) +
     labs(x = "Preservation", y = "Quantity (ng)") +
     scale_fill_viridis_d()
+
+weight_quant_plot <- uhplc_data_comb %>%
+  select(sample, contains("batch2")) %>% 
+  select(sample, batch2_weight, contains("calc")) %>% 
+  pivot_longer(
+    -c(sample, batch2_weight),
+    names_to = c("compound", "batch"),
+    names_pattern = "(.*)_(.*)",
+    values_to = c("conc")
+  ) %>% 
+  mutate(compound = str_remove(compound, "_calc")) %>%
+  #remove_missing() %>% 
+  filter(conc > 0) %>% 
+  ggplot(aes(x = batch2_weight, y = conc)) +
+    geom_point(col = viridisLite::viridis(1)) +
+    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
+    theme_bw() +
+    theme(axis.title.y = element_blank()) +
+    labs(y = "Quantity (ng)", x = "Weight (mg)")
+
+preservation_plot + weight_quant_plot + plot_annotation(tag_levels = "A")
 ```
@@
 #tobacco <- uhplc_calculus_demography %>%
+# should only be in males
 tobacco <- uhplc_calculus_long %>%
-  filter(compound %in% c("nicotine", "cotinine")) %>% 
-  mutate(
-    detection = if_else(quant == 0, 0, 1) # 0 = not detected; NA = not included in batch 2.
-
-    ) # 0 = not detected; NA = not included in batch 2.
+  filter(
+    compound %in% c("nicotine", "cotinine"),
+    sex == "m" | sex == "pm"
+    ) %>% 
+  mutate(detection = if_else(quant == 0, 0, 1)) # 0 = not detected; NA = not included in batch 2.
 
@@
 
-Additional results from dental and skeletal lesions can be found in the supplementary
-materials.
-
 ### Correlations between detected alkaloids and diseases
 
-For further statistical analyses, only the UHPLC-MS/MS results from batch 2 were
-used, as batch 1 had results that could not be replicated and may have contained
-contamination.
+For further statistical analyses, only the UHPLC-ESI-MS/MS results from batch 2
+were used, as batch 1 had multiple compounds that were not detected in batch 2
+and may have been contamination.
 
@@
 #| label: tbl-pearson
+#| tbl-cap: "Pearson correlation (*r*) on dichotomous skeletal lesions and compound concentrations from the second batch.
+Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation.
+OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid; PN = pipe notches."
+
 corr_tib <- conc_cor %>% 
@@
   filter(var != "age",
-         var != "periodont_status") %>% 
+         var != "periodont_status") #%>% 
+  #mutate(across(everything(), signif, 3))
+
+corr_tib_long <- corr_tib %>% 
+  pivot_longer(
+    -var,
+    names_to = "name",
+    values_to = "corr"
+  )
+# still need to remove everything under the diagonal (colour white?)
+#corr_tib[corr_tib == 1] <- kableExtra::cell_spec(corr_tib[corr_tib == 1], color = "transparent")
+corr_tib_out <- corr_tib %>%
   column_to_rownames("var") %>%
-  mutate(across(everything(), signif, 3))
-corr_tib %>% 
-  knitr::kable(caption = "Pearson correlation on dichotomous skeletal lesions and compound concentrations from batch 2.
-Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation.")
+  as.matrix() %>% 
+  signif(digits = 2)
+cols <- 0
+for(i in 7:nrow(corr_tib)){ # terrible solution for formatting table
+  cols <- cols + 1 
+  corr_tib_out[i,1:cols] <- ""
+} 
+corr_names <- c("Caries", "Nicotine", "SA", "Calculus", "PN", "Theophylline", "Caffeine", "Cotinine")
+row.names(corr_tib_out) <- c(row.names(corr_tib_out)[1:6], corr_names)
+corr_tib_out <- corr_tib_out[-nrow(corr_tib),] # remove last row because it has no values (all duplicate coefficients)
+corr_tib_out %>% 
+  knitr::kable(col.names = corr_names)
 ```
 
-Point-biserial (Pearson) correlation on caries and calculus versus
-presence/absence variables. We found only weak correlations between the calculus
-index and caries ratio and other variables.
+The strongest point-biserial (Pearson) correlation correlations were a near-perfect
+positive correlation between `r cor_statement(corr_tib_long, strength = "strong")`,
+and moderate correlations between
+`r cor_statement(corr_tib_long, strength = "moderate")` (@tbl-pearson).
 
@@
 Strong correlations were found between `r cor_statement(polycorr_tib, strength = "strong")`.
-Moderate correlations were found between `r cor_statement(polycorr_tib, strength = "moderate")`. Remaining
-correlations were weak or absent (@fig-polycorr).
+Moderate correlations were found between `r cor_statement(polycorr_tib, strength = "moderate")`.
+Remaining correlations were weak or absent (@fig-polycorr).
 Correlations with age will be depressed because age was largely controlled for
@@
 #| label: fig-polycorr
-#| fig-cap: "Plot of the polychoric correlations. Larger circles and stronger calculus indicate stronger correlation coefficient."
+#| fig-cap: "Plot of the polychoric correlations (*rho*). Larger circles and increased opacity indicates a stronger correlation coefficient."
 polycorr$rho %>%
@@
     method = "circle", type = "lower", 
-    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::plasma(1, begin = 1))
+    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::plasma(1, begin = 1)),
+    legend.title = expression(rho)
     ) +
-  scale_x_discrete(labels = compound_names) +
-  scale_y_discrete(position = "right") +
+  scale_x_discrete(labels = prettified_names) +
+  scale_y_discrete(position = "right", labels = prettified_names) +
   theme(
     legend.position = "left",
+    legend.title = element_text(face = "italic")
   )

### 2023-01-07 inst/paper/paper.qmd
incorporate feedback from coauthors

@@
----
-title: "Title Goes Here"
-author:
-  - Bjørn Peare Bartholdy:
-      correspondence: "yes"
-      email: seriousname@email.org
-      orcid: 0000-0003-1689-0557
-      institute:
-        - fosg
-  - Jørgen B. Hasselstrøm
-  - Lambert K. Sørensen
-  - Maia Casna
-  - Historisch Genootschap Beemster
-  - Amanda G. Henry
-institute:
-  - fosg:
-      name: ZZ9 Plural Z Alpha
-      address: 23 Science Street, Eureka, Mississippi, USA
-  - fop: Federation of Planets
-title-block-published: "Last updated"  
-#date: now
-#date-format: long
-format: 
-  html: 
-    theme: cosmo
-execute: 
-  freeze: true
-knitr: 
-  opts_chunk:
-    echo: false
-    warning: false
-    message: false
-    comment: "#>"
-    fig-path: "../figures/"
-    fig-dpi: 600
-filters:
-  - ../templates/scholarly-metadata.lua
-  - ../templates/author-info-blocks.lua
-  #- ../templates/pagebreak.lua
-bibliography: references.bib
-csl: "apa.csl" # Insert path for the bib-style
-abstract: |
-  This is an abstract.
-keywords: |
-  stuff; and; things
-highlights: |
-  These are the highlights. 
----
-
 ```{r}
@@
 
-<!-- The actual document text starts here: -->
-
 ## Introduction
@@
 
-## Conclusion
+## Conclusions
+
+{{< include _conclusions.qmd >}}
 
@@
 
-Cats and dogs
+We wish to thank Kirsten Ziesemer for helping track down the calculus samples
+from her studies. Additionally, we owe special thanks to Vincent Falger and Kees
+de Groot from the Middenbeemster Historical Society for their input on early
+draft manuscripts and a lovely guided tour of Middenbeemster.
+
+## Data Availability Statement
+
+All raw data, analysis scripts, and the source code for the manuscript and
+supplementary materials are available as a research compendium<!--Zenodo link-->
+using the structure defined in the **rrtools** R package [@Rrrtools].
 

### 2023-01-07 R/utils.R
add: extend cor_statement and term abbreviations

@@ -216,10 +216,27 @@ dentition_checker <- function(.data){
 #' @param strength character. Strength of correlation ("weak", "moderate", or "strong").
 #' @export
 cor_statement <- function(.data, strength){
-  if(nrow(filter(.data, strength == strength)) == 0){
+
+  corr_tib <- .data %>%
+    #as_tibble(rownames = "var") %>%
+    #pivot_longer(-var, values_to = "corr") %>%
+    mutate(
+      strength = case_when(
+        abs(corr) >= 0.8 ~ "strong",
+        abs(corr) < 0.8 & abs(corr) >= 0.4 ~ "moderate",
+        abs(corr) < 0.4 ~ "weak"
+      ),
+      direction = case_when(
+        corr > 0 ~ "positive",
+        corr < 0 ~ "negative"
+      )
+    ) %>%
+    filter(corr != 1)
+
+  if(nrow(filter(corr_tib, strength == strength)) == 0){
     correlations <- glue("No {strength} correlations were found")
   } else {
-    correlations <- .data %>%
+    correlations <- corr_tib %>%
       filter(strength == {{ strength }}) %>%
       #distinct(rho, .keep_all = T) %>%
       slice(seq(from = 2, to = nrow(.), by = 2)) %>%  # awkward solution to distinct not working
@@ -243,6 +260,25 @@ compound_names <- c(
   "thcva" = "THCVA",
   "thc" = "THC",
   "cbd" = "CBD"
+)
+
+prettified_names <- c(
+  "thca-a" = "THCA-A",
+  "cocaine" = "Cocaine",
+  "caffeine" = "Caffeine",
+  "theophyl" = "Theophylline",
+  "cotinine" = "Cotinine",
+  "nicotine" = "Nicotine",
+  "salicyl" = "Salicylic acid",
+  "cbn" = "CBN",
+  "thcva" = "THCVA",
+  "thc" = "THC",
+  "cbd" = "CBD",
+  "caries_ratio" = "Caries",
+  "periodont_status" = "Periodontitis",
+  "calc_index" = "Calculus",
+  "pipe_notches" = "PN",
+  "age" = "Age-at-death"
 )
 
 compound_name_abbrev <- data.frame(

### 2023-01-29 inst/paper/_discussion.qmd
add: co-author feedback

@@
 archaeological site. We applied ultra-high-performance liquid chromatography-tandem
-mass spectrometry method using pneumatically assisted electrospray ionisation
-(UHPLC-ESI-MS/MS), a method that was validated by co-occurrence
+mass spectrometry (UHPLC-MS/MS), a method that was validated by co-occurrence
 of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
@@
 <!-- compounds detected with potential sources, and those that were surprisingly missing -->
-
 <!-- nicotine and cotinine: tobacco, obviously...-->
-Nicotine and its principal metabolite, cotinine, were strongly positively correlated,
+Nicotine and its principal/main metabolite, cotinine, were strongly positively correlated,
 both in concentration and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
-The detection of nicotine and cotinine is not surprising, as tobacco-use and
-pipe-smoking in the Beemsterpolder is well-documented in the literature
+The detection of nicotine and cotinine is not surprising, as pipe-smoking in the
+Beemsterpolder is well-documented in the literature
 [@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
@@
 Caffeine and theophylline have very similar chemical structures, so we expect
-they would experience similar rates of incorporation and degradation
-degradation, allowing us to make interpretations the ratio and correlations
-between the compounds.
+they would experience similar rates of incorporation and degradation, allowing
+us to interpret the ratio and correlations between the compounds.
 Caffeine is present in coffee, tea, and cocoa beans, with concentrations slightly
@@
 and joint pains [@clarkeCannabisEvolution2013].
-The ability to detect cannabinoids in calculus may be limited by a reduced ability
+The ability to detect cannabinoids in calculus may be limited by their reduced ability
 to diffuse from serum to salivary glands due to an affinity for protein-binding,
-[@sorensenEffectAntioxidants2018; @sorensenDrugsCalculus2021],
-meaning detection would rely on oral consumption of the cannabinoids. Even then,
+[@coneInterpretationOral2007],
+meaning detection would rely on oral consumption. Even then,
 the overall instability of some cannabinoids could also affect detection
@@
 <!-- opioids: Damn... that would've been awesome -->
-Despite many of our sampled individuals living during the height of the opium era
-in the Netherlands, none of the targeted opioids
+Despite many of our sampled individuals having lived during the height of the opium era
+in the Netherlands [@machtHistoryOpium1915], none of the targeted opioids
 (morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
 were detected. 
-Sylvius de la Boe (1614–1672), a German-born physician who studied anatomy at
-Leiden University, the Netherlands, relied heavily on opium [@machtHistoryOpium1915].
 The absence of opioids could be a result of the people ascribing more to the
@@
 [@aten400JaarBeemster2012]. It was also generally
-considered a drug of the upper class [@scheltemaOpiumTrade1907], and may be
-more common in an urban center.
+considered a drug of the upper class [@scheltemaOpiumTrade1907], and may have been
+more common in urban centers.
 The absence could also be attributed to postmortem degradation.
@@
 indicators, such as pathological and dental lesions, preservation, and pipe notches.
-We found no evidence to suggest that preservation of the skeleton influences
-the recovery of compounds from the dental calculus. This could be due to
-the good overall preservation of both teeth and dental calculus, which often
-preserves better than the rest of the skeleton.  
+We found some evidence to suggest that preservation of the skeleton influences
+the recovery of compounds from the dental calculus, with well-preserved skeletons
+potentially serving as a better target for sampling.  
 We found no correlation between the number of pipe notches
@@
 as likely to have been heavy smokers.  
-<!-- inter-individual variability in calculus formation and degradation depending
-on microbiome, diet, etc. -->
 We found a positive correlation between CMS and nicotine, which may be
@@
 However, while the
-upper respiratory tract seems to benefit from chronic coffee consumption, it is
+lower respiratory tract seems to benefit from chronic coffee consumption, it is
 possible that elevated caffeine intake impacts mucosal moisture due to its
@@
 tested in archaeological dental calculus.
-Because we homogenised samples from multiple teeth of an indiviual, we were unable
+Because we homogenised samples from multiple teeth of an individual, we were unable
 to test the effect of oral biogeography.
@@
 more transient than for example bone, as it can be mechanically removed,
-intentionally or unintentionally, during life<!--cite-->. This would remove all
-trace of the alkaloids consumed prior to its removal, and provide a blank slate
-for newly formed calculus.  
+intentionally or unintentionally, during life, eliminating
+all trace of the alkaloids consumed prior to its removal.  
 Quantitation of the detected compounds may have limited value in archaeological
-samples, and will greatly affect correlations related to concentration.
-The detected quantity of a compound will depend on the quantity in dental
-calculus during life, which is largely dependent on quantity of consumption,
-how often the
-calculus was disrupted/removed, and inter-individual factors such as calculus
-formation, and metabolic breakdown of the compound
-[@mackiePreservationMetaproteome2017; @velskoAncientDental2022a].
+samples due to degradation, and will greatly affect our correlations related to
+concentration.
 Following burial, compound stability over
@@
 the soil environment, such as temperature, pH, and oxygen availability
-[@lindholstLongTerm2010].  
-Perhaps the main limitation, which is also true for many other analyses on
-archaeological materials, is that detecting the absence of a compound is not possible.
-The absence of a compound is
-not evidence of its absence, which complicates the interpretation of our results.
+[@mackiePreservationMetaproteome2017; @lindholstLongTerm2010].  
+The detected quantity of a compound will also depend on the quantity in dental
+calculus during life, which is largely controlled by quantity of consumption,
+how often the
+calculus was disrupted/removed, metabolic breakdown of the compound, and inter-
+and intra-individual factors related to stages of biofilm formation, maturation,
+and mineralisation
+[@lustmannScanningElectron1976; @velskoMicrobialDifferences2019; @zijngeBiofilmArchitecture2010].
+In short, this means it is not really possible to detect the absence of a compound.
+The absence of a compound is not evidence of absence of consumption. This complicates
+the interpretation of our results.
 We have attempted to minimise errors occurring due to this limitation by including
@@
 detection rate seen in tobacco, this remains a major limitation, and will likely
-been compounded by increasing antiquity of the samples.
+be compounded by increasing antiquity of the samples.
 The replication of samples increases the evidence of absence if the compound
@@
 Future studies should explore how sampling from various types of teeth
-and their position in the mouth effects the probability of a compound becoming
+and their position in the mouth affects the probability of a compound becoming
 entrapped in dental calculus.
@@
 incorporation and retention of alkaloids, as well as their long-term preservation
-in the burial environment. Here, *in vitro* studies
-may provide valuable insights<!--shameless self-citation?-->.
+in the burial environment.<!-- move to discussion in disseration: Here, *in vitro* studies
+may provide valuable insights-->
 
@@
 was mainly practised by males [@lemmersMiddenbeemster2013; @eerkensDentalCalculus2018],
-so chromatographic methods, such as the one presented here,
+so methods like the one presented here
 are suitable for exploring tobacco consumption in an entire society, rather

### 2023-01-29 inst/paper/_intro.qmd
add: co-author feedback

@@
 Dental calculus has proven to be an excellent source of a wide variety of information
-about our past. The increased accessibility and advancement of methods such as
-aDNA, paleoproteomics, and chromatography, has expanded our ability to identify
+about our past. The increased accessibility and advancement of methods in
+aDNA, paleoproteomics, and mass spectrometry, has expanded our ability to identify
 biomarkers of diet and disease on an increasingly large scale
@@
 detection, or detection of their metabolites, is of great interest to archaeologists.
-Previous studies have successfully recovered alkaloids archaeological
-contexts including ceramics [@smithDetectionOpium2018], pipes
+Previous studies have successfully recovered alkaloids in archaeological
+contexts, including ceramics [@smithDetectionOpium2018], pipes
 [@raffertyCurrentResearch2012], human hair
@@
 and (2) passive diffusion as alkaloids and other compounds are transferred from
-serum to saliva, and then into the oral cavity through the salivary glands
+plasma to saliva, and then into the oral cavity through the salivary glands
 in the hours-to-days following consumption [@coneInterpretationOral2007]. The
-relation to serum
+relation to plasma
 is why there is often a close correlation between presence (not concentration)
-of drugs in oral fluid and blood/serum
+of drugs in oral fluid and blood
 [@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
 The second pathway allows the
-detection of compounds that not only are consumed orally, but also any other mode
+detection of parent compounds and their metabolites that not only are consumed orally,
+but also any other mode
 of consumption or administration, as long as it is excreted through saliva<!--cite e.g. injection or topical The second pathway may also be more
@@
 In this study we use a ultra-high-performance liquid chromatography-tandem mass
-spectrometry method using pneumatically assisted electrospray ionisation
-(UHPLC-ESI-MS/MS) that was developed in a previous study on dental
+spectrometry (UHPLC-MS/MS) method that was developed in a previous study on dental
 calculus from cadavers and validated by comparing the results to compounds detected

### 2023-01-29 inst/paper/_materials.qmd
add: co-author feedback

@@
 a 19th century rural Dutch site.
+Of the `r nrow(demography)` individuals that were analysed,
+`r nrow(filter(metadata, replicated == TRUE))` individuals were subsequently
+replicated using leftover calculus from the original DNA study.
+<!--in the same lab? different lab? -->
 The village of Middenbeemster and the surrounding Beemsterpolder was established
@@
 create more farm land, mainly for the cultivation of cole seeds (de Vries 1978).
-In 1615, a decision was made to dedicate a church to Hendrick de Keyser, and
+In 1615, a decision was made to build a church, and
 construction started in 1618 (Hakvoort 2013). The excavated cemetery is associated
@@
 in the analysis, we preferentially selected males from the middle adult age category
-(35-49). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
+(35-49 years). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
 `r nrow(filter(demography, sex == "pm"))` probable males,
@@
 #| fig-cap: "Overview of sample demography.
-Left plot is the first batch and right plot is replication batch.
-eya = early young adult (18-24); lya = late young adult (25-34); ma = middle adult (35-49); old = old adult (50+)."
+Left plot is the first batch and right plot is the replication batch with 29 of the individuals from the first batch.
+eya = early young adult (18-24 years); lya = late young adult (25-34 years); ma = middle adult (35-49 years); old = old adult (50+ years).
+Male?
+= probable male; Female?
+= probable female."
 

### 2023-01-29 inst/paper/_methods.qmd
add: co-author feedback

@@
 The samples were sent to the Department of Forensic Medicine at Aarhus University
-for ultra-high-performance liquid chromatography-tandem mass spectrometry method
-using pneumatically assisted electrospray ionisation (UHPLC-ESI-MS/MS).
+for ultra-high-performance liquid chromatography-tandem mass spectrometry (UHPLC-MS/MS)
+analysis.
 
-The UHPLC-ESI-MS/MS method was validated in a separate study on cadavers received
-for forensic autopsy and toxicological analysis. Results from dental calculus were
-validated against compounds detected in whole blood samples from the same
-individuals.  
-In the original method, samples were washed three times to remove residual
-substances from the surface of the calculus that originated from oral fluids,
-and only extract substances from the calculus. In our samples the washes served
-to remove potential contaminants from the burial environment and post-excavation
-handling.  
 The list of targeted compounds included both naturally occurring compounds known
@@
 
-Briefly, dental calculus was treated with citric acid and the dissolution extracts
-were cleaned using weak and strong polymeric cation-exchange sorbents. Samples
-were washed with 0.5 mL MeOH for 10 seconds. Samples were weighed before and after
-each wash. Samples were air-dried for 24 hours at room temperature after each wash.
-Extracts from each wash were analysed by injecting 5 mL volume into the column
-of an Exion UHPLC system that consisted of two Exion AD pumps, an Exion AD
-multiplate autosampler set at 10 $\pm$ 2 &deg;C and an Exion AC column oven set
-at 40 $\pm$ 2 &deg;C (Sciex, Ontario, Canada).
-Separation was performed using a Raptor Biphenyl UHPLC column (2.7 mm, 2.1 mm I.D.
-$\times$100 mm) (Restek, Bellefonte, PA). The mass spectrometer was a Sciex QTRAP
-6500+ with a TurboIonSpray probe for electrospray ionisation.  
-The remaining calculus was dissolved using lysing tube beads in 800 $\mu$L of 0.5
-$\small{M}$ citric acid (CA) and 50 $\mu$L stable isotope-labelled analogue used
-as internal standards (SIL-IS) solution for 1 h at ambient
-temperature with gentle shaking. The suspension was then mixed with 800 $\mu$L
-MeOH and centrifuged at 10,000 $\times$ g for 5 mins, and analysed by the same
-method as the wash extracts.
-Data analysis was performed using Analyst 1.7 and MultiQuant 3.0.3 (Sciex).
-Raw quantities of compounds are presented in ng and concentrations as ng / mg.  
+Briefly, samples of dental calculus were washed three times each with one mL of
+methanol (MeOH), to remove surface contaminants. The wash solutions were collected
+separately. The solvent was
+evaporated and the residues were dissolved in 50 µL 30% MeOH.
+The washed calculus was homogenized in presence of 0.5 M citric acid using a lysing
+tube with stainless steel beads. Following one hour of incubation the dissolution
+extract was cleaned by weak and strong cation-exchange. After evaporation of the
+elution solvent the residue was dissolved in 50 µL 30% MeOH.
+The final extracts obtained from washing and dissolution of the dental calculus were
+analysed by UHPLC-MS/MS using a reversed-phase biphenyl column for chromatography.
+To obtain quantitative results, isotope dilution was applied.
 For more details about the method and validation, see the original study by
 Sørensen and colleagues [-@sorensenDrugsCalculus2021].
-
-Of the `r nrow(demography)` individuals that were analysed,
-`r nrow(filter(metadata, replicated == TRUE))` individuals were subsequently
-replicated using leftover calculus from the original DNA study.
-<!--in the same lab? different lab? -->
 
@@
 [@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985], and divided into the following categories:
-early young adult (EYA, 18-24), late young adult (LYA, 25-34), middle adult (MA, 
-35-49), old adult (Old, 50+).
+early young adult (EYA, 18-24 years), late young adult (LYA, 25-34 years), middle adult (MA, 
+35-49 years), old adult (Old, 50+ years).
 
@@
 Osteoarthritis (OA) was considered present in cases where eburnation was visible
-on one or more joint surface.
+on one or more joint surfaces.
 Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
@@
 Cribra orbitalia (CO) was diagnosed based on presence of pitting on the superior
-surface of the orbit. No distinction was made between active or <!-- ? --> lesions.
+surface of the orbit. No distinction was made between active or healing lesions.
 Degenerative disc disease (DDD), or spondylosis, is identified as a large diffuse
@@
 the vertebral body.
-Data on chronic maxillary sinusitis from @casnaUrbanizationRespiratory2021 were
+Data on chronic maxillary sinusitis (CMS) from @casnaUrbanizationRespiratory2021 were
 included in this study to assess the relationship between upper respiratory
 diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
-Lesions associated with chronic maxillary sinusitis (CMS) as defined by
+Lesions associated with CMS as defined by
 @boocockMaxillarySinusitis1995 were

### 2023-01-29 inst/paper/_results.qmd
add: co-author feedback

@@
 Multiple compounds were detected in the dental calculus samples. Compounds detected
-at a lower concentration than the LLOQ were considered not present. Not all the
-compounds detected in the first batch could be replicated in the second batch
-(@tbl-compound-detect). For a full list of targeted compounds, see Supplementary
-Material.
+at a lower concentration than the lower limit of quantitation (LLOQ) were considered
+not present. Not all the compounds detected in the first batch could be replicated
+in the second batch (@tbl-compound-detect).
+For a full list of targeted compounds, see Supplementary Material.
 
@@
 #| tbl-cap: "Target compound including whether it was detected (TRUE) or not (FALSE) in each batch, as well as the lower limit of quantitation (LLOQ) in ng.
-CBD = cannabidiol; CBN = cannabinol; THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid; THCVA = tetrahydrocannabivarin acid."
+CBD = cannabidiol; CBN = cannabinol; THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA = tetrahydrocannabivarin acid."
 uhplc_calculus_detect %>%
@@
 #| label: fig-auth-plot-batch2
-#| fig-cap: "(A) Number of samples in which the each compound was detected in the first and second batch.
-(B) Quantity (ng) of each compound extracted from each sample across the three washes and final calculus extraction (calc).
-Colours represent different calculus samples.
-Only results from batch 2 are shown."
+#| fig-cap: "(A) Number of samples in which each compound was detected in the first and second batch.
+(B) Quantity (ng) of each compound extracted from each sample in batch 2.
+The plot displays the extracted quantity across the three washes and final calculus extraction (calc).
+Each coloured line represents a different calculus sample."
 #| fig-width: 8
-#| fig-height: 7
+#| fig-height: 6
 
@@
 To see if preservation of the skeletal remains had any effect on the detection of
-compounds, concentrations of compounds were compared to the various levels of
-preservation. No apparent patterns were present between preservation and detection
-of compounds in the samples (@fig-detection-preservation).
+compounds, we compare extracted quantities of compounds to the various levels of
+skeletal preservation. Our results from batch 2 suggest that detection of a compound
+may be linked to the preservation of the skeleton, with better preservation leading
+to increased extraction quantity ([@fig-detection-preservation]A). We also
+find a weak positive correlation between the weight of the calculus sample and the
+quantity of compound extracted from the calculus ([@fig-detection-preservation]B).
 
@@
 #| label: fig-detection-preservation
-#| fig-cap: "(A) Distribution of extracted quantities of each compound, separated by state of preservation of the skeleton.
+#| fig-cap: "(A) Violin plot with overlaid box plots depicting the distribution of extracted quantities of each compound from batch 2 separated by state of preservation of the skeleton.
 (B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2."
@@
     !is.na(preservation),
-    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
+    quant > 0,
+    batch == "batch2",
+    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
     id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
@@
 
-weight_quant_plot <- uhplc_data_comb %>%
-  select(sample, contains("batch2")) %>% 
-  select(sample, batch2_weight, contains("calc")) %>% 
-  pivot_longer(
-    -c(sample, batch2_weight),
-    names_to = c("compound", "batch"),
-    names_pattern = "(.*)_(.*)",
-    values_to = c("conc")
+weight_quant_plot <- uhplc_calculus_long %>%
+  filter(
+    #!is.na(preservation),
+    batch == "batch2",
+    quant > 0,
+    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
+    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
     ) %>% 
-  mutate(compound = str_remove(compound, "_calc")) %>%
-  #remove_missing() %>% 
-  filter(conc > 0) %>% 
-  ggplot(aes(x = batch2_weight, y = conc)) +
+  ggplot(aes(x = weight, y = quant)) +
     geom_point(col = viridisLite::viridis(1)) +
@@
     labs(y = "Quantity (ng)", x = "Weight (mg)")
+
+# weight_quant_plot <- uhplc_data_comb %>%
+#   select(sample, contains("batch2")) %>% 
+#   select(sample, batch2_weight, contains("calc")) %>% 
+#   pivot_longer(
+#     -c(sample, batch2_weight),
+#     names_to = c("compound", "batch"),
+#     names_pattern = "(.*)_(.*)",
+#     values_to = c("conc")
+#   ) %>% 
+#   mutate(compound = str_remove(compound, "_calc")) %>%
+#   #remove_missing() %>% 
+#   filter(
+#     conc > 0,
+#     id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
+#     ) %>% 
+#   ggplot(aes(x = batch2_weight, y = conc)) +
+#     geom_point(col = viridisLite::viridis(1)) +
+#     facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
+#     theme_bw() +
+#     theme(axis.title.y = element_blank()) +
+#     labs(y = "Quantity (ng)", x = "Weight (mg)")
 
@@
 
-For further statistical analyses, only the UHPLC-ESI-MS/MS results from batch 2
+For further statistical analyses, only the UHPLC-MS/MS results from batch 2
 were used, as batch 1 had multiple compounds that were not detected in batch 2
-and may have been contamination.
+and may have been contaminated.
 
@@
 #| label: tbl-pearson
-#| tbl-cap: "Pearson correlation (*r*) on dichotomous skeletal lesions and compound concentrations from the second batch.
+#| tbl-cap: "Pearson correlation (*r*) on dichotomous skeletal lesions and compound concentrations (ng/mg) from the second batch.
 Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation.
@@
 
+```{r}
+prettified_names_lower <- sapply(prettified_names, str_to_lower)
+
+corr_tib_out <- corr_tib_long %>%
+  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F))
+
+polycorr_tib_out <- polycorr_tib %>%
+  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F))
+```
+
 The strongest point-biserial (Pearson) correlation correlations were a near-perfect
-positive correlation between `r cor_statement(corr_tib_long, strength = "strong")`,
+positive correlation between `r cor_statement(corr_tib_out, strength = "strong")`,
 and moderate correlations between
-`r cor_statement(corr_tib_long, strength = "moderate")` (@tbl-pearson).
+`r cor_statement(corr_tib_out, strength = "moderate")` (@tbl-pearson).
 
@@
 likely to cause spurious correlations.
-Strong correlations were found between `r cor_statement(polycorr_tib, strength = "strong")`.
-Moderate correlations were found between `r cor_statement(polycorr_tib, strength = "moderate")`.
+Strong correlations were found between `r cor_statement(polycorr_tib_out, strength = "strong")`.
+Moderate correlations were found between `r cor_statement(polycorr_tib_out, strength = "moderate")`.
 Remaining correlations were weak or absent (@fig-polycorr).

### 2023-02-07 inst/paper/_discussion.qmd
add co-author comments round 2

@@
 by extracting multiple different alkaloids from dental calculus, including
-cannabinoids, cocaine, nicotine, cotinine, caffeine, theophylline, and salicylic
-acid in multiple individuals. Of these compounds, the cannabinoids and cocaine
-could not be replicated in a second run on different samples from the same individuals.
+nicotine, cotinine, caffeine, theophylline, and salicylic
+acid in multiple individuals. The detection of these compounds was
+solidified in a replication analysis on different samples from the same
+individuals.
+Cocaine and multiple cannabinoids were also detected during the first
+analysis, but were not replicated.
 We discuss the implications of these findings in light of historical and
@@
 potentially serving as a better target for sampling.  
-We found no correlation between the number of pipe notches
-and the concentration of nicotine and cotinine, suggesting that our ability
-to detect tobacco consumption in dental calculus does not necessarily rely on
-targeting frequent smokers; here, we consider individuals with multiple pipe notches
-as likely to have been heavy smokers.  
 We found a positive correlation between CMS and nicotine, which may be
@@
 Tobacco smoke may play a significant role in diseases
-of the upper respiratory tract, including CMS [@rehImpactTobacco2012]. Although
+of the upper respiratory tract, including chronic maxillary sinusities
+[@rehImpactTobacco2012]. Although
 the mechanisms by which smoking increases the risk of infections is not fully
@@
 [@arcaviCigaretteSmoking2004]. Such changes, together with an altered immunologic
-response, are thought to predispose to the development of CMS
+response, are thought to predispose to the development of chronic maxillary sinusitis
 [@slavinDiagnosisManagement2005].  
-We also observed a moderate positive correlation between CMS and caffeine which
+We also observed a moderate positive correlation between chronic maxillary sinusitis and caffeine which
 contradicts previous research linking chronic coffee consumption with a positive
@@
 <!-- limitations of the platform show which questions would be unanswered -->
-Since we targeted individuals with moderate-to-large calculus deposits,
-it is likely a biased sample. The presence of calculus may increase risk of premature death
+Because we targeted individuals with moderate-to-large calculus deposits,
+it is likely a biased sample. The presence of calculus may increase the
+risk of premature death
 [@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
@@
 respiratory diseases, if periodontal and respiratory pathogens
-enter the blood stream
+enter the bloodstream
 [@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
-In our sample, the percentage of CMS
+In our sample, the percentage of chronic maxillary sinusitis
 (`r scales::percent(mean(explore_select$CMS, na.rm = T), accuracy = 0.1)`)
@@
 time will play a large role, as will microbial degradation of compounds by bacteria
-and funghi in soil [@liuNicotinedegradingMicroorganisms2015], as well as
+and fungi in soil [@liuNicotinedegradingMicroorganisms2015], as well as
 the soil environment, such as temperature, pH, and oxygen availability
@@
 We have attempted to minimise errors occurring due to this limitation by including
-a relatively large sample of individuals. Although given the relatively low
-detection rate seen in tobacco, this remains a major limitation, and will likely
-be compounded by increasing antiquity of the samples.
-The replication of samples increases the evidence of absence if the compound
-was absent in both batches.
+a relatively large sample of individuals and replicating our analysis. Although
+given the relatively low detection rate seen in tobacco, this remains a major
+limitation, and will likely be compounded by increasing antiquity of the samples.
 
@@
 and preserved in dental calculus.  
-We found a strong correlation between nicotine and cotinine, but
-a lesser correlation between caffeine and theophylline. This is likely because
-caffeine and theophylline have multiple plausible sources in our population, but
-more research is needed on the relationship between parent compounds and their
-metabolites in dental calculus.  
 There is an increasing interest in using oral fluid as a means of detecting alkaloids
@@
 incorporation and retention of alkaloids, as well as their long-term preservation
-in the burial environment.<!-- move to discussion in disseration: Here, *in vitro* studies
-may provide valuable insights-->
+in the burial environment.<!-- move to discussion in disseration: Here, *in vitro* studies may provide valuable insights-->
 
 <!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
+
+<!-- rephrase -->
 While a major limitation is the uncertainty surrounding whether or not a compound
-is actually absent, the power of the method is in detecting food and other items
-that are consumed in a manner that does not allow macroscopic detection. Taking
+is actually absent, the power of the method lies
+in the ability to detect dietary and other compounds that were incorporated
+via multiple
+consumption pathways that are not detected by other methods.
+Taking
 tobacco consumption as an example; while pipe
@@
 are suitable for exploring tobacco consumption in an entire society, rather
-than a trivial subset of past populations.<!--admittedly, this study did just that...-->
+than a trivial subset of past populations.<!--admittedly, our sample was just that...-->
+Combined with other methods, it can also give us a more complete picture
+of dietary patterns and medicinal/recreational plant-use in the past by capturing
+multiple possible incorporation pathways of dietary (and other) compounds.

### 2023-02-07 inst/paper/_intro.qmd
add co-author comments round 2

@@
 <!-- define alkaloids -->
-One such biomarker is alkaloids, a plant-derived group of compounds. Many alkaloids
+One such collection of biomarkers is alkaloids, a plant-derived group of compounds.
+Many alkaloids
 have important medicinal and psychoactive effects in humans, and their direct
@@
 plasma to saliva, and then into the oral cavity through the salivary glands
-in the hours-to-days following consumption [@coneInterpretationOral2007]. The
+in the hours to days following consumption [@coneInterpretationOral2007]. The
 relation to plasma
@@
 [@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
-The second pathway allows the
-detection of parent compounds and their metabolites that not only are consumed orally,
-but also any other mode
-of consumption or administration, as long as it is excreted through saliva<!--cite e.g. injection or topical The second pathway may also be more
+The second pathway allows the identification of parent compounds that are not
+consumed orally, as long as they, or their metabolites, are excreted through saliva.
+<!--cite e.g. injection or topical The second pathway may also be more
 detectable, as the excretion of compounds through saliva may take hours to days
@@
 calculus from cadavers and validated by comparing the results to compounds detected
-in the blood of the same individuals.
+in the blood of the same individuals [@sorensenDrugsCalculus2021].
 All compounds that were detected in the blood were also detected in dental calculus,

### 2023-02-07 inst/paper/_materials.qmd
add co-author comments round 2

@@
 a 19th century rural Dutch site.
-Of the `r nrow(demography)` individuals that were analysed,
-`r nrow(filter(metadata, replicated == TRUE))` individuals were subsequently
-replicated using leftover calculus from the original DNA study.
-<!--in the same lab? different lab? -->
 The village of Middenbeemster and the surrounding Beemsterpolder was established
 in the beginning of the 17th century, when the Beemster lake was drained to
-create more farm land, mainly for the cultivation of cole seeds (de Vries 1978).
+create more farmland, mainly for the cultivation of cole seeds (de Vries 1978).
 In 1615, a decision was made to build a church, and
@@
 
-To reduce the number of factors to account for
+To reduce the number of potentially confounding factors to account for
 in the analysis, we preferentially selected males from the middle adult age category
@@
 `r nrow(filter(demography, sex == "pf"))` probable females, and
-`r nrow(filter(demography, sex == "f"))` female [@fig-sample-demography].
+`r nrow(filter(demography, sex == "f"))` female (@fig-sample-demography).
 We selected males due to a higher
-occurrence of dental calculus deposits than females (unpublished observation).
+occurrence of pipe notches and dental calculus deposits than females (unpublished observation).
 

### 2023-02-07 inst/paper/_methods.qmd
add co-author comments round 2

@@
+### Skeletal analysis
+
+<!-- methods for age, sex, and stature estimation -->
+Demographic and pathological analyses were conducted in the Laboratory for Human
+Osteoarchaeology at Leiden University. Sex was estimated using cranial
+and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
+dental wear,
+auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
+[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985],
+and divided into the following categories:
+early young adult (18-24 years), late young adult (25-34 years), middle adult (
+35-49 years), old adult (50+ years).
+
+#### Paleopathology
+
+Pathological conditions and lesions that occur frequently in the population were
+included in the analysis. Data were
+dichotomised to presence/absence to allow statistical analysis.
+Osteoarthritis was considered present in cases where eburnation was visible
+on one or more joint surfaces.
+Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
+formation on the margin of the superior and inferior surfaces of the vertebral
+body.
+Cribra orbitalia was diagnosed based on the presence of pitting on the superior
+surface of the orbit. No distinction was made between active or healing lesions.
+Degenerative disc disease, or spondylosis, is identified as a large diffuse
+depression of the
+superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
+Schmorl's nodes are identified as any cortical depressions on the surface of
+the vertebral body.
+Data on chronic maxillary sinusitis from @casnaUrbanizationRespiratory2021 were
+included in this study to assess the relationship between upper respiratory
+diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
+Lesions associated with chronic maxillary sinusitis as defined by
+@boocockMaxillarySinusitis1995 were
+recorded for each individual and classified as "pitting", "spicule-type bone
+formation", "remodeled spicules", or "white pitted bone".
+chronic maxillary sinusitis was scored as absent
+when the sinus presented smooth surfaces with little or no associated pitting.
+
+#### Dental pathology
+
+Caries ratios were calculated by dividing the number of lesions by the number
+of teeth scored, resulting in a single caries ratio per individual. If the surface
+where the lesion originated is not visible, i.e. if the lesion covered multiple
+surfaces, this was scored as "crown".
+Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
+Calculus was scored with a four-stage scoring system (0-3) to score absent, slight,
+moderate, and heavy calculus deposits [@brothwellDiggingBones1981]
+on the lingual, buccal (and labial), and interproximal surfaces of each tooth. Only
+one score was used for the combined interproximal surfaces, resulting in three
+scores per tooth (when surfaces are intact), and four calculus indices per individual;
+upper anterior, upper posterior, lower anterior, lower posterior.
+Each index was calculated by dividing
+the sum of calculus scores for each surface by the total number of surfaces scored
+in each quadrant. If a tooth could not be scored on all three surfaces, the tooth
+was not included [@greeneQuantifyingCalculus2005].
+Periodontitis was scored on a visual four-stage (0-3) scoring system according to
+distance from cemento-enamel junction of each tooth to alveolar bone [@maatManualPhysical2005].
+
+### Calculus sampling
+
+Where possible, we used material that had already been sampled for a previous
+study to prevent unnecessary repeated sampling of individuals. 
+Calculus from the previous study was
+sampled in a dedicated ancient DNA laboratory at the Laboratories of Molecular
+Anthropology and Microbiome Research in Norman, Oklahoma, U.S.A,
+using established ancient DNA protocols. More details on the methods can be found
+in the published articles
+[@ziesemer16SChallenges2015; @ziesemerGenomeCalculus2018]. 
+Of the `r nrow(demography)` individuals that were originally included in our sample,
+`r nrow(filter(metadata, replicated == TRUE))`
+were replicated in a separate analysis only using calculus from the previous study.  
 New dental calculus samples were taken under sterile conditions in a positive
@@
 Eppendorf tubes.
-Where possible, previously sampled material was used to prevent repeated sampling
-of individuals. The sampling methods can be found in the published material
-[@ziesemer16SChallenges2015; @ziesemerGenomeCalculus2018]. 
-The samples were sent to the Department of Forensic Medicine at Aarhus University
+All calculus samples were sent to the Department of Forensic Medicine at Aarhus University
 for ultra-high-performance liquid chromatography-tandem mass spectrometry (UHPLC-MS/MS)
 analysis.
+
+### UHPLC-MS/MS
 
@@
 
-### Skeletal analysis
-
-<!-- methods for age, sex, and stature estimation -->
-Demographic and pathological analyses were conducted in the Laboratory for Human
-Osteoarchaeology at Leiden University. Sex was estimated using cranial
-and pelvic morphological traits [@Standards1994]. Age-at-death was estimated using
-dental wear,
-auricular and pubic surface appearance, cranial suture closure, and epiphyseal fusion 
-[@Standards1994; @SucheyBrooks1990; @buckberryAuricular2002; @lovejoyAuricular1985; @meindlSutureClosure1985], and divided into the following categories:
-early young adult (EYA, 18-24 years), late young adult (LYA, 25-34 years), middle adult (MA, 
-35-49 years), old adult (Old, 50+ years).
-
-#### Paleopathology
-
-Pathological conditions and lesions that occur frequently in the population were
-included in the analysis. Data were
-dichotomised to presence/absence to allow statistical analysis.
-Osteoarthritis (OA) was considered present in cases where eburnation was visible
-on one or more joint surfaces.
-Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
-formation on the margin of the superior and inferior surfaces of the vertebral
-body.
-Cribra orbitalia (CO) was diagnosed based on presence of pitting on the superior
-surface of the orbit. No distinction was made between active or healing lesions.
-Degenerative disc disease (DDD), or spondylosis, is identified as a large diffuse
-depression of the
-superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
-Schmorl's nodes (SN) are identified as any cortical depressions on the surface of
-the vertebral body.
-Data on chronic maxillary sinusitis (CMS) from @casnaUrbanizationRespiratory2021 were
-included in this study to assess the relationship between upper respiratory
-diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
-Lesions associated with CMS as defined by
-@boocockMaxillarySinusitis1995 were
-recorded for each individual and classified as "pitting", "spicule-type bone
-formation", "remodeled spicules", or "white pitted bone". CMS was scored as absent
-when the sinus presented smooth surfaces with little or no associated pitting.
-
-#### Dental pathology
-
-Caries ratios were calculated by dividing the number of lesions by the number
-of teeth scored, resulting in a single caries ratio per individual. If the surface
-where the lesion originated is not visible, i.e. if the lesion covered multiple
-surfaces, this was scored as "crown".
-Calculus indices were calculated according to Greene and colleagues [-@greeneQuantifyingCalculus2005].
-Calculus was scored with a four-stage scoring system (0-3) to score absent, slight,
-moderate, and heavy calculus deposits [@brothwellDiggingBones1981]
-on the lingual, buccal (labial), and interproximal surfaces of each tooth. Only
-one score was used for the combined interproximal surfaces, resulting in three
-scores per tooth (when surfaces are intact), and four calculus indices per individual;
-upper anterior (UA), upper posterior (UP), lower anterior (LA), lower posterior (LP).
-Each index was calculated by dividing
-the sum of calculus scores for each surface by the total number of surfaces scored
-in each quadrant. If a tooth could not be scored on all three surfaces, the tooth
-was not included [@greeneQuantifyingCalculus2005].
-Periodontitis was scored on a visual four-stage (0-3) scoring system according to
-distance from CEJ of each tooth to alveolar bone [@maatManualPhysical2005].
-
 ### Statistical analysis

### 2023-02-07 inst/paper/_results.qmd
add co-author comments round 2

@@
 
+The pattern we expect to see in authentic compounds representing compounds
+trapped within the dental calculus, 
+is a reduction in the quantity from wash 1 to wash 3 as potential surface contaminants
+are washed off, and then a
+spike in the final extraction when entrapped compounds are released and detected.
+
 Most plots show a large increase in extracted mass of a compound between the
@@
 The plot displays the extracted quantity across the three washes and final calculus extraction (calc).
-Each coloured line represents a different calculus sample."
+Each coloured line represents a different calculus sample.
+CBD = cannabidiol; CBN = cannabinol; THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA = tetrahydrocannabivarin acid."
 #| fig-width: 8
@@
       axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
-      legend.position = "none",
+      legend.position = "left",
       panel.grid.major.x = element_blank()
@@
 (B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2."
-#| fig-height: 6
-#| fig-width: 7
 
@@
     geom_boxplot(width = 0.2) +
-    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
+    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
     theme_bw() +
@@
     geom_point(col = viridisLite::viridis(1)) +
-    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
+    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
     theme_bw() +
@@
   filter(
+    batch == "batch2",
     compound %in% c("nicotine", "cotinine"),
@@
 The presence of pipe notch(es) in an individual and concurrent detection of nicotine
-and/or cotinine is used as a crude indicator of the accuracy of the method. When
-combining the results of both batches, the method was able to detect some form
+and/or cotinine is used as a crude indicator of the accuracy of the method. Only
+males were used in accuracy calculations, as pipe notches are ubiquitous in males,
+but not in females.
+In batch 2, the method was able to detect some form
 of tobacco in
@@
 Accuracy in the old adult age category is
-`r scales::percent(filter(accuracy_age, age == "old")$mean, accuracy = 0.1)`.
-
-In the replicated samples only, tobacco detection was successful in
-`r nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0))`
-out of
-`r nrow(filter(replicated_accuracy, pipe_notch == "Y"))`
-pipe smokers
-(`r scales::percent(nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(replicated_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
-Including individuals with absence of a pipe notch and concurrent absence of
-compounds as a correct identification, gives an overall accuracy of
-`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
+`r scales::percent(filter(accuracy_age, age == "old")$mean, accuracy = 0.1)`,
+but with only `r sum(tobacco_accuracy$age == "old")` individuals.
 
@@
 
+Point-biserial correlation was conducted on paired continuous and dichotomous
+variables, to see if any relationships exist between extracted concentrations
+and other variables.
 The strongest point-biserial (Pearson) correlation correlations were a near-perfect
@@
 #| label: fig-polycorr
-#| fig-cap: "Plot of the polychoric correlations (*rho*). Larger circles and increased opacity indicates a stronger correlation coefficient."
+#| fig-cap: "Plot of the polychoric correlations (*rho*).
+Larger circles and increased opacity indicates a stronger correlation coefficient.
+OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl’s nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid."
 polycorr$rho %>%

### 2023-02-07 inst/paper/_results.qmd
clean up code

@@
 #| label: detection-tbl-setup
-
 uhplc_calculus_detect <- uhplc_calculus_long %>%
@@
 (B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2."
+#| fig-width: 5
+#| fig-height: 7
 
@@
     batch == "batch2",
-    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
     id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
@@
   filter(
-    #!is.na(preservation),
     batch == "batch2",
     quant > 0,
-    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
     id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
@@
     labs(y = "Quantity (ng)", x = "Weight (mg)")
-
-# weight_quant_plot <- uhplc_data_comb %>%
-#   select(sample, contains("batch2")) %>% 
-#   select(sample, batch2_weight, contains("calc")) %>% 
-#   pivot_longer(
-#     -c(sample, batch2_weight),
-#     names_to = c("compound", "batch"),
-#     names_pattern = "(.*)_(.*)",
-#     values_to = c("conc")
-#   ) %>% 
-#   mutate(compound = str_remove(compound, "_calc")) %>%
-#   #remove_missing() %>% 
-#   filter(
-#     conc > 0,
-#     id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
-#     ) %>% 
-#   ggplot(aes(x = batch2_weight, y = conc)) +
-#     geom_point(col = viridisLite::viridis(1)) +
-#     facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
-#     theme_bw() +
-#     theme(axis.title.y = element_blank()) +
-#     labs(y = "Quantity (ng)", x = "Weight (mg)")
 
@@
     detection = sum(detection),
-    #compound = compound
     ) %>% 
@@
   filter(var != "age",
-         var != "periodont_status") #%>% 
-  #mutate(across(everything(), signif, 3))
+         var != "periodont_status")
 
@@
   )
-# still need to remove everything under the diagonal (colour white?)
-#corr_tib[corr_tib == 1] <- kableExtra::cell_spec(corr_tib[corr_tib == 1], color = "transparent")
+
 corr_tib_out <- corr_tib %>%
@@
 cols <- 0
-for(i in 7:nrow(corr_tib)){ # terrible solution for formatting table
+for(i in 7:nrow(corr_tib)){ # remove repeated values
   cols <- cols + 1 

### 2023-02-07 inst/scripts/01-data_prep.R

@@ -10,20 +10,12 @@ library(purrr)
 # Upload data -------------------------------------------------------------
 
 # See README for details
-metadata <- read_tsv("inst/data/raw_data/metadata.tsv") # ids need to be changed to match key
+metadata <- read_tsv("inst/data/raw_data/metadata.tsv")
 lloq <- read_tsv("inst/data/raw_data/lloq.tsv")
 uhplc_data_raw <- read_csv("inst/data/raw_data/uhplc-results.csv")
 uhplc_data_raw2 <- read_csv("inst/data/raw_data/uhplc-results_batch2.csv")
-dental_inv <- read_csv("inst/data/raw_data/dental-inv.csv")
 sinusitis <- read_csv("inst/data/raw_data/sinusitis.csv")
 path_cond <- read_csv("inst/data/raw_data/path-conditions.csv")
-caries <- read_csv("inst/data/raw_data/caries.csv")
-periodont <- read_csv("inst/data/raw_data/periodontitis.csv")
-periap <- read_csv("inst/data/raw_data/periapical.csv")
-calculus <- read_csv("inst/data/raw_data/calculus.csv")
-calculus_full <- read_csv("inst/data/raw_data/calculus_full.csv")
-demography <- read_csv("inst/data/raw_data/demography.csv")
-
 
 # CMS = chronic maxillary sinusitis
 # IPR = periosteal reaction on visceral surface of ribs
@@ -77,38 +69,6 @@ uhplc_data_comb <- uhplc_data_batch1 %>%
   #select(sample, contains("calc"))
 
 
-# Compounds presence/absence
-
-# uhplc_calculus_bin <- uhplc_calculus %>%
-#   mutate(
-#     across(
-#       where(is.numeric),
-#       function(x) if_else(!is.na(x), T, F) # convert concentration to boolean
-#       )
-#     ) %>%
-#   pivot_longer( # convert to long form to include only compounds present or absent in both batches
-#     cols = -sample,
-#     names_to = c("compound", "batch"),
-#     names_pattern = "(.*)_(.*)",
-#     values_to = c("value")
-#     ) %>%
-#     mutate(compound = str_remove(compound, "_calc"), # redundant suffix
-#            value = as.numeric(value)) %>% # convert to 0s and 1s
-#     group_by(sample, compound) %>%
-#     summarise(value = sum(value))
-
-# create data frame with replicated compounds and remove compounds absent in all individuals
[…]

### 2023-02-07 inst/scripts/02_data-explore.R

@@ -1,69 +1,0 @@
-# Soft-deprecated
-# should this all be done in the report?
-library(readr)
-library(dplyr)
-library(tidyr)
-library(purrr)
-
-# Upload data -------------------------------------------------------------
-
-metadata <- read_tsv("inst/data/raw_data/metadata.tsv")
-lloq <- read_tsv("inst/data/raw_data/lloq.tsv")
-uhplc_calculus <- read_tsv("inst/data/derived_data/uhplc-data_cleaned.csv")
-dental_inv <- read_csv("inst/data/raw_data/dental-inv.csv")
-caries <- read_csv("inst/data/raw_data/caries.csv")
-periodont <- read_csv("inst/data/raw_data/periodontitis.csv")
-periap <- read_csv("inst/data/raw_data/periapical.csv")
-calculus <- read_csv("inst/data/raw_data/calculus.csv")
-demography <- read_csv("inst/data/raw_data/demography.csv")
-sinusitis <- read_csv("inst/data/derived_data/sinusitis_cleaned.csv")
-
-
-# Convert to long ---------------------------------------------------------
-
-dental_inv_long <- dental_inv %>% # Inventory
-  right_join(demography) %>%
-  select(!c(occupation, own_grave, tax, pipe_notch)) %>%
-  pivot_longer(t11:t48, names_to = "tooth",
-               values_to = "status") %>%
-  mutate(status = case_when(status == "dna" ~ "m", # teeth missing due to DNA sampling
-                            TRUE ~ status))
-
-teeth_list <- list(caries, periodont, periap, calculus)
-
-test<- lapply(
-  teeth_list,
-  pivot_longer,
-  cols = t11:t48,
-  names_to = "tooth",
-  values_to = "score"
-)
-
-caries_long <- caries %>% # Caries
-  pivot_longer(t11:t48, names_to = "tooth",
-               values_to = "caries_score")
-
-periodont_long <- periodont %>% # Periodontitis
-  pivot_longer(cols = t11:t48,
-               names_to = "tooth",
-               values_to = "periodont_score")
-
-periap_long <- periap %>% # Periapical lesions
-  pivot_longer(cols = t11:t48,
-               names_to = "tooth",
-               values_to = "periap_score")
-
-calculus_long <- calculus %>% # Calculus
-  pivot_longer(cols = t11:t48,
-               names_to = "tooth",
-               values_to = "calculus_score")
-
-# prepare to merge data frames
[…]

### 2023-02-07 inst/scripts/03_data-analyse.R

@@ -1,104 +1,0 @@
-# data analysis
-library(psych)
-library(dplyr)
-library(readr)
-library(corrplot)
-
-# Upload data -------------------------------------------------------------
-
-#efa_data <- read_csv("inst/data/derived_data/efa-analysis_data.csv")
-
-# combine nicotine and cotinine to tobacco
-efa_comb <- efa_data %>%
-  mutate(tobacco = case_when(nicotine == 1 | cotinine == 1 ~ 1,
-                             TRUE ~ 0)) %>%
-  select(!c(nicotine, cotinine))
-
-# Exploratory factor analysis (psych package?)
-
-  # tetrachoric correlation for dichotomous variables
-
-
-# Missing values
-
-
-
-# selection of variables (remove very low and high frequencies)
-
-# dichotomise caries and calculus ratios for polychoric correlation
-  # caries ratios falling below the 25th percentile are considered absent
-  # discretisation of caries and calculus
-caries_discrete <- quantile(caries_rate$caries_rate)
-calculus_discrete <- quantile(calc_index$calc_index)
-
-caries_bin <- caries_rate %>%
-  mutate(caries_rate = case_when(caries_rate > quantile(caries_rate)[2] ~ 1,
-                                 TRUE ~ 0
-  ))
-
-
-efa_filter <- efa_comb %>%
-  select(!c(OD)) %>% # vertebral osteophytosis highly correlated with OA
-  summarise(across(where(is.numeric), function(x) sum(x) / length(x))) %>%
-  select(which(colSums(.) > 0.1))
-
-efa_select <- efa_comb %>%
-  select(names(efa_filter)) %>%
-  select(!c(periap_num)) %>%
-  filter(complete.cases(.)) #%>%
-  as.matrix()
-
-efa_discrete <- efa_select %>%
-  mutate(
-    calc_index = case_when(
-      calc_index <= calculus_discrete[1] ~ 0,
-      calc_index >= calculus_discrete[1] & calc_index < calculus_discrete[2] ~ 1,
-      calc_index >= calculus_discrete[2] & calc_index < calculus_discrete[3] ~ 2,
-      calc_index >= calculus_discrete[3] & calc_index < calculus_discrete[4] ~ 3,
-      calc_index >= calculus_discrete[4] & calc_index <= calculus_discrete[5] ~ 4,
-      ),
-    caries_rate = case_when(
-      caries_rate <= caries_discrete[1] ~ 0,
[…]

### 2023-02-07 inst/scripts/new-explore.R

@@ -1,118 +1,0 @@
-library(tidyverse)
-library(here)
-source(here("scripts/dental-plot.R"))
-
-dental_long <- readr::read_csv(here("data/derived_data/dental-full_long.csv"))
-
-
-
-# Data wrangling ----------------------------------------------------------
-
-# Caries rate
-
-# convert caries data from location to number of caries and caries rate
-# number of lesions per tooth observed
-# separate anterior and posterior teeth
-surface <- c("mes", "dis", "occ", "buc", "lin", "root", "crown") # here occ = incisal
-
-caries_rate <- dental_long %>%
-  filter(status == "p") %>%
-  separate_rows(caries_score, sep = ";") %>% # one lesion per row
-  mutate(caries_count = if_else(
-    caries_score == "none", 0L, 1L
-    )
-  ) %>%
-  group_by(id, tooth_type, position) %>%
-  summarise(
-    n_teeth = n(),
-         count = sum(caries_count, na.rm = T),
-         rate = count / n_teeth
-            )
-
-
-  # mutate(caries_count = if_else(caries_score == "none", 0L, 
-  #                               stringi::stri_count_words(caries_score)
-  #                               )
-  #       )
-  
-#caries_rate <- 
-# caries_long %>%
-#   group_by(id, position) %>%
-#   filter(status == "p") %>%
-#   summarise(n_teeth = n(),
-#             count = sum(caries_count, na.rm = T),
-#             rate = count / n_teeth
-#             ) %>%
-#   ungroup()
-#filter(status == "p") %>%
-#  summarise(caries_count = sum(caries_count, na.rm = T)) %>% #view
-#  group_by(id) %>%
-#  summarise(n_teeth = n(),
-#            caries_rate = sum(caries_count, na.rm = T) / n_teeth)
-
-# caries_rate_type <- dental_long %>%
-#   group_by(tooth_type) %>%
-#   summarise(n_teeth = sum(!is.na(caries_count)),
-#             n_caries = sum(caries_count, na.rm = T),
-#             caries_rate = n_caries / n_teeth) # shouldn't be any NA??
-
-# CARIES RATE CALCULATION
-
-# Caries longform data with row for each surface of each tooth?
-
-# convert periodontitis and calculus to mean/median score?  
-
-
-
-#full_data_long
-
-# compare replicated samples across batch 1 and 2
-
-caffeine_rep <- uhplc_data_comb %>%
-  select(sample, contains("caffeine")) %>%
[…]

### 2023-02-07 inst/scripts/new-prep.R

@@ -1,185 +1,0 @@
-library(tidyverse)
-library(here)
-
-# Upload data -------------------------------------------------------------
-
-# See README for details
-lloq <- readr::read_tsv(here("data/lloq.tsv"))
-uhplc_data_raw <- readr::read_csv(here("data/raw/anon/uhplc-results.csv"))
-uhplc_data_raw2 <- readr::read_csv(here("data/raw/anon/uhplc-results_batch2.csv"))
-dental_inv <- readr::read_csv(here("data/raw/anon/dental-inv.csv"))
-caries <- readr::read_csv(here("data/raw/anon/caries.csv"))
-periodont <- readr::read_csv(here("data/raw/anon/periodontitis.csv"))
-periap <- readr::read_csv(here("data/raw/anon/periapical.csv"))
-calculus <- readr::read_csv(here("data/raw/anon/calculus.csv"))
-demography <- readr::read_csv(here("data/raw/anon/demography.csv"))
-sinusitis <- readr::read_csv(here("data/raw/anon/sinusitis.csv"), na = "UNOBS")
-# CMS = chronic maxillary sinusitis
-# IPR = periosteal reaction on visceral surface of ribs 
-  # present only if active (pulmonary infection)
-
-# Helper objects ----------------------------------------------------------
-
-# vectors to categorise teeth by position and type (FDI notation)
-maxilla <- c(paste0("t", 11:18), paste0("t", 21:28))
-mandible <- c(paste0("t", 31:38), paste0("t", 41:48)) 
-left <- c(paste0("t", 21:28), paste0("t", 31:38))
-posterior <- c(paste0("t", 14:18), paste0("t", 24:28), 
-               paste0("t", 34:38), paste0("t", 44:48))
-molar <- c(paste0("t", 16:18), paste0("t", 26:28), 
-           paste0("t", 36:38), paste0("t", 46:48))
-incisor <- c(paste0("t", 11:12), paste0("t", 21:22), 
-             paste0("t", 31:32), paste0("t", 41:42))
-premolar <- c(paste0("t", 14:15), paste0("t", 24:25), 
-              paste0("t", 34:35), paste0("t", 44:45))
-canine <- c(paste0("t", 13), paste0("t", 23), 
-            paste0("t", 33), paste0("t", 43))
-
-# Data cleaning -----------------------------------------------------------
-
-# UHPLC-MS/MS results
-
-uhplc_data <- uhplc_data_raw %>%
[…]

### 2023-02-07 inst/scripts/setup-qmd.R

@@ -1,5 +1,3 @@
-library(here)
-library(readr)
 library(dplyr)
 library(tidyr)
 library(tibble)
@@ -10,7 +8,6 @@ library(glue)
 library(purrr)
 library(psych)
 library(patchwork)
-devtools::load_all()
 
 try(generate_bib())
 
@@ -70,45 +67,6 @@ uhplc_calculus_long <- uhplc_data_long %>%
   mutate(conc = quant / weight) %>%
   left_join(demography, by = "id") %>%
   mutate(preservation = factor(preservation, levels = c("fair", "good", "excellent")))
-
-# uhplc_calculus_demography <- uhplc_calculus_long %>%
-#   left_join(demography) %>%
-#   mutate(preservation = factor(preservation, levels = c("fair", "good", "excellent")))
-
-# presence/absence data frame
-uhplc_calculus_bin <- uhplc_calculus_long %>%
-  mutate(presence = if_else(quant > 0, 1, quant))
-
-# successfully replicated samples only
-uhplc_calculus_replicated <- uhplc_calculus_bin %>%
-  mutate(compound = str_remove(compound, "_calc")) %>%
-  group_by(id, sample, compound) %>% # combine batches
-  summarise(presence = sum(presence)) #%>%
-  # filter(presence == 0 | presence == 2) %>%
-  # group_by(compound) %>%
-  # mutate(presence = if_else(presence == 2, 1, 0)) # convert replications to presence/absence
-
-uhplc_calculus_replicated <- uhplc_calculus_bin %>%
-  filter(id %in% filter(metadata, replicated == TRUE)$id) %>%
-  #right_join(select(filter(metadata, replicated == TRUE)), by = c("id", "sample"))
-  #mutate(compound = str_remove(compound, "_calc")) %>%
-  group_by(id, sample, compound) %>% # combine batches
-  summarise(presence = sum(presence)) %>%
-  filter(presence == 0 | presence == 2) %>% # remove compounds only detected in one batch
-  group_by(compound) %>%
-  mutate(presence = if_else(presence == 0, 0, 1)) %>%  # convert replications to presence/absence
-  ungroup()
-
-uhplc_replicated_wide <- uhplc_calculus_replicated %>%
-  mutate(compound = case_when(compound == "nicotine" ~ "tobacco",
-                              compound == "cotinine" ~ "tobacco",
-                              TRUE ~ compound)) %>%
[…]

### 2023-02-07 analysis/paper/paper.qmd
rename analysis directory

--- 2023-01-07 inst/paper/paper.qmd
@@
 library(readr)
+devtools::load_all()
 
@@
 uhplc_data_comb <- read_csv(here("inst/data/derived_data/uhplc-data_combined.csv"))
-#uhplc_calculus <- read_csv(here("inst/data/derived_data/uhplc-calculus_cleaned.csv"))
 dental_inv <- read_csv(here("inst/data/raw_data/dental-inv.csv"))

### 2023-02-07 analysis/supplementary-materials/supp-mat.qmd
rename analysis directory

--- 2022-12-12 inst/supplementary-materials/supp-mat.qmd
@@
 title: "Supplementary Materials"
+execute: 
+  echo: false
+  warning: false
 knitr: 
   opts_chunk: 
-    warning: false
     message: false
@@
     toc: true
+    code-fold: true
 ---
@@
 
-source(here("inst/scripts/setup-qmd.R"), local = knitr::knit_global())
+source(here("inst/scripts/setup-qmd.R"))
 ```
@@
 
+<!-- id	sample	element	KZ_sample	KZ_element	replicated	batch1_weight	batch2_weight -->
+
 | variable | description |
@@
 | sample | sample number for the UHPLC-MS/MS analysis |
+| element | which tooth (FDI notation) was sampled for UHPLC-MS/MS analysis |
+| replicated | Whether the sample was included in the replication batch (TRUE/FALSE) |
+| batch1_weight | weight (mg) of sample in batch 1 |
+| batch2_weight | weight (mg) of sample in batch 2 |
 
@@
 |---|---|
-| feat | Identifier: Feature number of individual |
-| find | Identifier: Find number of individual |
+| id | unique identifier for the individual |
 | t11..t48 | Tooth number (FDI system) and status |
@@
 |---|---|
-| feat | Identifier: Feature number of individual |
-| find | Identifier: Find number of individual |
+| id | unique identifier for the individual |
 | t11..t48 | Tooth number (FDI system) and location of caries |
@@
 |---|---|
-| feat | Identifier: Feature number of individual |
-| find | Identifier: Find number of individual |
+| id | unique identifier for the individual |
 | t11..t48 | Tooth number (FDI system) and periodontitis score (0-3) |
@@
 |---|---|
-| feat | Identifier: Feature number of individual |
-| find | Identifier: Find number of individual |
+| id | unique identifier for the individual |
 | t11..t48 | Tooth number (FDI system) and periapical lesion |
@@
 |---|---|
-| feat | Identifier: Feature number of individual |
-| find | Identifier: Find number of individual |
+| id | unique identifier for the individual |
 | t11..t48 | Tooth number (FDI system) and deposit size (0-3) |
@@
 
-## Sample demographics
+## Sample
+
+### Selection
+
+```{r}
+#| label: tbl-sampling
+#| tbl-cap: "Table showing which individuals were sampled in this study (element) and which individuals were sampled in the previous study (KZ_element)."
+metadata %>% 
+  select(id, element, KZ_element) %>% 
+  mutate(
+    element = case_when(is.na(element) ~ FALSE,
+                   TRUE ~ TRUE),
+    KZ_element = case_when(is.na(KZ_element) ~ FALSE,
+                   TRUE ~ TRUE),
+    )
+```
+
+### Demographics
 
@@
 
-## Missing data
+### Missing data
 
@@
 ```{r}
-#| label: fig-missing-hex
+#| label: fig-missing-caries
 caries %>%
@@
     geom_tile()
-
 ```
@@
 #| label: fig-missing-dental
+#| fig-cap: "Number of missing values per individual and tooth across all dental scores (caries, periodontitis, periapical lesions, calculus)."
 dental_long %>%
@@
 ## UHPLC analysis
+
+```{r}
+#| label: setup-uhplc
+
+# presence/absence data frame
+uhplc_calculus_bin <- uhplc_calculus_long %>%
+  mutate(presence = if_else(quant > 0, 1, quant))
+
+# successfully replicated samples only
+uhplc_calculus_replicated <- uhplc_calculus_bin %>%
+  mutate(compound = str_remove(compound, "_calc")) %>%
+  group_by(id, sample, compound) %>% # combine batches
+  summarise(presence = sum(presence)) #%>%
+  # filter(presence == 0 | presence == 2) %>%
+  # group_by(compound) %>%
+  # mutate(presence = if_else(presence == 2, 1, 0)) # convert replications to presence/absence
+
+uhplc_calculus_replicated <- uhplc_calculus_bin %>%
+  filter(id %in% filter(metadata, replicated == TRUE)$id) %>%
+  #right_join(select(filter(metadata, replicated == TRUE)), by = c("id", "sample"))
+  #mutate(compound = str_remove(compound, "_calc")) %>%
+  group_by(id, sample, compound) %>% # combine batches
+  summarise(presence = sum(presence)) %>%
+  filter(presence == 0 | presence == 2) %>% # remove compounds only detected in one batch
+  group_by(compound) %>%
+  mutate(presence = if_else(presence == 0, 0, 1)) %>%  # convert replications to presence/absence
+  ungroup()
+
+uhplc_replicated_wide <- uhplc_calculus_replicated %>%
+  mutate(compound = case_when(compound == "nicotine" ~ "tobacco",
+                              compound == "cotinine" ~ "tobacco",
+                              TRUE ~ compound)) %>%
+  group_by(id, sample, compound) %>%
+  summarise(presence = sum(presence)) %>% # combine nicotine and cotinine
+  #remove_missing() %>%
+  mutate(presence = case_when(presence > 0 ~ TRUE,
+                              TRUE ~ FALSE)) %>%
+  pivot_wider(names_from = "compound", values_from = "presence")
+```
+
+The UHPLC-MS/MS method was validated in a separate study on cadavers received
+for forensic autopsy and toxicological analysis. Results from dental calculus were
+validated against compounds detected in whole blood samples from the same
+individuals [@sorensenDrugsCalculus2021].  
+In the original method, samples were washed three times to remove residual
+substances from the surface of the calculus that originated from oral fluids,
+and only extract substances from the calculus. In our samples the washes served
+to remove potential contaminants from the burial environment and post-excavation
+handling.
+
+Briefly, dental calculus was treated with citric acid and the dissolution extracts
+were cleaned using weak and strong polymeric cation-exchange sorbents. Samples
+were washed with 0.5 mL MeOH for 10 seconds. Samples were weighed before and after
+each wash. The wash solvent was evaporated to a residual volume of 10 µl and added
+50µl 30% methanol.
+Samples were air-dried for 24 hours at room temperature after each wash.
+Extracts from each wash were analysed by injecting 5 µL into the column
+on an Exion UHPLC system that consisted of two Exion AD pumps, an Exion AD
+multiplate autosampler set at 10 $\pm$ 2 &deg;C and an Exion AC column oven set
+at 40 $\pm$ 2 &deg;C (Sciex, Ontario, Canada).
+Separation was performed using a Raptor Biphenyl UHPLC column (2.7 mm, 2.1 mm I.D.
+$\times$ 100 mm) (Restek, Bellefonte, PA). The mass spectrometer was a Sciex QTRAP
+6500+ with a TurboIonSpray probe for electrospray ionisation.  
+The remaining calculus was dissolved using lysing tube beads in 800 $\mu$L of 0.5
+$\small{M}$ citric acid (CA) and 50 $\mu$L stable isotope-labelled analogue used
+as internal standards (SIL-IS) solution for 1 h at ambient
+temperature with gentle shaking. The suspension was then mixed with 800 $\mu$L
+MeOH and centrifuged at 10,000 $\times$ g for 5 mins, and analysed by the same
+method as the wash extracts.
+Data analysis was performed using Analyst 1.7 and MultiQuant 3.0.3 (Sciex).
+Raw quantities of compounds are presented in ng and concentrations as ng / mg.
+
+The samples in the replication batch were processed in the same way, but
+analysed on equipment used exclusively for oral samples.
 
@@
 will be similarly high (@fig-quant-weight-1 and @fig-quant-weight-1).
+
+The positive correlation between the weight of the calculus
+sample and recovered quantities of the compounds suggests sample weight may affect
+the ability to detect compounds; although, we were able to detect
+compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].
 
@@
     !is.na(preservation),
-    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected
+    batch == "batch1",
+    quant > 0,
+    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
     ) %>% 
-  ggplot(aes(x = preservation, y = quant, fill = preservation)) +
-    geom_violin() +
-    geom_jitter(width = 0.2) +
-    #geom_boxplot(alpha = 0.6, width = 0.2) +
-    facet_wrap(~ compound, scales = "free_y") +
-    theme_bw()
-# 
-# uhplc_calculus_demography %>%
-#   filter(
-#     !is.na(preservation),
-#     !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected
-#     ) %>% 
-#   ggplot(aes(x = preservation, y = quant, fill = preservation)) +
-#     geom_violin() +
-#     geom_jitter(width = 0.2) +
-#     #geom_boxplot(alpha = 0.6, width = 0.2) +
-#     facet_wrap(~ compound, scales = "free_y") +
-#     theme_bw()
+  ggplot(aes(x = preservation, y = quant)) +
+    geom_violin(aes(fill = preservation), alpha = 0.6) +
+    #geom_jitter(width = 0.2) +
+    geom_boxplot(width = 0.2) +
+    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
+    theme_bw() +
+    theme(
+      legend.position = "none",
+      panel.grid.major.x = element_blank(),
+      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
+      ) +
+    labs(x = "Preservation", y = "Quantity (ng)") +
+    scale_fill_viridis_d()
 ```
@@
 #| label: fig-detection-preservation2
-#uhplc_calculus_demography %>%
 uhplc_calculus_long %>%
-  filter(!is.na(preservation)) %>% 
-  ggplot(aes(x = preservation, y = quant, fill = preservation)) +
-    geom_boxplot() +
-    facet_wrap(~ compound, scales = "free_y") +
+  filter(
+    !is.na(preservation),
+    quant > 0,
+    batch == "batch2",
+    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
+    ) %>% 
+  ggplot(aes(x = preservation, y = quant)) +
+    geom_violin(aes(fill = preservation), alpha = 0.6) +
+    #geom_jitter(width = 0.2) +
+    geom_boxplot(width = 0.2) +
+    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
+    theme_bw() +
+    theme(
+      legend.position = "none",
+      panel.grid.major.x = element_blank(),
+      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
+      ) +
+    labs(x = "Preservation", y = "Quantity (ng)") +
+    scale_fill_viridis_d()
+
+```
+
+Distribution of state-of-preservation in batches 1 and 2 to make sure the number
+of skeletons are not affecting the relationships shown above. Given our sample
+contains a smaller number of individuals with fair preservation, this may bias
+our interpretations (@fig-preservation-detection).
+
+```{r}
+#| label: fig-preservation-detection
+uhplc_calculus_long %>% 
+  ungroup() %>% 
+  distinct(id, batch, .keep_all = TRUE) %>%
+  remove_missing(vars = c("preservation", "weight")) %>% 
+  ggplot(aes(x = preservation)) +
+    geom_bar() +
+    facet_wrap(~ batch) +
     theme_bw()
@@
   mutate(detection = if_else(is.na(conc), FALSE, TRUE))
+
+tobacco <- uhplc_calculus_long %>%
+  filter(
+    #batch == "batch2",
+    compound %in% c("nicotine", "cotinine"),
+    sex == "m" | sex == "pm"
+    ) %>% 
+  mutate(detection = if_else(quant == 0, 0, 1)) # 0 = not detected; NA = not included in batch 2.
+
+tobacco_accuracy <- tobacco %>%
+  remove_missing(vars = "quant") %>%
+  group_by(sample, id, .drop = F) %>% 
+  summarise(
+    detection = sum(detection),
+    #compound = compound
+    ) %>% 
+  left_join(select(demography, id, pipe_notch, preservation, age), by = "id") %>% 
+  mutate(
+    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
+    correct = case_when(
+      detection == 0 & pipe_notch == "N" ~ 1,
+      detection > 0 & pipe_notch == "Y" ~ 1,
+      detection > 0 & pipe_notch == "N" ~ NaN, # no way of knowing macroscopically if the person smoked without pipe
+      TRUE ~ 0
+    )
+  )
+
+accuracy_age <- tobacco_accuracy %>% # move to supplementary material
+  group_by(age) %>% 
+  summarise(mean = mean(correct, na.rm = T))
 
@@
 
+We found no correlation between the number of pipe notches
+and the concentration of nicotine and cotinine, suggesting that our ability
+to detect tobacco consumption in dental calculus does not necessarily rely on
+targeting frequent smokers; here, we consider individuals with multiple pipe notches
+as likely to have been heavy smokers.
+
 No apparent correlation between the number of pipe notches
 and the concentration of nicotine or cotinine [@fig-corr-notches].
+
+The presence of pipe notch(es) in an individual and concurrent detection of nicotine
+and/or cotinine is used as a crude indicator of the accuracy of the method. When
+combining the results of both batches, the method was able to detect some form
+of tobacco in
+`r nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0))`
+of `r nrow(filter(tobacco_accuracy, pipe_notch == "Y"))`
+individuals with a pipe notch
+(`r scales::percent(nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(tobacco_accuracy, pipe_notch == "Y")), accuracy = 0.1)`).
+When also considering correct the absence of a tobacco alkaloid together with the absence
+of a pipe notch, the accuracy of the method is
+`r scales::percent(mean(tobacco_accuracy$correct, na.rm = T), accuracy = 0.1)`.
+Accuracy in the old adult age category is
+`r scales::percent(filter(accuracy_age, age == "old")$mean, accuracy = 0.1)`.
+
+In the replicated samples only, tobacco detection was successful in
+`r nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0))`
+out of
+`r nrow(filter(replicated_accuracy, pipe_notch == "Y"))`
+pipe smokers
+(`r scales::percent(nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(replicated_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
+Including individuals with absence of a pipe notch and concurrent absence of
+compounds as a correct identification, gives an overall accuracy of
+`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
 
@@
 
+Pipe notches were identified by wear on the mesial and distal
+sides of the crowns between to teeth, resulting from the practice of clenching a
+pipe between adjacent and isomeric teeth, and which differs from the occlusal
+wear that occurs through mastication. Wear occurring between adjacent and isomeric
+teeth were counted as a single pipe notch.
+
 ```{r}
@@
 
-Caries were scored based on location in each individual. Multiple locations on
-a single tooth is possible. The size of caries was also
+Caries were scored as the location on each individual tooth. Multiple locations on
+a single tooth were separated with `;`. The size of caries was also
 recorded, but not used in further analysis. Large caries that cover multiple
-surfaces were recorded as 'crown'. Caries counts were extracted based on the
-number of locations recorded.
+surfaces with an unknown origin were recorded as 'crown'. Caries counts were
+extracted based on the number of locations recorded.
 
@@
 (`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`).
-This frequency has very little meaning, and was further broken down into tooth
-class and individuals.
+This frequency has very little meaning, and was further broken down into a ratio
+for each individual and each tooth class.
 
@@
 
+Data on chronic maxillary sinusitis from @casnaUrbanizationRespiratory2021 were
+included in this study to assess the relationship between upper respiratory
+diseases with environmental factors (i.e. tobacco smoke, caffeine consumption).
+Chronic maxillary sinusitis (CMS) is the inflammation of the lower paranasal
+sinuses, air-filled pockets located in the skull that defend the organism against
+inhaled particulate matter and pathogens. This occurs through the production of
+mucus carried by small hairs toward an opening situated on the superior part of
+the sinus, where pathogens are drained [@slavinDiagnosisManagement2005]. Without
+drainage, mucus begins to accumulate in the sinuses, providing an ideal environment
+for bacterial growth and thereby contributing to inflammation of the mucous
+membranes and subsequently of the bone surfaces [@jangBoneInvolvement2002].
+Lesions associated with CMS as defined by @boocockMaxillarySinusitis1995 were
+recorded for each individual and classified as "pitting", "spicule-type bone
+formation", "remodeled spicules", or "white pitted bone". CMS was scored as absent
+when the sinus presented smooth surfaces with little or no associated pitting.
+To facilitate inspection, fragmented sinuses were cleaned using a dry tooth-brush
+and water where necessary. If the sinuses were not observable with the naked eye,
+they were examined with a flexible medical endoscope (Pentax, model: FNL-10RBS,
+ø=4mm; view angle=30°) inserted through minor breaks naturally occurring on the
+inferior nasal conchae and palatine bone, where the bone tissue is thinner.
+
+<!-- description of other diseases
+Osteoarthritis
+Vertebral osteophytosis
+Cribra orbitalia
+Degenerative disc disease
+Schmorl's nodes
+-->
+
 ## Statistical analysis
@@
 ```
+
+The sample bias caused by targeting larger calculus deposits may also affect the detection of
+compounds, as caffeinated drinks are often
+acidic, and may cause a lower calculus formation.
+We only found a weak negative correlation
+between caffeine concentration and dental calculus index, but since we targeted individuals with
+calculus, this may inadvertently have been controlled for during sample collection.
+
+<!-- correlations on caries and calculus separated by quadrant/tooth type -->

### 2023-02-08 analysis/paper/paper.qmd
fix: broken links after renaming to analysis

@@
 # upload data
-metadata <- read_tsv(here("inst/data/raw_data/metadata.tsv"))
-demography <- read_csv(here("inst/data/raw_data/demography.csv"))
-lloq <- read_tsv(here("inst/data/raw_data/lloq.tsv"))
-uhplc_data_comb <- read_csv(here("inst/data/derived_data/uhplc-data_combined.csv"))
-dental_inv <- read_csv(here("inst/data/raw_data/dental-inv.csv"))
-caries <- read_csv(here("inst/data/raw_data/caries.csv"))
-periodont <- read_csv(here("inst/data/raw_data/periodontitis.csv"))
-periap <- read_csv(here("inst/data/raw_data/periapical.csv"))
-calculus <- read_csv(here("inst/data/raw_data/calculus.csv"))
-calculus_full <- read_csv(here("inst/data/raw_data/calculus_full.csv"))
-sinusitis_clean <- read_csv(here("inst/data/derived_data/sinusitis_cleaned.csv"))
-path_cond_clean <- read_csv(here("inst/data/derived_data/path-conditions_cleaned.csv"))
+metadata <- read_tsv(here("analysis/data/raw_data/metadata.tsv"))
+demography <- read_csv(here("analysis/data/raw_data/demography.csv"))
+lloq <- read_tsv(here("analysis/data/raw_data/lloq.tsv"))
+uhplc_data_comb <- read_csv(here("analysis/data/derived_data/uhplc-data_combined.csv"))
+dental_inv <- read_csv(here("analysis/data/raw_data/dental-inv.csv"))
+caries <- read_csv(here("analysis/data/raw_data/caries.csv"))
+periodont <- read_csv(here("analysis/data/raw_data/periodontitis.csv"))
+periap <- read_csv(here("analysis/data/raw_data/periapical.csv"))
+#calculus <- read_csv(here("analysis/data/raw_data/calculus.csv"))
+calculus_full <- read_csv(here("analysis/data/raw_data/calculus_full.csv"))
+sinusitis_clean <- read_csv(here("analysis/data/derived_data/sinusitis_cleaned.csv"))
+path_cond_clean <- read_csv(here("analysis/data/derived_data/path-conditions_cleaned.csv"))
 
-source(here("inst/scripts/setup-qmd.R"))
-#source(here("inst/supplementary-materials/supp-mat.qmd"), local = knitr::knit_global())
+source(here("analysis/scripts/setup-qmd.R"))
+#source(here("analysis/supplementary-materials/supp-mat.qmd"), local = knitr::knit_global())
 # set viridis as default pallette

### 2023-02-08 analysis/supplementary-materials/supp-mat.qmd
fix: broken links after renaming to analysis

@@
 # upload data
-metadata <- read_tsv(here("inst/data/raw_data/metadata.tsv"))
-demography <- read_csv(here("inst/data/raw_data/demography.csv"))
-lloq <- read_tsv(here("inst/data/raw_data/lloq.tsv"))
-uhplc_data_comb <- read_csv(here("inst/data/derived_data/uhplc-data_combined.csv"))
-#uhplc_calculus <- read_csv(here("inst/data/derived_data/uhplc-calculus_cleaned.csv"))
-dental_inv <- read_csv(here("inst/data/raw_data/dental-inv.csv"))
-caries <- read_csv(here("inst/data/raw_data/caries.csv"))
-periodont <- read_csv(here("inst/data/raw_data/periodontitis.csv"))
-periap <- read_csv(here("inst/data/raw_data/periapical.csv"))
-calculus <- read_csv(here("inst/data/raw_data/calculus.csv"))
-calculus_full <- read_csv(here("inst/data/raw_data/calculus_full.csv"))
-sinusitis_clean <- read_csv(here("inst/data/derived_data/sinusitis_cleaned.csv"))
-path_cond_clean <- read_csv(here("inst/data/derived_data/path-conditions_cleaned.csv"))
-
-source(here("inst/scripts/setup-qmd.R"))
+metadata <- read_tsv(here("analysis/data/raw_data/metadata.tsv"))
+demography <- read_csv(here("analysis/data/raw_data/demography.csv"))
+lloq <- read_tsv(here("analysis/data/raw_data/lloq.tsv"))
+uhplc_data_comb <- read_csv(here("analysis/data/derived_data/uhplc-data_combined.csv"))
+#uhplc_calculus <- read_csv(here("analysis/data/derived_data/uhplc-calculus_cleaned.csv"))
+dental_inv <- read_csv(here("analysis/data/raw_data/dental-inv.csv"))
+caries <- read_csv(here("analysis/data/raw_data/caries.csv"))
+periodont <- read_csv(here("analysis/data/raw_data/periodontitis.csv"))
+periap <- read_csv(here("analysis/data/raw_data/periapical.csv"))
+calculus <- read_csv(here("analysis/data/raw_data/calculus.csv"))
+calculus_full <- read_csv(here("analysis/data/raw_data/calculus_full.csv"))
+sinusitis_clean <- read_csv(here("analysis/data/derived_data/sinusitis_cleaned.csv"))
+path_cond_clean <- read_csv(here("analysis/data/derived_data/path-conditions_cleaned.csv"))
+
+source(here("analysis/scripts/setup-qmd.R"))
 ```
@@
 
-### calculus
+### calculus_full
 
@@
 | id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and deposit size (0-3) |
+| t11_bucc..t48_ip | Tooth number (FDI system) and surface (bucc = buccal; lin = lingual; ip = interproximal) and deposit size (0-3) |
 
@@
 
+Brief statement about amtl at site.
+
+```{r}
+dental_long %>% 
+  group_by(id) %>% 
+  amtl_ratio(.status = status, .add = T)
+```
+
 ### Caries
 
+<!--add a caries correction?-->
 Caries were scored as the location on each individual tooth. Multiple locations on

### 2023-02-08 analysis/scripts/setup-qmd.R

@@ -95,11 +95,11 @@ dental_inv_long <- dental_inv %>% # Inventory
   right_join(demography) %>%
   select(!c(occupation, own_grave, tax, pipe_notch)) %>%
   pivot_longer(t11:t48, names_to = "tooth",
-               values_to = "status") %>%
-  mutate(status = case_when(status == "dna" ~ "m", # teeth missing due to DNA sampling recoded as missing
-                            TRUE ~ status))
+               values_to = "status") #%>% no reason to recode dna...
+  #mutate(status = case_when(status == "dna" ~ "m", # teeth missing due to DNA sampling recoded as missing. WILL affect aml calculations...
+                            #TRUE ~ status))
 
-teeth_list <- list(caries, periodont, periap, calculus)
+teeth_list <- list(caries, periodont, periap)#, calculus) # calculus_full too wide - one column per surface
 
 # convert all data frames in list to long format
 dental_long_list <- lapply(
@@ -125,8 +125,8 @@ dental_long <- reduce(
   rename(
     caries = score.x,
     periodont = score.y,
-    periap = score.x.x,
-    calculus = score.y.y
+    periap = score,
+    #calculus = score.y.y
     )
 
 

### 2023-02-17 R/calculus_index.R
add: function to calculate antemortem tooth loss ratio

@@ -46,17 +46,40 @@ calculus_index <- function(.data, simple = FALSE){
 #'
 #' @param .data Data frame in long format, with one tooth surface per row, and one calculus score per surface.
 #' @param .caries Caries count variable.
-#' @param ... Can be used to add grouping variable(s). Group by tooth, tooth type, individual, etc.
+#' @param ... Can be used to add arguments to `group_by()`. Group by tooth, tooth type, individual, etc.
 #' @return Returns a data frame with caries ratio per group.
 #' @export
 
 caries_ratio <- function(.data, .caries, ...){
   .data %>%
-    group_by(..., .add = T) %>%
+    group_by(..., .add = T) %>% # .add = T is unnecessary. can be added with ... in call to function
     summarise(
       #n_teeth = n(),
       n_teeth = sum(!is.na( {{ .caries }} ), na.rm = T),
       count = sum( {{ .caries }} , na.rm = T),
       ratio = count / n_teeth
+    )
+}
+
+#' Calculate antemortem tooth loss
+#'
+#' @details Calculate the caries ratio for each tooth type. Calculated as the number
+#' of caries divided by the number of visible surfaces.
+#'
+#' @param .data Data frame in long format, with one tooth surface per row, and one calculus score per surface.
+#' @param .status character. dental status (p, m, aml, dna).
+#' @param ... Can be used to add grouping variable(s). Group by tooth, tooth type, individual, etc.
+#' @return Returns a data frame with amtl ratio per group.
+#' @export
+
+amtl_ratio <- function(.data, .status, ...){ # add .missing = "m", .loss = "aml" arguments
+  .data %>%
+    group_by(...) %>% # .add = T is required if using group_by prior to call to amtl_ratio()
+    # remove group_by and add elipses to summarise()?
+    summarise(
+      n_teeth = sum( {{ .status }} != "m"),
+      #n_teeth = sum(!is.na( {{ .amtl }} ), na.rm = T),
+      count = sum( {{ .status }} == "aml"),
+      ratio = count / n_teeth # need to remove the postmortem loss teeth? What if a tooth is missing? how to calculate??
     )
 }

### 2023-02-17 analysis/paper/_results.qmd
prepare preprint

--- 2023-02-07 inst/paper/_results.qmd
@@
       axis.title.x = element_blank(),
-      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
+      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
+      panel.grid.major.x = element_blank(),
+      panel.grid.minor.y = element_blank()
     )
@@
     scale_x_discrete(labels = compound_names) +
+    scale_fill_viridis_d(labels = c(1,2)) +
     theme(
@@
       ) +
-    labs(y = "Count")
+    labs(y = "Count", fill = "Batch")
 
@@
     theme_bw() +
-    theme(axis.title.y = element_blank()) +
+    theme(
+      axis.title.y = element_blank(),
+      axis.text.y = element_blank(),
+      axis.ticks.y = element_blank()
+    ) +
     labs(y = "Quantity (ng)", x = "Weight (mg)")

### 2023-02-17 analysis/paper/paper.qmd
prepare preprint

@@
 
+This research has received funding from the European Research Council under the
+European Union’s Horizon 2020 research and innovation program, grant agreement
+number STG–677576 (“HARVEST”).
+
 ## Data Availability Statement
 
-All raw data, analysis scripts, and the source code for the manuscript and
+All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.7648757>).
+Analysis scripts, and the source code for the manuscript and
 supplementary materials are available as a research compendium<!--Zenodo link-->
-using the structure defined in the **rrtools** R package [@Rrrtools].
+using the structure recommended by the **rrtools** R package [@Rrrtools].
 

### 2023-02-17 analysis/supplementary-materials/supp-mat.qmd
prepare preprint

@@
     code-fold: true
+  pdf:
+    toc: true
+bibliography: "../paper/references.bib"
 ---
@@
 
-### metadata
+All raw data are available for download from Zenodo.
+
+### metadata.csv
 
@@
 | element | which tooth (FDI notation) was sampled for UHPLC-MS/MS analysis |
+| KZ_element | which tooth (FDI notation) was sampled for aDNA in original study |
 | replicated | Whether the sample was included in the replication batch (TRUE/FALSE) |
@@
 
-### dental-inv
+### dental-inv.csv
 
@@
 - p = present
-- m = missing (for unknown reason)
+- m = missing (for unknown reason - likely postmortem loss)
 - aml = ante-mortem loss
@@
 
-### caries
+### caries.csv
 
@@
 - dis = distal
-- lin = lingual (and labial)
-- buc = buccal
+- lin = lingual
+- buc = buccal (and labial)
 - occ = occlusal
-- crown = crown (caries-related loss of much of/entire tooth crown)
+- crown = crown (large caries lesion covering multiple surfaces)
 - root = root
@@
 
-### periodontitis
+### periodontitis.csv
 
@@
 
-### periapical
+### periapical.csv
 
@@
 
-### calculus_full
+### calculus_full.csv
 
@@
 
-Scored according to Standards.
-
-### path-conditions
+### path-conditions.csv
 
@@
 #| label: tbl-sampling
-#| tbl-cap: "Table showing which individuals were sampled in this study (element) and which individuals were sampled in the previous study (KZ_element)."
+#| tbl-cap: "Table showing which individuals were sampled in this study and which individuals were sampled in the previous study. When both TRUE, the individual was sampled twice."
 metadata %>% 
@@
                    TRUE ~ TRUE),
-    )
+    ) %>%
+  knitr::kable(col.names = c("ID", "this study", "previous study"))
 ```
@@
 The sample consists of `r nrow(demography)` individuals, most of which are
-middle adult male individuals (@fig-age-distribution).
+middle adult male individuals (@fig-age-distribution). Middle adult males were
+preferentially targeted and the sample age and sex distribution is therefore not
+representative of the population. This was done to limit potential confounding
+factors, and because pipe notches, which served as a positive control for tobacco,
+are predominantly seen in male individuals at the site.
 
@@
 #| label: fig-age-distribution
+#| fig-cap: "Distribution of age and sex in the sample.
+f = female; pf = probable female; pm = probable male; m = male; eya = early young adult (18-24 years); lya = late young adult (25-34 years); ma = middle adult (35-49 years); old = old adult (50+ years)."
 demography %>%
@@
 
-Missing data across all dental datasets
+Missing data across all caries, periodontitis, and periapical datasets.
 
@@
 dental_long %>%
-  mutate(across(c(caries, periodont, periap, calculus), is.na)) %>%
-  select(c(caries, periodont, periap, calculus, id, tooth)) %>%
+  mutate(across(c(caries, periodont, periap), is.na)) %>%
+  select(c(caries, periodont, periap, id, tooth)) %>%
   #group_by(id, tooth) %>% 
   rowwise() %>% 
-  mutate(missing = sum(caries, periodont, periap, calculus)) %>% 
+  mutate(missing = sum(caries, periodont, periap)) %>% 
   ggplot(aes(x = tooth, y = id, fill = missing)) +
@@
 #| label: tbl-lloq
-#| tbl-cap: "Target compounds and lower limits of quantitation."
+#| tbl-cap: "Target compounds and lower limits of quantitation (LLOQ)."
 lloq %>%
-  arrange(compound)
+  arrange(compound) %>% 
+  knitr::kable(col.names = c("Target", "LLOQ"))
 ```
@@
   #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
+  select(!c(sample, conc, extraction, batch, presence)) %>%
   pivot_wider(names_from = "compound", values_from = "quant") %>% 
-  select(!c(extraction, batch))
+  knitr::kable()
 ```
@@
 uhplc_calculus_long %>%
+  filter(compound %in% c("caffeine", "cotinine", "nicotine", "salicyl", "theophyl")) %>% 
   ggplot(aes(x = pipe_notch, y = conc)) +
@@
                              TRUE ~ FALSE))
+
+replicated_accuracy <-
 
@@
   )
+```
+
+```{r}
+#| label: fig-notch-ratio
+#| fig-cap: "Relationship between ratio and number of pipe notches."
 
@@
 In the replicated samples only, tobacco detection was successful in
-`r nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0))`
+`r nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0))`
 out of
-`r nrow(filter(replicated_accuracy, pipe_notch == "Y"))`
+`r nrow(filter(uhplc_accuracy, pipe_notch == "Y"))`
 pipe smokers
-(`r scales::percent(nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(replicated_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
+(`r scales::percent(nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0)) / nrow(filter(uhplc_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
 Including individuals with absence of a pipe notch and concurrent absence of
 compounds as a correct identification, gives an overall accuracy of
-`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.
+`r scales::percent(mean(uhplc_accuracy$correct, na.rm = T), accuracy = 0.1)`.
 
@@
 
+Some of the teeth were missing because they have been sent elsewhere for DNA
+sampling. These teeth were considered present when determining antemortem loss
+ratios, and absent when scoring caries, periodontitis, and calculus.
+
 ```{r}
@@
 dental_long %>%
+  mutate(status = case_when(status == "dna" ~ "p")) %>% 
   dental_plot(fill = status)
@@
 
-Brief statement about amtl at site.
-
-```{r}
+Ratio of antemortem lost teeth per present teeth at the site.
+
+```{r}
+#| label: fig-aml-id
+#| fig-cap: "AMTL ratio per individual."
 dental_long %>% 
@@
 
+```{r}
+#| label: fig-aml-class
+#| fig-cap: "AMTL ratio per tooth class."
+dental_long %>% 
+  group_by(class) %>% 
+  amtl_ratio(.status = status, .add = T)
+```
+
+```{r}
+#| label: fig-aml-type
+#| fig-cap: "AMTL ratio per tooth type."
+dental_long %>% 
+  group_by(type) %>% 
+  amtl_ratio(.status = status, .add = T)
+```
+
 ### Caries
@@
 <!--add a caries correction?-->
-Caries were scored as the location on each individual tooth. Multiple locations on
-a single tooth were separated with `;`. The size of caries was also
+Caries were scored as the location on each individual tooth. Multiple locations
+on a single tooth were separated with `;`. The size of caries was also
 recorded, but not used in further analysis. Large caries that cover multiple
-surfaces with an unknown origin were recorded as 'crown'. Caries counts were
-extracted based on the number of locations recorded.
+surfaces with an unknown origin were recorded as 'crown'.
 
@@
     geom_violin(aes(fill = class)) +
-    geom_boxplot(width = 0.1)
-```
-
-```{r}
+    geom_boxplot(width = 0.1) +
+    theme_minimal()
+```
+
+```{r}
+#| label: fig-caries-ratio
 upper_order <- c(paste0("t", 18:11), paste0("t", 21:28))
@@
 
-```{r}
-#| label: fig-periodont
-dental_long %>%
-  filter(!is.na(periodont)) %>% 
-  dental_plot(fill = periodont)
-```
+Periodontitis was scored qualitatively on a scale from 0-3 as the amount of
+horizontal bone loss from the CEJ to the alveolar bone, accounting for ca. 2mm
+of gingival thickness.
 
@@
 ## Pathological conditions
+
+Pathological conditions and lesions that occur frequently in the population were
+included in the analysis. Data were
+dichotomised to presence/absence to allow statistical analysis. A conservative
+approach was taken, so when in doubt, absence of a disease was assumed.
+Osteoarthritis was considered present in cases where eburnation was visible
+on one or more joint surfaces.
+Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
+formation on the margin of the superior and inferior surfaces of the vertebral
+body.
+Cribra orbitalia was diagnosed based on the presence of pitting on the superior
+surface of the orbit. No distinction was made between active or healing lesions.
+Degenerative disc disease, or spondylosis, is identified as a large diffuse
+depression of the
+superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
+Schmorl's nodes are identified as any cortical depressions on the surface of
+the vertebral body. A note was made whether the lesion perforated the vertebral
+margin, but both perforating and non-perforating lesions were recorded as present.
 

### 2023-02-17 analysis/scripts/setup-qmd.R

@@ -82,7 +82,7 @@ uhplc_conc_wide <- uhplc_calculus_long %>%
   # mutate(presence = case_when(presence > 0 ~ TRUE,
   #                             TRUE ~ FALSE)) %>%
   select(
-    !c(presence, quant, weight, occupation, own_grave, tax, pipe_notch, sex, age, preservation, completeness)
+    !c(presence, quant, weight, pipe_notch, sex, age, preservation, completeness)
     ) %>%
   remove_missing() %>%
   pivot_wider(names_from = "compound", values_from = "conc") %>%
@@ -93,7 +93,7 @@ uhplc_conc_wide <- uhplc_calculus_long %>%
 
 dental_inv_long <- dental_inv %>% # Inventory
   right_join(demography) %>%
-  select(!c(occupation, own_grave, tax, pipe_notch)) %>%
+  select(!c(pipe_notch)) %>%
   pivot_longer(t11:t48, names_to = "tooth",
                values_to = "status") #%>% no reason to recode dna...
   #mutate(status = case_when(status == "dna" ~ "m", # teeth missing due to DNA sampling recoded as missing. WILL affect aml calculations...

### 2023-02-17 R/generate-bib.R
add function to generate bibliography

@@ -4,10 +4,10 @@
 #' @importFrom here here
 
 generate_bib <- function(){
-  cite_keys <- bbt_detect_citations(list.files(here("inst/paper/"), "*.qmd", full.names = T))
+  cite_keys <- bbt_detect_citations(list.files(here("analysis/paper/"), "*.qmd", full.names = T))
   ignore_keys <- grep("fig-|tbl-", cite_keys, value = T)
   bbt_write_bib(
-    here("inst/paper/references.bib"),
+    here("analysis/paper/references.bib"),
     keys = cite_keys,
     ignore = ignore_keys,
     overwrite = T

### 2023-02-17 README.Rmd
update binder link

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/mb11CalculusPilot/main?urlpath=rstudio)
 
@@
 
-  - [:file\_folder: paper](/inst/paper): Quarto source document
+  - [:file\_folder: paper](/analysis/paper): Quarto source document
     for manuscript. Includes code to reproduce the figures and tables
@@
     and tables in this file)
-  - [:file\_folder: data](/inst/data): Data used in the analysis.
-  - [:file\_folder: figures](/inst/figures): Plots and other
+  - [:file\_folder: data](/analysis/data): Data used in the analysis.
+  - [:file\_folder: figures](/analysis/figures): Plots and other
     illustrations
   - [:file\_folder:
-    supplementary-materials](/inst/supplementary-materials):
+    supplementary-materials](/analysis/supplementary-materials):
     Supplementary materials including notes and other documents

### 2023-02-17 analysis/paper/paper.qmd
add Zenodo links

@@
 Analysis scripts, and the source code for the manuscript and
-supplementary materials are available as a research compendium<!--Zenodo link-->
+supplementary materials are available as a research compendium
+(<https://doi.org/10.5281/zenodo.7649825>)
 using the structure recommended by the **rrtools** R package [@Rrrtools].

### 2023-02-17 README.Rmd
update README with preprint info

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population"
 ```
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/mb11CalculusPilot/main?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/mb11CalculusPilot/main?urlpath=rstudio) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.7649825.svg)](https://zenodo.org/badge/DOI/10.5281/zenodo.7649825)
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Bartholdy, B.P., Hasselstrøm, J.B., Sørensen, L.K., Casna, M., Hoogland, M., Historisch Genootschap Beemster, & Henry, A.G.
+(2023).
+_`r Title`_.
+Zenodo, Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/10.5281/zenodo.7649151>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Bartholdy, B.P., Hasselstrøm, J.B., Sørensen, L.K., Casna, M., Hoogland, M., Historisch Genootschap Beemster, & Henry, A.G., (`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/10.5281/zenodo.7649825>
 
@@
 You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:
+[main.zip](/archive/main.zip). After unzipping:
 - open the `.Rproj` file in RStudio
@@
 [DESCRIPTION](/DESCRIPTION) file).
-- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console
+- finally, open `analysis/paper/paper.qmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
 

### 2023-05-17 analysis/supplementary-materials/supp-mat.qmd
clean analysis code

@@
 # upload data
-metadata <- read_tsv(here("analysis/data/raw_data/metadata.tsv"))
-demography <- read_csv(here("analysis/data/raw_data/demography.csv"))
-lloq <- read_tsv(here("analysis/data/raw_data/lloq.tsv"))
-uhplc_data_comb <- read_csv(here("analysis/data/derived_data/uhplc-data_combined.csv"))
-#uhplc_calculus <- read_csv(here("analysis/data/derived_data/uhplc-calculus_cleaned.csv"))
-dental_inv <- read_csv(here("analysis/data/raw_data/dental-inv.csv"))
-caries <- read_csv(here("analysis/data/raw_data/caries.csv"))
-periodont <- read_csv(here("analysis/data/raw_data/periodontitis.csv"))
-periap <- read_csv(here("analysis/data/raw_data/periapical.csv"))
-calculus <- read_csv(here("analysis/data/raw_data/calculus.csv"))
-calculus_full <- read_csv(here("analysis/data/raw_data/calculus_full.csv"))
-sinusitis_clean <- read_csv(here("analysis/data/derived_data/sinusitis_cleaned.csv"))
-path_cond_clean <- read_csv(here("analysis/data/derived_data/path-conditions_cleaned.csv"))
+# metadata <- read_tsv(here("analysis/data/raw_data/metadata.tsv"))
+# demography <- read_csv(here("analysis/data/raw_data/demography.csv"))
+# lloq <- read_tsv(here("analysis/data/raw_data/lloq.tsv"))
+# uhplc_data_comb <- read_csv(here("analysis/data/derived_data/uhplc-data_combined.csv"))
+# #uhplc_calculus <- read_csv(here("analysis/data/derived_data/uhplc-calculus_cleaned.csv"))
+# dental_inv <- read_csv(here("analysis/data/raw_data/dental-inv.csv"))
+# caries <- read_csv(here("analysis/data/raw_data/caries.csv"))
+# periodont <- read_csv(here("analysis/data/raw_data/periodontitis.csv"))
+# periap <- read_csv(here("analysis/data/raw_data/periapical.csv"))
+# calculus <- read_csv(here("analysis/data/raw_data/calculus.csv"))
+# calculus_full <- read_csv(here("analysis/data/raw_data/calculus_full.csv"))
+# sinusitis_clean <- read_csv(here("analysis/data/derived_data/sinusitis_cleaned.csv"))
+# path_cond_clean <- read_csv(here("analysis/data/derived_data/path-conditions_cleaned.csv"))
 
@@
 
-## Session information
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r}
-print(sessionInfo(), locale = F)
-```
-
-## Data dictionary
-
-All raw data are available for download from Zenodo.
-
-### metadata.csv
-
-<!-- id	sample	element	KZ_sample	KZ_element	replicated	batch1_weight	batch2_weight -->
-
-| variable | description |
-|---|---|
-| id | unique identifier for the individual |
-| sample | sample number for the UHPLC-MS/MS analysis |
-| element | which tooth (FDI notation) was sampled for UHPLC-MS/MS analysis |
-| KZ_element | which tooth (FDI notation) was sampled for aDNA in original study |
-| replicated | Whether the sample was included in the replication batch (TRUE/FALSE) |
-| batch1_weight | weight (mg) of sample in batch 1 |
-| batch2_weight | weight (mg) of sample in batch 2 |
-
-### dental-inv.csv
-
-Dental inventory
-
-| variable | description |
-|---|---|
-| id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and status |
-
-Dental inventory key:
-
-- p = present
-- m = missing (for unknown reason - likely postmortem loss)
-- aml = ante-mortem loss
-- dna = removed for DNA sampling
-
-### caries.csv
-
-Caries lesions location
-
-| variable | description |
-|---|---|
-| id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and location of caries |
-
-Caries location key:
-
-- none = no caries present
-- mes = mesial
-- dis = distal
-- lin = lingual
-- buc = buccal (and labial)
-- occ = occlusal
-- crown = crown (large caries lesion covering multiple surfaces)
-- root = root
-- blank = tooth not present
-
-### periodontitis.csv
-
-| variable | description |
-|---|---|
-| id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and periodontitis score (0-3) |
-
-Periodontitis score:
-
-- 0 = none
-- 1 = slight
-- 2 = moderate
-- 3 = severe
-
-- blank = tooth not present
-
-### periapical.csv
-
-| variable | description |
-|---|---|
-| id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and periapical lesion |
-
-- none = no lesion
-- bucc = buccal
-- lin = lingual
-- perf = perforated alveolar bone
-- blank = not scoreable
-
-### calculus_full.csv
-
-| variable | description |
-|---|---|
-| id | unique identifier for the individual |
-| t11_bucc..t48_ip | Tooth number (FDI system) and surface (bucc = buccal; lin = lingual; ip = interproximal) and deposit size (0-3) |
-
-- 0 = no calculus
-- 1 = slight calculus
-- 2 = moderate calculus
-- 3 = heavy calculus
-
-### path-conditions.csv
-
-Pathological conditions
-
-- OA = osteoarthritis
-- IVDD = intervertebral disc disease
-- TB = tuberculosis
-- DISH = diffuse idiopathic skeletal hyperostosis
-- VOP = vertebral osteophytosis
-- SN = schmorl's nodes
-- DDD = degenerative disc disease
-- PNBF = periosteal new bone formation
-- OD = osteochondritis dissecans
-- CF = cribra femora
-- CO = cribra orbitalia
-
-## Sample
+## Samples
 
@@
     ) %>%
+  arrange(id) %>% 
   knitr::kable(col.names = c("ID", "this study", "previous study"))
@@
 ```{r}
+#| label: tobacco-accuracy-setup
 tobacco <- uhplc_calculus_long %>%
@@
   group_by(id) %>% 
-  arrange(desc(presence)) %>%
-  slice(which(compound == "nicotine" || compound == "cotinine")) %>%
+  arrange(desc(presence), .by_group = T) %>%
+  distinct(id, .keep_all = T) %>% 
+  #slice(which(str_detect(compound, "nicotine|cotinine"))) #%>%
+  # slice(which(compound == "nicotine" | compound == "cotinine")) %>%
   mutate(
@@
                              TRUE ~ FALSE))
-
-replicated_accuracy <-
 
@@
 <!-- correlations on caries and calculus separated by quadrant/tooth type -->
+
+## Data dictionary
+
+All raw data are available for download from Zenodo (<https://doi.org/10.5281/zenodo.7648757>).
+
+### lloq.tsv
+
+| variable | description |
+|---|---|
+| compound | name of target compound |
+| lloq | Lower limit of quantitation |
+
+### metadata.csv
+
+<!-- id	sample	element	KZ_sample	KZ_element	replicated	batch1_weight	batch2_weight -->
+
+| variable | description |
+|---|---|
+| id | unique identifier for the individual |
+| sample | sample number for the UHPLC-MS/MS analysis |
+| element | which tooth (FDI notation) was sampled for UHPLC-MS/MS analysis |
+| KZ_element | which tooth (FDI notation) was sampled for aDNA in original study |
+| replicated | Whether the sample was included in the replication batch (TRUE/FALSE) |
+| batch1_weight | weight (mg) of sample in batch 1 |
+| batch2_weight | weight (mg) of sample in batch 2 |
+
+### dental-inv.csv
+
+Dental inventory
+
+| variable | description |
+|---|---|
+| id | unique identifier for the individual |
+| t11..t48 | Tooth number (FDI system) and status |
+
+Dental inventory key:
+
+- p = present
+- m = missing (for unknown reason - likely postmortem loss)
+- aml = ante-mortem loss
+- dna = removed for DNA sampling
+
+### caries.csv
+
+Caries lesions location
+
+| variable | description |
+|---|---|
+| id | unique identifier for the individual |
+| t11..t48 | Tooth number (FDI system) and location of caries |
+
+Caries location key:
+
+- none = no caries present
+- mes = mesial
+- dis = distal
+- lin = lingual
+- buc = buccal (and labial)
+- occ = occlusal
+- crown = crown (large caries lesion covering multiple surfaces)
+- root = root
+- blank = tooth not present
+
+### periodontitis.csv
+
+| variable | description |
+|---|---|
+| id | unique identifier for the individual |
+| t11..t48 | Tooth number (FDI system) and periodontitis score (0-3) |
+
+Periodontitis score:
+
+- 0 = none
+- 1 = slight
+- 2 = moderate
+- 3 = severe
+
+- blank = tooth not present
+
+### periapical.csv
+
+| variable | description |
+|---|---|
+| id | unique identifier for the individual |
+| t11..t48 | Tooth number (FDI system) and periapical lesion |
+
+- none = no lesion
+- bucc = buccal
+- lin = lingual
+- perf = perforated alveolar bone
+- blank = not scoreable
+
+### calculus_full.csv
+
+| variable | description |
+|---|---|
+| id | unique identifier for the individual |
+| t11_bucc..t48_ip | Tooth number (FDI system) and surface (bucc = buccal; lin = lingual; ip = interproximal) and deposit size (0-3) |
+
+- 0 = no calculus
+- 1 = slight calculus
+- 2 = moderate calculus
+- 3 = heavy calculus
+
+### path-conditions.csv
+
+Pathological conditions
+
+- OA = osteoarthritis
+- IVDD = intervertebral disc disease
+- TB = tuberculosis
+- DISH = diffuse idiopathic skeletal hyperostosis
+- VOP = vertebral osteophytosis
+- SN = schmorl's nodes
+- DDD = degenerative disc disease
+- PNBF = periosteal new bone formation
+- OD = osteochondritis dissecans
+- CF = cribra femora
+- CO = cribra orbitalia
+
+## Session information
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r}
+print(sessionInfo(), locale = F)
+```

### 2023-05-17 analysis/scripts/setup-qmd.R

@@ -8,8 +8,30 @@ library(glue)
 library(purrr)
 library(psych)
 library(patchwork)
+library(readr)
 
+devtools::load_all()
 try(generate_bib())
+
+# upload data
+#metadata <- read_tsv(here("analysis/data/raw_data/metadata.tsv"))
+#demography <- read_csv(here("analysis/data/raw_data/demography.csv"))
+#lloq <- read_tsv(here("analysis/data/raw_data/lloq.tsv"))
+#uhplc_data_comb <- read_csv(here("analysis/data/derived_data/uhplc-data_combined.csv"))
+#dental_inv <- read_csv(here("analysis/data/raw_data/dental-inv.csv"))
+#caries <- read_csv(here("analysis/data/raw_data/caries.csv"))
+#periodont <- read_csv(here("analysis/data/raw_data/periodontitis.csv"))
+#periap <- read_csv(here("analysis/data/raw_data/periapical.csv"))
+#calculus <- read_csv(here("analysis/data/raw_data/calculus.csv"))
+#calculus_full <- read_csv(here("analysis/data/raw_data/calculus_full.csv"))
+#sinusitis_clean <- read_csv(here("analysis/data/derived_data/sinusitis_cleaned.csv"))
+#path_cond_clean <- read_csv(here("analysis/data/derived_data/path-conditions_cleaned.csv"))
+
+#source(here("analysis/scripts/setup-qmd.R"))
+#source(here("analysis/supplementary-materials/supp-mat.qmd"), local = knitr::knit_global())
+# set viridis as default pallette
+
+
 
 # UHPLC analysis ----------------------------------------------------------
 

### 2023-05-17 analysis/paper/_results.qmd

@@
   left_join(compound_name_abbrev, by = c("compound" = "abbrev")) %>% 
-  full_join(lloq, by = c("name" = "compound")) %>%
+  left_join(lloq, by = c("name" = "compound")) %>%
   mutate(across(batch1:batch2, ~ replace_na(.x, FALSE))) %>%

### 2023-05-17 analysis/paper/paper.qmd

@@
 library(here)
-library(readr)
-devtools::load_all()
-
-# upload data
-metadata <- read_tsv(here("analysis/data/raw_data/metadata.tsv"))
-demography <- read_csv(here("analysis/data/raw_data/demography.csv"))
-lloq <- read_tsv(here("analysis/data/raw_data/lloq.tsv"))
-uhplc_data_comb <- read_csv(here("analysis/data/derived_data/uhplc-data_combined.csv"))
-dental_inv <- read_csv(here("analysis/data/raw_data/dental-inv.csv"))
-caries <- read_csv(here("analysis/data/raw_data/caries.csv"))
-periodont <- read_csv(here("analysis/data/raw_data/periodontitis.csv"))
-periap <- read_csv(here("analysis/data/raw_data/periapical.csv"))
-#calculus <- read_csv(here("analysis/data/raw_data/calculus.csv"))
-calculus_full <- read_csv(here("analysis/data/raw_data/calculus_full.csv"))
-sinusitis_clean <- read_csv(here("analysis/data/derived_data/sinusitis_cleaned.csv"))
-path_cond_clean <- read_csv(here("analysis/data/derived_data/path-conditions_cleaned.csv"))
-
+#library(readr)
+#devtools::load_all()
 source(here("analysis/scripts/setup-qmd.R"))
-#source(here("analysis/supplementary-materials/supp-mat.qmd"), local = knitr::knit_global())
-# set viridis as default pallette
 

### 2023-05-17 analysis/paper/paper.tex
prepare paper for PCI submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{lineno}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy1,; Jørgen B. Hasselstrøm2; Lambert K. Sørensen2; Maia Casna1; Menno Hoogland1; Historisch Genootschap Beemster3; Amanda G. Henry1},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}
\author{Bjørn Peare Bartholdy\textsuperscript{1,*} \and Jørgen B. Hasselstrøm\textsuperscript{2} \and Lambert K. Sørensen\textsuperscript{2} \and Maia Casna\textsuperscript{1} \and Menno Hoogland\textsuperscript{1} \and Historisch Genootschap Beemster\textsuperscript{3} \and Amanda G. Henry\textsuperscript{1}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 60\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, children and women.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, sharp corners, enhanced]}{\end{tcolorbox}}\fi

\textsuperscript{1} Department of Archaeological Sciences, Leiden
University\\
\textsuperscript{2} Department of Forensic Medicine, Aarhus University\\
\textsuperscript{3} Historisch Genootschap Beemster

\textsuperscript{*} Correspondence:
\href{mailto:b.p.bartholdy@arch.leidenuniv.nl}{Bjørn Peare Bartholdy
\textless{}b.p.bartholdy@arch.leidenuniv.nl\textgreater{}}

Keywords: dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Dental calculus has proven to be an excellent source of a wide variety
of information about our past. The increased accessibility and
advancement of methods in aDNA, paleoproteomics, and mass spectrometry,
has expanded our ability to identify biomarkers of diet and disease on
an increasingly large scale (Gismondi et al., 2020; Velsko et al., 2017;
Warinner et al., 2014).

One such collection of biomarkers is alkaloids, a plant-derived group of
compounds. Many alkaloids have important medicinal and psychoactive
effects in humans, and their direct detection, or detection of their
metabolites, is of great interest to archaeologists. Previous studies
have successfully recovered alkaloids in archaeological contexts,
including ceramics (Smith et al., 2018), pipes (Rafferty et al., 2012),
human hair (Echeverría \& Niemeyer, 2013; Ogalde et al., 2009), and even
dental calculus employing both targeted (Eerkens et al., 2018) and
untargeted approaches (Buckley et al., 2014; Gismondi et al., 2020).
Especially nicotine, the principal alkaloid in tobacco leaves, has been
widely studied in the archaeological record due to its apparent
stability and ability to survive over long periods of time (Eerkens et
al., 2018; Rafferty et al., 2012; Tushingham et al., 2013).

Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through oral consumption of alkaloid-containing plants,
whether deliberate or accidental; and (2) passive diffusion as alkaloids
and other compounds are transferred from plasma to saliva, and then into
the oral cavity through the salivary glands in the hours to days
following consumption (Cone \& Huestis, 2007). The relation to plasma is
why there is often a close correlation between presence (not
concentration) of drugs in oral fluid and blood (Cone \& Huestis, 2007;
Milman et al., 2011; Wille et al., 2009). The second pathway allows the
identification of parent compounds that are not consumed orally, as long
as they, or their metabolites, are excreted through saliva. .

Many of the components involved in the formation and growth of dental
calculus originate from oral fluid. Proteins, bacteria, salts and other
compounds are transferred from saliva to biofilms on the tooth surface
(Jin \& Yip, 2002; White, 1997). This may also allow various alkaloids
of dietary and medicinal origin to become incorporated in dental plaque.
Dental plaque undergoes frequent mineralisation events, ultimately
causing the entrapped alkaloids and their metabolites to become
preserved within the dental calculus. Barring intentional or accidental
removal of the calculus during life, burial, excavation, and
post-excavation cleaning, the alkaloids can then be detected by various
methods to show a record of consumption during life.

In this study we use a ultra-high-performance liquid
chromatography-tandem mass spectrometry (UHPLC-MS/MS) method that was
developed in a previous study on dental calculus from cadavers and
validated by comparing the results to compounds detected in the blood of
the same individuals (Sørensen et al., 2021). All compounds that were
detected in the blood were also detected in dental calculus, with
additional compounds present in dental calculus that were not present in
blood, suggesting that dental calculus represents a comprehensive
history of consumption over a long period of time (Sørensen et al.,
2021). We were able to detect both parent compounds and metabolites,
including caffeine, nicotine, theophylline, and cotinine, in the dental
calculus of individuals from a 19th century Dutch population from
Middenbeemster. By detecting these compounds we are able to show the
consumption of tea and coffee and smoking of tobacco on an individual
scale, which is also confirmed by historic documentation and
identification of pipe notches in the dentition.

\hypertarget{materials}{%
\subsection{Materials}\label{materials}}

The sample consists of 41 individuals from Middenbeemster, a 19th
century rural Dutch site. The village of Middenbeemster and the
surrounding Beemsterpolder was established in the beginning of the 17th
century, when the Beemster lake was drained to create more farmland,
mainly for the cultivation of cole seeds (de Vries 1978). In 1615, a
decision was made to build a church, and construction started in 1618
(Hakvoort 2013). The excavated cemetery is associated with the
Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
(Lemmers et al., 2013). Archival documents are available for those
buried between AD 1829 and 1866, when the majority of individuals were
interred (Palmer et al., 2016). The main occupation of the inhabitants
was dairy farming, consisting largely of manual labour prior to the
industrial revolution (Aten et al., 2012; Palmer et al., 2016).

To reduce the number of potentially confounding factors to account for
in the analysis, we preferentially selected males from the middle adult
age category (35-49 years). The sample consists of 27 males, 11 probable
males, 2 probable females, and 1 female
(Figure~\ref{fig-sample-demography}). We selected males due to a higher
occurrence of pipe notches and dental calculus deposits than females
(unpublished observation).

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-sample-demography-1.pdf}

}

\caption{\label{fig-sample-demography}Overview of sample demography.
Left plot is the first batch and right plot is the replication batch
with 29 of the individuals from the first batch. eya = early young adult
(18-24 years); lya = late young adult (25-34 years); ma = middle adult
(35-49 years); old = old adult (50+ years). Male? = probable male;
Female? = probable female.}

\end{figure}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\hypertarget{skeletal-analysis}{%
\subsubsection{Skeletal analysis}\label{skeletal-analysis}}

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits (Buikstra \& Ubelaker, 1994).
Age-at-death was estimated using dental wear, auricular and pubic
surface appearance, cranial suture closure, and epiphyseal fusion
(Brooks \& Suchey, 1990; Buckberry \& Chamberlain, 2002; Buikstra \&
Ubelaker, 1994; Lovejoy et al., 1985; Meindl \& Lovejoy, 1985), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult ( 35-49 years), old adult
(50+ years).

\hypertarget{paleopathology}{%
\paragraph{Paleopathology}\label{paleopathology}}

Pathological conditions and lesions that occur frequently in the
population were included in the analysis. Data were dichotomised to
presence/absence to allow statistical analysis. Osteoarthritis was
considered present in cases where eburnation was visible on one or more
joint surfaces. Vertebral osteophytosis is identified by marginal
lipping and/or osteophyte formation on the margin of the superior and
inferior surfaces of the vertebral body. Cribra orbitalia was diagnosed
based on the presence of pitting on the superior surface of the orbit.
No distinction was made between active or healing lesions. Degenerative
disc disease, or spondylosis, is identified as a large diffuse
depression of the superior and/or inferior surfaces of the vertebral
body (Rogers, 2000). Schmorl's nodes are identified as any cortical
depressions on the surface of the vertebral body. Data on chronic
maxillary sinusitis from Casna et al. (2021) were included in this study
to assess the relationship between upper respiratory diseases with
environmental factors (i.e.~tobacco smoke, caffeine consumption).
Lesions associated with chronic maxillary sinusitis as defined by
Boocock et al. (1995) were recorded for each individual and classified
as ``pitting'', ``spicule-type bone formation'', ``remodeled spicules'',
or ``white pitted bone''. chronic maxillary sinusitis was scored as
absent when the sinus presented smooth surfaces with little or no
associated pitting.

\hypertarget{dental-pathology}{%
\paragraph{Dental pathology}\label{dental-pathology}}

Caries ratios were calculated by dividing the number of lesions by the
number of teeth scored, resulting in a single caries ratio per
individual. If the surface where the lesion originated is not visible,
i.e.~if the lesion covered multiple surfaces, this was scored as
``crown''. Calculus indices were calculated according to Greene and
colleagues (2005). Calculus was scored with a four-stage scoring system
(0-3) to score absent, slight, moderate, and heavy calculus deposits
(Brothwell, 1981) on the lingual, buccal (and labial), and interproximal
surfaces of each tooth. Only one score was used for the combined
interproximal surfaces, resulting in three scores per tooth (when
surfaces are intact), and four calculus indices per individual; upper
anterior, upper posterior, lower anterior, lower posterior. Each index
was calculated by dividing the sum of calculus scores for each surface
by the total number of surfaces scored in each quadrant. If a tooth
could not be scored on all three surfaces, the tooth was not included
(Greene et al., 2005). Periodontitis was scored on a visual four-stage
(0-3) scoring system according to distance from cemento-enamel junction
of each tooth to alveolar bone (Maat \& Mastwijk, 2005).

\hypertarget{calculus-sampling}{%
\subsubsection{Calculus sampling}\label{calculus-sampling}}

Where possible, we used material that had already been sampled for a
previous study to prevent unnecessary repeated sampling of individuals.
Calculus from the previous study was sampled in a dedicated ancient DNA
laboratory at the Laboratories of Molecular Anthropology and Microbiome
Research in Norman, Oklahoma, U.S.A, using established ancient DNA
protocols. More details on the methods can be found in the published
articles (Ziesemer et al., 2015, 2018). Of the 41 individuals that were
originally included in our sample, 29 were replicated in a separate
analysis only using calculus from the previous study.\\
New dental calculus samples were taken under sterile conditions in a
positive pressure laminar flow hood in a dedicated dental calculus lab
at Leiden University. The surface of the tooth was lightly brushed with
a sterile, disposable toothbrush to get rid of surface contaminants. A
sterile dental curette was then used to scrape calculus from the tooth
onto weighing paper, which was transferred to 1.5 ml Eppendorf tubes.
All calculus samples were sent to the Department of Forensic Medicine at
Aarhus University for ultra-high-performance liquid
chromatography-tandem mass spectrometry (UHPLC-MS/MS) analysis.

\hypertarget{uhplc-msms}{%
\subsubsection{UHPLC-MS/MS}\label{uhplc-msms}}

The list of targeted compounds included both naturally occurring
compounds known to have been used in the past, as well as synthetic
modern drugs that did not exist at the time (e.g.~Fentanyl, MDMA,
Amphetamine). These were part of the toxicology screening for the
original method (Sørensen et al., 2021), developed on cadavers. In our
study they serve as an authentication step, as their presence in
archaeological samples could only be the result of contamination.

Briefly, samples of dental calculus were washed three times each with
one mL of methanol (MeOH), to remove surface contaminants. The wash
solutions were collected separately. The solvent was evaporated and the
residues were dissolved in 50 µL 30\% MeOH. The washed calculus was
homogenized in presence of 0.5 M citric acid using a lysing tube with
stainless steel beads. Following one hour of incubation the dissolution
extract was cleaned by weak and strong cation-exchange. After
evaporation of the elution solvent the residue was dissolved in 50 µL
30\% MeOH. The final extracts obtained from washing and dissolution of
the dental calculus were analysed by UHPLC-MS/MS using a reversed-phase
biphenyl column for chromatography. To obtain quantitative results,
isotope dilution was applied. For more details about the method and
validation, see the original study by Sørensen and colleagues (2021).

\hypertarget{statistical-analysis}{%
\subsubsection{Statistical analysis}\label{statistical-analysis}}

All compounds and pathological conditions/lesions were converted to a
presence/absence score. Pearson product-moment correlation was applied
to the dichotomised pathological lesions (point-biserial correlation),
compound concentrations, calculus indices, and caries ratios to explore
relationships paired continuous-continuous variables and paired
continuous-binary variables. Compound concentrations were then
dichotomised to presence/absence, and the caries ratio and calculus
index for each individual were converted to an ordinal score from 0 to 4
by using quartiles. Polychoric correlation was applied to the paired
dichotomous variables and dichotomous-ordinal variables.

All statistical analysis was conducted in R version 4.3.0 (2023-04-21),
Already Tomorrow, (R Core Team, 2020). Data wrangling was conducted with
the \textbf{tidyverse} (Hadley Wickham et al., 2019) and visualisations
were created using \textbf{ggplot2} (H. Wickham, 2016). Polychoric
correlations were calculated with the \textbf{psych} package (Revelle,
2022).

\hypertarget{results}{%
\subsection{Results}\label{results}}

Multiple compounds were detected in the dental calculus samples.
Compounds detected at a lower concentration than the lower limit of
quantitation (LLOQ) were considered not present. Not all the compounds
detected in the first batch could be replicated in the second batch
(Table~\ref{tbl-compound-detect}). For a full list of targeted
compounds, see Supplementary Material.

\hypertarget{tbl-compound-detect}{}
\begin{longtable}[]{@{}lllr@{}}
\caption{\label{tbl-compound-detect}Target compound including whether it
was detected (TRUE) or not (FALSE) in each batch, as well as the lower
limit of quantitation (LLOQ) in ng. CBD = cannabidiol; CBN = cannabinol;
THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A;
THCVA = tetrahydrocannabivarin acid.}\tabularnewline
\toprule()
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule()
\endfirsthead
\toprule()
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule()
\endhead
CBD & TRUE & FALSE & 0.050 \\
CBN & TRUE & FALSE & 0.050 \\
Caffeine & TRUE & TRUE & 0.050 \\
Cocaine & TRUE & FALSE & 0.025 \\
Cotinine & TRUE & TRUE & 0.050 \\
Nicotine & TRUE & TRUE & 0.100 \\
Salicylic acid & TRUE & TRUE & 0.500 \\
THC & TRUE & FALSE & 0.100 \\
THCA-A & TRUE & FALSE & 0.025 \\
THCVA & TRUE & FALSE & 0.010 \\
Theophylline & TRUE & TRUE & 0.010 \\
\bottomrule()
\end{longtable}

The pattern we expect to see in authentic compounds representing
compounds trapped within the dental calculus, is a reduction in the
quantity from wash 1 to wash 3 as potential surface contaminants are
washed off, and then a spike in the final extraction when entrapped
compounds are released and detected.

Most plots show a large increase in extracted mass of a compound between
the calculus wash extracts (wash 1-3) and the dissolved calculus (calc).
Most samples containing theophylline and caffeine had the largest
quantity of the compound extracted from the first wash, then decreasing
in washes 2 and 3. There is an increase between wash 3 and the dissolved
calculus in all samples. The patterns are consistent across batches 1
and 2. Nicotine and cotinine have the same relative quantities in the
samples, i.e., the sample with the highest extracted quantity of
nicotine also had the highest extracted quantity of cotinine
Figure~\ref{fig-auth-plot-batch2}.

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-auth-plot-batch2-1.pdf}

}

\caption{\label{fig-auth-plot-batch2}(A) Number of samples in which each
compound was detected in the first and second batch. (B) Quantity (ng)
of each compound extracted from each sample in batch 2. The plot
displays the extracted quantity across the three washes and final
calculus extraction (calc). Each coloured line represents a different
calculus sample. CBD = cannabidiol; CBN = cannabinol; THC =
tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA =
tetrahydrocannabivarin acid.}

\end{figure}

To see if preservation of the skeletal remains had any effect on the
detection of compounds, we compare extracted quantities of compounds to
the various levels of skeletal preservation. Our results from batch 2
suggest that detection of a compound may be linked to the preservation
of the skeleton, with better preservation leading to increased
extraction quantity (Figure~\ref{fig-detection-preservation}A). We also
find a weak positive correlation between the weight of the calculus
sample and the quantity of compound extracted from the calculus
(Figure~\ref{fig-detection-preservation}B).

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-detection-preservation-1.pdf}

}

\caption{\label{fig-detection-preservation}(A) Violin plot with overlaid
box plots depicting the distribution of extracted quantities of each
compound from batch 2 separated by state of preservation of the
skeleton. (B) Extracted quantity (ng) of compound plotted against
weights of the calculus samples from batch 2.}

\end{figure}

The presence of pipe notch(es) in an individual and concurrent detection
of nicotine and/or cotinine is used as a crude indicator of the accuracy
of the method. Only males were used in accuracy calculations, as pipe
notches are ubiquitous in males, but not in females. In batch 2, the
method was able to detect some form of tobacco in 14 of 25 individuals
with a pipe notch (56.0\%). When also considering correct the absence of
a tobacco alkaloid together with the absence of a pipe notch, the
accuracy of the method is 59.3\%. Accuracy in the old adult age category
is 100.0\%, but with only 2 individuals.

One individual---an old adult, probable female---was positive for both
nicotine and cotinine, and had no signs of a pipe notch.

\hypertarget{correlations-between-detected-alkaloids-and-diseases}{%
\subsubsection{Correlations between detected alkaloids and
diseases}\label{correlations-between-detected-alkaloids-and-diseases}}

For further statistical analyses, only the UHPLC-MS/MS results from
batch 2 were used, as batch 1 had multiple compounds that were not
detected in batch 2 and may have been contaminated.

\hypertarget{tbl-pearson}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1566}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1566}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}@{}}
\caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
dichotomous skeletal lesions and compound concentrations (ng/mg) from
the second batch. Correlations between pairs of dichotomous variables
are removed due to incompatibility with a Pearson correlation. OA =
osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD
= degenerative disc disease; CO = cribra orbitalia; CMS = chronic
maxillary sinusitis; SA = salicylic acid; PN = pipe
notches.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Nicotine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
SA
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Calculus
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PN
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Theophylline
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Caffeine
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cotinine
\end{minipage} \\
\midrule()
\endhead
OA & -0.19 & -0.074 & 0.21 & 0.07 & 0.14 & 0.28 & 0.00098 & -0.067 \\
VOP & -0.061 & -0.16 & 0.34 & 0.061 & 0.25 & -0.06 & 0.013 & -0.13 \\
SN & -0.22 & 0.16 & 0.095 & 0.089 & 0.17 & 0.24 & 0.16 & 0.093 \\
DDD & 0.032 & 0.0037 & 0.19 & -0.39 & -0.077 & 0.31 & 0.06 & -0.0086 \\
CO & 0.14 & -0.051 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.065 \\
CMS & -0.18 & 0.28 & 0.0017 & -0.27 & 0.032 & 0.19 & 0.36 & 0.22 \\
Caries & & -0.13 & -0.27 & -0.19 & -0.037 & -0.16 & 0.079 & -0.16 \\
Nicotine & & & -0.21 & 0.01 & -0.014 & 0.43 & 0.14 & 0.98 \\
SA & & & & 0.14 & 0.37 & 0.038 & 0.17 & -0.17 \\
Calculus & & & & & 0.13 & -0.15 & -0.13 & 0.031 \\
PN & & & & & & -0.16 & 0.18 & -0.0068 \\
Theophylline & & & & & & & 0.51 & 0.36 \\
Caffeine & & & & & & & & 0.078 \\
\bottomrule()
\end{longtable}

Point-biserial correlation was conducted on paired continuous and
dichotomous variables, to see if any relationships exist between
extracted concentrations and other variables. The strongest
point-biserial (Pearson) correlation correlations were a near-perfect
positive correlation between cotinine and nicotine (0.982), and moderate
correlations between theophylline and nicotine (0.432), caffeine and
theophylline (0.507) (Table~\ref{tbl-pearson}).

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.851). Moderate correlations were
found between OA and DDD (0.484), VOP and periodontitis (0.491), SN and
cotinine (0.558), DDD and calculus (-0.421), CMS and caffeine (0.528),
caries and periodontitis (0.486), caries and theophylline (-0.491),
periodontitis and caries (0.486), periodontitis and caffeine (0.515),
nicotine and CMS (0.496), calculus and caries (0.502), age-at-death and
theophylline (-0.447), theophylline and age-at-death (-0.447), caffeine
and periodontitis (0.515), cotinine and CMS (0.425). Remaining
correlations were weak or absent (Figure~\ref{fig-polycorr}).
Correlations with age will be depressed because age was largely
controlled for in the sample selection.

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-polycorr-1.pdf}

}

\caption{\label{fig-polycorr}Plot of the polychoric correlations
(\emph{rho}). Larger circles and increased opacity indicates a stronger
correlation coefficient. OA = osteoarthritis; VOP = vertebral
osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO
= cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic
acid.}

\end{figure}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

In this study we were able to extract and identify multiple alkaloids
and salicylic acid from the dental calculus of individuals from
Middenbeemster, a 19th century Dutch archaeological site. We applied
ultra-high-performance liquid chromatography-tandem mass spectrometry
(UHPLC-MS/MS), a method that was validated by co-occurrence of drugs and
metabolites in dental calculus and blood (Sørensen et al., 2021). Here
we have shown that the method can also be successfully applied to
archaeological dental calculus. We extend findings from previous studies
on alkaloids in archaeological samples by extracting multiple different
alkaloids from dental calculus, including nicotine, cotinine, caffeine,
theophylline, and salicylic acid in multiple individuals. The detection
of these compounds was solidified in a replication analysis on different
samples from the same individuals. Cocaine and multiple cannabinoids
were also detected during the first analysis, but were not replicated.
We discuss the implications of these findings in light of historical and
archaeological evidence for the consumption of these drugs.

Nicotine and its principal/main metabolite, cotinine, were strongly
positively correlated, both in concentration and presence/absence in
individuals (Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). The
detection of nicotine and cotinine is not surprising, as pipe-smoking in
the Beemsterpolder is well-documented in the literature (Aten et al.,
2012; Bouman, 2017), and visible on the skeletal remains as pipe notches
(Lemmers et al., 2013). There is also documented medicinal use of
nicotine in the Beemsterpolder, where a tobacco-smoke enema was used for
headaches, respiratory problems, colds, and drowsiness from around 1780
to 1830 (Aten et al., 2012). In our sample, we also detected nicotine
and cotinine (replicated) in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the
dental calculus through pipe-smoking, as the individual had no visible
pipe notches; more likely the tobacco entered through an alternate mode
of consumption, secondhand smoke, or the aforementioned tobacco-smoke
enema.

Theophylline and caffeine were positively correlated in our samples,
though to a lesser extent than nicotine and cotinine, so we are unable
to determine if they originated from the same source
(Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). Caffeine and
theophylline have very similar chemical structures, so we expect they
would experience similar rates of incorporation and degradation,
allowing us to interpret the ratio and correlations between the
compounds. Caffeine is present in coffee, tea, and cocoa beans, with
concentrations slightly higher in coffee (Bispo et al., 2002; Chin et
al., 2008; Srdjenovic et al., 2008; Stavric et al., 1988). Theophylline
is present in both coffee beans and tea leaves, but in negligible
quantities (Stavric et al., 1988). It is also a primary metabolite of
caffeine produced by the liver. Given the low correlation, there are
likely multiple sources of caffeine and theophylline in the population,
with tea and coffee being the most obvious.\\
Tea consumption had become widespread in the Netherlands by 1820,
reaching all parts of society (Nierstrasz, 2015, p. 91). Historically,
we also know that both tea and coffee were consumed in the
Beemsterpolder during the 19th century. `Theegasten' (teatime) was a
special occasion occurring from 15.00-20.00 hours, where tea was served
along with the evening bread (Schuijtemaker, 2011). Many households also
owned at least one coffee pot and tea pot (Bouman, 2017). Distinguishing
between tea, coffee, and chocolate may be possible by also including
theobromine and comparing ratios of the compounds, as theobromine is
present in higher quantities in chocolate compared to caffeine and
theophylline (Alañón et al., 2016; Bispo et al., 2002; Stavric et al.,
1988). However, In addition to oral factors affecting alkaloid uptake in
dental calculus, there is some indication that theobromine does not
preserve well in the archaeological record (Velsko et al., 2017), and
frequent consumption of all three items would be difficult to parse.

Salicylic acid was found in all but one individual in our sample. It can
be extracted from the bark of willow trees, \emph{Salix alba}, and has
long been used for its pain-relieving properties (Bruinsma, 1872, p.
119). It is also present in many plant-based foods (Duthie \& Wood,
2011; Malakar et al., 2017), including potatoes, which were a staple of
the Beemsterpolder diet (Aten et al., 2012). The extracted quantity from
our samples decreased over the three washes, followed by a sharp
increase in the final calculus extraction, which is what we would expect
to see if the salicylic acid was incorporated during life
Figure~\ref{fig-auth-plot-batch2}. However, it has been shown that
salicilyc acid is a very mobile organic acid and the ubiquitous presence
may be due to environmental contamination, which would also explain the
high quantity in the washes (Badri \& Vivanco, 2009; Chen et al., 2001).
Given the multiple plausible sources of this residue, it will be
necessary to explore the extent to which salicylic acid can leach into
the dental calculus from the soil, and what the rate of degradation is
for salicylic acid when trapped in dental calculus.

Cannabinoids---specifically THC, THCA-A, THCVA, CBD, CBN---were found in
the first batch, but none were replicated in the second batch. Medicinal
use of cannabinoids has been well-established in Europe since
Medieval-times, and it was also grown in the Netherlands (Bruinsma,
1872). Administration was most common in the form of concoctions
containing various portions of the cannabis plant for ingestion; not
until the late 19th century did it become recommended to smoke it for
more immediate effects (Clarke, 2013). A Dutch medicinal use of hemp
involved an emulsion prepared from the seeds of the plants to treat pain
and various stomach ailments. Another preparation involving the roots of
the plants was used for inflammation, gout, and joint pains (Clarke,
2013). The ability to detect cannabinoids in calculus may be limited by
their reduced ability to diffuse from serum to salivary glands due to an
affinity for protein-binding, (Cone \& Huestis, 2007), meaning detection
would rely on oral consumption. Even then, the overall instability of
some cannabinoids could also affect detection (Lindholst, 2010; Sørensen
\& Hasselstrøm, 2018). However, given the lack of replication, we cannot
with security confirm that cannabis was used by the Beemster population.

Despite many of our sampled individuals having lived during the height
of the opium era in the Netherlands (Macht, 1915), none of the targeted
opioids (morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
were detected. The absence of opioids could be a result of the people
ascribing more to the ``traditional'' rather than ``scientific''
medicine, although laudanum and another opium containing concoction was
part of the ``traditional'' medicine in the Netherlands (Leuw \&
Marshall, 1994), including Middenbeemster (Aten et al., 2012). It was
also generally considered a drug of the upper class (Scheltema, 1907),
and may have been more common in urban centers. The absence could also
be attributed to postmortem degradation. It has been shown that, while
abundant in opium, morphine degrades rapidly, while thebaine and
papaverine are more resistant to various ageing processes (Chovanec et
al., 2012). The latter were also absent from our samples.

The only strictly modern compound (at least in a European context)
detected in the sample was cocaine, which was detected in the first
batch of samples. Our sample is derived from an early--mid 19th century
population, and cocaine was isolated in 1860 by Albert Niemann, and
entered popular medical practice in 1884. Coca arrived in Europe as
early as 1771, but as botanical specimens rather than for consumption,
and there were also issues importing enough viable specimens of coca for
cocaine extraction (Abduca, 2019, p. 108; Mortimer, 1901, p. 179). We
considered it possible that it would be present in a sample with most
individuals originating from the early- to mid-19th century. If
corroborated, this would have been the first case of
coca-leaf-consumption in Europe. In our replication batch, we included
all of the individuals who had been cocaine-positive in the first batch.
We were unable to replicate any of the cocaine results, and we were
unable to detect the principal metabolite, benzoylecgonine, in either
batch. We suspect that the original detection of cocaine was a result of
lab contamination during analysis.

We explored the relationship between detected compounds and various
skeletal indicators, such as pathological and dental lesions,
preservation, and pipe notches. We found some evidence to suggest that
preservation of the skeleton influences the recovery of compounds from
the dental calculus, with well-preserved skeletons potentially serving
as a better target for sampling.\\
We found a positive correlation between CMS and nicotine, which may be
indicative of the impact tobacco smoking had on the respiratory health
of the Beemster inhabitants. Tobacco smoke may play a significant role
in diseases of the upper respiratory tract, including chronic maxillary
sinusities (Reh et al., 2012). Although the mechanisms by which smoking
increases the risk of infections is not fully understood, solid evidence
has been presented linking tobacco smoke to increased mucosal
permeability and impairment of mucociliary clearance (Arcavi \&
Benowitz, 2004). Such changes, together with an altered immunologic
response, are thought to predispose to the development of chronic
maxillary sinusitis (Slavin et al., 2005).\\
We also observed a moderate positive correlation between chronic
maxillary sinusitis and caffeine which contradicts previous research
linking chronic coffee consumption with a positive effect on the
respiratory system, suggesting a preventive association between caffeine
intake and pneumonia (e.g. Alfaro et al., 2018; Kondo et al., 2021).
However, while the lower respiratory tract seems to benefit from chronic
coffee consumption, it is possible that elevated caffeine intake impacts
mucosal moisture due to its dehydrating effect (Maughan \& Griffin,
2003), thereby exposing individuals to greater risk of sinus infection.

The detection of nicotine in dental calculus has previously been
presented by Eerkens and colleagues (2018) in two individuals from
pre-contact California. They also targeted caffeine, cotinine, and
theophylline in their samples, but were unable to detect any of them. It
remains to be seen whether this is due to differences in methods used,
or due to our samples being more recent. They also suggest that the
choice of tooth for sampling may impact the detection of certain
compounds, as the incorporation in dental calculus may depend on the
mode of consumption. Tobacco smokers may have more nicotine present in
calculus on incisors, whereas tobacco chewers may have more on molars
(Eerkens et al., 2018). However, sampling may not be limited to mode of
consumption. The presence of cotinine suggests that the excretion of a
compound after being metabolised in the body is also a source of
deposition, and that deposition of alkaloids in dental calculus can
occur both on the way into the body, i.e.~during consumption, and on the
way out, i.e.~disposal of waste products via saliva secretion into the
mouth. Especially mucin-rich saliva from the sublingual and
submandibular glands preferentially binds toxins (Dodds et al., 2005),
and since these glands are located closest to the lower incisors, they
may be the most effective target for these studies. This has yet to be
systematically tested in archaeological dental calculus. Because we
homogenised samples from multiple teeth of an individual, we were unable
to test the effect of oral biogeography. It is also possible that
resident microflora within biofilms contribute to alkaloid breakdown and
that the presence of caffeine and nicotine metabolites following direct
ingestion can be explained by this pathway. However, the literature on
biofilm biodegradation of alkaloids is limited, and \emph{in vitro}
studies have only found minimal contributions by certain oral bacteria
in isolation (Cogo et al., 2008; Sun et al., 2016); it is possible that
a larger role is played by oral bacteria within larger, more
metabolically active communities, e.g.~biofilms (Takahashi, 2015).

Because we targeted individuals with moderate-to-large calculus
deposits, it is likely a biased sample. The presence of calculus may
increase the risk of premature death (Yaussy \& DeWitte, 2019), and
periodontal disease (which may or may not be associated with dental
calculus build-up) is a risk-factor for respiratory diseases, if
periodontal and respiratory pathogens enter the bloodstream (Azarpazhooh
\& Leake, 2006; Scannapieco, 1999; Scannapieco \& Ho, 2001). In our
sample, the percentage of chronic maxillary sinusitis (37.0\%) is lower
than in another (more representative) male sample (44.1\%) (Casna et
al., 2021), and the caries percentage is similarly lower in our sample
(12.7\%) than a more representative sample (22.9\%) (Lemmers et al.,
2013).\\
We used the presence/absence of a pipe notch and concurrent detection of
tobacco as a crude estimate of the accuracy of the method, which we
found to be around 59.3\%. This is a very rough estimate, as the
presence of a pipe notch is likely not a perfect indicator of whether or
not someone consumed tobacco. Dental calculus is also more transient
than for example bone, as it can be mechanically removed, intentionally
or unintentionally, during life, eliminating all trace of the alkaloids
consumed prior to its removal.\\
Quantitation of the detected compounds may have limited value in
archaeological samples due to degradation, and will greatly affect our
correlations related to concentration. Following burial, compound
stability over time will play a large role, as will microbial
degradation of compounds by bacteria and fungi in soil (Liu et al.,
2015), as well as the soil environment, such as temperature, pH, and
oxygen availability (Lindholst, 2010; Mackie et al., 2017).\\
The detected quantity of a compound will also depend on the quantity in
dental calculus during life, which is largely controlled by quantity of
consumption, how often the calculus was disrupted/removed, metabolic
breakdown of the compound, and inter- and intra-individual factors
related to stages of biofilm formation, maturation, and mineralisation
(Lustmann et al., 1976; Velsko et al., 2019; Zijnge et al., 2010). In
short, this means it is not really possible to detect the absence of a
compound. The absence of a compound is not evidence of absence of
consumption. This complicates the interpretation of our results. We have
attempted to minimise errors occurring due to this limitation by
including a relatively large sample of individuals and replicating our
analysis. Although given the relatively low detection rate seen in
tobacco, this remains a major limitation, and will likely be compounded
by increasing antiquity of the samples.

Future studies should explore how sampling from various types of teeth
and their position in the mouth affects the probability of a compound
becoming entrapped in dental calculus. This may also be related to
properties within the oral cavity, as well as chemical properties of the
compounds, which facilitate or reduce the incorporation-potential, and
which incorporation pathways are more likely for a given compound.\\
We only targeted drugs that were included in the forensic toxicological
screenings, and therefore only covered a limited number of the potential
compounds that could be of interest for exploring past diets and
medicinal treatments. The list of targeted compounds can be expanded as
we discover more potential targets based on which specific
compounds/metabolites are more likely to be incorporated and preserved
in dental calculus.\\
There is an increasing interest in using oral fluid as a means of
detecting alkaloids in living individuals due to the non-invasive nature
of the testing compared to blood and urine sampling (Cone, 1993; Valen
et al., 2017). These \emph{in vivo} studies are a valuable source of
method validation and can help determine the feasibility of detecting
certain alkaloids in oral fluid and, subsequently, dental calculus.
Archaeologists, though, will likely be responsible for exploring dental
calculus specific incorporation and retention of alkaloids, as well as
their long-term preservation in the burial environment.

While a major limitation is the uncertainty surrounding whether or not a
compound is actually absent, the power of the method lies in the ability
to detect dietary and other compounds that were incorporated via
multiple consumption pathways that are not detected by other methods.
Taking tobacco consumption as an example; while pipe notches are a
useful way to identify tobacco consumption, pipe smoking was not the
only mode of tobacco consumption, with others including chewing,
drinking, cigars, and snuff (Goodman, 1994, p. 67). Pipe-smoking was
mainly practised by males (Eerkens et al., 2018; Lemmers et al., 2013),
so methods like the one presented here are suitable for exploring
tobacco consumption in an entire society, rather than a trivial subset
of past populations. Combined with other methods, it can also give us a
more complete picture of dietary patterns and medicinal/recreational
plant-use in the past by capturing multiple possible incorporation
pathways of dietary (and other) compounds.

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

This preliminary study outlines the benefits of using calculus to target
a variety of compounds that could have been consumed as medicine or
diet. This method allows us to directly address specific individuals,
which can be especially useful in individuals that are not always
well-documented in historic documentation, such as rural communities,
children and women. We also show that there are many limitations that
will need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Another
limitation of dental calculus as a medium is the inter- and
intra-individual variability of its formation and the many factors that
can influence incorporation and retention of molecules and particles;
however, in the absence of hair and serum (quite uncommon in
archaeology), dental calculus represents an impressive long-term
reservoir of information regarding the consumption of various alkaloids,
whether dietary, medicinal, recreational, or otherwise.

\hypertarget{acknowledgements}{%
\subsection*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{subsection}{Acknowledgements}

We wish to thank Kirsten Ziesemer for helping track down the calculus
samples from her studies. Additionally, we owe special thanks to Vincent
Falger and Kees de Groot from the Middenbeemster Historical Society for
their input on early draft manuscripts and a lovely guided tour of
Middenbeemster.

This research has received funding from the European Research Council
under the European Union's Horizon 2020 research and innovation program,
grant agreement number STG--677576 (``HARVEST'').

\hypertarget{data-availability-statement}{%
\subsection*{Data Availability
Statement}\label{data-availability-statement}}
\addcontentsline{toc}{subsection}{Data Availability Statement}

All raw data is available on Zenodo
(\url{https://doi.org/10.5281/zenodo.7648757}). Analysis scripts, and
the source code for the manuscript and supplementary materials are
available as a research compendium
(\url{https://doi.org/10.5281/zenodo.7649825}) using the structure
recommended by the \textbf{rrtools} R package (Marwick, 2019).

\hypertarget{conflict-of-interest-disclosure}{%
\subsection*{Conflict of interest
disclosure}\label{conflict-of-interest-disclosure}}
\addcontentsline{toc}{subsection}{Conflict of interest disclosure}

The authors have no conflicts to declare.

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-abucaCocaTrade2019}{}}%
Abduca, R. (2019). Coca leaf transfers to {Europe}. {Effects} on the
consumption of coca in {North-western Argentina}. In M. Kaller \& F.
Jacob (Eds.), \emph{Transatlantic {Trade} and {Global Cultural Transfers
Since} 1492: {More} than {Commodities}}. {Routledge}.
\url{https://books.google.com?id=13imDwAAQBAJ}

\leavevmode\vadjust pre{\hypertarget{ref-alanonAssessmentFlavanol2016}{}}%
Alañón, M. E., Castle, S. M., Siswanto, P. J., Cifuentes-Gómez, T., \&
Spencer, J. P. E. (2016). Assessment of flavanol stereoisomers and
caffeine and theobromine content in commercial chocolates. \emph{Food
Chemistry}, \emph{208}, 177--184.
\url{https://doi.org/10.1016/j.foodchem.2016.03.116}

\leavevmode\vadjust pre{\hypertarget{ref-alfaroChronicCoffee2018}{}}%
Alfaro, T. M., Monteiro, R. A., Cunha, R. A., \& Cordeiro, C. R. (2018).
Chronic coffee consumption and respiratory disease: {A} systematic
review. \emph{The Clinical Respiratory Journal}, \emph{12}(3),
1283--1294. \url{https://doi.org/10.1111/crj.12662}

\leavevmode\vadjust pre{\hypertarget{ref-arcaviCigaretteSmoking2004}{}}%
Arcavi, L., \& Benowitz, N. L. (2004). Cigarette {Smoking} and
{Infection}. \emph{Archives of Internal Medicine}, \emph{164}(20),
2206--2216. \url{https://doi.org/10.1001/archinte.164.20.2206}

\leavevmode\vadjust pre{\hypertarget{ref-aten400JaarBeemster2012}{}}%
Aten, D., Bossaers, K. W. J. M., \& Misset, C. (2012). \emph{400 jaar
Beemster: 1612-2012}. {Stichting Uitgeverij Noord-Holland}.

\leavevmode\vadjust pre{\hypertarget{ref-azarpazhoohSystematicReview2006}{}}%
Azarpazhooh, A., \& Leake, J. L. (2006). Systematic {Review} of the
{Association Between Respiratory Diseases} and {Oral Health}.
\emph{Journal of Periodontology}, \emph{77}(9), 1465--1482.
\url{https://doi.org/10.1902/jop.2006.060010}

\leavevmode\vadjust pre{\hypertarget{ref-badriRegulationFunction2009}{}}%
Badri, D. V., \& Vivanco, J. M. (2009). Regulation and function of root
exudates. \emph{Plant, Cell \& Environment}, \emph{32}(6), 666--681.
\url{https://doi.org/10.1111/j.1365-3040.2009.01926.x}

\leavevmode\vadjust pre{\hypertarget{ref-bispoSimultaneousDetermination2002}{}}%
Bispo, M. S., Veloso, M. C. C., Pinheiro, H. L. C., De Oliveira, R. F.
S., Reis, J. O. N., \& De Andrade, J. B. (2002). Simultaneous
{Determination} of {Caffeine}, {Theobromine}, and {Theophylline} by
{High-Performance Liquid Chromatography}. \emph{Journal of
Chromatographic Science}, \emph{40}(1), 45--48.
\url{https://doi.org/10.1093/chromsci/40.1.45}

\leavevmode\vadjust pre{\hypertarget{ref-boocockMaxillarySinusitis1995}{}}%
Boocock, P., Roberts, C. A., \& Manchester, K. (1995). Maxillary
sinusitis in {Medieval Chichester}, {England}. \emph{American Journal of
Physical Anthropology}, \emph{98}(4), 483--495.
\url{https://doi.org/10.1002/ajpa.1330980408}

\leavevmode\vadjust pre{\hypertarget{ref-boumanBegravenis2017}{}}%
Bouman, J. (2017). De Begravenis. \emph{De Nieuwe Schouwschuit},
\emph{15}, 11--15.
\url{https://www.historischgenootschapbeemster.nl/wp-content/uploads/De_Nieuwe_Schouwschuit_15e_jaargang_november_2017.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-SucheyBrooks1990}{}}%
Brooks, S., \& Suchey, J. M. (1990). Skeletal age determination based on
the os pubis: {A} comparison of the {Acsádi-Nemeskéri} and
{Suchey-Brooks} methods. \emph{Human Evolution}, \emph{5}(3), 227--238.
\url{https://doi.org/10.1007/BF02437238}

\leavevmode\vadjust pre{\hypertarget{ref-brothwellDiggingBones1981}{}}%
Brothwell, D. (1981). \emph{Digging up {Bones}: {The} excavation,
treatment and study of human skeletal remains} (3rd ed.). {British
Museum (Natural History)}.

\leavevmode\vadjust pre{\hypertarget{ref-bruinsmaBijdragenTot1872}{}}%
Bruinsma, J. J. (1872). \emph{Bijdragen tot de {Geneeskundige
Plaatsbeschrijving} van {Nederland}}. {Van Weelden en Mingelen}.
\url{https://dlcs.io/pdf/wellcome/pdf-item/b24874140/0}

\leavevmode\vadjust pre{\hypertarget{ref-buckberryAuricular2002}{}}%
Buckberry, J. L., \& Chamberlain, A. T. (2002). Age estimation from the
auricular surface of the ilium: A revised method. \emph{American Journal
of Physical Anthropology}, \emph{119}(3), 231--239.
\url{https://doi.org/10.1002/ajpa.10130}

\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculus2014}{}}%
Buckley, S., Usai, D., Jakob, T., Radini, A., \& Hardy, K. (2014).
Dental {Calculus Reveals Unique Insights} into {Food Items}, {Cooking}
and {Plant Processing} in {Prehistoric Central Sudan}. \emph{PLOS ONE},
\emph{9}(7), e100808. \url{https://doi.org/10.1371/journal.pone.0100808}

\leavevmode\vadjust pre{\hypertarget{ref-Standards1994}{}}%
Buikstra, J. E., \& Ubelaker, D. H. (1994). Standards for data
collection from human skeletal remains: {Proceedings} of a seminar at
the {Field Museum} of {Natural History} ({Arkansas Archaeology Research
Series} 44). \emph{Fayetteville Arkansas Archaeological Survey}.

\leavevmode\vadjust pre{\hypertarget{ref-casnaUrbanizationRespiratory2021}{}}%
Casna, M., Burrell, C. L., Schats, R., Hoogland, M. L. P., \& Schrader,
S. A. (2021). Urbanization and respiratory stress in the {Northern Low
Countries}: {A} comparative study of chronic maxillary sinusitis in two
early modern sites from the {Netherlands} ({AD} 1626--1866).
\emph{International Journal of Osteoarchaeology}, \emph{31}(5),
891--901. \url{https://doi.org/10.1002/oa.3006}

\leavevmode\vadjust pre{\hypertarget{ref-chenCa2Dependent2001}{}}%
Chen, H., Hou, W., Kuć, J., \& Lin, Y. (2001). Ca2+‐dependent and
{Ca2}+‐independent excretion modes of salicylic acid in tobacco cell
suspension culture. \emph{Journal of Experimental Botany},
\emph{52}(359), 1219--1226.
\url{https://doi.org/10.1093/jexbot/52.359.1219}

\leavevmode\vadjust pre{\hypertarget{ref-chinCaffeineContent2008}{}}%
Chin, J. M., Merves, M. L., Goldberger, B. A., Sampson-Cone, A., \&
Cone, E. J. (2008). Caffeine {Content} of {Brewed Teas}. \emph{Journal
of Analytical Toxicology}, \emph{32}(8), 702--704.
\url{https://doi.org/10.1093/jat/32.8.702}

\leavevmode\vadjust pre{\hypertarget{ref-chovanecOpiumMasses2012}{}}%
Chovanec, Z., Rafferty, S., \& Swiny, S. (2012). Opium for the {Masses}.
\emph{Ethnoarchaeology}, \emph{4}(1), 5--36.
\url{https://doi.org/10.1179/eth.2012.4.1.5}

\leavevmode\vadjust pre{\hypertarget{ref-clarkeCannabisEvolution2013}{}}%
Clarke, R. (2013). \emph{Cannabis : {Evolution} and {Ethnobotany}}.
{University of California Press}.

\leavevmode\vadjust pre{\hypertarget{ref-cogoVitroEvaluation2008}{}}%
Cogo, K., Montan, M. F., Bergamaschi, C. de C., D. Andrade, E., Rosalen,
P. L., \& Groppo, F. C. (2008). In vitro evaluation of the effect of
nicotine, cotinine, and caffeine on oral microorganisms. \emph{Canadian
Journal of Microbiology}, \emph{54}(6), 501--508.
\url{https://doi.org/10.1139/W08-032}

\leavevmode\vadjust pre{\hypertarget{ref-coneSalivaTesting1993}{}}%
Cone, E. J. (1993). Saliva {Testing} for {Drugs} of {Abuse}.
\emph{Annals of the New York Academy of Sciences}, \emph{694}(1),
91--127. \url{https://doi.org/10.1111/j.1749-6632.1993.tb18346.x}

\leavevmode\vadjust pre{\hypertarget{ref-coneInterpretationOral2007}{}}%
Cone, E. J., \& Huestis, M. A. (2007). Interpretation of {Oral Fluid
Tests} for {Drugs} of {Abuse}. \emph{Annals of the New York Academy of
Sciences}, \emph{1098}, 51--103.
\url{https://doi.org/10.1196/annals.1384.037}

\leavevmode\vadjust pre{\hypertarget{ref-doddsHealthBenefits2005}{}}%
Dodds, M. W. J., Johnson, D. A., \& Yeh, C.-K. (2005). Health benefits
of saliva: A review. \emph{Journal of Dentistry}, \emph{33}(3),
223--233. \url{https://doi.org/10.1016/j.jdent.2004.10.009}

\leavevmode\vadjust pre{\hypertarget{ref-duthieNaturalSalicylates2011}{}}%
Duthie, G. G., \& Wood, A. D. (2011). Natural salicylates: Foods ,
functions and disease prevention. \emph{Food \& Function}, \emph{2}(9),
515--520. \url{https://doi.org/10.1039/C1FO10128E}

\leavevmode\vadjust pre{\hypertarget{ref-echeverriaNicotineHair2013}{}}%
Echeverría, J., \& Niemeyer, H. M. (2013). Nicotine in the hair of
mummies from {San Pedro} de {Atacama} ({Northern Chile}). \emph{Journal
of Archaeological Science}, \emph{40}(10), 3561--3568.
\url{https://doi.org/10.1016/j.jas.2013.04.030}

\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculus2018}{}}%
Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez,
K., Murga, E., Kaijankoski, P., Rosenthal, J. S., \& Gang, D. R. (2018).
Dental calculus as a source of ancient alkaloids: {Detection} of
nicotine by {LC-MS} in calculus samples from the {Americas}.
\emph{Journal of Archaeological Science: Reports}, \emph{18}, 509--515.
\url{https://doi.org/10.1016/j.jasrep.2018.02.004}

\leavevmode\vadjust pre{\hypertarget{ref-gismondiMultidisciplinaryApproach2020}{}}%
Gismondi, A., Baldoni, M., Gnes, M., Scorrano, G., D'Agostino, A.,
Marco, G. D., Calabria, G., Petrucci, M., Müldner, G., Tersch, M. V.,
Nardi, A., Enei, F., Canini, A., Rickards, O., Alexander, M., \&
Martínez-Labarga, C. (2020). A multidisciplinary approach for
investigating dietary and medicinal habits of the {Medieval} population
of {Santa Severa} (7th-15th centuries, {Rome}, {Italy}). \emph{PLOS
ONE}, \emph{15}(1), e0227433.
\url{https://doi.org/10.1371/journal.pone.0227433}

\leavevmode\vadjust pre{\hypertarget{ref-goodmanTobaccoHistory1994}{}}%
Goodman, J. (1994). \emph{Tobacco in history: The cultures of
dependence}. {Routledge}.

\leavevmode\vadjust pre{\hypertarget{ref-greeneQuantifyingCalculus2005}{}}%
Greene, T. R., Kuba, C. L., \& Irish, J. D. (2005). Quantifying
calculus: {A} suggested new approach for recording an important
indicator of diet and dental health. \emph{HOMO - Journal of Comparative
Human Biology}, \emph{56}(2), 119--132.
\url{https://doi.org/10.1016/j.jchb.2005.02.002}

\leavevmode\vadjust pre{\hypertarget{ref-jinSupragingivalCalculus2002}{}}%
Jin, Y., \& Yip, H.-K. (2002). Supragingival {Calculus}: {Formation} and
{Control}. \emph{Critical Reviews in Oral Biology \& Medicine}.
\url{https://doi.org/10.1177/154411130201300506}

\leavevmode\vadjust pre{\hypertarget{ref-kondoAssociationCoffee2021}{}}%
Kondo, K., Suzuki, K., Washio, M., Ohfuji, S., Adachi, S., Kan, S.,
Imai, S., Yoshimura, K., Miyashita, N., Fujisawa, N., Maeda, A.,
Fukushima, W., \& Hirota, Y. (2021). Association between coffee and
green tea intake and pneumonia among the {Japanese} elderly: A
case-control study. \emph{Scientific Reports}, \emph{11}(1, 1), 5570.
\url{https://doi.org/10.1038/s41598-021-84348-w}

\leavevmode\vadjust pre{\hypertarget{ref-lemmersMiddenbeemster2013}{}}%
Lemmers, S. A. M., Schats, R., Hoogland, M. L. P., \& Waters-Rist, A.
(2013). Fysisch antropologische analyse Middenbeemster. In \emph{De
begravingen bij de Keyserkerk te Middenbeemster} (pp. 35--60).

\leavevmode\vadjust pre{\hypertarget{ref-leuwProhibitionLegalizationDutch1994}{}}%
Leuw, E., \& Marshall, I. H. (1994). \emph{Between {Prohibition} and
{Legalization}: {The Dutch Experiment} in {Drug Policy}}. {Kugler
Publications}. \url{https://books.google.com?id=2mAVkStNG5EC}

\leavevmode\vadjust pre{\hypertarget{ref-lindholstLongTerm2010}{}}%
Lindholst, C. (2010). Long term stability of cannabis resin and cannabis
extracts. \emph{Australian Journal of Forensic Sciences}, \emph{42}(3),
181--190. \url{https://doi.org/10.1080/00450610903258144}

\leavevmode\vadjust pre{\hypertarget{ref-liuNicotinedegradingMicroorganisms2015}{}}%
Liu, J., Ma, G., Chen, T., Hou, Y., Yang, S., Zhang, K.-Q., \& Yang, J.
(2015). Nicotine-degrading microorganisms and their potential
applications. \emph{Applied Microbiology and Biotechnology},
\emph{99}(9), 3775--3785.
\url{https://doi.org/10.1007/s00253-015-6525-1}

\leavevmode\vadjust pre{\hypertarget{ref-lovejoyAuricular1985}{}}%
Lovejoy, C. O., Meindl, R. S., Pryzbeck, T. R., \& Mensforth, R. P.
(1985). Chronological metamorphosis of the auricular surface of the
ilium: {A} new method for the determination of adult skeletal age at
death. \emph{American Journal of Physical Anthropology}, \emph{68}(1),
15--28. \url{https://doi.org/10.1002/ajpa.1330680103}

\leavevmode\vadjust pre{\hypertarget{ref-lustmannScanningElectron1976}{}}%
Lustmann, J., Lewin-Epstein, J., \& Shteyer, A. (1976). Scanning
electron microscopy of dental calculus. \emph{Calcified Tissue
Research}, \emph{21}(1), 47--55.
\url{https://doi.org/10.1007/BF02547382}

\leavevmode\vadjust pre{\hypertarget{ref-maatManualPhysical2005}{}}%
Maat, G., \& Mastwijk, R. (2005). Manual for the physical
anthropological report. \emph{Barge's Anthropologica}, \emph{6}.

\leavevmode\vadjust pre{\hypertarget{ref-machtHistoryOpium1915}{}}%
Macht, D. I. (1915). The history of opium and some of its preparations
and alkaloids. \emph{The Journal of the American Medical Association},
\emph{LXIV}(6), 5.

\leavevmode\vadjust pre{\hypertarget{ref-mackiePreservationMetaproteome2017}{}}%
Mackie, M., Hendy, J., Lowe, A. D., Sperduti, A., Holst, M., Collins, M.
J., \& Speller, C. F. (2017). Preservation of the metaproteome:
Variability of protein preservation in ancient dental calculus.
\emph{STAR: Science \& Technology of Archaeological Research},
\emph{3}(1), 58--70. \url{https://doi.org/10.1080/20548923.2017.1361629}

\leavevmode\vadjust pre{\hypertarget{ref-malakarNaturallyOccurring2017}{}}%
Malakar, S., Gibson, P. R., Barrett, J. S., \& Muir, J. G. (2017).
Naturally occurring dietary salicylates: {A} closer look at common
{Australian} foods. \emph{Journal of Food Composition and Analysis},
\emph{57}, 31--39. \url{https://doi.org/10.1016/j.jfca.2016.12.008}

\leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research
compendium} {[}Manual{]}. \url{https://github.com/benmarwick/rrtools}

\leavevmode\vadjust pre{\hypertarget{ref-maughanCaffeineIngestion2003}{}}%
Maughan, R. J., \& Griffin, J. (2003). Caffeine ingestion and fluid
balance: A review. \emph{Journal of Human Nutrition and Dietetics},
\emph{16}(6), 411--420.
\url{https://doi.org/10.1046/j.1365-277X.2003.00477.x}

\leavevmode\vadjust pre{\hypertarget{ref-meindlSutureClosure1985}{}}%
Meindl, R. S., \& Lovejoy, C. O. (1985). Ectocranial suture closure: {A}
revised method for the determination of skeletal age at death based on
the lateral-anterior sutures. \emph{American Journal of Physical
Anthropology}, \emph{68}(1), 57--66.
\url{https://doi.org/10.1002/ajpa.1330680106}

\leavevmode\vadjust pre{\hypertarget{ref-milmanOralFluid2011}{}}%
Milman, G., Schwope, D. M., Schwilke, E. W., Darwin, W. D., Kelly, D.
L., Goodwin, R. S., Gorelick, D. A., \& Huestis, M. A. (2011). Oral
{Fluid} and {Plasma Cannabinoid Ratios} after {Around-the-Clock
Controlled Oral Δ9-Tetrahydrocannabinol Administration}. \emph{Clinical
Chemistry}, \emph{57}(11), 1597--1606.
\url{https://doi.org/10.1373/clinchem.2011.169490}

\leavevmode\vadjust pre{\hypertarget{ref-mortimerHistoryCoca1901}{}}%
Mortimer, W. G. (1901). \emph{Peru. {History} of coca, "the divine
plant" of the {Incas}; with an introductory account of the {Incas}, and
of the {Andean Indians} of to-day}. {New York, J. H. Vail \& Company}.
\url{http://archive.org/details/peruhistoryofcoc00mortrich}

\leavevmode\vadjust pre{\hypertarget{ref-nierstraszTeaTrade2015}{}}%
Nierstrasz, C. (2015). \emph{Rivalry for {Trade} in {Tea} and
{Textiles}: {The English} and {Dutch East India} companies
(1700--1800)}. {Springer}.
\url{https://books.google.com?id=uwtaCwAAQBAJ}

\leavevmode\vadjust pre{\hypertarget{ref-ogaldeIdentificationPsychoactive2009}{}}%
Ogalde, J. P., Arriaza, B. T., \& Soto, E. C. (2009). Identification of
psychoactive alkaloids in ancient {Andean} human hair by gas
chromatography/mass spectrometry. \emph{Journal of Archaeological
Science}, \emph{36}(2), 467--472.
\url{https://doi.org/10.1016/j.jas.2008.09.036}

\leavevmode\vadjust pre{\hypertarget{ref-palmerActivityReconstruction2016}{}}%
Palmer, J. L. A., Hoogland, M. H. L., \& Waters‐Rist, A. L. (2016).
Activity {Reconstruction} of {Post}‐{Medieval Dutch Rural Villagers}
from {Upper Limb Osteoarthritis} and {Entheseal Changes}.
\emph{International Journal of Osteoarchaeology}, \emph{26}(1), 78--92.
\url{https://doi.org/10.1002/oa.2397}

\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
R Core Team. (2020). \emph{R: {A} language and environment for
statistical computing} {[}Manual{]}. {R Foundation for Statistical
Computing}; {R Foundation for Statistical Computing}.
\url{https://www.R-project.org/}

\leavevmode\vadjust pre{\hypertarget{ref-raffertyCurrentResearch2012}{}}%
Rafferty, S. M., Lednev, I., Virkler, K., \& Chovanec, Z. (2012).
Current research on smoking pipe residues. \emph{Journal of
Archaeological Science}, \emph{39}(7), 1951--1959.
\url{https://doi.org/10.1016/j.jas.2012.02.001}

\leavevmode\vadjust pre{\hypertarget{ref-rehImpactTobacco2012}{}}%
Reh, D. D., Higgins, T. S., \& Smith, T. L. (2012). Impact of {Tobacco
Smoke} on {Chronic Rhinosinusitis} -- {A Review} of the {Literature}.
\emph{International Forum of Allergy \& Rhinology}, \emph{2}(5), 362.
\url{https://doi.org/10.1002/alr.21054}

\leavevmode\vadjust pre{\hypertarget{ref-Rpsych}{}}%
Revelle, W. (2022). \emph{Psych: {Procedures} for psychological,
psychometric, and personality research} {[}Manual{]}. {Northwestern
University}. \url{https://CRAN.R-project.org/package=psych}

\leavevmode\vadjust pre{\hypertarget{ref-rogersPalaeopathologyJoint2000}{}}%
Rogers, J. (2000). The palaeopathology of joint disease. In M. Cox \& S.
Mays (Eds.), \emph{Human osteology : {In} archaeology and forensic
science.} (1st ed, pp. 163--182). {Cambridge University Press}.
\url{https://login.ezproxy.leidenuniv.nl:2443/login?URL=https://search.ebscohost.com/login.aspx?direct=true\&db=e000xww\&AN=40641\&site=ehost-live}

\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoRoleOral1999}{}}%
Scannapieco, F. A. (1999). Role of {Oral Bacteria} in {Respiratory
Infection}. \emph{Journal of Periodontology}, \emph{70}(7), 793--802.
\url{https://doi.org/10.1902/jop.1999.70.7.793}

\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoPotentialAssociations2001}{}}%
Scannapieco, F. A., \& Ho, A. W. (2001). Potential {Associations Between
Chronic Respiratory Disease} and {Periodontal Disease}: {Analysis} of
{National Health} and {Nutrition Examination Survey III}. \emph{Journal
of Periodontology}, \emph{72}(1), 50--56.
\url{https://doi.org/10.1902/jop.2001.72.1.50}

\leavevmode\vadjust pre{\hypertarget{ref-scheltemaOpiumTrade1907}{}}%
Scheltema, J. F. (1907). The {Opium Trade} in the {Dutch East Indies}.
{I}. \emph{American Journal of Sociology}, \emph{13}(1), 79--112.

\leavevmode\vadjust pre{\hypertarget{ref-schuijtemakerTeTheegasten2011}{}}%
Schuijtemaker, D. (2011). Te Theegasten. \emph{De Nieuwe Schouwschuit},
\emph{9}, 16--17.

\leavevmode\vadjust pre{\hypertarget{ref-slavinDiagnosisManagement2005}{}}%
Slavin, R. G., Spector, S. L., Bernstein, I. L., Slavin, R. G., Kaliner,
M. A., Kennedy, D. W., Virant, F. S., Wald, E. R., Khan, D. A.,
Blessing-Moore, J., Lang, D. M., Nicklas, R. A., Oppenheimer, J. J.,
Portnoy, J. M., Schuller, D. E., Tilles, S. A., Borish, L., Nathan, R.
A., Smart, B. A., \& Vandewalker, M. L. (2005). The diagnosis and
management of sinusitis: {A} practice parameter update. \emph{Journal of
Allergy and Clinical Immunology}, \emph{116}, S13--S47.
\url{https://doi.org/10.1016/j.jaci.2005.09.048}

\leavevmode\vadjust pre{\hypertarget{ref-smithDetectionOpium2018}{}}%
Smith, R. K., Stacey, R. J., Bergström, E., \& Thomas-Oates, J. (2018).
Detection of opium alkaloids in a {Cypriot} base-ring juglet.
\emph{Analyst}, \emph{143}(21), 5127--5136.
\url{https://doi.org/10.1039/C8AN01040D}

\leavevmode\vadjust pre{\hypertarget{ref-sorensenEffectAntioxidants2018}{}}%
Sørensen, L. K., \& Hasselstrøm, J. B. (2018). The effect of
antioxidants on the long-term stability of {THC} and related
cannabinoids in sampled whole blood. \emph{Drug Testing and Analysis},
\emph{10}(2), 301--309. \url{https://doi.org/10.1002/dta.2221}

\leavevmode\vadjust pre{\hypertarget{ref-sorensenDrugsCalculus2021}{}}%
Sørensen, L. K., Hasselstrøm, J. B., Larsen, L. S., \& Bindslev, D. A.
(2021). Entrapment of drugs in dental calculus -- {Detection} validation
based on test results from post-mortem investigations. \emph{Forensic
Science International}, \emph{319}, 110647.
\url{https://doi.org/10.1016/j.forsciint.2020.110647}

\leavevmode\vadjust pre{\hypertarget{ref-srdjenovicSimultaneousHPLC2008}{}}%
Srdjenovic, B., Djordjevic-Milic, V., Grujic, N., Injac, R., \&
Lepojevic, Z. (2008). Simultaneous {HPLC Determination} of {Caffeine},
{Theobromine}, and {Theophylline} in {Food}, {Drinks}, and {Herbal
Products}. \emph{Journal of Chromatographic Science}, \emph{46}(2),
144--149. \url{https://doi.org/10.1093/chromsci/46.2.144}

\leavevmode\vadjust pre{\hypertarget{ref-stavricVariabilityCaffeine1988}{}}%
Stavric, B., Klassen, R., Watkinson, B., Karpinski, K., Stapley, R., \&
Fried, P. (1988). Variability in caffeine consumption from coffee and
tea: {Possible} significance for epidemiological studies. \emph{Food and
Chemical Toxicology}, \emph{26}(2), 111--118.
\url{https://doi.org/10.1016/0278-6915(88)90107-X}

\leavevmode\vadjust pre{\hypertarget{ref-sunMetabolomicsEvaluation2016}{}}%
Sun, J., Jin, J., Beger, R. D., Cerniglia, C. E., Yang, M., \& Chen, H.
(2016). Metabolomics evaluation of the impact of smokeless tobacco
exposure on the oral bacterium {Capnocytophaga} sputigena.
\emph{Toxicology in Vitro}, \emph{36}, 133--141.
\url{https://doi.org/10.1016/j.tiv.2016.07.020}

\leavevmode\vadjust pre{\hypertarget{ref-takahashiOralMicrobiome2015}{}}%
Takahashi, N. (2015). Oral {Microbiome Metabolism}: {From} {``{Who Are
They}?''} To {``{What Are They Doing}?''} \emph{Journal of Dental
Research}, \emph{94}(12), 1628--1637.
\url{https://doi.org/10.1177/0022034515606045}

\leavevmode\vadjust pre{\hypertarget{ref-tushinghamHuntergathererTobacco2013}{}}%
Tushingham, S., Ardura, D., Eerkens, J. W., Palazoglu, M., Shahbaz, S.,
\& Fiehn, O. (2013). Hunter-gatherer tobacco smoking: Earliest evidence
from the {Pacific Northwest Coast} of {North America}. \emph{Journal of
Archaeological Science}, \emph{40}(2), 1397--1407.
\url{https://doi.org/10.1016/j.jas.2012.09.019}

\leavevmode\vadjust pre{\hypertarget{ref-valenDetermination212017}{}}%
Valen, A., Leere Øiestad, Å. M., Strand, D. H., Skari, R., \& Berg, T.
(2017). Determination of 21 drugs in oral fluid using fully automated
supported liquid extraction and {UHPLC-MS}/{MS}. \emph{Drug Testing and
Analysis}, \emph{9}(5), 808--823. \url{https://doi.org/10.1002/dta.2045}

\leavevmode\vadjust pre{\hypertarget{ref-velskoMicrobialDifferences2019}{}}%
Velsko, I. M., Fellows Yates, J. A., Aron, F., Hagan, R. W., Frantz, L.
A. F., Loe, L., Martinez, J. B. R., Chaves, E., Gosden, C., Larson, G.,
\& Warinner, C. (2019). Microbial differences between dental plaque and
historic dental calculus are related to oral biofilm maturation stage.
\emph{Microbiome}, \emph{7}(1), 102.
\url{https://doi.org/10.1186/s40168-019-0717-3}

\leavevmode\vadjust pre{\hypertarget{ref-velskoDentalCalculus2017}{}}%
Velsko, I. M., Overmyer, K. A., Speller, C., Klaus, L., Collins, M. J.,
Loe, L., Frantz, L. A. F., Sankaranarayanan, K., Lewis, C. M., Martinez,
J. B. R., Chaves, E., Coon, J. J., Larson, G., \& Warinner, C. (2017).
The dental calculus metabolome in modern and historic samples.
\emph{Metabolomics}, \emph{13}(11), 134.
\url{https://doi.org/10.1007/s11306-017-1270-3}

\leavevmode\vadjust pre{\hypertarget{ref-warinnerEvidenceMilk2014}{}}%
Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R.,
Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O. E., Swallow, D. M.,
Fotakis, A., Christensen, R. J., Olsen, J. V., Liebert, A., Montalva,
N., Fiddyment, S., Charlton, S., Mackie, M., Canci, A., \ldots{}
Collins, M. J. (2014). Direct evidence of milk consumption from ancient
human dental calculus. \emph{Scientific Reports}, \emph{4}, 7104.
\url{https://doi.org/10.1038/srep07104}

\leavevmode\vadjust pre{\hypertarget{ref-whiteDentalCalculus1997}{}}%
White, D. J. (1997). Dental calculus: Recent insights into occurrence,
formation, prevention, removal and oral health effects of supragingival
and subgingival deposits. \emph{European Journal of Oral Sciences},
\emph{105}(5), 508--522.
\url{https://doi.org/10.1111/j.1600-0722.1997.tb00238.x}

\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
Wickham, H. (2016). \emph{Ggplot2: {Elegant Graphics} for {Data
Analysis}}. {Springer-Verlag}. \url{https://ggplot2.tidyverse.org}

\leavevmode\vadjust pre{\hypertarget{ref-tidyverse2019}{}}%
Wickham, Hadley, Averick, M., Bryan, J., Chang, W., McGowan, L. D.,
François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M.,
Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J.,
Robinson, D., Seidel, D. P., Spinu, V., \ldots{} Yutani, H. (2019).
Welcome to the {tidyverse}. \emph{Journal of Open Source Software},
\emph{4}(43), 1686. \url{https://doi.org/10.21105/joss.01686}

\leavevmode\vadjust pre{\hypertarget{ref-willeRelationshipOral2009}{}}%
Wille, S. M. R., Raes, E., Lillsunde, P., Gunnar, T., Laloup, M., Samyn,
N., Christophersen, A. S., Moeller, M. R., Hammer, K. P., \& Verstraete,
A. G. (2009). Relationship {Between Oral Fluid} and {Blood
Concentrations} of {Drugs} of {Abuse} in {Drivers Suspected} of {Driving
Under} the {Influence} of {Drugs}. \emph{Therapeutic Drug Monitoring},
\emph{31}(4), 511. \url{https://doi.org/10.1097/FTD.0b013e3181ae46ea}

\leavevmode\vadjust pre{\hypertarget{ref-yaussyCalculusSurvivorship2019}{}}%
Yaussy, S. L., \& DeWitte, S. N. (2019). Calculus and survivorship in
medieval {London}: {The} association between dental disease and a
demographic measure of general health. \emph{American Journal of
Physical Anthropology}, \emph{168}(3), 552--565.
\url{https://doi.org/10.1002/ajpa.23772}

\leavevmode\vadjust pre{\hypertarget{ref-ziesemer16SChallenges2015}{}}%
Ziesemer, K. A., Mann, A. E., Sankaranarayanan, K., Schroeder, H., Ozga,
A. T., Brandt, B. W., Zaura, E., Waters-Rist, A., Hoogland, M.,
Salazar-Garcia, D. C., Aldenderfer, M., Speller, C., Hendy, J., Weston,
D. A., MacDonald, S. J., Thomas, G. H., Collins, M. J., Lewis, C. M.,
Hofman, C., \& Warinner, C. (2015). Intrinsic challenges in ancient
microbiome reconstruction using {16S rRNA} gene amplification. \emph{Sci
Rep}, \emph{5}, 16498. \url{https://doi.org/10.1038/srep16498}

\leavevmode\vadjust pre{\hypertarget{ref-ziesemerGenomeCalculus2018}{}}%
Ziesemer, K. A., Ramos‐Madrigal, J., Mann, A. E., Brandt, B. W.,
Sankaranarayanan, K., Ozga, A. T., Hoogland, M., Hofman, C. A.,
Salazar‐García, D. C., Frohlich, B., Milner, G. R., Stone, A. C.,
Aldenderfer, M., Lewis, C. M., Hofman, C. L., Warinner, C., \&
Schroeder, H. (2018). The efficacy of whole human genome capture on
ancient dental calculus and dentin. \emph{American Journal of Physical
Anthropology}. \url{https://doi.org/10.1002/ajpa.23763}

\leavevmode\vadjust pre{\hypertarget{ref-zijngeBiofilmArchitecture2010}{}}%
Zijnge, V., van Leeuwen, M. B. M., Degener, J. E., Abbas, F., Thurnheer,
T., Gmür, R., \& M. Harmsen, H. J. (2010). Oral {Biofilm Architecture}
on {Natural Teeth}. \emph{PLoS ONE}, \emph{5}(2), e9321.
\url{https://doi.org/10.1371/journal.pone.0009321}

\end{CSLReferences}

\end{document}

### 2023-06-01 analysis/paper/_discussion.qmd
fix broken citekeys

--- 2023-02-07 inst/paper/_discussion.qmd
@@
 Beemsterpolder is well-documented in the literature
-[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
+[@aten400Jaar2012; @boumanBegravenis2017], and visible on the skeletal
 remains as pipe notches [@lemmersMiddenbeemster2013].
@@
 where a tobacco-smoke enema was used for headaches, respiratory problems, colds,
-and drowsiness from around 1780 to 1830 [@aten400JaarBeemster2012].
+and drowsiness from around 1780 to 1830 [@aten400Jaar2012].
 In our sample, we also detected nicotine and cotinine (replicated)
@@
 [@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017], including
-potatoes, which were a staple of the Beemsterpolder diet [@aten400JaarBeemster2012].
+potatoes, which were a staple of the Beemsterpolder diet [@aten400Jaar2012].
 The extracted quantity from our samples decreased over the three washes, followed
@@
 another opium containing concoction was part of the "traditional" medicine in
-the Netherlands [@leuwProhibitionLegalizationDutch1994], including Middenbeemster
-[@aten400JaarBeemster2012]. It was also generally
+the Netherlands [@leuwProhibitionLegalization1994], including Middenbeemster
+[@aten400Jaar2012]. It was also generally
 considered a drug of the upper class [@scheltemaOpiumTrade1907], and may have been

### 2023-06-01 analysis/paper/_materials.qmd
fix broken citekeys

--- 2023-02-07 inst/paper/_materials.qmd
@@
 was dairy farming, consisting largely of manual labour prior to the industrial
-revolution [@aten400JaarBeemster2012; @palmerActivityReconstruction2016].
+revolution [@aten400Jaar2012; @palmerActivityReconstruction2016].
 

### 2023-07-19 analysis/scripts/setup-qmd.R
fix: caries ratio calculations

@@ -161,19 +161,16 @@ dental_long <- reduce(
 surface <- c("mes", "dis", "occ", "buc", "lin", "root", "crown") # here occ = incisal
 
 caries_count <- dental_long %>%
-  # pivot_longer(t11:t48, names_to = "tooth",
-  #              values_to = "caries_score") %>%
-  #na.omit() %>%
   separate_rows(caries, sep = ";") %>% # one lesion per row
   mutate(
     caries_count = if_else(
-      caries == "none", 0L, 1L # convert lesion location to binary (present = 1; absent = 0)
+      caries == "none", 0L, 1L, missing = NA # convert lesion location to binary (present = 1; absent = 0)
     )
   ) %>%
   group_by(id, tooth) %>%
   summarise(
     #n_teeth = n(),
-    count = sum(caries_count, na.rm = T),
+    count = sum(caries_count),
   ) %>%
   tooth_position() %>%
   ungroup()

### 2023-07-19 R/data.R
update data source and documentation

@@ -1,5 +1,60 @@
 # can't produce doc if data aren\t exported by namespace
 
+#' Sample metadata
+#'
+#' Metadata for the samples analysed by UHPLC-MS/MS.
+#' @format ## `metadata`
+#' A data frame with 44 rows and 8 columns:
+#' \describe{
+#'   \item{id}{Skeletal ID}
+#'   \item{sample}{Dental calculus sample number}
+#'   \item{element}{FDI notation for tooth/teeth sampled for each individual.}
+#'   \item{KZ_sample}{Dental calculus sample number on tubes from previous study.}
+#'   \item{KZ_element}{FDI notation for tooth/teeth sampled from the previous study for each individual.}
+#'   \item{replicated}{Whether or not the sample was replicated in a separate UHPLC-MS/MS analysis.}
+#'   \item{batch1_weight}{weight (mg) of dental calculus sample analysed in the first batch.}
+#'   \item{batch2_weight}{weight (mg) of dental calculus sample analysed in the replication batch.}
+#'   ...
+#' }
+#' @source https://doi.org/10.5281/zenodo.8010630
+"metadata"
+
+#' Lower limits of quantitation (LLOQ)
+#'
+#' LLOQ values for the targeted compounds.
+#' @format ## `lloq`
+#' A data frame with 76 rows and 2 columns:
+#' \describe{
+#'   \item{compound}{Name of the compound.}
+#'   \item{lloq}{Lower limit of quantitation (ng)}
+#'   \item{element}{FDI notation for tooth/teeth sampled for each individual.}
+#'   \item{KZ_sample}{Dental calculus sample number on tubes from previous study.}
+#'   \item{KZ_element}{FDI notation for tooth/teeth sampled from the previous study for each individual.}
+#'   \item{replicated}{Whether or not the sample was replicated in a separate UHPLC-MS/MS analysis.}
+#'   \item{batch1_weight}{weight (mg) of dental calculus sample analysed in the first batch.}
+#'   \item{batch2_weight}{weight (mg) of dental calculus sample analysed in the replication batch.}
+#'   ...
+#' }
+#' @source https://doi.org/10.5281/zenodo.8010630
+"lloq"
+
+#' Middenbeemster sample demographics
+#'
+#' Age and sex of the individuals with dental calculus scored in `mb11_calculus`.
+#'
+#' @format ## `demography`
[…]

### 2023-07-26 R/generate-bib.R
add news and changelog files

@@ -4,7 +4,7 @@
 #' @importFrom here here
 
 generate_bib <- function(){
-  cite_keys <- bbt_detect_citations(list.files(here("analysis/paper/"), "*.qmd", full.names = T))
+  cite_keys <- bbt_detect_citations(list.files(here("analysis/"), "*.qmd", full.names = T, recursive = T))
   ignore_keys <- grep("fig-|tbl-", cite_keys, value = T)
   bbt_write_bib(
     here("analysis/paper/references.bib"),

### 2023-07-26 analysis/paper/paper.tex
prepare supp mat for submission

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
 \fi
@@
 \makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
 \makeatother
@@
 \@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
 \makeatother
@@
 population}
-\author{Bjørn Peare Bartholdy\textsuperscript{1,*} \and Jørgen B.
-Hasselstrøm\textsuperscript{2} \and Lambert K.
-Sørensen\textsuperscript{2} \and Maia Casna\textsuperscript{1} \and Menno Hoogland\textsuperscript{1} \and Historisch Genootschap Beemster\textsuperscript{3} \and Amanda G.
-Henry\textsuperscript{1}}
+\author{}
 \date{}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, sharp corners, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, boxrule=0pt, sharp corners, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-All statistical analysis was conducted in R version 4.3.0 (2023-04-21),
-Already Tomorrow, (R Core Team, 2020). Data wrangling was conducted with
+All statistical analysis was conducted in R version 4.3.1 (2023-06-16),
+Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
 the \textbf{tidyverse} (Hadley Wickham et al., 2019) and visualisations
@@
 THCVA = tetrahydrocannabivarin acid.}\tabularnewline
-\toprule()
+\toprule\noalign{}
 Compound & Batch 1 & Batch 2 & LLOQ \\
-\midrule()
+\midrule\noalign{}
 \endfirsthead
-\toprule()
+\toprule\noalign{}
 Compound & Batch 1 & Batch 2 & LLOQ \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 CBD & TRUE & FALSE & 0.050 \\
@@
 Theophylline & TRUE & TRUE & 0.010 \\
-\bottomrule()
 \end{longtable}
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1566}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0843}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1566}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1084}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0952}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}@{}}
 \caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
@@
 notches.}\tabularnewline
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endfirsthead
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
-OA & -0.19 & -0.074 & 0.21 & 0.07 & 0.14 & 0.28 & 0.00098 & -0.067 \\
-VOP & -0.061 & -0.16 & 0.34 & 0.061 & 0.25 & -0.06 & 0.013 & -0.13 \\
-SN & -0.22 & 0.16 & 0.095 & 0.089 & 0.17 & 0.24 & 0.16 & 0.093 \\
-DDD & 0.032 & 0.0037 & 0.19 & -0.39 & -0.077 & 0.31 & 0.06 & -0.0086 \\
-CO & 0.14 & -0.051 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.065 \\
-CMS & -0.18 & 0.28 & 0.0017 & -0.27 & 0.032 & 0.19 & 0.36 & 0.22 \\
-Caries & & -0.13 & -0.27 & -0.19 & -0.037 & -0.16 & 0.079 & -0.16 \\
+\bottomrule\noalign{}
+\endlastfoot
+OA & -0.12 & -0.074 & 0.21 & 0.07 & 0.14 & 0.28 & 0.00098 & -0.067 \\
+VOP & -0.088 & -0.16 & 0.34 & 0.061 & 0.25 & -0.06 & 0.013 & -0.13 \\
+SN & -0.24 & 0.16 & 0.095 & 0.089 & 0.17 & 0.24 & 0.16 & 0.093 \\
+DDD & -0.0038 & 0.0037 & 0.19 & -0.39 & -0.077 & 0.31 & 0.06 &
+-0.0086 \\
+CO & 0.064 & -0.051 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.065 \\
+CMS & -0.19 & 0.28 & 0.0017 & -0.27 & 0.032 & 0.19 & 0.36 & 0.22 \\
+Caries & & -0.2 & -0.36 & -0.15 & -0.17 & -0.21 & -0.0045 & -0.22 \\
 Nicotine & & & -0.21 & 0.01 & -0.014 & 0.43 & 0.14 & 0.98 \\
@@
 Caffeine & & & & & & & & 0.078 \\
-\bottomrule()
 \end{longtable}
@@
 and is likely to cause spurious correlations. Strong correlations were
-found between cotinine and nicotine (0.851). Moderate correlations were
-found between OA and DDD (0.484), VOP and periodontitis (0.491), SN and
-cotinine (0.558), DDD and calculus (-0.421), CMS and caffeine (0.528),
-caries and periodontitis (0.486), caries and theophylline (-0.491),
-periodontitis and caries (0.486), periodontitis and caffeine (0.515),
-nicotine and CMS (0.496), calculus and caries (0.502), age-at-death and
-theophylline (-0.447), theophylline and age-at-death (-0.447), caffeine
-and periodontitis (0.515), cotinine and CMS (0.425). Remaining
-correlations were weak or absent (Figure~\ref{fig-polycorr}).
-Correlations with age will be depressed because age was largely
-controlled for in the sample selection.
+found between cotinine and nicotine (0.847). Moderate correlations were
+found between OA and DDD (0.47), VOP and periodontitis (0.487), SN and
+cotinine (0.559), DDD and calculus (-0.416), CMS and caffeine (0.53),
+caries and periodontitis (0.523), periodontitis and VOP (0.487),
+periodontitis and age-at-death (0.407), nicotine and SN (0.53), calculus
+and DDD (-0.416), age-at-death and theophylline (-0.45), theophylline
+and age-at-death (-0.45), caffeine and periodontitis (0.494), cotinine
+and CMS (0.427). Remaining correlations were weak or absent
+(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
+because age was largely controlled for in the sample selection.
 
@@
 al., 2021), and the caries percentage is similarly lower in our sample
-(12.7\%) than a more representative sample (22.9\%) (Lemmers et al.,
+(17.6\%) than a more representative sample (22.9\%) (Lemmers et al.,
 2013).\\
@@
 All raw data is available on Zenodo
-(\url{https://doi.org/10.5281/zenodo.7648757}). Analysis scripts, and
+(\url{https://doi.org/10.5281/zenodo.8061483}). Analysis scripts, and
 the source code for the manuscript and supplementary materials are
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-aten400JaarBeemster2012}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-aten400Jaar2012}{}}%
 Aten, D., Bossaers, K. W. J. M., \& Misset, C. (2012). \emph{400 jaar
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-leuwProhibitionLegalizationDutch1994}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-leuwProhibitionLegalization1994}{}}%
 Leuw, E., \& Marshall, I. H. (1994). \emph{Between {Prohibition} and

### 2023-07-26 analysis/supplementary-materials/supp-mat.qmd
prepare supp mat for submission

@@
     code-fold: true
+    embed-resources: true
   pdf:
@@
 
+These supplementary figures and tables are a variable hodgepodge of things that
+didn't fit in the main manuscript, and additional things I thought might be
+useful. The best way to explore/verify the results and interpretations is to
+download all the data and code (<https://doi.org/10.5281/zenodo.7649824>)
+and just play around with it yourself. Enjoy!
+
 ## Samples
+
+Calculus samples are a combination of leftovers from a previous aDNA study and
+newly sampled individuals. In some cases individuals from the previous study were
+sampled again (if not enough calculus was left over from the previous study)
+(@tbl-sampling).
 
@@
 middle adult male individuals (@fig-age-distribution). Middle adult males were
-preferentially targeted and the sample age and sex distribution is therefore not
-representative of the population. This was done to limit potential confounding
+preferentially targeted due to larger calculus deposits (observation) and the
+sample age and sex distribution is therefore not
+representative of the population. This was also done to limit potential confounding
 factors, and because pipe notches, which served as a positive control for tobacco,
@@
 
-Missing data per dataset
-
-```{r}
-#| label: fig-missing-caries
+An overview of the missing teeth can be found in @fig-missing-teeth. Missing
+scores per tooth can be found in @tbl-missing-scores and @fig-missing-dental.
+
+```{r}
+#| label: fig-missing-teeth
+#| fig-cap: "Heatmap of missing teeth per individual in the sample. 1 = present, 2 = missing."
+dental_inv_long %>% 
+  mutate(status = if_else(status == "p", 1, 0)) %>% 
+  ggplot(aes(x = tooth, y = id, fill = status)) +
+    geom_tile()
+```
+
+```{r}
+#| label: missing-values-teeth
+caries_missing <- caries %>%
+  mutate(across(!id, ~ is.na(.x))) %>%
+  select(!id) %>%
+  colSums() %>% 
+  as_tibble_row()
+
+periodont_missing <- periodont %>%
+  mutate(across(!id, ~ is.na(.x))) %>%
+  select(!id) %>%
+  colSums() %>% 
+  as_tibble_row()
+
+periap_missing <- periap %>%
+  mutate(across(!id, ~ is.na(.x))) %>%
+  select(!id) %>%
+  colSums() %>% 
+  as_tibble_row()
+
+missing_tbl <- tibble(caries_missing) %>% 
+  add_case(periodont_missing) %>% 
+  add_case(periap_missing) %>% 
+  mutate(score = c("caries", "periodontitis", "periapical"), .before = 1)
+```
+
+```{r}
+#| label: tbl-missing-scores
+#| tbl-cap: "Table of missing scores by tooth."
+knitr::kable(missing_tbl)
+```
+
+```{r}
+#| label: fig-missing-dental
+#| fig-cap: "Plots"
+#| layout-ncol: 2
+#| fig-subcap: 
+#|   - "Caries"
+#|   - "Periodontitis"
+#|   - "Periapical lesions"
+#|   - "Combined caries, periodontitis, periapical."
 caries %>%
   mutate(across(!id, ~ is.na(.x))) %>% 
-  select(!id) %>%
-  colSums()
-
-caries %>%
+  pivot_longer(-id, names_to = "tooth") %>% 
+  mutate(value = if_else(value == TRUE, "missing", "present")) %>% 
+  ggplot(aes(x = tooth, y = id, fill = value)) +
+    geom_tile()
+
+periodont %>%
   mutate(across(!id, ~ is.na(.x))) %>% 
-  pivot_longer(-id) %>% 
-  ggplot(aes(x = name, y = id, fill = value)) +
+  pivot_longer(-id, names_to = "tooth") %>% 
+  mutate(value = if_else(value == TRUE, "missing", "present")) %>% 
+  ggplot(aes(x = tooth, y = id, fill = value)) +
     geom_tile()
-```
-
-Missing data across all caries, periodontitis, and periapical datasets.
-
-```{r}
-#| label: fig-missing-dental
-#| fig-cap: "Number of missing values per individual and tooth across all dental scores (caries, periodontitis, periapical lesions, calculus)."
+
+periap %>% 
+  mutate(across(!id, ~ is.na(.x))) %>% 
+  pivot_longer(-id, names_to = "tooth") %>% 
+  mutate(value = if_else(value == TRUE, "missing", "present")) %>% 
+  ggplot(aes(x = tooth, y = id, fill = value)) +
+    geom_tile()
+
 dental_long %>%
@@
   summarise(presence = sum(presence)) #%>%
-  # filter(presence == 0 | presence == 2) %>%
-  # group_by(compound) %>%
-  # mutate(presence = if_else(presence == 2, 1, 0)) # convert replications to presence/absence
 
@@
   filter(id %in% filter(metadata, replicated == TRUE)$id) %>%
-  #right_join(select(filter(metadata, replicated == TRUE)), by = c("id", "sample"))
-  #mutate(compound = str_remove(compound, "_calc")) %>%
   group_by(id, sample, compound) %>% # combine batches
@@
   filter(batch == "batch1") %>%
-  #select(sample, contains("batch1")) %>%
-  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
   select(!c(sample, conc, extraction, batch, presence)) %>%
@@
   filter(batch == "batch2") %>%
-  #select(sample, contains("batch1")) %>%
-  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
   pivot_wider(names_from = "compound", values_from = "quant") %>%
-  select(!c(extraction, batch))
+  select(!c(extraction, batch)) %>% 
+  knitr::kable()
 ```
@@
 in the replicated samples.
-
-```{r}
-#| label: tbl-uhplc-replicated
-uhplc_calculus_replicated %>%
-  filter(sum(presence) != 0) %>% # remove compounds absent in all individuals in sample
-  pivot_wider(names_from = "compound", values_from = "presence")
-```
-
-<!-- line plot with concentration across washes and to the final calculus sample -->
 
@@
 
-```{r auth-plot-batch1}
+```{r}
 #| label: fig-auth-plot-batch1
-
+#| fig-cap: "Plot of extracted quantities of each compound across the three washes and calculus extraction in batch 1. Each line represents an individual."
 uhplc_data_long %>%
@@
     geom_point(size = 0.2) +
-    facet_wrap(~ compound, scales = "free_y") +
+    facet_wrap(~ compound, scales = "free_y", ncol = 3) +
     theme_bw()
@@
 #| label: fig-auth-plot-batch2
+#| fig-cap: "Plot of extracted quantities of each compound across the three washes and calculus extraction in batch 2. Each line represents an individual."
 uhplc_data_long %>%
@@
     geom_point(size = 0.2) +
-    facet_wrap(~ compound, scales = "free_y") +
+    facet_wrap(~ compound, scales = "free_y", ncol = 2) +
     theme_bw()
@@
 the ability to detect compounds; although, we were able to detect
-compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].
+compounds in samples as small as 2 mg (@fig-quant-weight-1 and @fig-quant-weight-2).
 
@@
 #| label: fig-quant-weight-1
-#| #| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 1."
+#| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 1."
 uhplc_data_comb %>%
@@
     theme(legend.position = "none") +
-    labs(x = "", y = "Quantity (ng)")
+    labs(x = "Calculus weight (mg)", y = "Quantity (ng)")
 ```
@@
     theme_bw() +
-    labs(y = "Quantity (ng)", y = "", col = "Sample number")
+    labs(y = "Quantity (ng)", x = "Calculus weight (mg)", col = "Sample number")
 ```
@@
 
-# awkward wrangling required because the information on which individuals were included in replication was lost when combining batches 1 and 2
-#n_batch1 <- length(metadata$replicated)
-#n_batch2 <- sum(metadata$replicated) # sum boolean vector to get length of TRUE values
 uhplc_calculus_long %>%
-  #remove_missing() %>%
   filter(quant > 0) %>% 
-  #mutate(detect = replace_na(conc, 0)) %>% 
-  #mutate(detect = if_else(detect > 0, 1, 0)) %>%
-  #group_by(compound, batch) %>% 
-  # summarise(
-  #   n_detect = sum(detect)
-  #           ) %>% 
-  # mutate(n = case_when(batch == 1 ~ n_batch1,
-  #                      batch == 2 ~ n_batch2)) %>% 
-  # mutate(prop = n_detect / n) %>%
-
-ggplot(aes(x = compound, fill = compound)) +
-    #geom_jitter(width = 0.2) +
-    #geom_boxplot() +
-    geom_bar() +
-    facet_wrap(~ batch) +
-    theme_bw() +
-    theme(
-      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
-      legend.position = "none"
-      )
-```
-
-The replication showed that caffeine, theophylline, cotinine, nicotine, and
-salicylic acid could be consistently detected in the samples, although theophylline
-detection decreased between batches 1 and 2. CBD, CBN, cocaine, and THCA-A was not
-detected at all in the second batch.
-
-<!-- absolute counts in the replicated individuals -->
-
-```{r}
-#| label: fig-compounds-detect2
-#| fig-cap: "Number of individuals in which each compound was detected between batch 1 and 2. Only showing replicated individuals."
-
-uhplc_calculus_long %>%
-  filter(id %in% filter(metadata, replicated == T)$id,
-         quant > 0) %>%
-  #remove_missing() %>%
 ggplot(aes(x = compound, fill = compound)) +
@@
 
+The replication showed that caffeine, theophylline, cotinine, nicotine, and
+salicylic acid could be consistently detected in the samples, although theophylline
+detection decreased between batches 1 and 2. CBD, CBN, cocaine, and THCA-A was not
+detected at all in the second batch.
+
+<!-- absolute counts in the replicated individuals -->
+
+```{r}
+#| label: fig-compounds-detect2
+#| fig-cap: "Number of individuals in which each compound was detected between batch 1 and 2. Only showing replicated individuals."
+
+uhplc_calculus_long %>%
+  filter(id %in% filter(metadata, replicated == T)$id,
+         quant > 0) %>%
+  ggplot(aes(x = compound, fill = compound)) +
+    geom_bar() +
+    facet_wrap(~ batch) +
+    theme_bw() +
+    theme(
+      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
+      legend.position = "none"
+      )
+```
+
 ### Detection and preservation
@@
 #| label: fig-detection-preservation
+#| fig-cap: "Plot of relationship between the absolute quantity of a detected compound (ng) and the overall skeletal preservation of the individuals in which the compound was detected.
+Showing results for batch 1."
 uhplc_calculus_long %>%
@@
     quant > 0,
-    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
     ) %>% 
@@
     geom_violin(aes(fill = preservation), alpha = 0.6) +
-    #geom_jitter(width = 0.2) +
     geom_boxplot(width = 0.2) +
@@
 #| label: fig-detection-preservation2
+#| fig-cap: "Plot of relationship between the absolute quantity of a detected compound (ng) and the overall skeletal preservation of the individuals in which the compound was detected.
+Showing results for batch 2."
 uhplc_calculus_long %>%
@@
 #| label: fig-preservation-detection
+#| fig-cap: "Plot of the number of skeletons in each state of preservation separated by batch."
 uhplc_calculus_long %>% 
@@
 #| label: fig-corr-notches
+#| fig-cap: "Plot of relationships between the total number of pipe notches in an individual and the concentration of detected compounds.
+The only relevant comparisons in this case are nicotine and cotinine.
+The others are just included because I couldn't be bothered filtering them out."
 # correlation between nicotine and cotinine conc, and pipe notches
@@
     geom_point() +
-    facet_wrap(~ compound + batch, scales = "free_y")
+    facet_wrap(~ compound + batch, scales = "free_y") +
+    theme_bw() +
+    labs(x= "number of pipe notches", y = "detected compound concentration (ng/mg)")
 ```
@@
     detection = sum(detection), 
-    #sex, age, completeness, preservation, pipe_notch, 
     .groups = "keep" # combine compounds: 0 = none detected; 1 = 1 detected; 2 = both detected
@@
 # accuracy in replicated samples
-uhplc_accuracy <- uhplc_calculus_replicated %>%
-  left_join(select(demography, id, pipe_notch), by = "id") %>% 
-  filter(compound == "nicotine" | compound == "cotinine") %>%
+uhplc_accuracy_batch2 <- uhplc_calculus_long %>%
+  filter(compound == "nicotine" | compound == "cotinine",
+         batch == "batch2") %>%
   group_by(id) %>% 
@@
   distinct(id, .keep_all = T) %>% 
-  #slice(which(str_detect(compound, "nicotine|cotinine"))) #%>%
-  # slice(which(compound == "nicotine" | compound == "cotinine")) %>%
   mutate(
@@
 
-```{r}
-#| label: fig-notch-ratio
-#| fig-cap: "Relationship between ratio and number of pipe notches."
-
-tobacco_ratio %>%
-  ggplot(aes(y = ratio, x = pipe_notch)) +
-    #geom_boxplot()
-    geom_point()
-
-  #lm(ratio ~ age + sex + preservation + pipe_notch, tobacco_ratio)
-```
-
 We found no correlation between the number of pipe notches
@@
 No apparent correlation between the number of pipe notches
-and the concentration of nicotine or cotinine [@fig-corr-notches].
+and the concentration of nicotine or cotinine (@fig-corr-notches).
 
@@
 In the replicated samples only, tobacco detection was successful in
-`r nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0))`
+`r nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y", presence > 0))`
 out of
-`r nrow(filter(uhplc_accuracy, pipe_notch == "Y"))`
+`r nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y"))`
 pipe smokers
-(`r scales::percent(nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0)) / nrow(filter(uhplc_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
+(`r scales::percent(nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y", presence > 0)) / nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y")), accuracy = 0.1)`)
 Including individuals with absence of a pipe notch and concurrent absence of
 compounds as a correct identification, gives an overall accuracy of
-`r scales::percent(mean(uhplc_accuracy$correct, na.rm = T), accuracy = 0.1)`.
-
-Nicotine and cotinine detection successfully identified
-`r nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y", detection > 0))`
-smokers out of
-`r nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y"))`
-(`r scales::percent(nrow(filter(tobacco_comb, pipe_notch == "Y", batch == 1, detection > 0)) / nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`)
-individuals with pipe notches in the first analysis.
-
-Nicotine detection successfully identified
-`r nrow(filter(nicotine, batch == 1, pipe_notch == "Y", detection > 0))`
-smokers out of
-`r nrow(filter(nicotine, batch == 1, pipe_notch == "Y"))`
-individuals with pipe notches
-(`r scales::percent(nrow(filter(nicotine, batch == 1, pipe_notch == "Y", detection > 0)) / nrow(filter(nicotine, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`).
-
-Cotinine detection successfully identified
-`r nrow(filter(cotinine, batch == 1, pipe_notch == "Y", detection > 0))`
-smokers out of
-`r nrow(filter(cotinine, batch == 1, pipe_notch == "Y"))`
-individuals with pipe notches
-(`r scales::percent(nrow(filter(cotinine, batch == 1, pipe_notch == "Y", detection > 0)) / nrow(filter(cotinine, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`).
-
-In the replicated samples only, tobacco detection was successful in
-`r nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0))`
-out of
-`r nrow(filter(uhplc_accuracy, pipe_notch == "Y"))`
-smokers
-(`r scales::percent(mean(filter(uhplc_accuracy, pipe_notch == "Y")$correct, na.rm = T))`).
-Including individuals with absence of a pipe notch and concurrent absence of
-compounds as a correct identification, gives an overall accuracy of
-`r scales::percent(mean(uhplc_accuracy$correct, na.rm = T), accuracy = 0.1)`.
+`r scales::percent(mean(uhplc_accuracy_batch2$correct, na.rm = T), accuracy = 0.1)`.
 
@@
 
-```{r}
-#| label: fig-dental-plots
-#| layout-ncol: 2
-#| fig-cap: 
-#|   - "Antemortem tooth loss in sample."
-#|   - "Periodontal scores in the sample."
+An overview of available teeth can be seen in @fig-dental-inv.
+
+```{r}
+#| label: fig-dental-inv
+#| fig-cap: "Overview of the dental inventory of the sample. Teeth removed for DNA analysis considered 'present'."
 dental_long %>%
-  mutate(status = case_when(status == "dna" ~ "p")) %>% 
+  mutate(status = case_when(status == "dna" ~ "p", TRUE ~ status)) %>% 
   dental_plot(fill = status)
-
-dental_long %>%
-  filter(!is.na(periodont)) %>% 
-  dental_plot(fill = periodont)
 ```
@@
 
-Ratio of antemortem lost teeth per present teeth at the site.
-
-```{r}
-#| label: fig-aml-id
-#| fig-cap: "AMTL ratio per individual."
+Ratios of antemortem lost teeth per present teeth at the site. Calculated per
+individual (@tbl-aml-id), tooth class (@tbl-aml-class), and tooth type (@tbl-aml-type)
+
+```{r}
+#| label: tbl-aml-id
+#| tbl-cap: "AMTL ratio per individual."
 dental_long %>% 
   group_by(id) %>% 
-  amtl_ratio(.status = status, .add = T)
-```
-
-```{r}
-#| label: fig-aml-class
-#| fig-cap: "AMTL ratio per tooth class."
+  amtl_ratio(.status = status, .add = T) %>% 
+  knitr::kable()
+```
+
+```{r}
+#| label: tbl-aml-class
+#| tbl-cap: "AMTL ratio per tooth class."
 dental_long %>% 
   group_by(class) %>% 
-  amtl_ratio(.status = status, .add = T)
-```
-
-```{r}
-#| label: fig-aml-type
-#| fig-cap: "AMTL ratio per tooth type."
+  amtl_ratio(.status = status, .add = T) %>% 
+  knitr::kable()
+```
+
+```{r}
+#| label: tbl-aml-type
+#| tbl-cap: "AMTL ratio per tooth type."
 dental_long %>% 
   group_by(type) %>% 
-  amtl_ratio(.status = status, .add = T)
+  amtl_ratio(.status = status, .add = T) %>% 
+  knitr::kable()
 ```
@@
 
-<!--add a caries correction?-->
 Caries were scored as the location on each individual tooth. Multiple locations
@@
 
-```{r}
-surface <- c("mes", "dis", "occ", "buc", "lin", "root", "crown") # here occ = incisal
-
-# caries_count <- dental_long %>% 
-#   select(!c(periodont, periap, calculus)) %>% 
-#   #filter(status == "p") %>%
-#   separate_rows(caries, sep = ";") %>% # one lesion per row
-#   mutate(caries_count = if_else(
-#     caries == "none", 0L, 1L
-#     )
-#   ) %>%
-#   group_by(id, tooth) %>%
-#   summarise(
-#     caries_count = sum(caries_count, na.rm = F)
-#     ) %>%
-#   tooth_position()
-# 
-# site_wide_caries <- caries_count %>%
-#   caries_ratio(.caries = caries_count)
-```
-
 In the `r caries_ratio_site$n_teeth` that were examined, `r caries_ratio_site$count`
@@
 This frequency has very little meaning, and was further broken down into a ratio
-for each individual and each tooth class.
-
-```{r}
+for each individual and each tooth class (@tbl-caries-id-class and @fig-caries-class).
+As expected, the molars have a higher frequency of caries than the other teeth.
+
+```{r}
+#| label: tbl-caries-id-class
+#| tbl-cap: "Table of caries ratios per individual per tooth class."
 caries_count %>%
-  group_by(class) %>% 
-  caries_ratio(.caries = count)
-```
-
-As expected, the molars have a higher frequency of caries than the other teeth.
-
-```{r}
-caries_count %>% 
-  group_by(id, class) %>% 
-  caries_ratio(.caries = count)
-```
-
-```{r}
-caries_count %>%
-  caries_ratio(.caries = count, id, class)
-```
-
-```{r}
+  caries_ratio(.caries = count, id, class) %>% 
+  knitr::kable()
+```
+
+```{r}
+#| label: fig-caries-class
+#| fig-cap: "Plot of caries ratios calculated per individual per tooth class."
 caries_count %>%
@@
     geom_boxplot(width = 0.1) +
-    theme_minimal()
-```
-
-```{r}
-#| label: fig-caries-ratio
+    theme_minimal() +
+    theme(legend.position = "none") +
+    labs(x = "tooth class", y = "caries ratio")
+```
+
+```{r}
+#| label: fig-caries-fun
+#| fig-cap: "Plot of caries rate per tooth in pooled sample from all individuals. Teeth reordered along the x-axis to match position in the mouth (yes, the plot is supposed to resemble a mouth)."
 upper_order <- c(paste0("t", 18:11), paste0("t", 21:28))
@@
           axis.ticks.x = element_blank()) +
-    labs(y = "Caries rate")
+    labs(y = "caries ratio")
 
@@
 horizontal bone loss from the CEJ to the alveolar bone, accounting for ca. 2mm
-of gingival thickness.
+of gingival thickness. The distribution of scores in the pooled sample dentitions
+can be seen in @fig-periodont-scores.
+
+```{r}
+#| label: fig-periodont-scores
+#| fig-cap: "Distribution of periodontitis scores in each tooth (FDI notation) in the pooled sample."
+periodont %>% 
+  dental_longer(-id) %>% 
+  remove_missing(vars = "score") %>% 
+  dental_plot(fill = score)
+```
 
@@
   dental_longer(-id) %>% 
-    #   group_by(id, quadrant) %>%
-    # summarise(
-    #   n_surf = sum(!is.na(score)),
-    #   score_sum = sum(score, na.rm = T),
-    #   calc_index = score_sum / n_surf
-    # )
-
   calculus_index()
@@
 
-Distribution of calculus indices within the sample.
-<!-- it is not calculating per surface (I think) -->
-
-```{r}
-#| label: fig-calculus-index
+Calculus was scored on each tooth surface (interproximal surfaces were given a single score)
+on a scale of 0-3, representing absence of calculus (0) to heavy deposit (3).
+Distribution of individual calculus indices within the sample, separated by
+quadrant shows that the lower anterior quadrant had the largest deposits (@fig-calculus-quad).
+
+No apparent influence of lower anterior calculus index on the presence/absence
+of a compound (or vice versa) (@fig-calc-compound).
+
+```{r}
+#| label: fig-calculus-quad
+#| fig-cap: "Calculus index per quadrant. LA = lower anterior, LP = lower posterior, UA = upper anterior, UP = upper posterior."
 calc_index %>%
@@
 ```{r}
+#| label: fig-calc-compound
+#| fig-cap: "Relationship between the presence (1) or abesence (0) of a compound and the calculus index of the lower anterior quadrant of an individual."
 calc_index %>%
@@
     geom_boxplot(width = 0.1) +
-    facet_wrap(~ compound)
-```
-
-No apparent influence of lower anterior calculus index on the presence/absence
-of a compound (or vice versa).
+    facet_wrap(~ compound) +
+    labs(x = "Presence/absence", y = "Calculus index (LA)") +
+    theme_bw() +
+    theme(legend.position = "none")
+```
 
@@
 Point-biserial (Pearson) correlation was conducted on compound concentrations,
-calculus index, caries ratio, and binary variables.
+calculus index, caries ratio, and binary variables (@fig-pearson-corr).
 This is done to see if any correlations exist prior to discretisation of continuous
-variables.
+variables. Irrelevant correlations (anything not between two continuous or a
+continuous and binary variable) are removed from the plot.
 
@@
 #| label: fig-pearson-corr
+#| fig-cap: "Pearson correlation plot."
 conc_cor %>%
@@
 
-The calculus index and caries ratio for each individual was converted to an
+Before analysing the sample with a polychoric correlation (@tbl-polycorr and @fig-polycorr),
+the calculus index and caries ratio for each individual was converted to an
 ordinal variable by using quartiles, providing a score from 0--4.
@@
 
-```{r polycorr}
-polycorr$rho
-```
-
-```{r fig-polycorr}
+```{r}
+#| label: tbl-polycorr
+#| tbl-cap: "Table of polychoric correlations (rho)."
+knitr::kable(polycorr$rho, digits = 3)
+```
+
+```{r}
+#| label: fig-polycorr
+#| fig-cap: "Heatmap of polychoric correlations (rho)."
 ggcorrplot::ggcorrplot(polycorr$rho, type = "lower", show.diag = F)
@@
 We only found a weak negative correlation
-between caffeine concentration and dental calculus index, but since we targeted individuals with
+between caffeine concentration and dental calculus index (@fig-pearson-corr),
+but since we targeted individuals with
 calculus, this may inadvertently have been controlled for during sample collection.
@@
 
-All raw data are available for download from Zenodo (<https://doi.org/10.5281/zenodo.7648757>).
-
-### lloq.tsv
-
-| variable | description |
-|---|---|
-| compound | name of target compound |
-| lloq | Lower limit of quantitation |
+All raw data are available for download from Zenodo (<https://zenodo.org/record/8061483>).
 
@@
 
+### lloq.tsv
+
+| variable | description |
+|---|---|
+| compound | name of target compound |
+| lloq | Lower limit of quantitation |
+
+### uhplc-results(_batch2).csv
+
+| variable | description |
+|---|---|
+| sample | UHPLC-MS/MS sample number |
+| weight | weight of calculus sample in ng before washes |
+| weight_wash<1..3> | weight of calculus sample in ng following each wash |
+| weight_avg | mean of weights |
+| <compound>_wash<1..3> | Extracted quantity of compound in ng from washes |
+| <compound>_calc | Extracted quantity of compound in ng from calculus |
+
 ### dental-inv.csv
@@
 | id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and status |
+| t11..t48 | status of tooth (FDI notation for variable name) |
 
@@
 - aml = ante-mortem loss
-- dna = removed for DNA sampling
+- dna = previously removed for DNA sampling
 
@@
 | id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and location of caries |
+| t11..t48 | location of caries lesion(s) |
 
@@
 | id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and periodontitis score (0-3) |
+| t11..t48 | periodontitis score (0-3) |
 
@@
 | id | unique identifier for the individual |
-| t11..t48 | Tooth number (FDI system) and periapical lesion |
+| t11..t48 | location of periapical lesion |
 
@@
 | id | unique identifier for the individual |
-| t11_bucc..t48_ip | Tooth number (FDI system) and surface (bucc = buccal; lin = lingual; ip = interproximal) and deposit size (0-3) |
+| t11_bucc..t48_ip | calculus deposit size (0-3) per tooth surface (bucc = buccal; lin = lingual; ip = interproximal) |
 
@@
 
+### sinusitis.csv
+
+| variable | description |
+|---|---|
+| id | Individual ID |
+| CMS | Presence (YES) or absence (NO) of chronic maxillary sinusitis |
+| IPR | Presence (YES) or absence (NO) of periosteal reaction on visceral surface of ribs |
+
+### path-conditions.csv
+
+| variable | description |
+|----------|-------------|
+| id | Individual ID |
+| OA | Presence/absence of lesions related to osteoarthritis |
+| IVDD | Presence/absence of lesions related to inter-vertebral disc disease |
+|	TB | Presence/absence of lesions related to tuberculosis |
+| Mastoiditis | Presence/absence of lesions related to mastoiditis |
+| DISH | Presence/absence of lesions related to diffuse idiopathic skeletal hyperostosis |
+| VOP | Presence/absence of lesions related to vertebral osteophytosis |
+| SN | Presence/absence of lesions related to Schmorl's node(s) |
+| DDD | Presence/absence of lesions related to degenerative disc disease |
+| PNBF | Presence/absence of periosteal new bone formation |
+| OD | Presence/absence of lesions related to osteochondritis dissecans |
+| CF | Presence/absence of lesions related to cribra femora |
+| CO | Presence/absence of lesions related to cribra orbitalia |
+
 ## Session information
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+This report was generated on `r Sys.Date()` using the following computational environment and dependencies:
 
@@
 ```
+
+## References {-}

### 2023-07-30 R/calculus_index.R
clean up functions and analysis code

@@ -3,6 +3,8 @@
 #' @details Calculate the calculus index for each quadrant of an individual.
 #' Based on Greene et al. 2005.
 #' @param .data Data frame in long format, with one tooth surface per row, and one calculus score per surface.
+#' @param simple Whether data should be grouped by id (TRUE) or id and quadrant (FALSE).
+#' @importFrom dplyr group_by mutate summarise ungroup
 #' @return Returns a data frame with four rows per individual containing the calculated calculus
 #' index for each quadrant (CI-UP, CI-LP, CI-UA, CI-LA).
 #' @export
@@ -47,6 +49,7 @@ calculus_index <- function(.data, simple = FALSE){
 #' @param .data Data frame in long format, with one tooth surface per row, and one calculus score per surface.
 #' @param .caries Caries count variable.
 #' @param ... Can be used to add arguments to `group_by()`. Group by tooth, tooth type, individual, etc.
+#' @importFrom dplyr group_by summarise
 #' @return Returns a data frame with caries ratio per group.
 #' @export
 
@@ -69,6 +72,7 @@ caries_ratio <- function(.data, .caries, ...){
 #' @param .data Data frame in long format, with one tooth surface per row, and one calculus score per surface.
 #' @param .status character. dental status (p, m, aml, dna).
 #' @param ... Can be used to add grouping variable(s). Group by tooth, tooth type, individual, etc.
+#' @importFrom dplyr group_by summarise
 #' @return Returns a data frame with amtl ratio per group.
 #' @export
 

### 2023-07-30 R/data.R

@@ -27,16 +27,27 @@
 #' \describe{
 #'   \item{compound}{Name of the compound.}
 #'   \item{lloq}{Lower limit of quantitation (ng)}
-#'   \item{element}{FDI notation for tooth/teeth sampled for each individual.}
-#'   \item{KZ_sample}{Dental calculus sample number on tubes from previous study.}
-#'   \item{KZ_element}{FDI notation for tooth/teeth sampled from the previous study for each individual.}
-#'   \item{replicated}{Whether or not the sample was replicated in a separate UHPLC-MS/MS analysis.}
-#'   \item{batch1_weight}{weight (mg) of dental calculus sample analysed in the first batch.}
-#'   \item{batch2_weight}{weight (mg) of dental calculus sample analysed in the replication batch.}
 #'   ...
 #' }
 #' @source https://doi.org/10.5281/zenodo.8010630
 "lloq"
+
+#' UHPLC analysis data from both batches
+#'
+#' Contains the results for each sample from each extraction step (washes, and dissolution)
+#' analysed by UHPLC in the first and second (replication) batch.
+#' @format ## `uhplc_data_comb`
+#' A data frame with 42 rows and 91 columns:
+#' \describe{
+#'   \item{sample}{number on the sample tube.}
+#'   \item{<compound>_wash{1..3}_batch{1..2}}{Absolute quantity (ng) of compound detected in sample washes.}
+#'   \item{<compound>_calc_batch{1..2}}{Absolute quantity (ng) of compound detected after sample dissolution.}
+#'   \item{batch{1..2}_weight}{Weight of the calculus sample in mg.}
+#'   ...
+#' }
+#' @source https://doi.org/10.5281/zenodo.8010630
+"uhplc_data_comb"
+
 
 #' Middenbeemster sample demographics
 #'

### 2023-07-30 R/dental-plot.R

@@ -3,6 +3,8 @@
 #' @param .data Data frame containing...
 #' @param fill What variable is used for the 'fill' argument. Passed to ggplot().
 #' @param notation Currently not in use. May be used to support different notation in the future.
+#' @importFrom dplyr filter mutate
+#' @import patchwork
 #' @export dental_plot
 
 dental_plot <- function(.data, fill, notation){
@@ -36,6 +38,6 @@ dental_plot <- function(.data, fill, notation){
     theme(axis.title.x = element_blank(),
           panel.grid = element_blank())
 
-  require(patchwork)
+  #requireNamespace(patchwork)
   maxilla / mandible + plot_layout(guides = "collect")
 }

### 2023-07-30 R/utils.R

@@ -16,6 +16,8 @@ NULL
 #' Produce detailed tooth region information from tooth number (FDI notation)
 #'
 #' @param .data data frame containing tooth number (t + FDI notation)
+#' @importFrom dplyr if_else case_when
+#' @importFrom stringr str_detect
 #' @return Returns a data frame with tooth number, region (maxilla, mandible),
 #' position (anterior, posterior), side, tooth_type (incisor, canine, etc.)
 #' @export tooth_position
@@ -112,6 +114,7 @@ dental_longer <- function(.data, cols){
 #'
 #' @param .data A data frame containing one column with a unique identifier and
 #' multiple columns with calculus scores from each surface of a tooth.
+#' @importFrom dplyr group_by summarise arrange slice select filter anti_join
 #' @return Returns a warning if data frame does not fulfill criterion.
 #' @export dentition_checker
 dentition_checker <- function(.data){
@@ -141,58 +144,6 @@ dentition_checker <- function(.data){
 
 #' Function to convert compound abbreviations to full name
 
-# compound_name_repair <- function(.data){
-#   # repair abbreviated compound names
-#     .data %>%
-#     rename(
-#       "THCA-A" = "thca-a",
-#       "Cocaine" = "cocaine",
-#       "Caffeine" = "caffeine",
-#       "Theophylline" = "theophyl",
-#       "Cotinine" = "cotinine",
-#       "Nicotine" = "nicotine",
-#       "Salicylic acid" = "salicyl",
-#       "CBN" = "cbn",
-#       "THCVA" = "thcva",
-#       "THC" = "thc",
-#       "CBD" = "cbd"
-#     )
-# }
-
-# compound_name_repair <- function(.data, ...){
-#   # repair abbreviated compound names
-#   .data %>%
-#     pivot_wider( {{ ... }} ) %>%
-#     rename(
-#       "THCA-A" = "thca-a",
-#       "Cocaine" = "cocaine",
-#       "Caffeine" = "caffeine",
-#       "Theophylline" = "theophyl",
-#       "Cotinine" = "cotinine",
-#       "Nicotine" = "nicotine",
-#       "Salicylic acid" = "salicyl",
-#       "CBN" = "cbn",
-#       "THCVA" = "thcva",
-#       "THC" = "thc",
-#       "CBD" = "cbd"
-#     )
-# }
-
-# compound_name_repair <- function(.data, var){
-#
[…]

### 2023-07-30 analysis/scripts/setup-qmd.R

@@ -13,25 +13,6 @@ library(readr)
 devtools::load_all()
 try(generate_bib())
 
-# upload data
-#metadata <- read_tsv(here("analysis/data/raw_data/metadata.tsv"))
-#demography <- read_csv(here("analysis/data/raw_data/demography.csv"))
-#lloq <- read_tsv(here("analysis/data/raw_data/lloq.tsv"))
-#uhplc_data_comb <- read_csv(here("analysis/data/derived_data/uhplc-data_combined.csv"))
-#dental_inv <- read_csv(here("analysis/data/raw_data/dental-inv.csv"))
-#caries <- read_csv(here("analysis/data/raw_data/caries.csv"))
-#periodont <- read_csv(here("analysis/data/raw_data/periodontitis.csv"))
-#periap <- read_csv(here("analysis/data/raw_data/periapical.csv"))
-#calculus <- read_csv(here("analysis/data/raw_data/calculus.csv"))
-#calculus_full <- read_csv(here("analysis/data/raw_data/calculus_full.csv"))
-#sinusitis_clean <- read_csv(here("analysis/data/derived_data/sinusitis_cleaned.csv"))
-#path_cond_clean <- read_csv(here("analysis/data/derived_data/path-conditions_cleaned.csv"))
-
-#source(here("analysis/scripts/setup-qmd.R"))
-#source(here("analysis/supplementary-materials/supp-mat.qmd"), local = knitr::knit_global())
-# set viridis as default pallette
-
-
 
 # UHPLC analysis ----------------------------------------------------------
 
@@ -69,20 +50,8 @@ quant_filter <- uhplc_data_long %>%
   filter(sum > 0) %>%
   select(!sum)
 
-# uhplc_calculus_id <- uhplc_calculus %>%
-#   mutate(sample = as.character(sample)) %>%
-#   left_join(select(metadata, id, sample), by = "sample") # add individual IDs to data frame
-
 uhplc_calculus_long <- uhplc_data_long %>%
-  # pivot_longer(
-  #   -c(sample, id),
-  #   names_to = c("compound", "batch"),
-  #   names_pattern = "(.*)_(.*)",
-  #   values_to = c("quant")
-  # ) %>%
   filter(extraction == "calc") %>%
-  #mutate(compound = str_remove(compound, "_calc")) %>%
-  #remove_missing() %>%
   left_join(weight) %>%  # add weight of calculus samples
[…]

### 2023-07-30 README.Rmd
re-build README

@@
 This repository contains the data and code for our paper:
-
+<!--
 > Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
@@
 Our pre-print is online here:
-
+-->
 > Bartholdy, B.P., Hasselstrøm, J.B., Sørensen, L.K., Casna, M., Hoogland, M., Historisch Genootschap Beemster, & Henry, A.G.
@@
 
+Use the above paper to cite this compendium.
+<!--
 Please cite this compendium as:
@@
 Online at <https://doi.org/10.5281/zenodo.7649825>
+-->
 
@@
     and tables in this file)
-  - [:file\_folder: data](/analysis/data): Data used in the analysis.
-  - [:file\_folder: figures](/analysis/figures): Plots and other
+  - [:file\_folder: data](/data/): Data output from `data-raw/*.R`, used in the analysis (*.rda*).
+  - [:file\_folder: analysis/figures](/analysis/figures): Plots and other
     illustrations
@@
     prepared and collected during the analysis.
+  - [:file\_folder: R](/R/): R functions.
+  - [:file\_folder: data-raw](/data-raw/): R script to download and process the
+    raw data from Zenodo (https://zenodo.org/record/8061483).
 
@@
 [DESCRIPTION](/DESCRIPTION) file).
-- finally, open `analysis/paper/paper.qmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
+- finally, open `analysis/paper/paper.qmd` and knit to produce the PDF manuscript (`_output/`),
+  or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console.
 

### 2023-07-31 analysis/supplementary-materials/supp-mat.qmd
remove unused files and add code-folding to supp mat

@@
   html:
+    format-links: false
+    echo: true
     toc: true

### 2023-08-15 README.Rmd
add github actions to build README.md on push

@@
 This repository contains the data and code for our paper:
+
 <!--
@@
 -->
+
 > Bartholdy, B.P., Hasselstrøm, J.B., Sørensen, L.K., Casna, M., Hoogland, M., Historisch Genootschap Beemster, & Henry, A.G.
@@
 Use the above paper to cite this compendium.
+
 <!--
@@
   - [:file\_folder: R](/R/): R functions.
-  - [:file\_folder: data-raw](/data-raw/): R script to download and process the
-    raw data from Zenodo (https://zenodo.org/record/8061483).
+  - [:file\_folder: data-raw](/data-raw/): R script to download and process the raw data from Zenodo (https://zenodo.org/record/8061483).
 
@@
 [DESCRIPTION](/DESCRIPTION) file).
-- finally, open `analysis/paper/paper.qmd` and knit to produce the PDF manuscript (`_output/`),
+- finally, open `analysis/paper/paper.qmd` and **render** to produce the PDF manuscript (`_output/`),
   or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console.

### 2023-08-15 README.Rmd
Update README.Rmd

@@
     for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
-    `paper.pdf`, suitable for reading (the code is replaced by figures
-    and tables in this file)
+    generated by the analysis.
   - [:file\_folder: data](/data/): Data output from `data-raw/*.R`, used in the analysis (*.rda*).
@@
     prepared and collected during the analysis.
-  - [:file\_folder: R](/R/): R functions.
+  - [:file\_folder: R](/R/): R functions required in the analysis.
   - [:file\_folder: data-raw](/data-raw/): R script to download and process the raw data from Zenodo (https://zenodo.org/record/8061483).
@@
 installed on your computer the [R software](https://cloud.r-project.org/)
-itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).
+itself, optionally (but recommended) [RStudio Desktop](https://rstudio.com/products/rstudio/download/), and [Quarto](https://quarto.org/docs/get-started/).
 

### 2023-09-12 analysis/paper/paper.tex
add R package check workflow

@@
 \usepackage{lineno}
-\linenumbers
 \makeatletter
@@
   pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
-  pdfauthor={Bjørn Peare Bartholdy1,; Jørgen B. Hasselstrøm2; Lambert K. Sørensen2; Maia Casna1; Menno Hoogland1; Historisch Genootschap Beemster3; Amanda G. Henry1},
+  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
   pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
@@
 population}
-\author{}
+\author{Bjørn Peare Bartholdy \and Jørgen B. Hasselstrøm \and Lambert K.
+Sørensen \and Maia Casna \and Menno Hoogland \and Historisch Genootschap
+Beemster \and Amanda G. Henry}
 \date{}
@@
 \begin{document}
-\maketitle
+%% modified from: https://github.com/quarto-journals/plos/blob/main/_extensions/plos/partials/before-body.tex
+\begin{flushleft}
+{\LARGE
+\textbf\newline{Multiproxy analysis exploring patterns of diet and
+disease in dental calculus and skeletal remains from a 19th century
+Dutch population}
+}
+\newline
+
+\vspace{12pt}
+% Insert author names, affiliations and corresponding author email
+
+\normalsize{Bjørn Peare
+Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
+Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
+Sørensen}\textsuperscript{2}, \normalsize{Maia
+Casna}\textsuperscript{1}, \normalsize{Menno
+Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
+Beemster}\textsuperscript{3}, \normalsize{Amanda G.
+Henry}\textsuperscript{1}
+\\
+\bigskip
+\small{1} \small{Department of Archaeological Sciences, Leiden
+University}\\ \small{2} \small{Department of Forensic Medicine, Aarhus
+University}\\ \small{3} \small{Historisch Genootschap Beemster}
+\newline
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* b.p.bartholdy@arch.leidenuniv.nl
+
+\end{flushleft}
+
+\linenumbers
+
 \begin{abstract}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, boxrule=0pt, sharp corners, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
-
-\textsuperscript{1} Department of Archaeological Sciences, Leiden
-University\\
-\textsuperscript{2} Department of Forensic Medicine, Aarhus University\\
-\textsuperscript{3} Historisch Genootschap Beemster
-
-\textsuperscript{*} Correspondence:
-\href{mailto:b.p.bartholdy@arch.leidenuniv.nl}{Bjørn Peare Bartholdy
-\textless{}b.p.bartholdy@arch.leidenuniv.nl\textgreater{}}
 
@@
 sinusitis; caffeine; tobacco 
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
 

### 2023-09-20 README.Rmd
trigger render-readme action
update README and dependencies - issues during CODECHECK

@@
 
+The **root** directory contains:
+
+  - [:file\_folder: data-raw](/data-raw/): R script to download and process the raw data from Zenodo (https://zenodo.org/record/8061483).
+  - [:file\_folder: data](/data/): Data output from `data-raw/*.R`, used in the analysis (*.rda*).
+  - [:file\_folder: R](/R/): R functions required in the analysis.
+
 The **analysis** directory contains:
@@
     generated by the analysis.
-  - [:file\_folder: data](/data/): Data output from `data-raw/*.R`, used in the analysis (*.rda*).
-  - [:file\_folder: analysis/figures](/analysis/figures): Plots and other
+  - [:file\_folder: figures](/analysis/figures): Plots and other
     illustrations
@@
     prepared and collected during the analysis.
-  - [:file\_folder: R](/R/): R functions required in the analysis.
-  - [:file\_folder: data-raw](/data-raw/): R script to download and process the raw data from Zenodo (https://zenodo.org/record/8061483).
 
 ## How to run in your browser or download and run locally
+
+**If you are running Linux, you will need to install `gfortran`.**
+
+e.g. `sudo apt install gfortran` for Ubuntu.
 
@@
 You can download the compendium as a zip from from this URL:
-[main.zip](/archive/main.zip). After unzipping:
+
+[main.zip](/archive/refs/heads/main.zip).
+
+or clone the repository:
+
+`git clone git@github.com:bbartholdy/mb11CalculusPilot.git`
+
+Then:
+
 - open the `.Rproj` file in RStudio
@@
 - finally, open `analysis/paper/paper.qmd` and **render** to produce the PDF manuscript (`_output/`),
-  or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console.
+  or run `quarto::quarto_render("analysis/paper/paper.qmd")` in the R console.
 

### 2023-09-20 analysis/paper/paper.tex
update README and dependencies - issues during CODECHECK

@@
 sinusitis; caffeine; tobacco 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi
 

### 2023-09-20 README.Rmd
more descriptive title

@@
 )
-# Please put your title here to include it in the file below.
+
 Title <- "Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population"
@@
 
-# mb11CalculusPilot
+# Research compendium for Bartholdy et al. 2023
 

### 2023-10-27 analysis/paper/_discussion.qmd
round 1 of modifications based on reviewer comments

@@
 archaeological site. We applied ultra-high-performance liquid chromatography-tandem
-mass spectrometry (UHPLC-MS/MS), a method that was validated by co-occurrence
+mass spectrometry (UHPLC-MS/MS) using a method that was validated by co-occurrence
 of drugs and metabolites in dental calculus and blood [@sorensenDrugsCalculus2021].
@@
 We extend findings from previous studies on alkaloids in archaeological samples
-by extracting multiple different alkaloids from dental calculus, including
-nicotine, cotinine, caffeine, theophylline, and salicylic
-acid in multiple individuals. The detection of these compounds was
-solidified in a replication analysis on different samples from the same
-individuals.
+by detecting multiple different alkaloids in dental calculus, including
+nicotine, cotinine, caffeine, theophylline, and salicylic acid. The detection of
+these compounds was solidified in a replication analysis on different samples
+from the same individuals.
 Cocaine and multiple cannabinoids were also detected during the first
 analysis, but were not replicated.
-We discuss the implications of these findings in light of historical and
-archaeological evidence for the consumption of these drugs.
+We contextualize these findings within the historical and
+archaeological evidence for consumption of these drugs and dietary compounds.
 
@@
 
-<!-- theophylline and caffeine: coffee and tea? -->
+<!-- theophylline and caffeine: anyone fancy a cuppa? -->
 Theophylline and caffeine were positively correlated in our samples, though to a
@@
 [@velskoDentalCalculus2017], and frequent consumption of all three items would
-be difficult to parse.
+be difficult to parse. Additionally, we do not understand well enough the
+effect of the burial on these specific compounds, and the original concentration of
+the compounds in plants can be quite variable [@kingCautionaryTales2017].
 
-<!-- salicylic acid: diet, medicine, or environmental??? -->
+<!-- salicylic acid: diet, medicine, or environment??? -->
 Salicylic acid was found in all but one individual in our sample.
@@
 by a sharp increase in the final calculus extraction, which is what we would expect
-to see if the salicylic acid was incorporated during life [@fig-auth-plot-batch2].
-However, it has been shown that salicilyc acid is a very mobile organic acid and
-the ubiquitous presence may be due to environmental contamination, which would
-also explain the high quantity in the washes
-[@chenCa2Dependent2001; @badriRegulationFunction2009].
-Given the multiple plausible sources of this residue, it will be necessary to
-explore the extent to which salicylic acid can leach
-into the dental calculus from the soil, and what the rate of degradation is for
-salicylic acid when trapped in dental calculus.
-<!-- perhaps using an oral biofilm model ;) -->
+to see if the salicylic acid was incorporated during life (@fig-auth-plot-batch2).
+It is important to note that, especially with salicylic acid, there is a possibility
+for the compound to enter the calculus through contact with the surrounding soil.
+Salicylic acid is a very mobile organic acid
+[@chenCa2Dependent2001; @badriRegulationFunction2009] and the ubiqutous presence
+in our samples may be explained by the compound leching into the dental calculus
+from the burial environment.
+We can therefore not confidently rule out environmental contamination without
+analysing samples from the surrounding soil.
 
@@
 recommended to smoke it for more immediate effects [@clarkeCannabisEvolution2013].
-A Dutch medicinal use of hemp involved an emulsion prepared from the seeds of
-the plants to treat pain and various stomach ailments. Another
-preparation involving the roots of the plants was used for inflammation, gout,
-and joint pains [@clarkeCannabisEvolution2013].
-The ability to detect cannabinoids in calculus may be limited by their reduced ability
-to diffuse from serum to salivary glands due to an affinity for protein-binding,
-[@coneInterpretationOral2007],
-meaning detection would rely on oral consumption. Even then,
-the overall instability of some cannabinoids could also affect detection
+Dutch medicinal preparations were used to treat a variety of ailments and symptoms,
+including pain, inflammation, various stomach ailments, gout, and joint pains
+[@clarkeCannabisEvolution2013].
+Because cannabinoids have an affinity for protein-binding, they are less likely
+to diffuse from serum to saliva [@coneInterpretationOral2007]. This may make them
+difficult to detect in dental calculus unless the cannibinoids were consumed
+orally; even then,
+the overall instability of some cannabinoids could also limit detection
 [@sorensenEffectAntioxidants2018; @lindholstLongTerm2010].
-However, given the lack of replication, we cannot with security confirm that
+Given the lack of replication, we cannot with security confirm that
 cannabis was used by the Beemster population.
@@
 The absence could also be attributed to postmortem degradation.
-It has been shown that, while abundant in opium, morphine
-degrades rapidly, while thebaine and papaverine are more resistant to various
-ageing processes [@chovanecOpiumMasses2012]. The latter were also absent from
+It has been shown that, while morphine is abundant in opium, it
+degrades rapidly. Thebaine and papaverine are more resistant to various
+ageing processes [@chovanecOpiumMasses2012], however, these were also absent from
 our samples.
@@
 [@mortimerHistoryCoca1901 p. 179; @abucaCocaTrade2019, p.108].
-We considered it possible that it would be present in a sample with
-most individuals originating from the early- to mid-19th century. If corroborated,
-this would have been the first case of coca-leaf-consumption in Europe.
-In our replication batch, we included all
-of the individuals who had been cocaine-positive in the first batch. We were
-unable to replicate any of the cocaine results, and we were unable to detect the
-principal metabolite, benzoylecgonine, in either batch.
+This would have been the first case of coca-leaf-consumption in Europe; however,
+we were unable to replicate any of the cocaine results in the second batch.
 We suspect that the original detection of cocaine was a result of lab contamination
@@
 Tobacco smoke may play a significant role in diseases
-of the upper respiratory tract, including chronic maxillary sinusities
+of the upper respiratory tract, including chronic maxillary sinusitis
 [@rehImpactTobacco2012]. Although
@@
 [@slavinDiagnosisManagement2005].  
-We also observed a moderate positive correlation between chronic maxillary sinusitis and caffeine which
+We also observed a moderate positive correlation between chronic maxillary sinusitis
+and caffeine which
 contradicts previous research linking chronic coffee consumption with a positive
@@
 <!-- limitations of the platform show which questions would be unanswered -->
-Because we targeted individuals with moderate-to-large calculus deposits,
-it is likely a biased sample. The presence of calculus may increase the
+Targeting individuals with moderate-to-large calculus deposits likely biased our
+sample, as the presence of calculus may increase the
 risk of premature death
-[@yaussyCalculusSurvivorship2019], and periodontal disease (which may or may not
-be associated with dental calculus build-up) is a risk-factor for
-respiratory diseases, if periodontal and respiratory pathogens
-enter the bloodstream
+[@yaussyCalculusSurvivorship2019]. Additionally, periodontal disease (often linked
+to the presence of calculus)
+is a risk-factor for respiratory diseases, if periodontal and respiratory
+pathogens enter the bloodstream
 [@azarpazhoohSystematicReview2006; @scannapiecoRoleOral1999; @scannapiecoPotentialAssociations2001].
@@
 Dental calculus is also
-more transient than for example bone, as it can be mechanically removed,
-intentionally or unintentionally, during life, eliminating
-all trace of the alkaloids consumed prior to its removal.  
-Quantitation of the detected compounds may have limited value in archaeological
-samples due to degradation, and will greatly affect our correlations related to
-concentration.
+more transient than for example bone, as it can become dislodged during life,
+intentionally or unintentionally, eliminating all trace of the alkaloids consumed
+prior to its removal.  
 Following burial, compound stability over
@@
 [@mackiePreservationMetaproteome2017; @lindholstLongTerm2010].  
+Due to this, quantitation of the detected compounds may have limited value in
+archaeological samples due to degradation, and will greatly affect our correlations
+related to concentration.
 The detected quantity of a compound will also depend on the quantity in dental
-calculus during life, which is largely controlled by quantity of consumption,
+calculus during life, which is largely controlled by the quantity consumed,
 how often the
 calculus was disrupted/removed, metabolic breakdown of the compound, and inter-
-and intra-individual factors related to stages of biofilm formation, maturation,
-and mineralisation
+and intra-individual factors related to stages of biofilm maturation
 [@lustmannScanningElectron1976; @velskoMicrobialDifferences2019; @zijngeBiofilmArchitecture2010].
-In short, this means it is not really possible to detect the absence of a compound.
-The absence of a compound is not evidence of absence of consumption. This complicates
+In short, this means it is not really possible to equate the absence of a compound
+as evidence for the absence of consumption, which complicates
 the interpretation of our results.
 We have attempted to minimise errors occurring due to this limitation by including
-a relatively large sample of individuals and replicating our analysis. Although
+a relatively large sample of individuals and replicating our analysis. Although,
 given the relatively low detection rate seen in tobacco, this remains a major
@@
 certain alkaloids in oral fluid and, subsequently, dental calculus. Archaeologists,
-though, will likely be responsible for exploring dental calculus specific
+though, will likely be responsible for exploring dental-calculus-specific
 incorporation and retention of alkaloids, as well as their long-term preservation
-in the burial environment.<!-- move to discussion in disseration: Here, *in vitro* studies may provide valuable insights-->
+in the burial environment. Finally, following our experience with salicylic acid,
+we encourage all future studies to ensure that a control sample is taken from the
+soil, either from the soil surrounding the individual, or, ideally, directly from
+the skeletal remains. This should preferably happen before cleaning, but there will
+often be soil left over in cavities (e.g. nasal cavity, orbit, auditory meatus).
 

### 2023-10-27 analysis/paper/_intro.qmd
round 1 of modifications based on reviewer comments

--- 2023-02-07 inst/paper/_intro.qmd
@@
 Alkaloids may enter the oral cavity via two pathways: (1) direct
-incorporation through oral consumption of alkaloid-containing plants, whether
+incorporation through ingestion of alkaloid-containing plants, whether
 deliberate or accidental;
 and (2) passive diffusion as alkaloids and other compounds are transferred from
-plasma to saliva, and then into the oral cavity through the salivary glands
-in the hours to days following consumption [@coneInterpretationOral2007]. The
-relation to plasma
-is why there is often a close correlation between presence (not concentration)
-of drugs in oral fluid and blood
-[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
-The second pathway allows the identification of parent compounds that are not
-consumed orally, as long as they, or their metabolites, are excreted through saliva.
+plasma to saliva, and then gradually secreted into the oral cavity through the
+salivary glands in the hours-to-days following ingestion [@coneInterpretationOral2007].
+The second pathway allows the identification of parent compounds that do not enter
+the mouth (e.g. injection), as long as they, or their metabolites, are excreted
+through the saliva, thus eventually entering the oral cavity.
+
 <!--cite e.g. injection or topical The second pathway may also be more
 detectable, as the excretion of compounds through saliva may take hours to days
-whereas compounds spend less time in the oral cavity during direct consumption-->.
+whereas compounds spend less time in the oral cavity during direct consumption.-->
 
@@
 by various methods to show a record of consumption during life.
+Because drugs may be transferred from plasma to saliva,
+there is often a close correlation between drugs detected in
+oral fluid and blood, though there are differences in detected concentrations
+[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
+This was also shown to be true for dental calculus and blood [@sorensenDrugsCalculus2021],
+making dental calculus a potentially useful substance for detecting ancient
+alkaloids and other dietary compounds.
 

### 2023-10-27 analysis/paper/_methods.qmd
round 1 of modifications based on reviewer comments

--- 2023-02-07 inst/paper/_methods.qmd
@@
 early young adult (18-24 years), late young adult (25-34 years), middle adult (
-35-49 years), old adult (50+ years).
+35-49 years), old adult (50+ years). Preservation was visually scored on a
+four-stage scale (excellent, good, fair, poor) based on the surface condition of
+the bones and the extent of taphonomic degradation.
 
@@
 included in the analysis. Data were
-dichotomised to presence/absence to allow statistical analysis.
+dichotomised to presence/absence to allow for statistical analysis.
 Osteoarthritis was considered present in cases where eburnation was visible

### 2023-10-27 analysis/paper/_results.qmd
round 1 of modifications based on reviewer comments

@@
   mutate(across(!compound, ~ if_else(!is.na(.x), TRUE, FALSE)))
+
+corr_weight_quant <- uhplc_calculus_long %>% 
+  ungroup() %>% 
+  filter(batch == "batch2",
+    quant > 0) %>% 
+  select(id, weight, compound, quant) %>% 
+  pivot_wider(names_from = compound, values_from = quant) %>% 
+  select(where(is.numeric)) %>% 
+  cor(use = "pairwise.complete.obs") %>% 
+  .[1,]
+
+max_quant <- uhplc_calculus_long %>% 
+  filter(batch == "batch2",
+    id != "MB329"
+    ) %>%
+  group_by(compound) %>% 
+  summarise(
+    max_quant = max(quant, na.rm = T),
+    pos_y = 0.8 * max_quant
+    )
+
+corr_labels <- tibble(
+  compound = names(corr_weight_quant)[-1],
+  corr = signif(corr_weight_quant[-1], 2),
+) %>% 
+  inner_join(max_quant) %>% 
+  arrange(compound) 
+#%>% 
+#  mutate(corr = as.character(corr))
+#corr_labels$compounds <- prettified_names[corr_labels$compounds]
 ```
@@
 sample with the highest extracted quantity of nicotine also had the highest extracted
-quantity of cotinine [@fig-auth-plot-batch2].
+quantity of cotinine (@fig-auth-plot-batch2).
 
@@
     theme_bw() +
-    scale_fill_viridis_d() +
+    #scale_fill_viridis_d() +
     scale_x_discrete(labels = compound_names) +
-    scale_fill_viridis_d(labels = c(1,2)) +
+    scale_fill_viridis_d(labels = c(1,2), end = 0.6) +
     theme(
@@
 #| fig-cap: "(A) Violin plot with overlaid box plots depicting the distribution of extracted quantities of each compound from batch 2 separated by state of preservation of the skeleton.
-(B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2."
-#| fig-width: 5
-#| fig-height: 7
+(B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2.
+r = Pearson correlation coefficient."
+#| fig-width: 4.2
+#| fig-asp: 1.6
 
@@
     batch == "batch2",
-    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
+    #id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
     ) %>% 
@@
     quant > 0,
-    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
+    #id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
     ) %>% 
   ggplot(aes(x = weight, y = quant)) +
+    geom_smooth(method = "lm", se = F, col = viridisLite::viridis(1, begin = 0.6)) +
     geom_point(col = viridisLite::viridis(1)) +
-    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
+    geom_text(
+      data = corr_labels, 
+      aes(x = 24, y = pos_y, label = paste("r =", corr)),
+      #vjust = -1, hjust = -0.1) +
+      ) +
+
+    facet_wrap(
+      ~ compound,
+      scales = "free_y",
+      labeller = labeller(compound = compound_names),
+      dir = "v",
+      ncol = 1
+      ) +
+
     theme_bw() +
@@
 Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation.
+Moderate and strong correlations in **bold**.
 OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid; PN = pipe notches."
@@
   as.matrix() %>% 
-  signif(digits = 2)
+  round(digits = 2)
 cols <- 0
@@
 corr_tib_out <- corr_tib_out[-nrow(corr_tib),] # remove last row because it has no values (all duplicate coefficients)
+#mode(corr_tib_out) <- "numeric"
+options(knitr.kable.NA = "")
 corr_tib_out %>% 
+  as_tibble(rownames = "var") %>% 
+  mutate(across(!var, as.numeric)) %>% 
+  tibble::column_to_rownames("var") %>% 
+  mutate(across(everything(), ~kableExtra::cell_spec(.x, bold = if_else(.x >= 0.4, T, F)))) %>%   mutate(across(everything(), ~ str_remove_all(.x, "NA"))) %>% 
+  #kableExtra::cell_spec(bold = if_else(corr_tib_out >= 0.4, T, F)) #%>% 
+  #kableExtra::spec_color(na_color = "fff") %>% 
   knitr::kable(col.names = corr_names)
+  #kableExtra::kbl(col.names = corr_names) %>%
+
 ```
@@
 corr_tib_out <- corr_tib_long %>%
-  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F))
+  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F)) %>% 
+  mutate(corr = signif(corr, 2))
 
 polycorr_tib_out <- polycorr_tib %>%
-  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F))
+  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F)) %>% 
+  mutate(corr = signif(corr, 2))
 ```
@@
 OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl’s nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid."
+#| fig-width: 5
+#| fig-height: 4
 polycorr$rho %>%
@@
     method = "circle", type = "lower", 
-    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::plasma(1, begin = 1)),
+    outline.color = "grey40",
+    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::inferno(1, begin = 0.7)),
     legend.title = expression(rho)

### 2023-10-27 analysis/paper/paper.qmd
prepare paper for PCI submission
update data source and documentation
tweak first page
round 1 of modifications based on reviewer comments

@@
 library(here)
+library(knitr)
 #library(readr)
@@
 
-Keywords: {{< meta keywords >}}
+<!-- Keywords: {{< meta keywords >}} -->
 
@@
 de Groot from the Middenbeemster Historical Society for their input on early
-draft manuscripts and a lovely guided tour of Middenbeemster.
+draft manuscripts and a lovely guided tour of Middenbeemster. We also thank
+Louise Le Meillour for taking on our preprint and providing helpful comments,
+and Mario Zimmerman and two other anonymous reviewers for their time and invaluable
+insights.
 
@@
 
-## Data Availability Statement
+## Data Availability Statement {-}
 
-All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.7648757>).
+All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.8061483>).
 Analysis scripts, and the source code for the manuscript and
@@
 
+## Conflict of interest disclosure {-}
+
+The authors have no conflicts to declare.
+
 ## References {-}

### 2023-10-27 analysis/paper/paper.tex
round 1 of modifications based on reviewer comments

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \usepackage{lineno}
@@
 historic documentation and identification of pipe notches in the
-dentition. Nicotine and/or cotinine was present in 60\% of individuals
+dentition. Nicotine and/or cotinine was present in 56\% of individuals
 with at least one visible pipe notch. We find some influence of skeletal
@@
 especially useful in individuals that are not always well-documented in
-historic documentation, such as rural populations, children and women.
+historic documentation, such as rural populations, and especially
+children and women.
 \end{abstract}
@@
 sinusitis; caffeine; tobacco 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Alkaloids may enter the oral cavity via two pathways: (1) direct
-incorporation through oral consumption of alkaloid-containing plants,
-whether deliberate or accidental; and (2) passive diffusion as alkaloids
-and other compounds are transferred from plasma to saliva, and then into
-the oral cavity through the salivary glands in the hours to days
-following consumption (Cone \& Huestis, 2007). The relation to plasma is
-why there is often a close correlation between presence (not
-concentration) of drugs in oral fluid and blood (Cone \& Huestis, 2007;
-Milman et al., 2011; Wille et al., 2009). The second pathway allows the
-identification of parent compounds that are not consumed orally, as long
-as they, or their metabolites, are excreted through saliva. .
+incorporation through ingestion of alkaloid-containing plants, whether
+deliberate or accidental; and (2) passive diffusion as alkaloids and
+other compounds are transferred from plasma to saliva, and then
+gradually secreted into the oral cavity through the salivary glands in
+the hours-to-days following ingestion (Cone \& Huestis, 2007). The
+second pathway allows the identification of parent compounds that do not
+enter the mouth (e.g.~injection), as long as they, or their metabolites,
+are excreted through the saliva, thus eventually entering the oral
+cavity.
 
@@
 post-excavation cleaning, the alkaloids can then be detected by various
-methods to show a record of consumption during life.
+methods to show a record of consumption during life. Because drugs may
+be transferred from plasma to saliva, there is often a close correlation
+between drugs detected in oral fluid and blood, though there are
+differences in detected concentrations (Cone \& Huestis, 2007; Milman et
+al., 2011; Wille et al., 2009). This was also shown to be true for
+dental calculus and blood (Sørensen et al., 2021), making dental
+calculus a potentially useful substance for detecting ancient alkaloids
+and other dietary compounds.
 
@@
 buried between AD 1829 and 1866, when the majority of individuals were
-interred (Palmer et al., 2016). The main occupation of the inhabitants
-was dairy farming, consisting largely of manual labour prior to the
-industrial revolution (Aten et al., 2012; Palmer et al., 2016).
-
-To reduce the number of potentially confounding factors to account for
-in the analysis, we preferentially selected males from the middle adult
-age category (35-49 years). The sample consists of 27 males, 11 probable
-males, 2 probable females, and 1 female
-(Figure~\ref{fig-sample-demography}). We selected males due to a higher
-occurrence of pipe notches and dental calculus deposits than females
-(unpublished observation).
+interred. The main occupation of the inhabitants was dairy farming,
+consisting largely of manual labour prior to the industrial revolution
+(Aten et al., 2012; Palmer et al., 2016).
+
+For our sample, we preferentially selected males from the middle adult
+age category (35-49 years) to minimise the effect of confounding
+cultural and biological factors. Previous research on Middenbeemster has
+shown a gendered division of labour (Palmer et al., 2016), and there are
+biological differences in dental calculus formation and drug metabolism
+that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
+White, 1997). The sample consists of 27 males, 11 probable males, 2
+probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
+selected males due to a higher occurrence of pipe notches and dental
+calculus deposits than females (unpublished observation).
 
@@
 late young adult (25-34 years), middle adult ( 35-49 years), old adult
-(50+ years).
+(50+ years). Preservation was visually scored on a four-stage scale
+(excellent, good, fair, poor) based on the surface condition of the
+bones and the extent of taphonomic degradation.
 
@@
 population were included in the analysis. Data were dichotomised to
-presence/absence to allow statistical analysis. Osteoarthritis was
+presence/absence to allow for statistical analysis. Osteoarthritis was
 considered present in cases where eburnation was visible on one or more
@@
 Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
-the \textbf{tidyverse} (Hadley Wickham et al., 2019) and visualisations
-were created using \textbf{ggplot2} (H. Wickham, 2016). Polychoric
-correlations were calculated with the \textbf{psych} package (Revelle,
-2022).
+the \textbf{tidyverse} (Wickham et al., 2019) and visualisations were
+created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
+were calculated with the \textbf{psych} package (Revelle, 2022).
 
@@
 nicotine also had the highest extracted quantity of cotinine
-Figure~\ref{fig-auth-plot-batch2}.
+(Figure~\ref{fig-auth-plot-batch2}).
 
@@
 skeleton. (B) Extracted quantity (ng) of compound plotted against
-weights of the calculus samples from batch 2.}
+weights of the calculus samples from batch 2. r = Pearson correlation
+coefficient.}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0952}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1548}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1071}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1413}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0761}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0978}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0652}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0978}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0652}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}@{}}
 \caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
@@
 the second batch. Correlations between pairs of dichotomous variables
-are removed due to incompatibility with a Pearson correlation. OA =
-osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD
-= degenerative disc disease; CO = cribra orbitalia; CMS = chronic
-maxillary sinusitis; SA = salicylic acid; PN = pipe
-notches.}\tabularnewline
+are removed due to incompatibility with a Pearson correlation. Moderate
+and strong correlations in \textbf{bold}. OA = osteoarthritis; VOP =
+vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc
+disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA =
+salicylic acid; PN = pipe notches.}\tabularnewline
 \toprule\noalign{}
@@
 \endlastfoot
-OA & -0.12 & -0.074 & 0.21 & 0.07 & 0.14 & 0.28 & 0.00098 & -0.067 \\
-VOP & -0.088 & -0.16 & 0.34 & 0.061 & 0.25 & -0.06 & 0.013 & -0.13 \\
-SN & -0.24 & 0.16 & 0.095 & 0.089 & 0.17 & 0.24 & 0.16 & 0.093 \\
-DDD & -0.0038 & 0.0037 & 0.19 & -0.39 & -0.077 & 0.31 & 0.06 &
--0.0086 \\
-CO & 0.064 & -0.051 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.065 \\
-CMS & -0.19 & 0.28 & 0.0017 & -0.27 & 0.032 & 0.19 & 0.36 & 0.22 \\
-Caries & & -0.2 & -0.36 & -0.15 & -0.17 & -0.21 & -0.0045 & -0.22 \\
-Nicotine & & & -0.21 & 0.01 & -0.014 & 0.43 & 0.14 & 0.98 \\
-SA & & & & 0.14 & 0.37 & 0.038 & 0.17 & -0.17 \\
-Calculus & & & & & 0.13 & -0.15 & -0.13 & 0.031 \\
-PN & & & & & & -0.16 & 0.18 & -0.0068 \\
-Theophylline & & & & & & & 0.51 & 0.36 \\
-Caffeine & & & & & & & & 0.078 \\
+OA & -0.12 & -0.07 & 0.21 & 0.07 & 0.14 & 0.28 & 0 & -0.07 \\
+VOP & -0.09 & -0.16 & 0.34 & 0.06 & 0.25 & -0.06 & 0.01 & -0.13 \\
+SN & -0.24 & 0.16 & 0.09 & 0.09 & 0.17 & 0.24 & 0.16 & 0.09 \\
+DDD & 0 & 0 & 0.19 & -0.39 & -0.08 & 0.31 & 0.06 & -0.01 \\
+CO & 0.06 & -0.05 & 0.2 & 0.14 & -0.2 & -0.11 & 0.19 & -0.06 \\
+CMS & -0.19 & 0.28 & 0 & -0.27 & 0.03 & 0.19 & 0.36 & 0.22 \\
+Caries & & -0.2 & -0.36 & -0.15 & -0.17 & -0.21 & 0 & -0.22 \\
+Nicotine & & & -0.21 & 0.01 & -0.01 & \textbf{0.43} & 0.14 &
+\textbf{0.98} \\
+SA & & & & 0.14 & 0.37 & 0.04 & 0.17 & -0.17 \\
+Calculus & & & & & 0.13 & -0.15 & -0.13 & 0.03 \\
+PN & & & & & & -0.16 & 0.18 & -0.01 \\
+Theophylline & & & & & & & \textbf{0.51} & 0.36 \\
+Caffeine & & & & & & & & 0.08 \\
 \end{longtable}
@@
 point-biserial (Pearson) correlation correlations were a near-perfect
-positive correlation between cotinine and nicotine (0.982), and moderate
-correlations between theophylline and nicotine (0.432), caffeine and
-theophylline (0.507) (Table~\ref{tbl-pearson}).
+positive correlation between cotinine and nicotine (0.98), and moderate
+correlations between theophylline and nicotine (0.43), caffeine and
+theophylline (0.51) (Table~\ref{tbl-pearson}).
 
@@
 and is likely to cause spurious correlations. Strong correlations were
-found between cotinine and nicotine (0.847). Moderate correlations were
-found between OA and DDD (0.47), VOP and periodontitis (0.487), SN and
-cotinine (0.559), DDD and calculus (-0.416), CMS and caffeine (0.53),
-caries and periodontitis (0.523), periodontitis and VOP (0.487),
-periodontitis and age-at-death (0.407), nicotine and SN (0.53), calculus
-and DDD (-0.416), age-at-death and theophylline (-0.45), theophylline
-and age-at-death (-0.45), caffeine and periodontitis (0.494), cotinine
-and CMS (0.427). Remaining correlations were weak or absent
+found between cotinine and nicotine (0.85). Moderate correlations were
+found between OA and DDD (0.47), VOP and periodontitis (0.49), SN and
+cotinine (0.56), DDD and calculus (-0.42), CMS and caffeine (0.53),
+caries and periodontitis (0.52), periodontitis and VOP (0.49),
+periodontitis and age-at-death (0.41), nicotine and SN (0.53), calculus
+and DDD (-0.42), age-at-death and theophylline (-0.45), theophylline and
+age-at-death (-0.45), caffeine and periodontitis (0.49), cotinine and
+CMS (0.43). Remaining correlations were weak or absent
 (Figure~\ref{fig-polycorr}). Correlations with age will be depressed
@@
 ultra-high-performance liquid chromatography-tandem mass spectrometry
-(UHPLC-MS/MS), a method that was validated by co-occurrence of drugs and
-metabolites in dental calculus and blood (Sørensen et al., 2021). Here
-we have shown that the method can also be successfully applied to
-archaeological dental calculus. We extend findings from previous studies
-on alkaloids in archaeological samples by extracting multiple different
-alkaloids from dental calculus, including nicotine, cotinine, caffeine,
-theophylline, and salicylic acid in multiple individuals. The detection
-of these compounds was solidified in a replication analysis on different
+(UHPLC-MS/MS) using a method that was validated by co-occurrence of
+drugs and metabolites in dental calculus and blood (Sørensen et al.,
+2021). Here we have shown that the method can also be successfully
+applied to archaeological dental calculus. We extend findings from
+previous studies on alkaloids in archaeological samples by detecting
+multiple different alkaloids in dental calculus, including nicotine,
+cotinine, caffeine, theophylline, and salicylic acid. The detection of
+these compounds was solidified in a replication analysis on different
 samples from the same individuals. Cocaine and multiple cannabinoids
 were also detected during the first analysis, but were not replicated.
-We discuss the implications of these findings in light of historical and
-archaeological evidence for the consumption of these drugs.
+We contextualize these findings within the historical and archaeological
+evidence for consumption of these drugs and dietary compounds.
 
@@
 frequent consumption of all three items would be difficult to parse.
+Additionally, we do not understand well enough the effect of the burial
+on these specific compounds, and the original concentration of the
+compounds in plants can be quite variable (King et al., 2017).
 
@@
 to see if the salicylic acid was incorporated during life
-Figure~\ref{fig-auth-plot-batch2}. However, it has been shown that
-salicilyc acid is a very mobile organic acid and the ubiquitous presence
-may be due to environmental contamination, which would also explain the
-high quantity in the washes (Badri \& Vivanco, 2009; Chen et al., 2001).
-Given the multiple plausible sources of this residue, it will be
-necessary to explore the extent to which salicylic acid can leach into
-the dental calculus from the soil, and what the rate of degradation is
-for salicylic acid when trapped in dental calculus.
+(Figure~\ref{fig-auth-plot-batch2}). It is important to note that,
+especially with salicylic acid, there is a possibility for the compound
+to enter the calculus through contact with the surrounding soil.
+Salicylic acid is a very mobile organic acid (Badri \& Vivanco, 2009;
+Chen et al., 2001) and the ubiqutous presence in our samples may be
+explained by the compound leching into the dental calculus from the
+burial environment. We can therefore not confidently rule out
+environmental contamination without analysing samples from the
+surrounding soil.
 
@@
 until the late 19th century did it become recommended to smoke it for
-more immediate effects (Clarke, 2013). A Dutch medicinal use of hemp
-involved an emulsion prepared from the seeds of the plants to treat pain
-and various stomach ailments. Another preparation involving the roots of
-the plants was used for inflammation, gout, and joint pains (Clarke,
-2013). The ability to detect cannabinoids in calculus may be limited by
-their reduced ability to diffuse from serum to salivary glands due to an
-affinity for protein-binding, (Cone \& Huestis, 2007), meaning detection
-would rely on oral consumption. Even then, the overall instability of
-some cannabinoids could also affect detection (Lindholst, 2010; Sørensen
-\& Hasselstrøm, 2018). However, given the lack of replication, we cannot
-with security confirm that cannabis was used by the Beemster population.
+more immediate effects (Clarke, 2013). Dutch medicinal preparations were
+used to treat a variety of ailments and symptoms, including pain,
+inflammation, various stomach ailments, gout, and joint pains (Clarke,
+2013). Because cannabinoids have an affinity for protein-binding, they
+are less likely to diffuse from serum to saliva (Cone \& Huestis, 2007).
+This may make them difficult to detect in dental calculus unless the
+cannibinoids were consumed orally; even then, the overall instability of
+some cannabinoids could also limit detection (Lindholst, 2010; Sørensen
+\& Hasselstrøm, 2018). Given the lack of replication, we cannot with
+security confirm that cannabis was used by the Beemster population.
 
@@
 be attributed to postmortem degradation. It has been shown that, while
-abundant in opium, morphine degrades rapidly, while thebaine and
+morphine is abundant in opium, it degrades rapidly. Thebaine and
 papaverine are more resistant to various ageing processes (Chovanec et
-al., 2012). The latter were also absent from our samples.
+al., 2012), however, these were also absent from our samples.
 
@@
 and there were also issues importing enough viable specimens of coca for
-cocaine extraction (Abduca, 2019, p. 108; Mortimer, 1901, p. 179). We
-considered it possible that it would be present in a sample with most
-individuals originating from the early- to mid-19th century. If
-corroborated, this would have been the first case of
-coca-leaf-consumption in Europe. In our replication batch, we included
-all of the individuals who had been cocaine-positive in the first batch.
-We were unable to replicate any of the cocaine results, and we were
-unable to detect the principal metabolite, benzoylecgonine, in either
-batch. We suspect that the original detection of cocaine was a result of
-lab contamination during analysis.
+cocaine extraction (Abduca, 2019, p. 108; Mortimer, 1901, p. 179). This
+would have been the first case of coca-leaf-consumption in Europe;
+however, we were unable to replicate any of the cocaine results in the
+second batch. We suspect that the original detection of cocaine was a
+result of lab contamination during analysis.
 
@@
 in diseases of the upper respiratory tract, including chronic maxillary
-sinusities (Reh et al., 2012). Although the mechanisms by which smoking
+sinusitis (Reh et al., 2012). Although the mechanisms by which smoking
 increases the risk of infections is not fully understood, solid evidence
@@
 
-Because we targeted individuals with moderate-to-large calculus
-deposits, it is likely a biased sample. The presence of calculus may
-increase the risk of premature death (Yaussy \& DeWitte, 2019), and
-periodontal disease (which may or may not be associated with dental
-calculus build-up) is a risk-factor for respiratory diseases, if
-periodontal and respiratory pathogens enter the bloodstream (Azarpazhooh
-\& Leake, 2006; Scannapieco, 1999; Scannapieco \& Ho, 2001). In our
-sample, the percentage of chronic maxillary sinusitis (37.0\%) is lower
-than in another (more representative) male sample (44.1\%) (Casna et
-al., 2021), and the caries percentage is similarly lower in our sample
-(17.6\%) than a more representative sample (22.9\%) (Lemmers et al.,
-2013).\\
+Targeting individuals with moderate-to-large calculus deposits likely
+biased our sample, as the presence of calculus may increase the risk of
+premature death (Yaussy \& DeWitte, 2019). Additionally, periodontal
+disease (often linked to the presence of calculus) is a risk-factor for
+respiratory diseases, if periodontal and respiratory pathogens enter the
+bloodstream (Azarpazhooh \& Leake, 2006; Scannapieco, 1999; Scannapieco
+\& Ho, 2001). In our sample, the percentage of chronic maxillary
+sinusitis (37.0\%) is lower than in another (more representative) male
+sample (44.1\%) (Casna et al., 2021), and the caries percentage is
+similarly lower in our sample (17.6\%) than a more representative sample
+(22.9\%) (Lemmers et al., 2013).\\
 We used the presence/absence of a pipe notch and concurrent detection of
@@
 not someone consumed tobacco. Dental calculus is also more transient
-than for example bone, as it can be mechanically removed, intentionally
-or unintentionally, during life, eliminating all trace of the alkaloids
+than for example bone, as it can become dislodged during life,
+intentionally or unintentionally, eliminating all trace of the alkaloids
 consumed prior to its removal.\\
-Quantitation of the detected compounds may have limited value in
-archaeological samples due to degradation, and will greatly affect our
-correlations related to concentration. Following burial, compound
-stability over time will play a large role, as will microbial
-degradation of compounds by bacteria and fungi in soil (Liu et al.,
-2015), as well as the soil environment, such as temperature, pH, and
-oxygen availability (Lindholst, 2010; Mackie et al., 2017).\\
-The detected quantity of a compound will also depend on the quantity in
-dental calculus during life, which is largely controlled by quantity of
-consumption, how often the calculus was disrupted/removed, metabolic
-breakdown of the compound, and inter- and intra-individual factors
-related to stages of biofilm formation, maturation, and mineralisation
-(Lustmann et al., 1976; Velsko et al., 2019; Zijnge et al., 2010). In
-short, this means it is not really possible to detect the absence of a
-compound. The absence of a compound is not evidence of absence of
-consumption. This complicates the interpretation of our results. We have
-attempted to minimise errors occurring due to this limitation by
-including a relatively large sample of individuals and replicating our
-analysis. Although given the relatively low detection rate seen in
-tobacco, this remains a major limitation, and will likely be compounded
-by increasing antiquity of the samples.
+Following burial, compound stability over time will play a large role,
+as will microbial degradation of compounds by bacteria and fungi in soil
+(Liu et al., 2015), as well as the soil environment, such as
+temperature, pH, and oxygen availability (Lindholst, 2010; Mackie et
+al., 2017).\\
+Due to this, quantitation of the detected compounds may have limited
+value in archaeological samples due to degradation, and will greatly
+affect our correlations related to concentration. The detected quantity
+of a compound will also depend on the quantity in dental calculus during
+life, which is largely controlled by the quantity consumed, how often
+the calculus was disrupted/removed, metabolic breakdown of the compound,
+and inter- and intra-individual factors related to stages of biofilm
+maturation (Lustmann et al., 1976; Velsko et al., 2019; Zijnge et al.,
+2010). In short, this means it is not really possible to equate the
+absence of a compound as evidence for the absence of consumption, which
+complicates the interpretation of our results. We have attempted to
+minimise errors occurring due to this limitation by including a
+relatively large sample of individuals and replicating our analysis.
+Although, given the relatively low detection rate seen in tobacco, this
+remains a major limitation, and will likely be compounded by increasing
+antiquity of the samples.
 
@@
 certain alkaloids in oral fluid and, subsequently, dental calculus.
-Archaeologists, though, will likely be responsible for exploring dental
-calculus specific incorporation and retention of alkaloids, as well as
-their long-term preservation in the burial environment.
+Archaeologists, though, will likely be responsible for exploring
+dental-calculus-specific incorporation and retention of alkaloids, as
+well as their long-term preservation in the burial environment. Finally,
+following our experience with salicylic acid, we encourage all future
+studies to ensure that a control sample is taken from the soil, either
+from the soil surrounding the individual, or, ideally, directly from the
+skeletal remains. This should preferably happen before cleaning, but
+there will often be soil left over in cavities (e.g.~nasal cavity,
+orbit, auditory meatus).
 
@@
 their input on early draft manuscripts and a lovely guided tour of
-Middenbeemster.
+Middenbeemster. We also thank Louise Le Meillour for taking on our
+preprint and providing helpful comments, and Mario Zimmerman and two
+other anonymous reviewers for their time and invaluable insights.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-huangDecipheringGenetic2023}{}}%
+Huang, Y., Shan, Y., Zhang, W., Lee, A. M., Li, F., Stranger, B. E., \&
+Huang, R. S. (2023). Deciphering genetic causes for sex differences in
+human health through drug metabolism and transporter genes. \emph{Nature
+Communications}, \emph{14}(1, 1), 175.
+\url{https://doi.org/10.1038/s41467-023-35808-6}
+
 \leavevmode\vadjust pre{\hypertarget{ref-jinSupragingivalCalculus2002}{}}%
@@
 \url{https://doi.org/10.1177/154411130201300506}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kingCautionaryTales2017}{}}%
+King, A., Powis, T. G., Cheong, K. F., \& Gaikwad, N. W. (2017).
+Cautionary tales on the identification of caffeinated beverages in
+{North America}. \emph{Journal of Archaeological Science}, \emph{85},
+30--40. \url{https://doi.org/10.1016/j.jas.2017.06.006}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-unoSexAgedependent2017}{}}%
+Uno, Y., Takata, R., Kito, G., Yamazaki, H., Nakagawa, K., Nakamura, Y.,
+Kamataki, T., \& Katagiri, T. (2017). Sex- and age-dependent gene
+expression in human liver: {An} implication for drug-metabolizing
+enzymes. \emph{Drug Metabolism and Pharmacokinetics}, \emph{32}(1),
+100--107. \url{https://doi.org/10.1016/j.dmpk.2016.10.409}
+
 \leavevmode\vadjust pre{\hypertarget{ref-valenDetermination212017}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tidyverse2019}{}}%
-Wickham, Hadley, Averick, M., Bryan, J., Chang, W., McGowan, L. D.,
+Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D.,
 François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M.,

### 2023-10-27 analysis/supplementary-materials/supp-mat.qmd
round 1 of modifications based on reviewer comments

@@
 The samples in the replication batch were processed in the same way, but
-analysed on equipment used exclusively for oral samples.
+analysed on different equipment used exclusively for oral samples.
+
+Raw quantities of compounds detected in the dissolved calculus from batches
+1 and 2 are presented in @tbl-uhplc-batch-1 and @tbl-uhplc-batch-2. Since
+these tables may or may not be legible in PDF format, not to mention that they
+don't adhere to FAIR principles in this format, the raw data can be downloaded
+from Zenodo (https://doi.org/10.5281/zenodo.8061483).
 
@@
 #| label: tbl-uhplc-batch-1
-#| tbl-cap: "Results from the UHPLC analysis first batch. Quantity of compound in the dissolved calculus, represented in ng."
+#| tbl-cap: "Results from the UHPLC analysis first batch. Quantity of compound in the dissolved calculus, represented in ng and rounded to 3 digits after the decimal."
 uhplc_calculus_long %>%
   filter(batch == "batch1") %>%
+  mutate(quant = round(quant, 3)) %>% 
   select(!c(sample, conc, extraction, batch, presence)) %>%
@@
 #| label: tbl-uhplc-batch-2
-#| tbl-cap: "Results from the UHPLC analysis second batch. Quantity of compound in calculus after third wash, represented in ng."
+#| tbl-cap: "Results from the UHPLC analysis second batch. Quantity of compound in calculus after third wash, represented in ng and rounded to 3 digits after the decimal."
 uhplc_calculus_long %>%
   filter(batch == "batch2") %>%
+  mutate(quant = round(quant, 3)) %>% 
   pivot_wider(names_from = "compound", values_from = "quant") %>%

### 2023-11-03 analysis/paper/paper.tex
feat: formatting for PCI_Archaeo

@@
 \usepackage{xcolor}
-\usepackage[top=30mm,left=30mm,heightrounded]{geometry}
+\usepackage[top=10mm,bottom=10mm,left=30mm,heightrounded]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 \usepackage{xcolor}
-\usepackage{lineno}
+\usepackage{fancyhdr, graphicx}
+\usepackage[hidelinks]{hyperref}
+
+\fancyhead{\vspace{70pt}}
+\renewcommand{\headrulewidth}{1.8pt}
+
+\makeatletter
+\def\@maketitle{%
+  \newpage
+  \hfill\href{https://doi.org/10.24072/pci.archaeo.100389}{\includegraphics[width=3.5cm]{badge_PCI_Archaeology.png}}\hfill
+  \null
+ \begin{center}
+{\LARGE
+\textbf\newline{\@title}
+}
+%\newline
+
+\end{center}
+  \par}
+\makeatother
 \makeatletter
@@
 %% modified from: https://github.com/quarto-journals/plos/blob/main/_extensions/plos/partials/before-body.tex
-\begin{flushleft}
-{\LARGE
-\textbf\newline{Multiproxy analysis exploring patterns of diet and
-disease in dental calculus and skeletal remains from a 19th century
-Dutch population}
-}
-\newline
-
-\vspace{12pt}
+
+\maketitle
+
+\begin{center}
+
+\thispagestyle{fancy}
+
 % Insert author names, affiliations and corresponding author email
@@
 Beemster}\textsuperscript{3}, \normalsize{Amanda G.
-Henry}\textsuperscript{1}
-\\
-\bigskip
-\small{1} \small{Department of Archaeological Sciences, Leiden
-University}\\ \small{2} \small{Department of Forensic Medicine, Aarhus
-University}\\ \small{3} \small{Historisch Genootschap Beemster}
-\newline
+Henry}\textsuperscript{1}\vspace{12pt}
+
+\rule{\textwidth}{1.8pt}\vspace{8pt}
+
+\end{center}
+
+\textsuperscript{\footnotesize{1}}\textit{\small{Department of
+Archaeological Sciences, Leiden
+University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Department
+of Forensic Medicine, Aarhus
+University}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Historisch
+Genootschap Beemster}}
 
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* b.p.bartholdy@arch.leidenuniv.nl
-
-\end{flushleft}
-
-\linenumbers
+* \small{b.p.bartholdy@arch.leidenuniv.nl}
+
+\vspace{12pt}
 
@@
 
-Keywords:  dental calculus; LC-MS/MS; alkaloids; dental pathology;
-sinusitis; caffeine; tobacco 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\small \textbf{Keywords:}  dental calculus; LC-MS/MS; alkaloids; dental
+pathology; sinusitis; caffeine; tobacco   \\
+
+\vfill
+
+\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
 \hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}}
 
@@
 \hypertarget{materials}{%
-\subsection{Materials}\label{materials}}
+\section{Materials}\label{materials}}
 
@@
 \hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+\section{Methods}\label{methods}}
 
 \hypertarget{skeletal-analysis}{%
-\subsubsection{Skeletal analysis}\label{skeletal-analysis}}
+\subsection{Skeletal analysis}\label{skeletal-analysis}}
 
@@
 \hypertarget{paleopathology}{%
-\paragraph{Paleopathology}\label{paleopathology}}
+\subsubsection{Paleopathology}\label{paleopathology}}
 
@@
 \hypertarget{dental-pathology}{%
-\paragraph{Dental pathology}\label{dental-pathology}}
+\subsubsection{Dental pathology}\label{dental-pathology}}
 
@@
 \hypertarget{calculus-sampling}{%
-\subsubsection{Calculus sampling}\label{calculus-sampling}}
+\subsection{Calculus sampling}\label{calculus-sampling}}
 
@@
 \hypertarget{uhplc-msms}{%
-\subsubsection{UHPLC-MS/MS}\label{uhplc-msms}}
+\subsection{UHPLC-MS/MS}\label{uhplc-msms}}
 
@@
 \hypertarget{statistical-analysis}{%
-\subsubsection{Statistical analysis}\label{statistical-analysis}}
+\subsection{Statistical analysis}\label{statistical-analysis}}
 
@@
 \hypertarget{results}{%
-\subsection{Results}\label{results}}
+\section{Results}\label{results}}
 
@@
 \hypertarget{correlations-between-detected-alkaloids-and-diseases}{%
-\subsubsection{Correlations between detected alkaloids and
+\subsection{Correlations between detected alkaloids and
 diseases}\label{correlations-between-detected-alkaloids-and-diseases}}
@@
 \hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}}
 
@@
 \hypertarget{conclusions}{%
-\subsection{Conclusions}\label{conclusions}}
+\section{Conclusions}\label{conclusions}}
 
@@
 \hypertarget{acknowledgements}{%
-\subsection*{Acknowledgements}\label{acknowledgements}}
-\addcontentsline{toc}{subsection}{Acknowledgements}
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
 
@@
 
+\hypertarget{funding}{%
+\section*{Funding}\label{funding}}
+\addcontentsline{toc}{section}{Funding}
+
 This research has received funding from the European Research Council
@@
 
+\hypertarget{conflict-of-interest-disclosure}{%
+\section*{Conflict of interest
+disclosure}\label{conflict-of-interest-disclosure}}
+\addcontentsline{toc}{section}{Conflict of interest disclosure}
+
+The authors have no conflicts to declare.
+
 \hypertarget{data-availability-statement}{%
-\subsection*{Data Availability
+\section*{Data Availability
 Statement}\label{data-availability-statement}}
-\addcontentsline{toc}{subsection}{Data Availability Statement}
+\addcontentsline{toc}{section}{Data Availability Statement}
 
@@
 
-\hypertarget{conflict-of-interest-disclosure}{%
-\subsection*{Conflict of interest
-disclosure}\label{conflict-of-interest-disclosure}}
-\addcontentsline{toc}{subsection}{Conflict of interest disclosure}
-
-The authors have no conflicts to declare.
-
 \hypertarget{references}{%
-\subsection*{References}\label{references}}
-\addcontentsline{toc}{subsection}{References}
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 

### 2023-11-03 analysis/paper/paper.tex
tweak figure

@@
 
-\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 

### 2023-11-03 analysis/paper/paper.qmd
feat: formatting for PCI_Archaeo
bump version to 0.1.0

@@
 
+## Funding {-}
+
 This research has received funding from the European Research Council under the
@@
 number STG–677576 (“HARVEST”).
+
+## Conflict of interest disclosure {-}
+
+The authors have no conflicts to declare.
 
@@
 supplementary materials are available as a research compendium
-(<https://doi.org/10.5281/zenodo.7649825>)
+(<https://doi.org/10.5281/zenodo.7649824>)
 using the structure recommended by the **rrtools** R package [@Rrrtools].
-
-## Conflict of interest disclosure {-}
-
-The authors have no conflicts to declare.
 

### 2023-11-10 analysis/paper/paper.tex
update NEWS.md

@@
 
-\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 available as a research compendium
-(\url{https://doi.org/10.5281/zenodo.7649825}) using the structure
+(\url{https://doi.org/10.5281/zenodo.7649824}) using the structure
 recommended by the \textbf{rrtools} R package (Marwick, 2019).

### 2023-11-20 analysis/paper/_materials.qmd
submitted to PLOS

@@
 [@lemmersMiddenbeemster2013]. Archival documents are available for those buried
-between AD 1829 and 1866, when the majority of individuals were interred
-[@palmerActivityReconstruction2016]. The main occupation of the inhabitants
+between AD 1829 and 1866, when the majority of individuals were interred.
+The main occupation of the inhabitants
 was dairy farming, consisting largely of manual labour prior to the industrial
@@
 
-To reduce the number of potentially confounding factors to account for
-in the analysis, we preferentially selected males from the middle adult age category
-(35-49 years). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
+For our sample, we preferentially selected males from the middle adult age category
+(35-49 years) to minimise the effect of confounding cultural and biological factors.
+Previous research on Middenbeemster has shown a gendered division of labour
+[@palmerActivityReconstruction2016], and there are biological differences in
+dental calculus formation and drug metabolism that are related to age and sex
+[@huangDecipheringGenetic2023; @unoSexAgedependent2017; @whiteDentalCalculus1997].
+The sample consists of `r nrow(filter(demography, sex == "m"))` males,
 `r nrow(filter(demography, sex == "pm"))` probable males,
@@
 occurrence of pipe notches and dental calculus deposits than females (unpublished observation).
+
+*Fig 1 (fig-sample-demography): Overview of sample demography.
+Left plot is the first batch and right plot is the replication batch with 29 of the individuals from the first batch.
+eya = early young adult (18-24 years); lya = late young adult (25-34 years); ma = middle adult (35-49 years); old = old adult (50+ years).
+Male?
+= probable male; Female?
+= probable female.*
 

### 2023-11-20 analysis/paper/_results.qmd
tweak figure
submitted to PLOS

@@
 
+*Fig 2 (fig-auth-plot-batch2): (A) Number of samples in which each compound was detected in the first and second batch.
+(B) Quantity (ng) of each compound extracted from each sample in batch 2.
+The plot displays the extracted quantity across the three washes and final calculus extraction (calc).
+Each coloured line represents a different calculus sample.
+CBD = cannabidiol; CBN = cannabinol; THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA = tetrahydrocannabivarin acid.*
+
 ```{r}
@@
 CBD = cannabidiol; CBN = cannabinol; THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA = tetrahydrocannabivarin acid."
-#| fig-width: 8
-#| fig-height: 6
+#| fig-width: 7
+#| fig-asp: 0.75
 
@@
 
+*Fig 3 (fig-detection-preservation): (A) Violin plot with overlaid box plots depicting the distribution of extracted quantities of each compound from batch 2 separated by state of preservation of the skeleton.
+(B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2.
+r = Pearson correlation coefficient.*
+
 ```{r}
@@
 r = Pearson correlation coefficient."
-#| fig-width: 4.2
+#| fig-width: 4.8
 #| fig-asp: 1.6
@@
 
+*Fig 4 (fig-polycorr): Plot of the polychoric correlations (*rho*).
+Larger circles and increased opacity indicates a stronger correlation coefficient.
+OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl’s nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid.*
+
 ```{r}
@@
 #| fig-width: 5
-#| fig-height: 4
+#| fig-asp: 0.8
 polycorr$rho %>%

### 2023-11-20 analysis/paper/paper.qmd
submitted to PLOS

@@
 options(ggplot.discrete.colour = function() scale_fill_viridis_d())
+knitr::opts_chunk$set(
+  fig.show = "hide",
+  fig.path = "../figures/plos-",
+  dev = "tiff",
+  dev.args = list("tiff" = list(compression = "lzw"))
+  #fig.keep = "none"
+)
 ```

### 2023-11-20 analysis/paper/paper.tex
submitted to PLOS

@@
 \usepackage{xcolor}
-\usepackage[top=10mm,bottom=10mm,left=30mm,heightrounded]{geometry}
+\usepackage[top=30mm,bottom=30mm,left=30mm,heightrounded]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \usepackage{xcolor}
-\usepackage{fancyhdr, graphicx}
-\usepackage[hidelinks]{hyperref}
-
-\fancyhead{\vspace{70pt}}
-\renewcommand{\headrulewidth}{1.8pt}
-
-\makeatletter
-\def\@maketitle{%
-  \newpage
-  \hfill\href{https://doi.org/10.24072/pci.archaeo.100389}{\includegraphics[width=3.5cm]{badge_PCI_Archaeology.png}}\hfill
-  \null
- \begin{center}
-{\LARGE
-\textbf\newline{\@title}
-}
-%\newline
-
-\end{center}
-  \par}
-\makeatother
+\usepackage{lineno}
 \makeatletter
@@
 \ifdefined\figurename
-  \renewcommand*\figurename{Figure}
+  \renewcommand*\figurename{Fig}
 \else
-  \newcommand\figurename{Figure}
+  \newcommand\figurename{Fig}
 \fi
@@
 
-\maketitle
+%\maketitle
 
@@
 
-\thispagestyle{fancy}
+%\thispagestyle{fancy}
+
+\LARGE{Multiproxy analysis exploring patterns of diet and disease in
+dental calculus and skeletal remains from a 19th century Dutch
+population}
 
@@
 \normalsize{Bjørn Peare
-Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
-Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
-Sørensen}\textsuperscript{2}, \normalsize{Maia
+Bartholdy}\textsuperscript{1,2*}, \normalsize{Jørgen B.
+Hasselstrøm}\textsuperscript{3}, \normalsize{Lambert K.
+Sørensen}\textsuperscript{3}, \normalsize{Maia
 Casna}\textsuperscript{1}, \normalsize{Menno
 Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
-Beemster}\textsuperscript{3}, \normalsize{Amanda G.
+Beemster}\textsuperscript{4}, \normalsize{Amanda G.
 Henry}\textsuperscript{1}\vspace{12pt}
 
-\rule{\textwidth}{1.8pt}\vspace{8pt}
+%\rule{\textwidth}{1.8pt}\vspace{8pt}
 
@@
 Archaeological Sciences, Leiden
-University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Department
+University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Faculty
+of Mechanical, Maritime and Materials Engineering, Delft University of
+Technology}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Department
 of Forensic Medicine, Aarhus
-University}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Historisch
+University}}\\ \textsuperscript{\footnotesize{4}}\textit{\small{Historisch
 Genootschap Beemster}}
@@
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* \small{b.p.bartholdy@arch.leidenuniv.nl}
+* \small{b.p.bartholdy@tudelft.nl}
 
 \vspace{12pt}
+
+\linenumbers
 
@@
 archaeological site. Skeletal and dental analysis was performed to
-explore potential relationships between pathological conditions/lesions
-and the presence of alkaloids. We also explored other factors
-potentially affecting the detection of alkaloids, including sample
-weight and skeletal preservation. Dental calculus was sampled and
-analysed using ultra-high-performance liquid chromatography-tandem mass
-spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
-cotinine, caffeine, theophylline, and salicylic acid. By detecting these
-compounds we are able to show the consumption of tea and coffee and
-smoking of tobacco on an individual scale, which is also confirmed by
-historic documentation and identification of pipe notches in the
-dentition. Nicotine and/or cotinine was present in 56\% of individuals
-with at least one visible pipe notch. We find some influence of skeletal
-preservation on the detection of alkaloids and salicylic acid, with
+explore potential relationships between pathological lesions and
+presence of alkaloids. Dental calculus was analysed using
+ultra-high-performance liquid chromatography-tandem mass spectrometry
+(UHPLC-ESI-MS/MS). We were able to detect nicotine, cotinine, caffeine,
+theophylline, and salicylic acid, suggesting the consumption of tea and
+coffee and smoking of tobacco on an individual scale, which is also
+confirmed by historic documentation and identification of pipe notches
+in the dentition. Nicotine and/or cotinine was present in 56\% of
+individuals with at least one visible pipe notch. There is some
+influence of skeletal preservation on the detection of alkaloids, with
 higher quantities of compounds extracted from well-preserved
-individuals, and also observe a relationship between weight of the
-calculus sample and raw quantity of the detected compounds, and we were
-able to detect alkaloids in samples as small as 2 mg. We found
-correlations between chronic maxillary sinusitis and the presence of
-multiple alkaloids. We show that there are many limitations that will
-need to be addressed going forward with this type of analysis, and
-stress the need for more systematic research on the consumption of
-alkaloid-containing items and their subsequent concentration and
-preservation in dental calculus, in addition to how mode of consumption
-may affect concentrations on different parts of the dentition. Despite
-the limitations, this preliminary study illustrates the many benefits of
-using calculus to target a variety of compounds that could have been
-ingested as medicine or diet, or consumed in a different manner. This
-method allows us to directly address specific individuals, which can be
-especially useful in individuals that are not always well-documented in
-historic documentation, such as rural populations, and especially
-children and women.
+individuals, and we observe a positive relationship between weight of
+the calculus sample and quantity of detected compounds, as well as
+between chronic maxillary sinusitis and the presence of multiple
+alkaloids. There are many limitations that will need to be addressed
+going forward with this type of analysis; we stress the need for more
+systematic research on the consumption of alkaloid-containing items and
+their subsequent concentration and preservation in dental calculus, in
+addition to how mode of consumption may affect concentrations in the
+dentition. Despite the limitations, this preliminary study illustrates
+many benefits of using calculus to target a variety of compounds that
+could have been consumed as medicine or diet. This method allows us to
+directly address specific individuals, which can be especially useful in
+individuals that are not always well-documented in historic
+documentation, such as rural populations, and especially children and
+women.
 \end{abstract}
@@
 
-\vfill
-
-\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
+%\vfill
+
+%\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
@@
 White, 1997). The sample consists of 27 males, 11 probable males, 2
-probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
+probable females, and 1 female (\textbf{?@fig-sample-demography}). We
 selected males due to a higher occurrence of pipe notches and dental
@@
 
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-pdf/fig-sample-demography-1.pdf}
-
-}
-
-\caption{\label{fig-sample-demography}Overview of sample demography.
-Left plot is the first batch and right plot is the replication batch
-with 29 of the individuals from the first batch. eya = early young adult
-(18-24 years); lya = late young adult (25-34 years); ma = middle adult
-(35-49 years); old = old adult (50+ years). Male? = probable male;
-Female? = probable female.}
-
-\end{figure}
+\emph{Fig 1 (fig-sample-demography): Overview of sample demography. Left
+plot is the first batch and right plot is the replication batch with 29
+of the individuals from the first batch. eya = early young adult (18-24
+years); lya = late young adult (25-34 years); ma = middle adult (35-49
+years); old = old adult (50+ years). Male? = probable male; Female? =
+probable female.}
 
@@
 
-All statistical analysis was conducted in R version 4.3.1 (2023-06-16),
-Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
-the \textbf{tidyverse} (Wickham et al., 2019) and visualisations were
+All statistical analysis was conducted in R version 4.3.2 (2023-10-31),
+Eye Holes, (R Core Team, 2020). Data wrangling was conducted with the
+\textbf{tidyverse} (Wickham et al., 2019) and visualisations were
 created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
@@
 nicotine also had the highest extracted quantity of cotinine
-(Figure~\ref{fig-auth-plot-batch2}).
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-pdf/fig-auth-plot-batch2-1.pdf}
-
-}
-
-\caption{\label{fig-auth-plot-batch2}(A) Number of samples in which each
+(\textbf{?@fig-auth-plot-batch2}).
+
+\emph{Fig 2 (fig-auth-plot-batch2): (A) Number of samples in which each
 compound was detected in the first and second batch. (B) Quantity (ng)
@@
 
-\end{figure}
-
 To see if preservation of the skeletal remains had any effect on the
@@
 of the skeleton, with better preservation leading to increased
-extraction quantity (Figure~\ref{fig-detection-preservation}A). We also
+extraction quantity (\textbf{?@fig-detection-preservation}A). We also
 find a weak positive correlation between the weight of the calculus
 sample and the quantity of compound extracted from the calculus
-(Figure~\ref{fig-detection-preservation}B).
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-pdf/fig-detection-preservation-1.pdf}
-
-}
-
-\caption{\label{fig-detection-preservation}(A) Violin plot with overlaid
+(\textbf{?@fig-detection-preservation}B).
+
+\emph{Fig 3 (fig-detection-preservation): (A) Violin plot with overlaid
 box plots depicting the distribution of extracted quantities of each
@@
 coefficient.}
-
-\end{figure}
 
@@
 CMS (0.43). Remaining correlations were weak or absent
-(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
+(\textbf{?@fig-polycorr}). Correlations with age will be depressed
 because age was largely controlled for in the sample selection.
 
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-pdf/fig-polycorr-1.pdf}
-
-}
-
-\caption{\label{fig-polycorr}Plot of the polychoric correlations
-(\emph{rho}). Larger circles and increased opacity indicates a stronger
+\emph{Fig 4 (fig-polycorr): Plot of the polychoric correlations
+(}rho\emph{). Larger circles and increased opacity indicates a stronger
 correlation coefficient. OA = osteoarthritis; VOP = vertebral
@@
 acid.}
-
-\end{figure}
 
@@
 positively correlated, both in concentration and presence/absence in
-individuals (Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). The
+individuals (Table~\ref{tbl-pearson} and \textbf{?@fig-polycorr}). The
 detection of nicotine and cotinine is not surprising, as pipe-smoking in
@@
 to determine if they originated from the same source
-(Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). Caffeine and
+(Table~\ref{tbl-pearson} and \textbf{?@fig-polycorr}). Caffeine and
 theophylline have very similar chemical structures, so we expect they
@@
 to see if the salicylic acid was incorporated during life
-(Figure~\ref{fig-auth-plot-batch2}). It is important to note that,
+(\textbf{?@fig-auth-plot-batch2}). It is important to note that,
 especially with salicylic acid, there is a possibility for the compound

### 2024-04-16 analysis/paper/paper.tex
add tex file for submission

--- 2023-11-10 analysis/paper/paper.tex
@@
 \documentclass[
-]{article}
+  11.5pt,
+  leqno]{scrartcl}
 
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \usepackage{xcolor}
-\usepackage[top=10mm,bottom=10mm,left=30mm,heightrounded]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 
@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
+\usepackage[sfdefault]{roboto}
+\usepackage{unicode-math}
+\usepackage[a4paper,headheight=13pt,footskip=20pt,margin=2.54cm]{geometry}
+\usepackage{indentfirst}
+
+% Modify section headings
+\usepackage{titlesec}
+\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{}
+\titleformat{\subsection}{\normalfont\bfseries}{}{.5em}{}
+\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{\addperiod}
+\newcommand{\addperiod}[1]{#1.} %https://tex.stackexchange.com/questions/137193/adding-a-period-after-section
+
+\setlength{\parindent}{1.5em}
+\setlength{\parskip}{0pt plus 1pt}
+\linespread{1.1}
+
+% Adjust caption style
+\usepackage[labelfont=bf,
+    labelsep=endash,
+    singlelinecheck=off,
+    format=plain,
+    margin=1.5cm,
+    aboveskip=12pt,
+    belowskip=12pt,
+    font=footnotesize,
+    justification=justified]{caption}
+
+% Defining document colors
+\usepackage{xcolor}
+\definecolor{darkgray}{HTML}{808080}
+\definecolor{mediumgray}{HTML}{6D6E70}
+\definecolor{ligthgray}{HTML}{d9d9d9}
+\definecolor{pciblue}{HTML}{346E92}
+\definecolor{opengreen}{HTML}{77933c}
+
+%\usepackage[colorlinks=true,linkcolor=pciblue,urlcolor=pciblue,citecolor=pciblue]{hyperref} % Not working - links in pdf not changing colour. 
+                                                         %Probably because hyperref is no the last package to be loaded, which it needs to be.
+%\urlstyle{same}
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
-\usepackage{fancyhdr, graphicx}
-\usepackage[hidelinks]{hyperref}
-
-\fancyhead{\vspace{70pt}}
-\renewcommand{\headrulewidth}{1.8pt}
-
-\makeatletter
-\def\@maketitle{%
-  \newpage
-  \hfill\href{https://doi.org/10.24072/pci.archaeo.100389}{\includegraphics[width=3.5cm]{badge_PCI_Archaeology.png}}\hfill
-  \null
- \begin{center}
-{\LARGE
-\textbf\newline{\@title}
-}
-%\newline
-
-\end{center}
-  \par}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
   colorlinks=true,
-  linkcolor={blue},
+  linkcolor={pciblue},
   filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
+  citecolor={pciblue},
+  urlcolor={pciblue},
   pdfcreator={LaTeX via pandoc}}
 
-\title{Multiproxy analysis exploring patterns of diet and disease in
-dental calculus and skeletal remains from a 19th century Dutch
-population}
-\author{Bjørn Peare Bartholdy \and Jørgen B. Hasselstrøm \and Lambert K.
-Sørensen \and Maia Casna \and Menno Hoogland \and Historisch Genootschap
-Beemster \and Amanda G. Henry}
+\title{}
+\author{}
 \date{}
@@
 \begin{document}
-%% modified from: https://github.com/quarto-journals/plos/blob/main/_extensions/plos/partials/before-body.tex
-
-\maketitle
-
-\begin{center}
-
-\thispagestyle{fancy}
-
-% Insert author names, affiliations and corresponding author email
-
-\normalsize{Bjørn Peare
-Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
-Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
-Sørensen}\textsuperscript{2}, \normalsize{Maia
-Casna}\textsuperscript{1}, \normalsize{Menno
-Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
-Beemster}\textsuperscript{3}, \normalsize{Amanda G.
-Henry}\textsuperscript{1}\vspace{12pt}
-
-\rule{\textwidth}{1.8pt}\vspace{8pt}
-
-\end{center}
-
-\textsuperscript{\footnotesize{1}}\textit{\small{Department of
-Archaeological Sciences, Leiden
-University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Department
-of Forensic Medicine, Aarhus
-University}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Historisch
-Genootschap Beemster}}
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* \small{b.p.bartholdy@arch.leidenuniv.nl}
-
-\vspace{12pt}
-
-\begin{abstract}
-Dental calculus is an excellent source of information on the dietary
-patterns of past populations, including consumption of plant-based
-items. The detection of plant-derived residues such as alkaloids and
-their metabolites in dental calculus provides direct evidence of
-consumption by individuals within a population. We conducted a study on
-41 individuals from Middenbeemster, a 19th century rural Dutch
-archaeological site. Skeletal and dental analysis was performed to
-explore potential relationships between pathological conditions/lesions
-and the presence of alkaloids. We also explored other factors
-potentially affecting the detection of alkaloids, including sample
-weight and skeletal preservation. Dental calculus was sampled and
-analysed using ultra-high-performance liquid chromatography-tandem mass
-spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
-cotinine, caffeine, theophylline, and salicylic acid. By detecting these
-compounds we are able to show the consumption of tea and coffee and
-smoking of tobacco on an individual scale, which is also confirmed by
-historic documentation and identification of pipe notches in the
-dentition. Nicotine and/or cotinine was present in 56\% of individuals
-with at least one visible pipe notch. We find some influence of skeletal
-preservation on the detection of alkaloids and salicylic acid, with
-higher quantities of compounds extracted from well-preserved
-individuals, and also observe a relationship between weight of the
-calculus sample and raw quantity of the detected compounds, and we were
-able to detect alkaloids in samples as small as 2 mg. We found
-correlations between chronic maxillary sinusitis and the presence of
-multiple alkaloids. We show that there are many limitations that will
-need to be addressed going forward with this type of analysis, and
-stress the need for more systematic research on the consumption of
-alkaloid-containing items and their subsequent concentration and
-preservation in dental calculus, in addition to how mode of consumption
-may affect concentrations on different parts of the dentition. Despite
-the limitations, this preliminary study illustrates the many benefits of
-using calculus to target a variety of compounds that could have been
-ingested as medicine or diet, or consumed in a different manner. This
-method allows us to directly address specific individuals, which can be
-especially useful in individuals that are not always well-documented in
-historic documentation, such as rural populations, and especially
-children and women.
-\end{abstract}
-
-\small \textbf{Keywords:}  dental calculus; LC-MS/MS; alkaloids; dental
-pathology; sinusitis; caffeine; tobacco   \\
-
-\vfill
-
-\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+
+\setstretch{1.5}
+\section{Introduction}\label{introduction}
 
@@
 has expanded our ability to identify biomarkers of diet and disease on
-an increasingly large scale (Gismondi et al., 2020; Velsko et al., 2017;
-Warinner et al., 2014).
+an increasingly large scale
+(\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
+2020}; \citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017};
+\citeproc{ref-warinnerEvidenceMilk2014}{Warinner et al., 2014}).
 
@@
 have successfully recovered alkaloids in archaeological contexts,
-including ceramics (Smith et al., 2018), pipes (Rafferty et al., 2012),
-human hair (Echeverría \& Niemeyer, 2013; Ogalde et al., 2009), and even
-dental calculus employing both targeted (Eerkens et al., 2018) and
-untargeted approaches (Buckley et al., 2014; Gismondi et al., 2020).
-Especially nicotine, the principal alkaloid in tobacco leaves, has been
-widely studied in the archaeological record due to its apparent
-stability and ability to survive over long periods of time (Eerkens et
-al., 2018; Rafferty et al., 2012; Tushingham et al., 2013).
+including ceramics (\citeproc{ref-smithDetectionOpium2018}{Smith et al.,
+2018}), pipes (\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et
+al., 2012}), human hair
+(\citeproc{ref-echeverriaNicotineHair2013}{Echeverría \& Niemeyer,
+2013}; \citeproc{ref-ogaldeIdentificationPsychoactive2009}{Ogalde et
+al., 2009}), and even dental calculus employing both targeted
+(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}) and
+untargeted approaches (\citeproc{ref-buckleyDentalCalculus2014}{Buckley
+et al., 2014};
+\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
+2020}). Especially nicotine, the principal alkaloid in tobacco leaves,
+has been widely studied in the archaeological record due to its apparent
+stability and ability to survive over long periods of time
+(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018};
+\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et al., 2012};
+\citeproc{ref-tushinghamHuntergathererTobacco2013}{Tushingham et al.,
+2013}).
 
@@
 gradually secreted into the oral cavity through the salivary glands in
-the hours-to-days following ingestion (Cone \& Huestis, 2007). The
+the hours-to-days following ingestion
+(\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007}). The
 second pathway allows the identification of parent compounds that do not
@@
 compounds are transferred from saliva to biofilms on the tooth surface
-(Jin \& Yip, 2002; White, 1997). This may also allow various alkaloids
-of dietary and medicinal origin to become incorporated in dental plaque.
-Dental plaque undergoes frequent mineralisation events, ultimately
-causing the entrapped alkaloids and their metabolites to become
-preserved within the dental calculus. Barring intentional or accidental
-removal of the calculus during life, burial, excavation, and
-post-excavation cleaning, the alkaloids can then be detected by various
-methods to show a record of consumption during life. Because drugs may
-be transferred from plasma to saliva, there is often a close correlation
-between drugs detected in oral fluid and blood, though there are
-differences in detected concentrations (Cone \& Huestis, 2007; Milman et
-al., 2011; Wille et al., 2009). This was also shown to be true for
-dental calculus and blood (Sørensen et al., 2021), making dental
-calculus a potentially useful substance for detecting ancient alkaloids
-and other dietary compounds.
+(\citeproc{ref-jinSupragingivalCalculus2002}{Jin \& Yip, 2002};
+\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). This may also
+allow various alkaloids of dietary and medicinal origin to become
+incorporated in dental plaque. Dental plaque undergoes frequent
+mineralisation events, ultimately causing the entrapped alkaloids and
+their metabolites to become preserved within the dental calculus.
+Barring intentional or accidental removal of the calculus during life,
+burial, excavation, and post-excavation cleaning, the alkaloids can then
+be detected by various methods to show a record of consumption during
+life. Because drugs may be transferred from plasma to saliva, there is
+often a close correlation between drugs detected in oral fluid and
+blood, though there are differences in detected concentrations
+(\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007};
+\citeproc{ref-milmanOralFluid2011}{Milman et al., 2011};
+\citeproc{ref-willeRelationshipOral2009}{Wille et al., 2009}). This was
+also shown to be true for dental calculus and blood
+(\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}),
+making dental calculus a potentially useful substance for detecting
+ancient alkaloids and other dietary compounds.
 
@@
 validated by comparing the results to compounds detected in the blood of
-the same individuals (Sørensen et al., 2021). All compounds that were
-detected in the blood were also detected in dental calculus, with
-additional compounds present in dental calculus that were not present in
-blood, suggesting that dental calculus represents a comprehensive
-history of consumption over a long period of time (Sørensen et al.,
-2021). We were able to detect both parent compounds and metabolites,
-including caffeine, nicotine, theophylline, and cotinine, in the dental
-calculus of individuals from a 19th century Dutch population from
-Middenbeemster. By detecting these compounds we are able to show the
-consumption of tea and coffee and smoking of tobacco on an individual
-scale, which is also confirmed by historic documentation and
-identification of pipe notches in the dentition.
-
-\hypertarget{materials}{%
-\section{Materials}\label{materials}}
+the same individuals (\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen
+et al., 2021}). All compounds that were detected in the blood were also
+detected in dental calculus, with additional compounds present in dental
+calculus that were not present in blood, suggesting that dental calculus
+represents a comprehensive history of consumption over a long period of
+time (\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}).
+We were able to detect both parent compounds and metabolites, including
+caffeine, nicotine, theophylline, and cotinine, in the dental calculus
+of individuals from a 19th century Dutch population from Middenbeemster.
+By detecting these compounds we are able to show the consumption of tea
+and coffee and smoking of tobacco on an individual scale, which is also
+confirmed by historic documentation and identification of pipe notches
+in the dentition.
+
+\section{Materials}\label{materials}
 
@@
 and the surrounding Beemsterpolder were buried between AD 1615 and 1866
-(Lemmers et al., 2013). Archival documents are available for those
-buried between AD 1829 and 1866, when the majority of individuals were
-interred. The main occupation of the inhabitants was dairy farming,
-consisting largely of manual labour prior to the industrial revolution
-(Aten et al., 2012; Palmer et al., 2016).
+(\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}).
+Archival documents are available for those buried between AD 1829 and
+1866, when the majority of individuals were interred. The main
+occupation of the inhabitants was dairy farming, consisting largely of
+manual labour prior to the industrial revolution
+(\citeproc{ref-aten400Jaar2012}{Aten et al., 2012};
+\citeproc{ref-palmerActivityReconstruction2016}{Palmer et al., 2016}).
 
@@
 cultural and biological factors. Previous research on Middenbeemster has
-shown a gendered division of labour (Palmer et al., 2016), and there are
-biological differences in dental calculus formation and drug metabolism
-that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
-White, 1997). The sample consists of 27 males, 11 probable males, 2
-probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
-selected males due to a higher occurrence of pipe notches and dental
-calculus deposits than females (unpublished observation).
+shown a gendered division of labour
+(\citeproc{ref-palmerActivityReconstruction2016}{Palmer et al., 2016}),
+and there are biological differences in dental calculus formation and
+drug metabolism that are related to age and sex
+(\citeproc{ref-huangDecipheringGenetic2023}{Huang et al., 2023};
+\citeproc{ref-unoSexAgedependent2017}{Uno et al., 2017};
+\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). The sample
+consists of 27 males, 11 probable males, 2 probable females, and 1
+female (Figure~\ref{fig-sample-demography}). We selected males due to a
+higher occurrence of pipe notches and dental calculus deposits than
+females (unpublished observation).
 
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-sample-demography-1.pdf}
+\centering{
+
+\includegraphics{../figures/fig-sample-demography-1.pdf}
 
@@
 
-\end{figure}
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-\hypertarget{skeletal-analysis}{%
-\subsection{Skeletal analysis}\label{skeletal-analysis}}
+\end{figure}%
+
+\section{Methods}\label{methods}
+
+\subsection{Skeletal analysis}\label{skeletal-analysis}
 
@@
 for Human Osteoarchaeology at Leiden University. Sex was estimated using
-cranial and pelvic morphological traits (Buikstra \& Ubelaker, 1994).
-Age-at-death was estimated using dental wear, auricular and pubic
-surface appearance, cranial suture closure, and epiphyseal fusion
-(Brooks \& Suchey, 1990; Buckberry \& Chamberlain, 2002; Buikstra \&
-Ubelaker, 1994; Lovejoy et al., 1985; Meindl \& Lovejoy, 1985), and
+cranial and pelvic morphological traits
+(\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994}). Age-at-death
+was estimated using dental wear, auricular and pubic surface appearance,
+cranial suture closure, and epiphyseal fusion
+(\citeproc{ref-SucheyBrooks1990}{Brooks \& Suchey, 1990};
+\citeproc{ref-buckberryAuricular2002}{Buckberry \& Chamberlain, 2002};
+\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994};
+\citeproc{ref-lovejoyAuricular1985}{Lovejoy et al., 1985};
+\citeproc{ref-meindlSutureClosure1985}{Meindl \& Lovejoy, 1985}), and
 divided into the following categories: early young adult (18-24 years),
@@
 
-\hypertarget{paleopathology}{%
-\subsubsection{Paleopathology}\label{paleopathology}}
+\subsubsection{Paleopathology}\label{paleopathology}
 
@@
 depression of the superior and/or inferior surfaces of the vertebral
-body (Rogers, 2000). Schmorl's nodes are identified as any cortical
-depressions on the surface of the vertebral body. Data on chronic
-maxillary sinusitis from Casna et al. (2021) were included in this study
-to assess the relationship between upper respiratory diseases with
-environmental factors (i.e.~tobacco smoke, caffeine consumption).
-Lesions associated with chronic maxillary sinusitis as defined by
-Boocock et al. (1995) were recorded for each individual and classified
-as ``pitting'', ``spicule-type bone formation'', ``remodeled spicules'',
-or ``white pitted bone''. chronic maxillary sinusitis was scored as
-absent when the sinus presented smooth surfaces with little or no
-associated pitting.
-
-\hypertarget{dental-pathology}{%
-\subsubsection{Dental pathology}\label{dental-pathology}}
+body (\citeproc{ref-rogersPalaeopathologyJoint2000}{Rogers, 2000}).
+Schmorl's nodes are identified as any cortical depressions on the
+surface of the vertebral body. Data on chronic maxillary sinusitis from
+Casna et al. (\citeproc{ref-casnaUrbanizationRespiratory2021}{2021})
+were included in this study to assess the relationship between upper
+respiratory diseases with environmental factors (i.e.~tobacco smoke,
+caffeine consumption). Lesions associated with chronic maxillary
+sinusitis as defined by Boocock et al.
+(\citeproc{ref-boocockMaxillarySinusitis1995}{1995}) were recorded for
+each individual and classified as ``pitting'', ``spicule-type bone
+formation'', ``remodeled spicules'', or ``white pitted bone''. chronic
+maxillary sinusitis was scored as absent when the sinus presented smooth
+surfaces with little or no associated pitting.
+
+\subsubsection{Dental pathology}\label{dental-pathology}
 
@@
 ``crown''. Calculus indices were calculated according to Greene and
-colleagues (2005). Calculus was scored with a four-stage scoring system
-(0-3) to score absent, slight, moderate, and heavy calculus deposits
-(Brothwell, 1981) on the lingual, buccal (and labial), and interproximal
-surfaces of each tooth. Only one score was used for the combined
-interproximal surfaces, resulting in three scores per tooth (when
-surfaces are intact), and four calculus indices per individual; upper
-anterior, upper posterior, lower anterior, lower posterior. Each index
-was calculated by dividing the sum of calculus scores for each surface
-by the total number of surfaces scored in each quadrant. If a tooth
-could not be scored on all three surfaces, the tooth was not included
-(Greene et al., 2005). Periodontitis was scored on a visual four-stage
-(0-3) scoring system according to distance from cemento-enamel junction
-of each tooth to alveolar bone (Maat \& Mastwijk, 2005).
-
-\hypertarget{calculus-sampling}{%
-\subsection{Calculus sampling}\label{calculus-sampling}}
+colleagues (\citeproc{ref-greeneQuantifyingCalculus2005}{2005}).
+Calculus was scored with a four-stage scoring system (0-3) to score
+absent, slight, moderate, and heavy calculus deposits
+(\citeproc{ref-brothwellDiggingBones1981}{Brothwell, 1981}) on the
+lingual, buccal (and labial), and interproximal surfaces of each tooth.
+Only one score was used for the combined interproximal surfaces,
+resulting in three scores per tooth (when surfaces are intact), and four
+calculus indices per individual; upper anterior, upper posterior, lower
+anterior, lower posterior. Each index was calculated by dividing the sum
+of calculus scores for each surface by the total number of surfaces
+scored in each quadrant. If a tooth could not be scored on all three
+surfaces, the tooth was not included
+(\citeproc{ref-greeneQuantifyingCalculus2005}{Greene et al., 2005}).
+Periodontitis was scored on a visual four-stage (0-3) scoring system
+according to distance from cemento-enamel junction of each tooth to
+alveolar bone (\citeproc{ref-maatManualPhysical2005}{Maat \& Mastwijk,
+2005}).
+
+\subsection{Calculus sampling}\label{calculus-sampling}
 
@@
 protocols. More details on the methods can be found in the published
-articles (Ziesemer et al., 2015, 2018). Of the 41 individuals that were
-originally included in our sample, 29 were replicated in a separate
-analysis only using calculus from the previous study.\\
+articles (\citeproc{ref-ziesemer16SChallenges2015}{Ziesemer et al.,
+2015}, \citeproc{ref-ziesemerGenomeCalculus2018}{2018}). Of the 41
+individuals that were originally included in our sample, 29 were
+replicated in a separate analysis only using calculus from the previous
+study.\\
 New dental calculus samples were taken under sterile conditions in a
@@
 
-\hypertarget{uhplc-msms}{%
-\subsection{UHPLC-MS/MS}\label{uhplc-msms}}
+\subsection{UHPLC-MS/MS}\label{uhplc-msms}
 
@@
 Amphetamine). These were part of the toxicology screening for the
-original method (Sørensen et al., 2021), developed on cadavers. In our
-study they serve as an authentication step, as their presence in
-archaeological samples could only be the result of contamination.
+original method (\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et
+al., 2021}), developed on cadavers. In our study they serve as an
+authentication step, as their presence in archaeological samples could
+only be the result of contamination.
 
@@
 isotope dilution was applied. For more details about the method and
-validation, see the original study by Sørensen and colleagues (2021).
-
-\hypertarget{statistical-analysis}{%
-\subsection{Statistical analysis}\label{statistical-analysis}}
+validation, see the original study by Sørensen and colleagues
+(\citeproc{ref-sorensenDrugsCalculus2021}{2021}).
+
+\subsection{Statistical analysis}\label{statistical-analysis}
 
@@
 
-All statistical analysis was conducted in R version 4.3.1 (2023-06-16),
-Beagle Scouts, (R Core Team, 2020). Data wrangling was conducted with
-the \textbf{tidyverse} (Wickham et al., 2019) and visualisations were
-created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
-were calculated with the \textbf{psych} package (Revelle, 2022).
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
+All statistical analysis was conducted in R version 4.3.3 (2024-02-29),
+Angel Food Cake, (\citeproc{ref-Rbase}{R Core Team, 2020}). Data
+wrangling was conducted with the \textbf{tidyverse}
+(\citeproc{ref-tidyverse2019}{Wickham et al., 2019}) and visualisations
+were created using \textbf{ggplot2} (\citeproc{ref-ggplot2}{Wickham,
+2016}). Polychoric correlations were calculated with the \textbf{psych}
+package (\citeproc{ref-Rpsych}{Revelle, 2022}).
+
+\section{Results}\label{results}
 
@@
 
-\hypertarget{tbl-compound-detect}{}
 \begin{longtable}[]{@{}lllr@{}}
+
 \caption{\label{tbl-compound-detect}Target compound including whether it
@@
 THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A;
-THCVA = tetrahydrocannabivarin acid.}\tabularnewline
-\toprule\noalign{}
-Compound & Batch 1 & Batch 2 & LLOQ \\
-\midrule\noalign{}
-\endfirsthead
+THCVA = tetrahydrocannabivarin acid.}
+
+\tabularnewline
+
 \toprule\noalign{}
@@
 Theophylline & TRUE & TRUE & 0.010 \\
+
 \end{longtable}
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-auth-plot-batch2-1.pdf}
+\centering{
+
+\includegraphics{../figures/fig-auth-plot-batch2-1.pdf}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-detection-preservation-1.pdf}
+\centering{
+
+\includegraphics{../figures/fig-detection-preservation-1.pdf}
 
@@
 
-\hypertarget{correlations-between-detected-alkaloids-and-diseases}{%
 \subsection{Correlations between detected alkaloids and
-diseases}\label{correlations-between-detected-alkaloids-and-diseases}}
+diseases}\label{correlations-between-detected-alkaloids-and-diseases}
 
@@
 
-\hypertarget{tbl-pearson}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}@{}}
+
 \caption{\label{tbl-pearson}Pearson correlation (\emph{r}) on
@@
 disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA =
-salicylic acid; PN = pipe notches.}\tabularnewline
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Caries
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Nicotine
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-SA
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Calculus
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-PN
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Theophylline
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Caffeine
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Cotinine
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
+salicylic acid; PN = pipe notches.}
+
+\tabularnewline
+
 \toprule\noalign{}
@@
 Caffeine & & & & & & & & 0.08 \\
+
 \end{longtable}
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-polycorr-1.pdf}
+\centering{
+
+\includegraphics{../figures/fig-polycorr-1.pdf}
 
@@
 
-\end{figure}
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\end{figure}%
+
+\section{Discussion}\label{discussion}
 
@@
 (UHPLC-MS/MS) using a method that was validated by co-occurrence of
-drugs and metabolites in dental calculus and blood (Sørensen et al.,
-2021). Here we have shown that the method can also be successfully
-applied to archaeological dental calculus. We extend findings from
-previous studies on alkaloids in archaeological samples by detecting
-multiple different alkaloids in dental calculus, including nicotine,
-cotinine, caffeine, theophylline, and salicylic acid. The detection of
-these compounds was solidified in a replication analysis on different
-samples from the same individuals. Cocaine and multiple cannabinoids
-were also detected during the first analysis, but were not replicated.
-We contextualize these findings within the historical and archaeological
-evidence for consumption of these drugs and dietary compounds.
+drugs and metabolites in dental calculus and blood
+(\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}). Here
+we have shown that the method can also be successfully applied to
+archaeological dental calculus. We extend findings from previous studies
+on alkaloids in archaeological samples by detecting multiple different
+alkaloids in dental calculus, including nicotine, cotinine, caffeine,
+theophylline, and salicylic acid. The detection of these compounds was
+solidified in a replication analysis on different samples from the same
+individuals. Cocaine and multiple cannabinoids were also detected during
+the first analysis, but were not replicated. We contextualize these
+findings within the historical and archaeological evidence for
+consumption of these drugs and dietary compounds.
 
@@
 detection of nicotine and cotinine is not surprising, as pipe-smoking in
-the Beemsterpolder is well-documented in the literature (Aten et al.,
-2012; Bouman, 2017), and visible on the skeletal remains as pipe notches
-(Lemmers et al., 2013). There is also documented medicinal use of
-nicotine in the Beemsterpolder, where a tobacco-smoke enema was used for
-headaches, respiratory problems, colds, and drowsiness from around 1780
-to 1830 (Aten et al., 2012). In our sample, we also detected nicotine
-and cotinine (replicated) in an old adult, probable female individual.
-In this particular case it is unlikely that the compounds entered the
-dental calculus through pipe-smoking, as the individual had no visible
-pipe notches; more likely the tobacco entered through an alternate mode
-of consumption, secondhand smoke, or the aforementioned tobacco-smoke
-enema.
+the Beemsterpolder is well-documented in the literature
+(\citeproc{ref-aten400Jaar2012}{Aten et al., 2012};
+\citeproc{ref-boumanBegravenis2017}{Bouman, 2017}), and visible on the
+skeletal remains as pipe notches
+(\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}). There
+is also documented medicinal use of nicotine in the Beemsterpolder,
+where a tobacco-smoke enema was used for headaches, respiratory
+problems, colds, and drowsiness from around 1780 to 1830
+(\citeproc{ref-aten400Jaar2012}{Aten et al., 2012}). In our sample, we
+also detected nicotine and cotinine (replicated) in an old adult,
+probable female individual. In this particular case it is unlikely that
+the compounds entered the dental calculus through pipe-smoking, as the
+individual had no visible pipe notches; more likely the tobacco entered
+through an alternate mode of consumption, secondhand smoke, or the
+aforementioned tobacco-smoke enema.
 
@@
 compounds. Caffeine is present in coffee, tea, and cocoa beans, with
-concentrations slightly higher in coffee (Bispo et al., 2002; Chin et
-al., 2008; Srdjenovic et al., 2008; Stavric et al., 1988). Theophylline
-is present in both coffee beans and tea leaves, but in negligible
-quantities (Stavric et al., 1988). It is also a primary metabolite of
-caffeine produced by the liver. Given the low correlation, there are
-likely multiple sources of caffeine and theophylline in the population,
-with tea and coffee being the most obvious.\\
+concentrations slightly higher in coffee
+(\citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al., 2002};
+\citeproc{ref-chinCaffeineContent2008}{Chin et al., 2008};
+\citeproc{ref-srdjenovicSimultaneousHPLC2008}{Srdjenovic et al., 2008};
+\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988}).
+Theophylline is present in both coffee beans and tea leaves, but in
+negligible quantities
+(\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988}).
+It is also a primary metabolite of caffeine produced by the liver. Given
+the low correlation, there are likely multiple sources of caffeine and
+theophylline in the population, with tea and coffee being the most
+obvious.\\
 Tea consumption had become widespread in the Netherlands by 1820,
-reaching all parts of society (Nierstrasz, 2015, p. 91). Historically,
-we also know that both tea and coffee were consumed in the
+reaching all parts of society
+(\citeproc{ref-nierstraszTeaTrade2015}{Nierstrasz, 2015, p. 91}).
+Historically, we also know that both tea and coffee were consumed in the
 Beemsterpolder during the 19th century. `Theegasten' (teatime) was a
 special occasion occurring from 15.00-20.00 hours, where tea was served
-along with the evening bread (Schuijtemaker, 2011). Many households also
-owned at least one coffee pot and tea pot (Bouman, 2017). Distinguishing
+along with the evening bread
+(\citeproc{ref-schuijtemakerTeTheegasten2011}{Schuijtemaker, 2011}).
+Many households also owned at least one coffee pot and tea pot
+(\citeproc{ref-boumanBegravenis2017}{Bouman, 2017}). Distinguishing
 between tea, coffee, and chocolate may be possible by also including
@@
 present in higher quantities in chocolate compared to caffeine and
-theophylline (Alañón et al., 2016; Bispo et al., 2002; Stavric et al.,
-1988). However, In addition to oral factors affecting alkaloid uptake in
-dental calculus, there is some indication that theobromine does not
-preserve well in the archaeological record (Velsko et al., 2017), and
+theophylline (\citeproc{ref-alanonAssessmentFlavanol2016}{Alañón et al.,
+2016}; \citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al.,
+2002}; \citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al.,
+1988}). However, In addition to oral factors affecting alkaloid uptake
+in dental calculus, there is some indication that theobromine does not
+preserve well in the archaeological record
+(\citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017}), and
 frequent consumption of all three items would be difficult to parse.
@@
 on these specific compounds, and the original concentration of the
-compounds in plants can be quite variable (King et al., 2017).
+compounds in plants can be quite variable
+(\citeproc{ref-kingCautionaryTales2017}{King et al., 2017}).
 
@@
 be extracted from the bark of willow trees, \emph{Salix alba}, and has
-long been used for its pain-relieving properties (Bruinsma, 1872, p.
-119). It is also present in many plant-based foods (Duthie \& Wood,
-2011; Malakar et al., 2017), including potatoes, which were a staple of
-the Beemsterpolder diet (Aten et al., 2012). The extracted quantity from
-our samples decreased over the three washes, followed by a sharp
-increase in the final calculus extraction, which is what we would expect
-to see if the salicylic acid was incorporated during life
+long been used for its pain-relieving properties
+(\citeproc{ref-bruinsmaBijdragenTot1872}{Bruinsma, 1872, p. 119}). It is
+also present in many plant-based foods
+(\citeproc{ref-duthieNaturalSalicylates2011}{Duthie \& Wood, 2011};
+\citeproc{ref-malakarNaturallyOccurring2017}{Malakar et al., 2017}),
+including potatoes, which were a staple of the Beemsterpolder diet
+(\citeproc{ref-aten400Jaar2012}{Aten et al., 2012}). The extracted
+quantity from our samples decreased over the three washes, followed by a
+sharp increase in the final calculus extraction, which is what we would
+expect to see if the salicylic acid was incorporated during life
 (Figure~\ref{fig-auth-plot-batch2}). It is important to note that,
@@
 to enter the calculus through contact with the surrounding soil.
-Salicylic acid is a very mobile organic acid (Badri \& Vivanco, 2009;
-Chen et al., 2001) and the ubiqutous presence in our samples may be
-explained by the compound leching into the dental calculus from the
-burial environment. We can therefore not confidently rule out
-environmental contamination without analysing samples from the
-surrounding soil.
+Salicylic acid is a very mobile organic acid
+(\citeproc{ref-badriRegulationFunction2009}{Badri \& Vivanco, 2009};
+\citeproc{ref-chenCa2Dependent2001}{Chen et al., 2001}) and the
+ubiqutous presence in our samples may be explained by the compound
+leching into the dental calculus from the burial environment. We can
+therefore not confidently rule out environmental contamination without
+analysing samples from the surrounding soil.
 
@@
 use of cannabinoids has been well-established in Europe since
-Medieval-times, and it was also grown in the Netherlands (Bruinsma,
-1872). Administration was most common in the form of concoctions
-containing various portions of the cannabis plant for ingestion; not
-until the late 19th century did it become recommended to smoke it for
-more immediate effects (Clarke, 2013). Dutch medicinal preparations were
-used to treat a variety of ailments and symptoms, including pain,
-inflammation, various stomach ailments, gout, and joint pains (Clarke,
-2013). Because cannabinoids have an affinity for protein-binding, they
-are less likely to diffuse from serum to saliva (Cone \& Huestis, 2007).
-This may make them difficult to detect in dental calculus unless the
+Medieval-times, and it was also grown in the Netherlands
+(\citeproc{ref-bruinsmaBijdragenTot1872}{Bruinsma, 1872}).
+Administration was most common in the form of concoctions containing
+various portions of the cannabis plant for ingestion; not until the late
+19th century did it become recommended to smoke it for more immediate
+effects (\citeproc{ref-clarkeCannabisEvolution2013}{Clarke, 2013}).
+Dutch medicinal preparations were used to treat a variety of ailments
+and symptoms, including pain, inflammation, various stomach ailments,
+gout, and joint pains
+(\citeproc{ref-clarkeCannabisEvolution2013}{Clarke, 2013}). Because
+cannabinoids have an affinity for protein-binding, they are less likely
+to diffuse from serum to saliva
+(\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007}). This
+may make them difficult to detect in dental calculus unless the
 cannibinoids were consumed orally; even then, the overall instability of
-some cannabinoids could also limit detection (Lindholst, 2010; Sørensen
-\& Hasselstrøm, 2018). Given the lack of replication, we cannot with
-security confirm that cannabis was used by the Beemster population.
+some cannabinoids could also limit detection
+(\citeproc{ref-lindholstLongTerm2010}{Lindholst, 2010};
+\citeproc{ref-sorensenEffectAntioxidants2018}{Sørensen \& Hasselstrøm,
+2018}). Given the lack of replication, we cannot with security confirm
+that cannabis was used by the Beemster population.
 
 Despite many of our sampled individuals having lived during the height
-of the opium era in the Netherlands (Macht, 1915), none of the targeted
-opioids (morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
-were detected. The absence of opioids could be a result of the people
-ascribing more to the ``traditional'' rather than ``scientific''
-medicine, although laudanum and another opium containing concoction was
-part of the ``traditional'' medicine in the Netherlands (Leuw \&
-Marshall, 1994), including Middenbeemster (Aten et al., 2012). It was
-also generally considered a drug of the upper class (Scheltema, 1907),
-and may have been more common in urban centers. The absence could also
-be attributed to postmortem degradation. It has been shown that, while
-morphine is abundant in opium, it degrades rapidly. Thebaine and
-papaverine are more resistant to various ageing processes (Chovanec et
-al., 2012), however, these were also absent from our samples.
+of the opium era in the Netherlands
+(\citeproc{ref-machtHistoryOpium1915}{Macht, 1915}), none of the
+targeted opioids (morphine, codeine, thebaine, papaverine, norcodeine,
+noscapine) were detected. The absence of opioids could be a result of
+the people ascribing more to the ``traditional'' rather than
+``scientific'' medicine, although laudanum and another opium containing
+concoction was part of the ``traditional'' medicine in the Netherlands
+(\citeproc{ref-leuwProhibitionLegalization1994}{Leuw \& Marshall,
+1994}), including Middenbeemster (\citeproc{ref-aten400Jaar2012}{Aten et
+al., 2012}). It was also generally considered a drug of the upper class
+(\citeproc{ref-scheltemaOpiumTrade1907}{Scheltema, 1907}), and may have
+been more common in urban centers. The absence could also be attributed
+to postmortem degradation. It has been shown that, while morphine is
+abundant in opium, it degrades rapidly. Thebaine and papaverine are more
+resistant to various ageing processes
+(\citeproc{ref-chovanecOpiumMasses2012}{Chovanec et al., 2012}),
+however, these were also absent from our samples.
 
@@
 and there were also issues importing enough viable specimens of coca for
-cocaine extraction (Abduca, 2019, p. 108; Mortimer, 1901, p. 179). This
-would have been the first case of coca-leaf-consumption in Europe;
+cocaine extraction (\citeproc{ref-abucaCocaTrade2019}{Abduca, 2019, p.
+108}; \citeproc{ref-mortimerHistoryCoca1901}{Mortimer, 1901, p. 179}).
+This would have been the first case of coca-leaf-consumption in Europe;
 however, we were unable to replicate any of the cocaine results in the
@@
 in diseases of the upper respiratory tract, including chronic maxillary
-sinusitis (Reh et al., 2012). Although the mechanisms by which smoking
-increases the risk of infections is not fully understood, solid evidence
-has been presented linking tobacco smoke to increased mucosal
-permeability and impairment of mucociliary clearance (Arcavi \&
-Benowitz, 2004). Such changes, together with an altered immunologic
-response, are thought to predispose to the development of chronic
-maxillary sinusitis (Slavin et al., 2005).\\
+sinusitis (\citeproc{ref-rehImpactTobacco2012}{Reh et al., 2012}).
+Although the mechanisms by which smoking increases the risk of
+infections is not fully understood, solid evidence has been presented
+linking tobacco smoke to increased mucosal permeability and impairment
+of mucociliary clearance
+(\citeproc{ref-arcaviCigaretteSmoking2004}{Arcavi \& Benowitz, 2004}).
+Such changes, together with an altered immunologic response, are thought
+to predispose to the development of chronic maxillary sinusitis
+(\citeproc{ref-slavinDiagnosisManagement2005}{Slavin et al., 2005}).\\
 We also observed a moderate positive correlation between chronic
@@
 respiratory system, suggesting a preventive association between caffeine
-intake and pneumonia (e.g. Alfaro et al., 2018; Kondo et al., 2021).
-However, while the lower respiratory tract seems to benefit from chronic
-coffee consumption, it is possible that elevated caffeine intake impacts
-mucosal moisture due to its dehydrating effect (Maughan \& Griffin,
-2003), thereby exposing individuals to greater risk of sinus infection.
+intake and pneumonia (e.g. \citeproc{ref-alfaroChronicCoffee2018}{Alfaro
+et al., 2018}; \citeproc{ref-kondoAssociationCoffee2021}{Kondo et al.,
+2021}). However, while the lower respiratory tract seems to benefit from
+chronic coffee consumption, it is possible that elevated caffeine intake
+impacts mucosal moisture due to its dehydrating effect
+(\citeproc{ref-maughanCaffeineIngestion2003}{Maughan \& Griffin, 2003}),
+thereby exposing individuals to greater risk of sinus infection.
 
 The detection of nicotine in dental calculus has previously been
-presented by Eerkens and colleagues (2018) in two individuals from
+presented by Eerkens and colleagues
+(\citeproc{ref-eerkensDentalCalculus2018}{2018}) in two individuals from
 pre-contact California. They also targeted caffeine, cotinine, and
@@
 calculus on incisors, whereas tobacco chewers may have more on molars
-(Eerkens et al., 2018). However, sampling may not be limited to mode of
-consumption. The presence of cotinine suggests that the excretion of a
-compound after being metabolised in the body is also a source of
-deposition, and that deposition of alkaloids in dental calculus can
-occur both on the way into the body, i.e.~during consumption, and on the
-way out, i.e.~disposal of waste products via saliva secretion into the
-mouth. Especially mucin-rich saliva from the sublingual and
-submandibular glands preferentially binds toxins (Dodds et al., 2005),
-and since these glands are located closest to the lower incisors, they
-may be the most effective target for these studies. This has yet to be
+(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}).
+However, sampling may not be limited to mode of consumption. The
+presence of cotinine suggests that the excretion of a compound after
+being metabolised in the body is also a source of deposition, and that
+deposition of alkaloids in dental calculus can occur both on the way
+into the body, i.e.~during consumption, and on the way out,
+i.e.~disposal of waste products via saliva secretion into the mouth.
+Especially mucin-rich saliva from the sublingual and submandibular
+glands preferentially binds toxins
+(\citeproc{ref-doddsHealthBenefits2005}{Dodds et al., 2005}), and since
+these glands are located closest to the lower incisors, they may be the
+most effective target for these studies. This has yet to be
 systematically tested in archaeological dental calculus. Because we
@@
 studies have only found minimal contributions by certain oral bacteria
-in isolation (Cogo et al., 2008; Sun et al., 2016); it is possible that
-a larger role is played by oral bacteria within larger, more
-metabolically active communities, e.g.~biofilms (Takahashi, 2015).
+in isolation (\citeproc{ref-cogoVitroEvaluation2008}{Cogo et al., 2008};
+\citeproc{ref-sunMetabolomicsEvaluation2016}{Sun et al., 2016}); it is
+possible that a larger role is played by oral bacteria within larger,
+more metabolically active communities, e.g.~biofilms
+(\citeproc{ref-takahashiOralMicrobiome2015}{Takahashi, 2015}).
 
@@
 biased our sample, as the presence of calculus may increase the risk of
-premature death (Yaussy \& DeWitte, 2019). Additionally, periodontal
-disease (often linked to the presence of calculus) is a risk-factor for
-respiratory diseases, if periodontal and respiratory pathogens enter the
-bloodstream (Azarpazhooh \& Leake, 2006; Scannapieco, 1999; Scannapieco
-\& Ho, 2001). In our sample, the percentage of chronic maxillary
-sinusitis (37.0\%) is lower than in another (more representative) male
-sample (44.1\%) (Casna et al., 2021), and the caries percentage is
-similarly lower in our sample (17.6\%) than a more representative sample
-(22.9\%) (Lemmers et al., 2013).\\
+premature death (\citeproc{ref-yaussyCalculusSurvivorship2019}{Yaussy \&
+DeWitte, 2019}). Additionally, periodontal disease (often linked to the
+presence of calculus) is a risk-factor for respiratory diseases, if
+periodontal and respiratory pathogens enter the bloodstream
+(\citeproc{ref-azarpazhoohSystematicReview2006}{Azarpazhooh \& Leake,
+2006}; \citeproc{ref-scannapiecoRoleOral1999}{Scannapieco, 1999};
+\citeproc{ref-scannapiecoPotentialAssociations2001}{Scannapieco \& Ho,
+2001}). In our sample, the percentage of chronic maxillary sinusitis
+(37.0\%) is lower than in another (more representative) male sample
+(44.1\%) (\citeproc{ref-casnaUrbanizationRespiratory2021}{Casna et al.,
+2021}), and the caries percentage is similarly lower in our sample
+(17.6\%) than a more representative sample (22.9\%)
+(\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}).\\
 We used the presence/absence of a pipe notch and concurrent detection of
@@
 as will microbial degradation of compounds by bacteria and fungi in soil
-(Liu et al., 2015), as well as the soil environment, such as
-temperature, pH, and oxygen availability (Lindholst, 2010; Mackie et
-al., 2017).\\
+(\citeproc{ref-liuNicotinedegradingMicroorganisms2015}{Liu et al.,
+2015}), as well as the soil environment, such as temperature, pH, and
+oxygen availability (\citeproc{ref-lindholstLongTerm2010}{Lindholst,
+2010}; \citeproc{ref-mackiePreservationMetaproteome2017}{Mackie et al.,
+2017}).\\
 Due to this, quantitation of the detected compounds may have limited
@@
 and inter- and intra-individual factors related to stages of biofilm
-maturation (Lustmann et al., 1976; Velsko et al., 2019; Zijnge et al.,
-2010). In short, this means it is not really possible to equate the
+maturation (\citeproc{ref-lustmannScanningElectron1976}{Lustmann et al.,
+1976}; \citeproc{ref-velskoMicrobialDifferences2019}{Velsko et al.,
+2019}; \citeproc{ref-zijngeBiofilmArchitecture2010}{Zijnge et al.,
+2010}). In short, this means it is not really possible to equate the
 absence of a compound as evidence for the absence of consumption, which
@@
 detecting alkaloids in living individuals due to the non-invasive nature
-of the testing compared to blood and urine sampling (Cone, 1993; Valen
-et al., 2017). These \emph{in vivo} studies are a valuable source of
-method validation and can help determine the feasibility of detecting
-certain alkaloids in oral fluid and, subsequently, dental calculus.
-Archaeologists, though, will likely be responsible for exploring
-dental-calculus-specific incorporation and retention of alkaloids, as
-well as their long-term preservation in the burial environment. Finally,
-following our experience with salicylic acid, we encourage all future
-studies to ensure that a control sample is taken from the soil, either
-from the soil surrounding the individual, or, ideally, directly from the
-skeletal remains. This should preferably happen before cleaning, but
-there will often be soil left over in cavities (e.g.~nasal cavity,
-orbit, auditory meatus).
+of the testing compared to blood and urine sampling
+(\citeproc{ref-coneSalivaTesting1993}{Cone, 1993};
+\citeproc{ref-valenDetermination212017}{Valen et al., 2017}). These
+\emph{in vivo} studies are a valuable source of method validation and
+can help determine the feasibility of detecting certain alkaloids in
+oral fluid and, subsequently, dental calculus. Archaeologists, though,
+will likely be responsible for exploring dental-calculus-specific
+incorporation and retention of alkaloids, as well as their long-term
+preservation in the burial environment. Finally, following our
+experience with salicylic acid, we encourage all future studies to
+ensure that a control sample is taken from the soil, either from the
+soil surrounding the individual, or, ideally, directly from the skeletal
+remains. This should preferably happen before cleaning, but there will
+often be soil left over in cavities (e.g.~nasal cavity, orbit, auditory
+meatus).
 
@@
 only mode of tobacco consumption, with others including chewing,
-drinking, cigars, and snuff (Goodman, 1994, p. 67). Pipe-smoking was
-mainly practised by males (Eerkens et al., 2018; Lemmers et al., 2013),
-so methods like the one presented here are suitable for exploring
-tobacco consumption in an entire society, rather than a trivial subset
-of past populations. Combined with other methods, it can also give us a
-more complete picture of dietary patterns and medicinal/recreational
+drinking, cigars, and snuff
+(\citeproc{ref-goodmanTobaccoHistory1994}{Goodman, 1994, p. 67}).
+Pipe-smoking was mainly practised by males
+(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018};
+\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}), so
+methods like the one presented here are suitable for exploring tobacco
+consumption in an entire society, rather than a trivial subset of past
+populations. Combined with other methods, it can also give us a more
+complete picture of dietary patterns and medicinal/recreational
 plant-use in the past by capturing multiple possible incorporation
@@
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
+\section{Conclusions}\label{conclusions}
 
@@
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
+\section*{Acknowledgements}\label{acknowledgements}
 \addcontentsline{toc}{section}{Acknowledgements}
@@
 
-\hypertarget{funding}{%
-\section*{Funding}\label{funding}}
+\section*{Funding}\label{funding}
 \addcontentsline{toc}{section}{Funding}
@@
 
-\hypertarget{conflict-of-interest-disclosure}{%
 \section*{Conflict of interest
-disclosure}\label{conflict-of-interest-disclosure}}
+disclosure}\label{conflict-of-interest-disclosure}
 \addcontentsline{toc}{section}{Conflict of interest disclosure}
@@
 
-\hypertarget{data-availability-statement}{%
 \section*{Data Availability
-Statement}\label{data-availability-statement}}
+Statement}\label{data-availability-statement}
 \addcontentsline{toc}{section}{Data Availability Statement}
@@
 (\url{https://doi.org/10.5281/zenodo.7649824}) using the structure
-recommended by the \textbf{rrtools} R package (Marwick, 2019).
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
+recommended by the \textbf{rrtools} R package
+(\citeproc{ref-Rrrtools}{Marwick, 2019}).
+
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-abucaCocaTrade2019}{}}%
+\bibitem[\citeproctext]{ref-abucaCocaTrade2019}
 Abduca, R. (2019). Coca leaf transfers to {Europe}. {Effects} on the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-alanonAssessmentFlavanol2016}{}}%
+\bibitem[\citeproctext]{ref-alanonAssessmentFlavanol2016}
 Alañón, M. E., Castle, S. M., Siswanto, P. J., Cifuentes-Gómez, T., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-alfaroChronicCoffee2018}{}}%
+\bibitem[\citeproctext]{ref-alfaroChronicCoffee2018}
 Alfaro, T. M., Monteiro, R. A., Cunha, R. A., \& Cordeiro, C. R. (2018).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-arcaviCigaretteSmoking2004}{}}%
+\bibitem[\citeproctext]{ref-arcaviCigaretteSmoking2004}
 Arcavi, L., \& Benowitz, N. L. (2004). Cigarette {Smoking} and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-aten400Jaar2012}{}}%
+\bibitem[\citeproctext]{ref-aten400Jaar2012}
 Aten, D., Bossaers, K. W. J. M., \& Misset, C. (2012). \emph{400 jaar
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-azarpazhoohSystematicReview2006}{}}%
+\bibitem[\citeproctext]{ref-azarpazhoohSystematicReview2006}
 Azarpazhooh, A., \& Leake, J. L. (2006). Systematic {Review} of the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-badriRegulationFunction2009}{}}%
+\bibitem[\citeproctext]{ref-badriRegulationFunction2009}
 Badri, D. V., \& Vivanco, J. M. (2009). Regulation and function of root
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bispoSimultaneousDetermination2002}{}}%
+\bibitem[\citeproctext]{ref-bispoSimultaneousDetermination2002}
 Bispo, M. S., Veloso, M. C. C., Pinheiro, H. L. C., De Oliveira, R. F.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-boocockMaxillarySinusitis1995}{}}%
+\bibitem[\citeproctext]{ref-boocockMaxillarySinusitis1995}
 Boocock, P., Roberts, C. A., \& Manchester, K. (1995). Maxillary
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-boumanBegravenis2017}{}}%
+\bibitem[\citeproctext]{ref-boumanBegravenis2017}
 Bouman, J. (2017). De Begravenis. \emph{De Nieuwe Schouwschuit},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-SucheyBrooks1990}{}}%
+\bibitem[\citeproctext]{ref-SucheyBrooks1990}
 Brooks, S., \& Suchey, J. M. (1990). Skeletal age determination based on
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-brothwellDiggingBones1981}{}}%
+\bibitem[\citeproctext]{ref-brothwellDiggingBones1981}
 Brothwell, D. (1981). \emph{Digging up {Bones}: {The} excavation,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bruinsmaBijdragenTot1872}{}}%
+\bibitem[\citeproctext]{ref-bruinsmaBijdragenTot1872}
 Bruinsma, J. J. (1872). \emph{Bijdragen tot de {Geneeskundige
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-buckberryAuricular2002}{}}%
+\bibitem[\citeproctext]{ref-buckberryAuricular2002}
 Buckberry, J. L., \& Chamberlain, A. T. (2002). Age estimation from the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculus2014}{}}%
+\bibitem[\citeproctext]{ref-buckleyDentalCalculus2014}
 Buckley, S., Usai, D., Jakob, T., Radini, A., \& Hardy, K. (2014).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Standards1994}{}}%
+\bibitem[\citeproctext]{ref-Standards1994}
 Buikstra, J. E., \& Ubelaker, D. H. (1994). Standards for data
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-casnaUrbanizationRespiratory2021}{}}%
+\bibitem[\citeproctext]{ref-casnaUrbanizationRespiratory2021}
 Casna, M., Burrell, C. L., Schats, R., Hoogland, M. L. P., \& Schrader,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chenCa2Dependent2001}{}}%
+\bibitem[\citeproctext]{ref-chenCa2Dependent2001}
 Chen, H., Hou, W., Kuć, J., \& Lin, Y. (2001). Ca2+‐dependent and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chinCaffeineContent2008}{}}%
+\bibitem[\citeproctext]{ref-chinCaffeineContent2008}
 Chin, J. M., Merves, M. L., Goldberger, B. A., Sampson-Cone, A., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chovanecOpiumMasses2012}{}}%
+\bibitem[\citeproctext]{ref-chovanecOpiumMasses2012}
 Chovanec, Z., Rafferty, S., \& Swiny, S. (2012). Opium for the {Masses}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-clarkeCannabisEvolution2013}{}}%
+\bibitem[\citeproctext]{ref-clarkeCannabisEvolution2013}
 Clarke, R. (2013). \emph{Cannabis : {Evolution} and {Ethnobotany}}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cogoVitroEvaluation2008}{}}%
+\bibitem[\citeproctext]{ref-cogoVitroEvaluation2008}
 Cogo, K., Montan, M. F., Bergamaschi, C. de C., D. Andrade, E., Rosalen,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-coneSalivaTesting1993}{}}%
+\bibitem[\citeproctext]{ref-coneSalivaTesting1993}
 Cone, E. J. (1993). Saliva {Testing} for {Drugs} of {Abuse}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-coneInterpretationOral2007}{}}%
+\bibitem[\citeproctext]{ref-coneInterpretationOral2007}
 Cone, E. J., \& Huestis, M. A. (2007). Interpretation of {Oral Fluid
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-doddsHealthBenefits2005}{}}%
+\bibitem[\citeproctext]{ref-doddsHealthBenefits2005}
 Dodds, M. W. J., Johnson, D. A., \& Yeh, C.-K. (2005). Health benefits
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-duthieNaturalSalicylates2011}{}}%
+\bibitem[\citeproctext]{ref-duthieNaturalSalicylates2011}
 Duthie, G. G., \& Wood, A. D. (2011). Natural salicylates: Foods ,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-echeverriaNicotineHair2013}{}}%
+\bibitem[\citeproctext]{ref-echeverriaNicotineHair2013}
 Echeverría, J., \& Niemeyer, H. M. (2013). Nicotine in the hair of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculus2018}{}}%
+\bibitem[\citeproctext]{ref-eerkensDentalCalculus2018}
 Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gismondiMultidisciplinaryApproach2020}{}}%
+\bibitem[\citeproctext]{ref-gismondiMultidisciplinaryApproach2020}
 Gismondi, A., Baldoni, M., Gnes, M., Scorrano, G., D'Agostino, A.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-goodmanTobaccoHistory1994}{}}%
+\bibitem[\citeproctext]{ref-goodmanTobaccoHistory1994}
 Goodman, J. (1994). \emph{Tobacco in history: The cultures of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-greeneQuantifyingCalculus2005}{}}%
+\bibitem[\citeproctext]{ref-greeneQuantifyingCalculus2005}
 Greene, T. R., Kuba, C. L., \& Irish, J. D. (2005). Quantifying
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-huangDecipheringGenetic2023}{}}%
+\bibitem[\citeproctext]{ref-huangDecipheringGenetic2023}
 Huang, Y., Shan, Y., Zhang, W., Lee, A. M., Li, F., Stranger, B. E., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-jinSupragingivalCalculus2002}{}}%
+\bibitem[\citeproctext]{ref-jinSupragingivalCalculus2002}
 Jin, Y., \& Yip, H.-K. (2002). Supragingival {Calculus}: {Formation} and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kingCautionaryTales2017}{}}%
+\bibitem[\citeproctext]{ref-kingCautionaryTales2017}
 King, A., Powis, T. G., Cheong, K. F., \& Gaikwad, N. W. (2017).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kondoAssociationCoffee2021}{}}%
+\bibitem[\citeproctext]{ref-kondoAssociationCoffee2021}
 Kondo, K., Suzuki, K., Washio, M., Ohfuji, S., Adachi, S., Kan, S.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lemmersMiddenbeemster2013}{}}%
+\bibitem[\citeproctext]{ref-lemmersMiddenbeemster2013}
 Lemmers, S. A. M., Schats, R., Hoogland, M. L. P., \& Waters-Rist, A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-leuwProhibitionLegalization1994}{}}%
+\bibitem[\citeproctext]{ref-leuwProhibitionLegalization1994}
 Leuw, E., \& Marshall, I. H. (1994). \emph{Between {Prohibition} and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lindholstLongTerm2010}{}}%
+\bibitem[\citeproctext]{ref-lindholstLongTerm2010}
 Lindholst, C. (2010). Long term stability of cannabis resin and cannabis
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-liuNicotinedegradingMicroorganisms2015}{}}%
+\bibitem[\citeproctext]{ref-liuNicotinedegradingMicroorganisms2015}
 Liu, J., Ma, G., Chen, T., Hou, Y., Yang, S., Zhang, K.-Q., \& Yang, J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lovejoyAuricular1985}{}}%
+\bibitem[\citeproctext]{ref-lovejoyAuricular1985}
 Lovejoy, C. O., Meindl, R. S., Pryzbeck, T. R., \& Mensforth, R. P.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lustmannScanningElectron1976}{}}%
+\bibitem[\citeproctext]{ref-lustmannScanningElectron1976}
 Lustmann, J., Lewin-Epstein, J., \& Shteyer, A. (1976). Scanning
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-maatManualPhysical2005}{}}%
+\bibitem[\citeproctext]{ref-maatManualPhysical2005}
 Maat, G., \& Mastwijk, R. (2005). Manual for the physical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-machtHistoryOpium1915}{}}%
+\bibitem[\citeproctext]{ref-machtHistoryOpium1915}
 Macht, D. I. (1915). The history of opium and some of its preparations
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mackiePreservationMetaproteome2017}{}}%
+\bibitem[\citeproctext]{ref-mackiePreservationMetaproteome2017}
 Mackie, M., Hendy, J., Lowe, A. D., Sperduti, A., Holst, M., Collins, M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-malakarNaturallyOccurring2017}{}}%
+\bibitem[\citeproctext]{ref-malakarNaturallyOccurring2017}
 Malakar, S., Gibson, P. R., Barrett, J. S., \& Muir, J. G. (2017).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
+\bibitem[\citeproctext]{ref-Rrrtools}
 Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-maughanCaffeineIngestion2003}{}}%
+\bibitem[\citeproctext]{ref-maughanCaffeineIngestion2003}
 Maughan, R. J., \& Griffin, J. (2003). Caffeine ingestion and fluid
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-meindlSutureClosure1985}{}}%
+\bibitem[\citeproctext]{ref-meindlSutureClosure1985}
 Meindl, R. S., \& Lovejoy, C. O. (1985). Ectocranial suture closure: {A}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-milmanOralFluid2011}{}}%
+\bibitem[\citeproctext]{ref-milmanOralFluid2011}
 Milman, G., Schwope, D. M., Schwilke, E. W., Darwin, W. D., Kelly, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mortimerHistoryCoca1901}{}}%
+\bibitem[\citeproctext]{ref-mortimerHistoryCoca1901}
 Mortimer, W. G. (1901). \emph{Peru. {History} of coca, "the divine
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nierstraszTeaTrade2015}{}}%
+\bibitem[\citeproctext]{ref-nierstraszTeaTrade2015}
 Nierstrasz, C. (2015). \emph{Rivalry for {Trade} in {Tea} and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ogaldeIdentificationPsychoactive2009}{}}%
+\bibitem[\citeproctext]{ref-ogaldeIdentificationPsychoactive2009}
 Ogalde, J. P., Arriaza, B. T., \& Soto, E. C. (2009). Identification of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-palmerActivityReconstruction2016}{}}%
+\bibitem[\citeproctext]{ref-palmerActivityReconstruction2016}
 Palmer, J. L. A., Hoogland, M. H. L., \& Waters‐Rist, A. L. (2016).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
+\bibitem[\citeproctext]{ref-Rbase}
 R Core Team. (2020). \emph{R: {A} language and environment for
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-raffertyCurrentResearch2012}{}}%
+\bibitem[\citeproctext]{ref-raffertyCurrentResearch2012}
 Rafferty, S. M., Lednev, I., Virkler, K., \& Chovanec, Z. (2012).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rehImpactTobacco2012}{}}%
+\bibitem[\citeproctext]{ref-rehImpactTobacco2012}
 Reh, D. D., Higgins, T. S., \& Smith, T. L. (2012). Impact of {Tobacco
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Rpsych}{}}%
+\bibitem[\citeproctext]{ref-Rpsych}
 Revelle, W. (2022). \emph{Psych: {Procedures} for psychological,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rogersPalaeopathologyJoint2000}{}}%
+\bibitem[\citeproctext]{ref-rogersPalaeopathologyJoint2000}
 Rogers, J. (2000). The palaeopathology of joint disease. In M. Cox \& S.
@@
 science.} (1st ed, pp. 163--182). {Cambridge University Press}.
-\url{https://login.ezproxy.leidenuniv.nl:2443/login?URL=https://search.ebscohost.com/login.aspx?direct=true\&db=e000xww\&AN=40641\&site=ehost-live}
-
-\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoRoleOral1999}{}}%
+\url{https://login.ezproxy.leidenuniv.nl:2443/login?URL=https://search.ebscohost.com/login.aspx?direct=true&db=e000xww&AN=40641&site=ehost-live}
+
+\bibitem[\citeproctext]{ref-scannapiecoRoleOral1999}
 Scannapieco, F. A. (1999). Role of {Oral Bacteria} in {Respiratory
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoPotentialAssociations2001}{}}%
+\bibitem[\citeproctext]{ref-scannapiecoPotentialAssociations2001}
 Scannapieco, F. A., \& Ho, A. W. (2001). Potential {Associations Between
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-scheltemaOpiumTrade1907}{}}%
+\bibitem[\citeproctext]{ref-scheltemaOpiumTrade1907}
 Scheltema, J. F. (1907). The {Opium Trade} in the {Dutch East Indies}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-schuijtemakerTeTheegasten2011}{}}%
+\bibitem[\citeproctext]{ref-schuijtemakerTeTheegasten2011}
 Schuijtemaker, D. (2011). Te Theegasten. \emph{De Nieuwe Schouwschuit},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-slavinDiagnosisManagement2005}{}}%
+\bibitem[\citeproctext]{ref-slavinDiagnosisManagement2005}
 Slavin, R. G., Spector, S. L., Bernstein, I. L., Slavin, R. G., Kaliner,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-smithDetectionOpium2018}{}}%
+\bibitem[\citeproctext]{ref-smithDetectionOpium2018}
 Smith, R. K., Stacey, R. J., Bergström, E., \& Thomas-Oates, J. (2018).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sorensenEffectAntioxidants2018}{}}%
+\bibitem[\citeproctext]{ref-sorensenEffectAntioxidants2018}
 Sørensen, L. K., \& Hasselstrøm, J. B. (2018). The effect of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sorensenDrugsCalculus2021}{}}%
+\bibitem[\citeproctext]{ref-sorensenDrugsCalculus2021}
 Sørensen, L. K., Hasselstrøm, J. B., Larsen, L. S., \& Bindslev, D. A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-srdjenovicSimultaneousHPLC2008}{}}%
+\bibitem[\citeproctext]{ref-srdjenovicSimultaneousHPLC2008}
 Srdjenovic, B., Djordjevic-Milic, V., Grujic, N., Injac, R., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-stavricVariabilityCaffeine1988}{}}%
+\bibitem[\citeproctext]{ref-stavricVariabilityCaffeine1988}
 Stavric, B., Klassen, R., Watkinson, B., Karpinski, K., Stapley, R., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sunMetabolomicsEvaluation2016}{}}%
+\bibitem[\citeproctext]{ref-sunMetabolomicsEvaluation2016}
 Sun, J., Jin, J., Beger, R. D., Cerniglia, C. E., Yang, M., \& Chen, H.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-takahashiOralMicrobiome2015}{}}%
+\bibitem[\citeproctext]{ref-takahashiOralMicrobiome2015}
 Takahashi, N. (2015). Oral {Microbiome Metabolism}: {From} {``{Who Are
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tushinghamHuntergathererTobacco2013}{}}%
+\bibitem[\citeproctext]{ref-tushinghamHuntergathererTobacco2013}
 Tushingham, S., Ardura, D., Eerkens, J. W., Palazoglu, M., Shahbaz, S.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-unoSexAgedependent2017}{}}%
+\bibitem[\citeproctext]{ref-unoSexAgedependent2017}
 Uno, Y., Takata, R., Kito, G., Yamazaki, H., Nakagawa, K., Nakamura, Y.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-valenDetermination212017}{}}%
+\bibitem[\citeproctext]{ref-valenDetermination212017}
 Valen, A., Leere Øiestad, Å. M., Strand, D. H., Skari, R., \& Berg, T.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-velskoMicrobialDifferences2019}{}}%
+\bibitem[\citeproctext]{ref-velskoMicrobialDifferences2019}
 Velsko, I. M., Fellows Yates, J. A., Aron, F., Hagan, R. W., Frantz, L.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-velskoDentalCalculus2017}{}}%
+\bibitem[\citeproctext]{ref-velskoDentalCalculus2017}
 Velsko, I. M., Overmyer, K. A., Speller, C., Klaus, L., Collins, M. J.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-warinnerEvidenceMilk2014}{}}%
+\bibitem[\citeproctext]{ref-warinnerEvidenceMilk2014}
 Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-whiteDentalCalculus1997}{}}%
+\bibitem[\citeproctext]{ref-whiteDentalCalculus1997}
 White, D. J. (1997). Dental calculus: Recent insights into occurrence,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+\bibitem[\citeproctext]{ref-ggplot2}
 Wickham, H. (2016). \emph{Ggplot2: {Elegant Graphics} for {Data
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tidyverse2019}{}}%
+\bibitem[\citeproctext]{ref-tidyverse2019}
 Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-willeRelationshipOral2009}{}}%
+\bibitem[\citeproctext]{ref-willeRelationshipOral2009}
 Wille, S. M. R., Raes, E., Lillsunde, P., Gunnar, T., Laloup, M., Samyn,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-yaussyCalculusSurvivorship2019}{}}%
+\bibitem[\citeproctext]{ref-yaussyCalculusSurvivorship2019}
 Yaussy, S. L., \& DeWitte, S. N. (2019). Calculus and survivorship in
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ziesemer16SChallenges2015}{}}%
+\bibitem[\citeproctext]{ref-ziesemer16SChallenges2015}
 Ziesemer, K. A., Mann, A. E., Sankaranarayanan, K., Schroeder, H., Ozga,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ziesemerGenomeCalculus2018}{}}%
+\bibitem[\citeproctext]{ref-ziesemerGenomeCalculus2018}
 Ziesemer, K. A., Ramos‐Madrigal, J., Mann, A. E., Brandt, B. W.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zijngeBiofilmArchitecture2010}{}}%
+\bibitem[\citeproctext]{ref-zijngeBiofilmArchitecture2010}
 Zijnge, V., van Leeuwen, M. B. M., Degener, J. E., Abbas, F., Thurnheer,

### 2024-04-16 R/generate-bib.R
fix latex unicode error in bibliography

@@ -10,6 +10,7 @@ generate_bib <- function(){
     here("analysis/paper/references.bib"),
     keys = cite_keys,
     ignore = ignore_keys,
+    translator = "bibtex",
     overwrite = T
   )
 }

### 2024-04-17 analysis/paper/paper.tex
update pcj quarto extension

@@
 \documentclass[
-  11.5pt,
+  11pt,
   leqno]{scrartcl}
@@
 \usepackage[a4paper,headheight=13pt,footskip=20pt,margin=2.54cm]{geometry}
-\usepackage{indentfirst}
+%\usepackage{indentfirst}
 
 % Modify section headings
-\usepackage{titlesec}
-\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{}
-\titleformat{\subsection}{\normalfont\bfseries}{}{.5em}{}
-\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{\addperiod}
-\newcommand{\addperiod}[1]{#1.} %https://tex.stackexchange.com/questions/137193/adding-a-period-after-section
+\usepackage[indentafter,explicit]{titlesec}
+\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{#1}
+%\titleformat{\section}{\normalfont\large\bfseries\centering}{\cdr@ifeq{\sectionname}{}{}{\sectionname~}\thesection.}{.5em}{#1}
+\titleformat{\subsection}{\normalfont\bfseries}{\thesubsection.}{.5em}{#1}
+\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{#1. }
 
@@
 \setlength{\parskip}{0pt plus 1pt}
-\linespread{1.1}
 
@@
     belowskip=12pt,
-    font=footnotesize,
+    %font=footnotesize,
+    font={footnotesize,stretch=1},
     justification=justified]{caption}
@@
 \usepackage{xcolor}
-\definecolor{darkgray}{HTML}{808080}
-\definecolor{mediumgray}{HTML}{6D6E70}
-\definecolor{ligthgray}{HTML}{d9d9d9}
-\definecolor{pciblue}{HTML}{346E92}
-\definecolor{opengreen}{HTML}{77933c}
-
-%\usepackage[colorlinks=true,linkcolor=pciblue,urlcolor=pciblue,citecolor=pciblue]{hyperref} % Not working - links in pdf not changing colour. 
-                                                         %Probably because hyperref is no the last package to be loaded, which it needs to be.
-%\urlstyle{same}
+\definecolor{PCIdark}{HTML}{346E92}
+\definecolor{PCImedium}{HTML}{77A4BF}
+\definecolor{PCIlight}{HTML}{9FBFD2}
 \usepackage{booktabs}
@@
   colorlinks=true,
-  linkcolor={pciblue},
+  linkcolor={PCIdark},
   filecolor={Maroon},
-  citecolor={pciblue},
-  urlcolor={pciblue},
+  citecolor={PCIdark},
+  urlcolor={PCIdark},
   pdfcreator={LaTeX via pandoc}}
@@
 
-\setstretch{1.5}
+\setstretch{1.1}
 \section{Introduction}\label{introduction}
@@
 an increasingly large scale
-(\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
-2020}; \citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017};
-\citeproc{ref-warinnerEvidenceMilk2014}{Warinner et al., 2014}).
+(\citeproc{ref-warinnerEvidenceMilk2014}{Warinner et al., 2014};
+\citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017};
+\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
+2020}).
 
@@
 al., 2012}), human hair
-(\citeproc{ref-echeverriaNicotineHair2013}{Echeverría \& Niemeyer,
-2013}; \citeproc{ref-ogaldeIdentificationPsychoactive2009}{Ogalde et
-al., 2009}), and even dental calculus employing both targeted
+(\citeproc{ref-ogaldeIdentificationPsychoactive2009}{Ogalde et al.,
+2009}; \citeproc{ref-echeverriaNicotineHair2013}{Echeverría \& Niemeyer,
+2013}), and even dental calculus employing both targeted
 (\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}) and
@@
 stability and ability to survive over long periods of time
-(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018};
-\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et al., 2012};
+(\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et al., 2012};
 \citeproc{ref-tushinghamHuntergathererTobacco2013}{Tushingham et al.,
-2013}).
+2013}; \citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}).
 
@@
 compounds are transferred from saliva to biofilms on the tooth surface
-(\citeproc{ref-jinSupragingivalCalculus2002}{Jin \& Yip, 2002};
-\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). This may also
-allow various alkaloids of dietary and medicinal origin to become
+(\citeproc{ref-whiteDentalCalculus1997}{White, 1997};
+\citeproc{ref-jinSupragingivalCalculus2002}{Jin \& Yip, 2002}). This may
+also allow various alkaloids of dietary and medicinal origin to become
 incorporated in dental plaque. Dental plaque undergoes frequent
@@
 (\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007};
-\citeproc{ref-milmanOralFluid2011}{Milman et al., 2011};
-\citeproc{ref-willeRelationshipOral2009}{Wille et al., 2009}). This was
-also shown to be true for dental calculus and blood
+\citeproc{ref-willeRelationshipOral2009}{Wille et al., 2009};
+\citeproc{ref-milmanOralFluid2011}{Milman et al., 2011}). This was also
+shown to be true for dental calculus and blood
 (\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}),
@@
 drug metabolism that are related to age and sex
-(\citeproc{ref-huangDecipheringGenetic2023}{Huang et al., 2023};
+(\citeproc{ref-whiteDentalCalculus1997}{White, 1997};
 \citeproc{ref-unoSexAgedependent2017}{Uno et al., 2017};
-\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). The sample
-consists of 27 males, 11 probable males, 2 probable females, and 1
-female (Figure~\ref{fig-sample-demography}). We selected males due to a
-higher occurrence of pipe notches and dental calculus deposits than
+\citeproc{ref-huangDecipheringGenetic2023}{Huang et al., 2023}). The
+sample consists of 27 males, 11 probable males, 2 probable females, and
+1 female (Figure~\ref{fig-sample-demography}). We selected males due to
+a higher occurrence of pipe notches and dental calculus deposits than
 females (unpublished observation).
@@
 
-\includegraphics{../figures/fig-sample-demography-1.pdf}
+\includegraphics{../figures/pcj-fig-sample-demography-1.pdf}
 
@@
 cranial suture closure, and epiphyseal fusion
-(\citeproc{ref-SucheyBrooks1990}{Brooks \& Suchey, 1990};
-\citeproc{ref-buckberryAuricular2002}{Buckberry \& Chamberlain, 2002};
+(\citeproc{ref-lovejoyAuricular1985}{Lovejoy et al., 1985};
+\citeproc{ref-meindlSutureClosure1985}{Meindl \& Lovejoy, 1985};
+\citeproc{ref-SucheyBrooks1990}{Brooks \& Suchey, 1990};
 \citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994};
-\citeproc{ref-lovejoyAuricular1985}{Lovejoy et al., 1985};
-\citeproc{ref-meindlSutureClosure1985}{Meindl \& Lovejoy, 1985}), and
-divided into the following categories: early young adult (18-24 years),
-late young adult (25-34 years), middle adult ( 35-49 years), old adult
-(50+ years). Preservation was visually scored on a four-stage scale
-(excellent, good, fair, poor) based on the surface condition of the
-bones and the extent of taphonomic degradation.
+\citeproc{ref-buckberryAuricular2002}{Buckberry \& Chamberlain, 2002}),
+and divided into the following categories: early young adult (18-24
+years), late young adult (25-34 years), middle adult ( 35-49 years), old
+adult (50+ years). Preservation was visually scored on a four-stage
+scale (excellent, good, fair, poor) based on the surface condition of
+the bones and the extent of taphonomic degradation.
 
@@
 
-\includegraphics{../figures/fig-auth-plot-batch2-1.pdf}
+\includegraphics{../figures/pcj-fig-auth-plot-batch2-1.pdf}
 
@@
 
-\includegraphics{../figures/fig-detection-preservation-1.pdf}
+\includegraphics{../figures/pcj-fig-detection-preservation-1.pdf}
 
@@
 
-\includegraphics{../figures/fig-polycorr-1.pdf}
+\includegraphics{../figures/pcj-fig-polycorr-1.pdf}
 
@@
 concentrations slightly higher in coffee
-(\citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al., 2002};
-\citeproc{ref-chinCaffeineContent2008}{Chin et al., 2008};
+(\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988};
+\citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al., 2002};
 \citeproc{ref-srdjenovicSimultaneousHPLC2008}{Srdjenovic et al., 2008};
-\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988}).
-Theophylline is present in both coffee beans and tea leaves, but in
-negligible quantities
-(\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al., 1988}).
-It is also a primary metabolite of caffeine produced by the liver. Given
-the low correlation, there are likely multiple sources of caffeine and
-theophylline in the population, with tea and coffee being the most
-obvious.\\
+\citeproc{ref-chinCaffeineContent2008}{Chin et al., 2008}). Theophylline
+is present in both coffee beans and tea leaves, but in negligible
+quantities (\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et
+al., 1988}). It is also a primary metabolite of caffeine produced by the
+liver. Given the low correlation, there are likely multiple sources of
+caffeine and theophylline in the population, with tea and coffee being
+the most obvious.\\
 Tea consumption had become widespread in the Netherlands by 1820,
@@
 present in higher quantities in chocolate compared to caffeine and
-theophylline (\citeproc{ref-alanonAssessmentFlavanol2016}{Alañón et al.,
-2016}; \citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et al.,
-2002}; \citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et al.,
-1988}). However, In addition to oral factors affecting alkaloid uptake
+theophylline (\citeproc{ref-stavricVariabilityCaffeine1988}{Stavric et
+al., 1988}; \citeproc{ref-bispoSimultaneousDetermination2002}{Bispo et
+al., 2002}; \citeproc{ref-alanonAssessmentFlavanol2016}{Alañón et al.,
+2016}). However, In addition to oral factors affecting alkaloid uptake
 in dental calculus, there is some indication that theobromine does not
@@
 Salicylic acid is a very mobile organic acid
-(\citeproc{ref-badriRegulationFunction2009}{Badri \& Vivanco, 2009};
-\citeproc{ref-chenCa2Dependent2001}{Chen et al., 2001}) and the
-ubiqutous presence in our samples may be explained by the compound
+(\citeproc{ref-chenCa2Dependent2001}{Chen et al., 2001};
+\citeproc{ref-badriRegulationFunction2009}{Badri \& Vivanco, 2009}) and
+the ubiqutous presence in our samples may be explained by the compound
 leching into the dental calculus from the burial environment. We can
@@
 and there were also issues importing enough viable specimens of coca for
-cocaine extraction (\citeproc{ref-abucaCocaTrade2019}{Abduca, 2019, p.
-108}; \citeproc{ref-mortimerHistoryCoca1901}{Mortimer, 1901, p. 179}).
+cocaine extraction (\citeproc{ref-mortimerHistoryCoca1901}{Mortimer,
+1901, p. 179}; \citeproc{ref-abucaCocaTrade2019}{Abduca, 2019, p. 108}).
 This would have been the first case of coca-leaf-consumption in Europe;
@@
 periodontal and respiratory pathogens enter the bloodstream
-(\citeproc{ref-azarpazhoohSystematicReview2006}{Azarpazhooh \& Leake,
-2006}; \citeproc{ref-scannapiecoRoleOral1999}{Scannapieco, 1999};
+(\citeproc{ref-scannapiecoRoleOral1999}{Scannapieco, 1999};
 \citeproc{ref-scannapiecoPotentialAssociations2001}{Scannapieco \& Ho,
-2001}). In our sample, the percentage of chronic maxillary sinusitis
-(37.0\%) is lower than in another (more representative) male sample
-(44.1\%) (\citeproc{ref-casnaUrbanizationRespiratory2021}{Casna et al.,
-2021}), and the caries percentage is similarly lower in our sample
-(17.6\%) than a more representative sample (22.9\%)
+2001}; \citeproc{ref-azarpazhoohSystematicReview2006}{Azarpazhooh \&
+Leake, 2006}). In our sample, the percentage of chronic maxillary
+sinusitis (37.0\%) is lower than in another (more representative) male
+sample (44.1\%) (\citeproc{ref-casnaUrbanizationRespiratory2021}{Casna
+et al., 2021}), and the caries percentage is similarly lower in our
+sample (17.6\%) than a more representative sample (22.9\%)
 (\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}).\\
@@
 maturation (\citeproc{ref-lustmannScanningElectron1976}{Lustmann et al.,
-1976}; \citeproc{ref-velskoMicrobialDifferences2019}{Velsko et al.,
-2019}; \citeproc{ref-zijngeBiofilmArchitecture2010}{Zijnge et al.,
-2010}). In short, this means it is not really possible to equate the
+1976}; \citeproc{ref-zijngeBiofilmArchitecture2010}{Zijnge et al.,
+2010}; \citeproc{ref-velskoMicrobialDifferences2019}{Velsko et al.,
+2019}). In short, this means it is not really possible to equate the
 absence of a compound as evidence for the absence of consumption, which
@@
 Pipe-smoking was mainly practised by males
-(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018};
-\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}), so
+(\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013};
+\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}), so
 methods like the one presented here are suitable for exploring tobacco
@@
 \bibitem[\citeproctext]{ref-abucaCocaTrade2019}
-Abduca, R. (2019). Coca leaf transfers to {Europe}. {Effects} on the
-consumption of coca in {North-western Argentina}. In M. Kaller \& F.
-Jacob (Eds.), \emph{Transatlantic {Trade} and {Global Cultural Transfers
-Since} 1492: {More} than {Commodities}}. {Routledge}.
-\url{https://books.google.com?id=13imDwAAQBAJ}
+Abduca R (2019) \href{https://books.google.com?id=13imDwAAQBAJ}{Coca
+leaf transfers to {Europe}. {Effects} on the consumption of coca in
+{North-western Argentina}.} In: \emph{Transatlantic {Trade} and {Global
+Cultural Transfers Since} 1492: {More} than {Commodities}} (eds Kaller
+M, Jacob F). {Routledge}.
 
 \bibitem[\citeproctext]{ref-alanonAssessmentFlavanol2016}
-Alañón, M. E., Castle, S. M., Siswanto, P. J., Cifuentes-Gómez, T., \&
-Spencer, J. P. E. (2016). Assessment of flavanol stereoisomers and
-caffeine and theobromine content in commercial chocolates. \emph{Food
-Chemistry}, \emph{208}, 177--184.
-\url{https://doi.org/10.1016/j.foodchem.2016.03.116}
+Alañón ME, Castle SM, Siswanto PJ, Cifuentes-Gómez T, Spencer JPE (2016)
+Assessment of flavanol stereoisomers and caffeine and theobromine
+content in commercial chocolates. \emph{Food Chemistry}, \textbf{208},
+177--184.
+https://doi.org/\href{https://doi.org/10.1016/j.foodchem.2016.03.116}{10.1016/j.foodchem.2016.03.116}
 
 \bibitem[\citeproctext]{ref-alfaroChronicCoffee2018}
-Alfaro, T. M., Monteiro, R. A., Cunha, R. A., \& Cordeiro, C. R. (2018).
-Chronic coffee consumption and respiratory disease: {A} systematic
-review. \emph{The Clinical Respiratory Journal}, \emph{12}(3),
-1283--1294. \url{https://doi.org/10.1111/crj.12662}
+Alfaro TM, Monteiro RA, Cunha RA, Cordeiro CR (2018) Chronic coffee
+consumption and respiratory disease: {A} systematic review. \emph{The
+Clinical Respiratory Journal}, \textbf{12}, 1283--1294.
+https://doi.org/\href{https://doi.org/10.1111/crj.12662}{10.1111/crj.12662}
 
 \bibitem[\citeproctext]{ref-arcaviCigaretteSmoking2004}
-Arcavi, L., \& Benowitz, N. L. (2004). Cigarette {Smoking} and
-{Infection}. \emph{Archives of Internal Medicine}, \emph{164}(20),
-2206--2216. \url{https://doi.org/10.1001/archinte.164.20.2206}
+Arcavi L, Benowitz NL (2004) Cigarette {Smoking} and {Infection}.
+\emph{Archives of Internal Medicine}, \textbf{164}, 2206--2216.
+https://doi.org/\href{https://doi.org/10.1001/archinte.164.20.2206}{10.1001/archinte.164.20.2206}
 
 \bibitem[\citeproctext]{ref-aten400Jaar2012}
-Aten, D., Bossaers, K. W. J. M., \& Misset, C. (2012). \emph{400 jaar
-Beemster: 1612-2012}. {Stichting Uitgeverij Noord-Holland}.
+Aten D, Bossaers KWJM, Misset C (2012) \emph{400 jaar Beemster:
+1612-2012}. {Stichting Uitgeverij Noord-Holland}, {Wormer}.
 
 \bibitem[\citeproctext]{ref-azarpazhoohSystematicReview2006}
-Azarpazhooh, A., \& Leake, J. L. (2006). Systematic {Review} of the
-{Association Between Respiratory Diseases} and {Oral Health}.
-\emph{Journal of Periodontology}, \emph{77}(9), 1465--1482.
-\url{https://doi.org/10.1902/jop.2006.060010}
+Azarpazhooh A, Leake JL (2006) Systematic {Review} of the {Association
+Between Respiratory Diseases} and {Oral Health}. \emph{Journal of
+Periodontology}, \textbf{77}, 1465--1482.
+https://doi.org/\href{https://doi.org/10.1902/jop.2006.060010}{10.1902/jop.2006.060010}
 
 \bibitem[\citeproctext]{ref-badriRegulationFunction2009}
-Badri, D. V., \& Vivanco, J. M. (2009). Regulation and function of root
-exudates. \emph{Plant, Cell \& Environment}, \emph{32}(6), 666--681.
-\url{https://doi.org/10.1111/j.1365-3040.2009.01926.x}
+Badri DV, Vivanco JM (2009) Regulation and function of root exudates.
+\emph{Plant, Cell \& Environment}, \textbf{32}, 666--681.
+https://doi.org/\href{https://doi.org/10.1111/j.1365-3040.2009.01926.x}{10.1111/j.1365-3040.2009.01926.x}
 
 \bibitem[\citeproctext]{ref-bispoSimultaneousDetermination2002}
-Bispo, M. S., Veloso, M. C. C., Pinheiro, H. L. C., De Oliveira, R. F.
-S., Reis, J. O. N., \& De Andrade, J. B. (2002). Simultaneous
-{Determination} of {Caffeine}, {Theobromine}, and {Theophylline} by
-{High-Performance Liquid Chromatography}. \emph{Journal of
-Chromatographic Science}, \emph{40}(1), 45--48.
-\url{https://doi.org/10.1093/chromsci/40.1.45}
+Bispo MS, Veloso MCC, Pinheiro HLC, De Oliveira RFS, Reis JON, De
+Andrade JB (2002) Simultaneous {Determination} of {Caffeine},
+{Theobromine}, and {Theophylline} by {High-Performance Liquid
+Chromatography}. \emph{Journal of Chromatographic Science}, \textbf{40},
+45--48.
+https://doi.org/\href{https://doi.org/10.1093/chromsci/40.1.45}{10.1093/chromsci/40.1.45}
 
 \bibitem[\citeproctext]{ref-boocockMaxillarySinusitis1995}
-Boocock, P., Roberts, C. A., \& Manchester, K. (1995). Maxillary
-sinusitis in {Medieval Chichester}, {England}. \emph{American Journal of
-Physical Anthropology}, \emph{98}(4), 483--495.
-\url{https://doi.org/10.1002/ajpa.1330980408}
+Boocock P, Roberts CA, Manchester K (1995) Maxillary sinusitis in
+{Medieval Chichester}, {England}. \emph{American Journal of Physical
+Anthropology}, \textbf{98}, 483--495.
+https://doi.org/\href{https://doi.org/10.1002/ajpa.1330980408}{10.1002/ajpa.1330980408}
 
 \bibitem[\citeproctext]{ref-boumanBegravenis2017}
-Bouman, J. (2017). De Begravenis. \emph{De Nieuwe Schouwschuit},
-\emph{15}, 11--15.
-\url{https://www.historischgenootschapbeemster.nl/wp-content/uploads/De_Nieuwe_Schouwschuit_15e_jaargang_november_2017.pdf}
+Bouman J (2017)
+\href{https://www.historischgenootschapbeemster.nl/wp-content/uploads/De_Nieuwe_Schouwschuit_15e_jaargang_november_2017.pdf}{De
+Begravenis}. \emph{De Nieuwe Schouwschuit}, \textbf{15}, 11--15.
 
 \bibitem[\citeproctext]{ref-SucheyBrooks1990}
-Brooks, S., \& Suchey, J. M. (1990). Skeletal age determination based on
-the os pubis: {A} comparison of the {Acsádi-Nemeskéri} and
-{Suchey-Brooks} methods. \emph{Human Evolution}, \emph{5}(3), 227--238.
-\url{https://doi.org/10.1007/BF02437238}
+Brooks S, Suchey JM (1990) Skeletal age determination based on the os
+pubis: {A} comparison of the {Acsádi-Nemeskéri} and {Suchey-Brooks}
+methods. \emph{Human Evolution}, \textbf{5}, 227--238.
+https://doi.org/\href{https://doi.org/10.1007/BF02437238}{10.1007/BF02437238}
 
 \bibitem[\citeproctext]{ref-brothwellDiggingBones1981}
-Brothwell, D. (1981). \emph{Digging up {Bones}: {The} excavation,
-treatment and study of human skeletal remains} (3rd ed.). {British
-Museum (Natural History)}.
+Brothwell D (1981) \emph{Digging up {Bones}: {The} excavation, treatment
+and study of human skeletal remains}. {British Museum (Natural
+History)}, {London}.
 
 \bibitem[\citeproctext]{ref-bruinsmaBijdragenTot1872}
-Bruinsma, J. J. (1872). \emph{Bijdragen tot de {Geneeskundige
-Plaatsbeschrijving} van {Nederland}}. {Van Weelden en Mingelen}.
-\url{https://dlcs.io/pdf/wellcome/pdf-item/b24874140/0}
+Bruinsma JJ (1872)
+\emph{\href{https://dlcs.io/pdf/wellcome/pdf-item/b24874140/0}{Bijdragen
+tot de {Geneeskundige Plaatsbeschrijving} van {Nederland}}}. {Van
+Weelden en Mingelen}, {s'Gravenhage}.
 
 \bibitem[\citeproctext]{ref-buckberryAuricular2002}
-Buckberry, J. L., \& Chamberlain, A. T. (2002). Age estimation from the
-auricular surface of the ilium: A revised method. \emph{American Journal
-of Physical Anthropology}, \emph{119}(3), 231--239.
-\url{https://doi.org/10.1002/ajpa.10130}
+Buckberry JL, Chamberlain AT (2002) Age estimation from the auricular
+surface of the ilium: A revised method. \emph{American Journal of
+Physical Anthropology}, \textbf{119}, 231--239.
+https://doi.org/\href{https://doi.org/10.1002/ajpa.10130}{10.1002/ajpa.10130}
 
 \bibitem[\citeproctext]{ref-buckleyDentalCalculus2014}
-Buckley, S., Usai, D., Jakob, T., Radini, A., \& Hardy, K. (2014).
-Dental {Calculus Reveals Unique Insights} into {Food Items}, {Cooking}
-and {Plant Processing} in {Prehistoric Central Sudan}. \emph{PLOS ONE},
-\emph{9}(7), e100808. \url{https://doi.org/10.1371/journal.pone.0100808}
+Buckley S, Usai D, Jakob T, Radini A, Hardy K (2014) Dental {Calculus
+Reveals Unique Insights} into {Food Items}, {Cooking} and {Plant
+Processing} in {Prehistoric Central Sudan}. \emph{PLOS ONE}, \textbf{9},
+e100808.
+https://doi.org/\href{https://doi.org/10.1371/journal.pone.0100808}{10.1371/journal.pone.0100808}
 
 \bibitem[\citeproctext]{ref-Standards1994}
-Buikstra, J. E., \& Ubelaker, D. H. (1994). Standards for data
-collection from human skeletal remains: {Proceedings} of a seminar at
-the {Field Museum} of {Natural History} ({Arkansas Archaeology Research
-Series} 44). \emph{Fayetteville Arkansas Archaeological Survey}.
+Buikstra JE, Ubelaker DH (1994) Standards for data collection from human
+skeletal remains: {Proceedings} of a seminar at the {Field Museum} of
+{Natural History} ({Arkansas Archaeology Research Series} 44).
+\emph{Fayetteville Arkansas Archaeological Survey}.
 
 \bibitem[\citeproctext]{ref-casnaUrbanizationRespiratory2021}
-Casna, M., Burrell, C. L., Schats, R., Hoogland, M. L. P., \& Schrader,
-S. A. (2021). Urbanization and respiratory stress in the {Northern Low
-Countries}: {A} comparative study of chronic maxillary sinusitis in two
-early modern sites from the {Netherlands} ({AD} 1626--1866).
-\emph{International Journal of Osteoarchaeology}, \emph{31}(5),
-891--901. \url{https://doi.org/10.1002/oa.3006}
+Casna M, Burrell CL, Schats R, Hoogland MLP, Schrader SA (2021)
+Urbanization and respiratory stress in the {Northern Low Countries}: {A}
+comparative study of chronic maxillary sinusitis in two early modern
+sites from the {Netherlands} ({AD} 1626--1866). \emph{International
+Journal of Osteoarchaeology}, \textbf{31}, 891--901.
+https://doi.org/\href{https://doi.org/10.1002/oa.3006}{10.1002/oa.3006}
 
 \bibitem[\citeproctext]{ref-chenCa2Dependent2001}
-Chen, H., Hou, W., Kuć, J., \& Lin, Y. (2001). Ca2+‐dependent and
-{Ca2}+‐independent excretion modes of salicylic acid in tobacco cell
-suspension culture. \emph{Journal of Experimental Botany},
-\emph{52}(359), 1219--1226.
-\url{https://doi.org/10.1093/jexbot/52.359.1219}
+Chen H, Hou W, Kuć J, Lin Y (2001) Ca2+‐dependent and {Ca2}+‐independent
+excretion modes of salicylic acid in tobacco cell suspension culture.
+\emph{Journal of Experimental Botany}, \textbf{52}, 1219--1226.
+https://doi.org/\href{https://doi.org/10.1093/jexbot/52.359.1219}{10.1093/jexbot/52.359.1219}
 
 \bibitem[\citeproctext]{ref-chinCaffeineContent2008}
-Chin, J. M., Merves, M. L., Goldberger, B. A., Sampson-Cone, A., \&
-Cone, E. J. (2008). Caffeine {Content} of {Brewed Teas}. \emph{Journal
-of Analytical Toxicology}, \emph{32}(8), 702--704.
-\url{https://doi.org/10.1093/jat/32.8.702}
+Chin JM, Merves ML, Goldberger BA, Sampson-Cone A, Cone EJ (2008)
+Caffeine {Content} of {Brewed Teas}. \emph{Journal of Analytical
+Toxicology}, \textbf{32}, 702--704.
+https://doi.org/\href{https://doi.org/10.1093/jat/32.8.702}{10.1093/jat/32.8.702}
 
 \bibitem[\citeproctext]{ref-chovanecOpiumMasses2012}
-Chovanec, Z., Rafferty, S., \& Swiny, S. (2012). Opium for the {Masses}.
-\emph{Ethnoarchaeology}, \emph{4}(1), 5--36.
-\url{https://doi.org/10.1179/eth.2012.4.1.5}
+Chovanec Z, Rafferty S, Swiny S (2012) Opium for the {Masses}.
+\emph{Ethnoarchaeology}, \textbf{4}, 5--36.
+https://doi.org/\href{https://doi.org/10.1179/eth.2012.4.1.5}{10.1179/eth.2012.4.1.5}
 
 \bibitem[\citeproctext]{ref-clarkeCannabisEvolution2013}
-Clarke, R. (2013). \emph{Cannabis : {Evolution} and {Ethnobotany}}.
-{University of California Press}.
+Clarke R (2013) \emph{Cannabis : {Evolution} and {Ethnobotany}}.
+{University of California Press}, {{[}N.p.{]}}.
 
 \bibitem[\citeproctext]{ref-cogoVitroEvaluation2008}
-Cogo, K., Montan, M. F., Bergamaschi, C. de C., D. Andrade, E., Rosalen,
-P. L., \& Groppo, F. C. (2008). In vitro evaluation of the effect of
-nicotine, cotinine, and caffeine on oral microorganisms. \emph{Canadian
-Journal of Microbiology}, \emph{54}(6), 501--508.
-\url{https://doi.org/10.1139/W08-032}
+Cogo K, Montan MF, Bergamaschi C de C, D. Andrade E, Rosalen PL, Groppo
+FC (2008) In vitro evaluation of the effect of nicotine, cotinine, and
+caffeine on oral microorganisms. \emph{Canadian Journal of
+Microbiology}, \textbf{54}, 501--508.
+https://doi.org/\href{https://doi.org/10.1139/W08-032}{10.1139/W08-032}
 
 \bibitem[\citeproctext]{ref-coneSalivaTesting1993}
-Cone, E. J. (1993). Saliva {Testing} for {Drugs} of {Abuse}.
-\emph{Annals of the New York Academy of Sciences}, \emph{694}(1),
-91--127. \url{https://doi.org/10.1111/j.1749-6632.1993.tb18346.x}
+Cone EJ (1993) Saliva {Testing} for {Drugs} of {Abuse}. \emph{Annals of
+the New York Academy of Sciences}, \textbf{694}, 91--127.
+https://doi.org/\href{https://doi.org/10.1111/j.1749-6632.1993.tb18346.x}{10.1111/j.1749-6632.1993.tb18346.x}
 
 \bibitem[\citeproctext]{ref-coneInterpretationOral2007}
-Cone, E. J., \& Huestis, M. A. (2007). Interpretation of {Oral Fluid
-Tests} for {Drugs} of {Abuse}. \emph{Annals of the New York Academy of
-Sciences}, \emph{1098}, 51--103.
-\url{https://doi.org/10.1196/annals.1384.037}
+Cone EJ, Huestis MA (2007) Interpretation of {Oral Fluid Tests} for
+{Drugs} of {Abuse}. \emph{Annals of the New York Academy of Sciences},
+\textbf{1098}, 51--103.
+https://doi.org/\href{https://doi.org/10.1196/annals.1384.037}{10.1196/annals.1384.037}
 
 \bibitem[\citeproctext]{ref-doddsHealthBenefits2005}
-Dodds, M. W. J., Johnson, D. A., \& Yeh, C.-K. (2005). Health benefits
-of saliva: A review. \emph{Journal of Dentistry}, \emph{33}(3),
-223--233. \url{https://doi.org/10.1016/j.jdent.2004.10.009}
+Dodds MWJ, Johnson DA, Yeh C-K (2005) Health benefits of saliva: A
+review. \emph{Journal of Dentistry}, \textbf{33}, 223--233.
+https://doi.org/\href{https://doi.org/10.1016/j.jdent.2004.10.009}{10.1016/j.jdent.2004.10.009}
 
 \bibitem[\citeproctext]{ref-duthieNaturalSalicylates2011}
-Duthie, G. G., \& Wood, A. D. (2011). Natural salicylates: Foods ,
-functions and disease prevention. \emph{Food \& Function}, \emph{2}(9),
-515--520. \url{https://doi.org/10.1039/C1FO10128E}
+Duthie GG, Wood AD (2011) Natural salicylates: Foods , functions and
+disease prevention. \emph{Food \& Function}, \textbf{2}, 515--520.
+https://doi.org/\href{https://doi.org/10.1039/C1FO10128E}{10.1039/C1FO10128E}
 
 \bibitem[\citeproctext]{ref-echeverriaNicotineHair2013}
-Echeverría, J., \& Niemeyer, H. M. (2013). Nicotine in the hair of
-mummies from {San Pedro} de {Atacama} ({Northern Chile}). \emph{Journal
-of Archaeological Science}, \emph{40}(10), 3561--3568.
-\url{https://doi.org/10.1016/j.jas.2013.04.030}
+Echeverría J, Niemeyer HM (2013) Nicotine in the hair of mummies from
+{San Pedro} de {Atacama} ({Northern Chile}). \emph{Journal of
+Archaeological Science}, \textbf{40}, 3561--3568.
+https://doi.org/\href{https://doi.org/10.1016/j.jas.2013.04.030}{10.1016/j.jas.2013.04.030}
 
 \bibitem[\citeproctext]{ref-eerkensDentalCalculus2018}
-Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez,
-K., Murga, E., Kaijankoski, P., Rosenthal, J. S., \& Gang, D. R. (2018).
-Dental calculus as a source of ancient alkaloids: {Detection} of
-nicotine by {LC-MS} in calculus samples from the {Americas}.
-\emph{Journal of Archaeological Science: Reports}, \emph{18}, 509--515.
-\url{https://doi.org/10.1016/j.jasrep.2018.02.004}
+Eerkens JW, Tushingham S, Brownstein KJ, Garibay R, Perez K, Murga E,
+Kaijankoski P, Rosenthal JS, Gang DR (2018) Dental calculus as a source
+of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus
+samples from the {Americas}. \emph{Journal of Archaeological Science:
+Reports}, \textbf{18}, 509--515.
+https://doi.org/\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}
 
 \bibitem[\citeproctext]{ref-gismondiMultidisciplinaryApproach2020}
-Gismondi, A., Baldoni, M., Gnes, M., Scorrano, G., D'Agostino, A.,
-Marco, G. D., Calabria, G., Petrucci, M., Müldner, G., Tersch, M. V.,
-Nardi, A., Enei, F., Canini, A., Rickards, O., Alexander, M., \&
-Martínez-Labarga, C. (2020). A multidisciplinary approach for
-investigating dietary and medicinal habits of the {Medieval} population
-of {Santa Severa} (7th-15th centuries, {Rome}, {Italy}). \emph{PLOS
-ONE}, \emph{15}(1), e0227433.
-\url{https://doi.org/10.1371/journal.pone.0227433}
+Gismondi A, Baldoni M, Gnes M, Scorrano G, D'Agostino A, Marco GD,
+Calabria G, Petrucci M, Müldner G, Tersch MV, Nardi A, Enei F, Canini A,
+Rickards O, Alexander M, Martínez-Labarga C (2020) A multidisciplinary
+approach for investigating dietary and medicinal habits of the
+{Medieval} population of {Santa Severa} (7th-15th centuries, {Rome},
+{Italy}). \emph{PLOS ONE}, \textbf{15}, e0227433.
+https://doi.org/\href{https://doi.org/10.1371/journal.pone.0227433}{10.1371/journal.pone.0227433}
 
 \bibitem[\citeproctext]{ref-goodmanTobaccoHistory1994}
-Goodman, J. (1994). \emph{Tobacco in history: The cultures of
-dependence}. {Routledge}.
+Goodman J (1994) \emph{Tobacco in history: The cultures of dependence}.
+{Routledge}, {London}.
 
 \bibitem[\citeproctext]{ref-greeneQuantifyingCalculus2005}
-Greene, T. R., Kuba, C. L., \& Irish, J. D. (2005). Quantifying
-calculus: {A} suggested new approach for recording an important
-indicator of diet and dental health. \emph{HOMO - Journal of Comparative
-Human Biology}, \emph{56}(2), 119--132.
-\url{https://doi.org/10.1016/j.jchb.2005.02.002}
+Greene TR, Kuba CL, Irish JD (2005) Quantifying calculus: {A} suggested
+new approach for recording an important indicator of diet and dental
+health. \emph{HOMO - Journal of Comparative Human Biology}, \textbf{56},
+119--132.
+https://doi.org/\href{https://doi.org/10.1016/j.jchb.2005.02.002}{10.1016/j.jchb.2005.02.002}
 
 \bibitem[\citeproctext]{ref-huangDecipheringGenetic2023}
-Huang, Y., Shan, Y., Zhang, W., Lee, A. M., Li, F., Stranger, B. E., \&
-Huang, R. S. (2023). Deciphering genetic causes for sex differences in
-human health through drug metabolism and transporter genes. \emph{Nature
-Communications}, \emph{14}(1, 1), 175.
-\url{https://doi.org/10.1038/s41467-023-35808-6}
+Huang Y, Shan Y, Zhang W, Lee AM, Li F, Stranger BE, Huang RS (2023)
+Deciphering genetic causes for sex differences in human health through
+drug metabolism and transporter genes. \emph{Nature Communications},
+\textbf{14}, 175.
+https://doi.org/\href{https://doi.org/10.1038/s41467-023-35808-6}{10.1038/s41467-023-35808-6}
 
 \bibitem[\citeproctext]{ref-jinSupragingivalCalculus2002}
-Jin, Y., \& Yip, H.-K. (2002). Supragingival {Calculus}: {Formation} and
+Jin Y, Yip H-K (2002) Supragingival {Calculus}: {Formation} and
 {Control}. \emph{Critical Reviews in Oral Biology \& Medicine}.
-\url{https://doi.org/10.1177/154411130201300506}
+https://doi.org/\href{https://doi.org/10.1177/154411130201300506}{10.1177/154411130201300506}
 
 \bibitem[\citeproctext]{ref-kingCautionaryTales2017}
-King, A., Powis, T. G., Cheong, K. F., \& Gaikwad, N. W. (2017).
-Cautionary tales on the identification of caffeinated beverages in
-{North America}. \emph{Journal of Archaeological Science}, \emph{85},
-30--40. \url{https://doi.org/10.1016/j.jas.2017.06.006}
+King A, Powis TG, Cheong KF, Gaikwad NW (2017) Cautionary tales on the
+identification of caffeinated beverages in {North America}.
+\emph{Journal of Archaeological Science}, \textbf{85}, 30--40.
+https://doi.org/\href{https://doi.org/10.1016/j.jas.2017.06.006}{10.1016/j.jas.2017.06.006}
 
 \bibitem[\citeproctext]{ref-kondoAssociationCoffee2021}
-Kondo, K., Suzuki, K., Washio, M., Ohfuji, S., Adachi, S., Kan, S.,
-Imai, S., Yoshimura, K., Miyashita, N., Fujisawa, N., Maeda, A.,
-Fukushima, W., \& Hirota, Y. (2021). Association between coffee and
-green tea intake and pneumonia among the {Japanese} elderly: A
-case-control study. \emph{Scientific Reports}, \emph{11}(1, 1), 5570.
-\url{https://doi.org/10.1038/s41598-021-84348-w}
+Kondo K, Suzuki K, Washio M, Ohfuji S, Adachi S, Kan S, Imai S,
+Yoshimura K, Miyashita N, Fujisawa N, Maeda A, Fukushima W, Hirota Y
+(2021) Association between coffee and green tea intake and pneumonia
+among the {Japanese} elderly: A case-control study. \emph{Scientific
+Reports}, \textbf{11}, 5570.
+https://doi.org/\href{https://doi.org/10.1038/s41598-021-84348-w}{10.1038/s41598-021-84348-w}
 
 \bibitem[\citeproctext]{ref-lemmersMiddenbeemster2013}
-Lemmers, S. A. M., Schats, R., Hoogland, M. L. P., \& Waters-Rist, A.
-(2013). Fysisch antropologische analyse Middenbeemster. In \emph{De
-begravingen bij de Keyserkerk te Middenbeemster} (pp. 35--60).
+Lemmers SAM, Schats R, Hoogland MLP, Waters-Rist A (2013) Fysisch
+antropologische analyse Middenbeemster. In: \emph{De begravingen bij de
+Keyserkerk te Middenbeemster}, pp. 35--60.
 
 \bibitem[\citeproctext]{ref-leuwProhibitionLegalization1994}
-Leuw, E., \& Marshall, I. H. (1994). \emph{Between {Prohibition} and
-{Legalization}: {The Dutch Experiment} in {Drug Policy}}. {Kugler
-Publications}. \url{https://books.google.com?id=2mAVkStNG5EC}
+Leuw E, Marshall IH (1994)
+\emph{\href{https://books.google.com?id=2mAVkStNG5EC}{Between
+{Prohibition} and {Legalization}: {The Dutch Experiment} in {Drug
+Policy}}}. {Kugler Publications}.
 
 \bibitem[\citeproctext]{ref-lindholstLongTerm2010}
-Lindholst, C. (2010). Long term stability of cannabis resin and cannabis
-extracts. \emph{Australian Journal of Forensic Sciences}, \emph{42}(3),
-181--190. \url{https://doi.org/10.1080/00450610903258144}
+Lindholst C (2010) Long term stability of cannabis resin and cannabis
+extracts. \emph{Australian Journal of Forensic Sciences}, \textbf{42},
+181--190.
+https://doi.org/\href{https://doi.org/10.1080/00450610903258144}{10.1080/00450610903258144}
 
 \bibitem[\citeproctext]{ref-liuNicotinedegradingMicroorganisms2015}
-Liu, J., Ma, G., Chen, T., Hou, Y., Yang, S., Zhang, K.-Q., \& Yang, J.
-(2015). Nicotine-degrading microorganisms and their potential
-applications. \emph{Applied Microbiology and Biotechnology},
-\emph{99}(9), 3775--3785.
-\url{https://doi.org/10.1007/s00253-015-6525-1}
+Liu J, Ma G, Chen T, Hou Y, Yang S, Zhang K-Q, Yang J (2015)
+Nicotine-degrading microorganisms and their potential applications.
+\emph{Applied Microbiology and Biotechnology}, \textbf{99}, 3775--3785.
+https://doi.org/\href{https://doi.org/10.1007/s00253-015-6525-1}{10.1007/s00253-015-6525-1}
 
 \bibitem[\citeproctext]{ref-lovejoyAuricular1985}
-Lovejoy, C. O., Meindl, R. S., Pryzbeck, T. R., \& Mensforth, R. P.
-(1985). Chronological metamorphosis of the auricular surface of the
-ilium: {A} new method for the determination of adult skeletal age at
-death. \emph{American Journal of Physical Anthropology}, \emph{68}(1),
-15--28. \url{https://doi.org/10.1002/ajpa.1330680103}
+Lovejoy CO, Meindl RS, Pryzbeck TR, Mensforth RP (1985) Chronological
+metamorphosis of the auricular surface of the ilium: {A} new method for
+the determination of adult skeletal age at death. \emph{American Journal
+of Physical Anthropology}, \textbf{68}, 15--28.
+https://doi.org/\href{https://doi.org/10.1002/ajpa.1330680103}{10.1002/ajpa.1330680103}
 
 \bibitem[\citeproctext]{ref-lustmannScanningElectron1976}
-Lustmann, J., Lewin-Epstein, J., \& Shteyer, A. (1976). Scanning
-electron microscopy of dental calculus. \emph{Calcified Tissue
-Research}, \emph{21}(1), 47--55.
-\url{https://doi.org/10.1007/BF02547382}
+Lustmann J, Lewin-Epstein J, Shteyer A (1976) Scanning electron
+microscopy of dental calculus. \emph{Calcified Tissue Research},
+\textbf{21}, 47--55.
+https://doi.org/\href{https://doi.org/10.1007/BF02547382}{10.1007/BF02547382}
 
 \bibitem[\citeproctext]{ref-maatManualPhysical2005}
-Maat, G., \& Mastwijk, R. (2005). Manual for the physical
-anthropological report. \emph{Barge's Anthropologica}, \emph{6}.
+Maat G, Mastwijk R (2005) Manual for the physical anthropological
+report. \emph{Barge's Anthropologica}.
 
 \bibitem[\citeproctext]{ref-machtHistoryOpium1915}
-Macht, D. I. (1915). The history of opium and some of its preparations
-and alkaloids. \emph{The Journal of the American Medical Association},
-\emph{LXIV}(6), 5.
+Macht DI (1915) The history of opium and some of its preparations and
+alkaloids. \emph{The Journal of the American Medical Association},
+\textbf{LXIV}, 5.
 
 \bibitem[\citeproctext]{ref-mackiePreservationMetaproteome2017}
-Mackie, M., Hendy, J., Lowe, A. D., Sperduti, A., Holst, M., Collins, M.
-J., \& Speller, C. F. (2017). Preservation of the metaproteome:
-Variability of protein preservation in ancient dental calculus.
-\emph{STAR: Science \& Technology of Archaeological Research},
-\emph{3}(1), 58--70. \url{https://doi.org/10.1080/20548923.2017.1361629}
+Mackie M, Hendy J, Lowe AD, Sperduti A, Holst M, Collins MJ, Speller CF
+(2017) Preservation of the metaproteome: Variability of protein
+preservation in ancient dental calculus. \emph{STAR: Science \&
+Technology of Archaeological Research}, \textbf{3}, 58--70.
+https://doi.org/\href{https://doi.org/10.1080/20548923.2017.1361629}{10.1080/20548923.2017.1361629}
 
 \bibitem[\citeproctext]{ref-malakarNaturallyOccurring2017}
-Malakar, S., Gibson, P. R., Barrett, J. S., \& Muir, J. G. (2017).
-Naturally occurring dietary salicylates: {A} closer look at common
-{Australian} foods. \emph{Journal of Food Composition and Analysis},
-\emph{57}, 31--39. \url{https://doi.org/10.1016/j.jfca.2016.12.008}
+Malakar S, Gibson PR, Barrett JS, Muir JG (2017) Naturally occurring
+dietary salicylates: {A} closer look at common {Australian} foods.
+\emph{Journal of Food Composition and Analysis}, \textbf{57}, 31--39.
+https://doi.org/\href{https://doi.org/10.1016/j.jfca.2016.12.008}{10.1016/j.jfca.2016.12.008}
 
 \bibitem[\citeproctext]{ref-Rrrtools}
-Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research
-compendium} {[}Manual{]}. \url{https://github.com/benmarwick/rrtools}
+Marwick B (2019)
+\emph{\href{https://github.com/benmarwick/rrtools}{Rrtools: {Creates} a
+reproducible research compendium}}.
 
 \bibitem[\citeproctext]{ref-maughanCaffeineIngestion2003}
-Maughan, R. J., \& Griffin, J. (2003). Caffeine ingestion and fluid
-balance: A review. \emph{Journal of Human Nutrition and Dietetics},
-\emph{16}(6), 411--420.
-\url{https://doi.org/10.1046/j.1365-277X.2003.00477.x}
+Maughan RJ, Griffin J (2003) Caffeine ingestion and fluid balance: A
+review. \emph{Journal of Human Nutrition and Dietetics}, \textbf{16},
+411--420.
+https://doi.org/\href{https://doi.org/10.1046/j.1365-277X.2003.00477.x}{10.1046/j.1365-277X.2003.00477.x}
 
 \bibitem[\citeproctext]{ref-meindlSutureClosure1985}
-Meindl, R. S., \& Lovejoy, C. O. (1985). Ectocranial suture closure: {A}
-revised method for the determination of skeletal age at death based on
-the lateral-anterior sutures. \emph{American Journal of Physical
-Anthropology}, \emph{68}(1), 57--66.
-\url{https://doi.org/10.1002/ajpa.1330680106}
+Meindl RS, Lovejoy CO (1985) Ectocranial suture closure: {A} revised
+method for the determination of skeletal age at death based on the
+lateral-anterior sutures. \emph{American Journal of Physical
+Anthropology}, \textbf{68}, 57--66.
+https://doi.org/\href{https://doi.org/10.1002/ajpa.1330680106}{10.1002/ajpa.1330680106}
 
 \bibitem[\citeproctext]{ref-milmanOralFluid2011}
-Milman, G., Schwope, D. M., Schwilke, E. W., Darwin, W. D., Kelly, D.
-L., Goodwin, R. S., Gorelick, D. A., \& Huestis, M. A. (2011). Oral
-{Fluid} and {Plasma Cannabinoid Ratios} after {Around-the-Clock
-Controlled Oral Δ9-Tetrahydrocannabinol Administration}. \emph{Clinical
-Chemistry}, \emph{57}(11), 1597--1606.
-\url{https://doi.org/10.1373/clinchem.2011.169490}
+Milman G, Schwope DM, Schwilke EW, Darwin WD, Kelly DL, Goodwin RS,
+Gorelick DA, Huestis MA (2011) Oral {Fluid} and {Plasma Cannabinoid
+Ratios} after {Around-the-Clock Controlled Oral Δ9-Tetrahydrocannabinol
+Administration}. \emph{Clinical Chemistry}, \textbf{57}, 1597--1606.
+https://doi.org/\href{https://doi.org/10.1373/clinchem.2011.169490}{10.1373/clinchem.2011.169490}
 
 \bibitem[\citeproctext]{ref-mortimerHistoryCoca1901}
-Mortimer, W. G. (1901). \emph{Peru. {History} of coca, "the divine
-plant" of the {Incas}; with an introductory account of the {Incas}, and
-of the {Andean Indians} of to-day}. {New York, J. H. Vail \& Company}.
-\url{http://archive.org/details/peruhistoryofcoc00mortrich}
+Mortimer WG (1901)
+\emph{\href{http://archive.org/details/peruhistoryofcoc00mortrich}{Peru.
+{History} of coca, "the divine plant" of the {Incas}; with an
+introductory account of the {Incas}, and of the {Andean Indians} of
+to-day}}. {New York, J. H. Vail \& Company}.
 
 \bibitem[\citeproctext]{ref-nierstraszTeaTrade2015}
-Nierstrasz, C. (2015). \emph{Rivalry for {Trade} in {Tea} and
-{Textiles}: {The English} and {Dutch East India} companies
-(1700--1800)}. {Springer}.
-\url{https://books.google.com?id=uwtaCwAAQBAJ}
+Nierstrasz C (2015)
+\emph{\href{https://books.google.com?id=uwtaCwAAQBAJ}{Rivalry for
+{Trade} in {Tea} and {Textiles}: {The English} and {Dutch East India}
+companies (1700--1800)}}. {Springer}.
 
 \bibitem[\citeproctext]{ref-ogaldeIdentificationPsychoactive2009}
-Ogalde, J. P., Arriaza, B. T., \& Soto, E. C. (2009). Identification of
-psychoactive alkaloids in ancient {Andean} human hair by gas
-chromatography/mass spectrometry. \emph{Journal of Archaeological
-Science}, \emph{36}(2), 467--472.
-\url{https://doi.org/10.1016/j.jas.2008.09.036}
+Ogalde JP, Arriaza BT, Soto EC (2009) Identification of psychoactive
+alkaloids in ancient {Andean} human hair by gas chromatography/mass
+spectrometry. \emph{Journal of Archaeological Science}, \textbf{36},
+467--472.
+https://doi.org/\href{https://doi.org/10.1016/j.jas.2008.09.036}{10.1016/j.jas.2008.09.036}
 
 \bibitem[\citeproctext]{ref-palmerActivityReconstruction2016}
-Palmer, J. L. A., Hoogland, M. H. L., \& Waters‐Rist, A. L. (2016).
-Activity {Reconstruction} of {Post}‐{Medieval Dutch Rural Villagers}
-from {Upper Limb Osteoarthritis} and {Entheseal Changes}.
-\emph{International Journal of Osteoarchaeology}, \emph{26}(1), 78--92.
-\url{https://doi.org/10.1002/oa.2397}
+Palmer JLA, Hoogland MHL, Waters‐Rist AL (2016) Activity
+{Reconstruction} of {Post}‐{Medieval Dutch Rural Villagers} from {Upper
+Limb Osteoarthritis} and {Entheseal Changes}. \emph{International
+Journal of Osteoarchaeology}, \textbf{26}, 78--92.
+https://doi.org/\href{https://doi.org/10.1002/oa.2397}{10.1002/oa.2397}
 
 \bibitem[\citeproctext]{ref-Rbase}
-R Core Team. (2020). \emph{R: {A} language and environment for
-statistical computing} {[}Manual{]}. {R Foundation for Statistical
-Computing}; {R Foundation for Statistical Computing}.
-\url{https://www.R-project.org/}
+R Core Team (2020) \emph{\href{https://www.R-project.org/}{R: {A}
+language and environment for statistical computing}}. {R Foundation for
+Statistical Computing}; {R Foundation for Statistical Computing},
+{Vienna, Austria}.
 
 \bibitem[\citeproctext]{ref-raffertyCurrentResearch2012}
-Rafferty, S. M., Lednev, I., Virkler, K., \& Chovanec, Z. (2012).
-Current research on smoking pipe residues. \emph{Journal of
-Archaeological Science}, \emph{39}(7), 1951--1959.
-\url{https://doi.org/10.1016/j.jas.2012.02.001}
+Rafferty SM, Lednev I, Virkler K, Chovanec Z (2012) Current research on
+smoking pipe residues. \emph{Journal of Archaeological Science},
+\textbf{39}, 1951--1959.
+https://doi.org/\href{https://doi.org/10.1016/j.jas.2012.02.001}{10.1016/j.jas.2012.02.001}
 
 \bibitem[\citeproctext]{ref-rehImpactTobacco2012}
-Reh, D. D., Higgins, T. S., \& Smith, T. L. (2012). Impact of {Tobacco
-Smoke} on {Chronic Rhinosinusitis} -- {A Review} of the {Literature}.
-\emph{International Forum of Allergy \& Rhinology}, \emph{2}(5), 362.
-\url{https://doi.org/10.1002/alr.21054}
+Reh DD, Higgins TS, Smith TL (2012) Impact of {Tobacco Smoke} on
+{Chronic Rhinosinusitis} -- {A Review} of the {Literature}.
+\emph{International forum of allergy \& rhinology}, \textbf{2}, 362.
+https://doi.org/\href{https://doi.org/10.1002/alr.21054}{10.1002/alr.21054}
 
 \bibitem[\citeproctext]{ref-Rpsych}
-Revelle, W. (2022). \emph{Psych: {Procedures} for psychological,
-psychometric, and personality research} {[}Manual{]}. {Northwestern
-University}. \url{https://CRAN.R-project.org/package=psych}
+Revelle W (2022)
+\emph{\href{https://CRAN.R-project.org/package=psych}{Psych:
+{Procedures} for psychological, psychometric, and personality
+research}}. {Northwestern University}, {Evanston, Illinois}.
 
 \bibitem[\citeproctext]{ref-rogersPalaeopathologyJoint2000}
-Rogers, J. (2000). The palaeopathology of joint disease. In M. Cox \& S.
-Mays (Eds.), \emph{Human osteology : {In} archaeology and forensic
-science.} (1st ed, pp. 163--182). {Cambridge University Press}.
-\url{https://login.ezproxy.leidenuniv.nl:2443/login?URL=https://search.ebscohost.com/login.aspx?direct=true&db=e000xww&AN=40641&site=ehost-live}
+Rogers J (2000)
+\href{https://login.ezproxy.leidenuniv.nl:2443/login?URL=https://search.ebscohost.com/login.aspx?direct=true&db=e000xww&AN=40641&site=ehost-live}{The
+palaeopathology of joint disease}. In: \emph{Human osteology : {In}
+archaeology and forensic science.} (eds Cox M, Mays S), pp. 163--182.
+{Cambridge University Press}, {Cambridge}.
 
 \bibitem[\citeproctext]{ref-scannapiecoRoleOral1999}
-Scannapieco, F. A. (1999). Role of {Oral Bacteria} in {Respiratory
-Infection}. \emph{Journal of Periodontology}, \emph{70}(7), 793--802.
-\url{https://doi.org/10.1902/jop.1999.70.7.793}
+Scannapieco FA (1999) Role of {Oral Bacteria} in {Respiratory
+Infection}. \emph{Journal of Periodontology}, \textbf{70}, 793--802.
+https://doi.org/\href{https://doi.org/10.1902/jop.1999.70.7.793}{10.1902/jop.1999.70.7.793}
 
 \bibitem[\citeproctext]{ref-scannapiecoPotentialAssociations2001}
-Scannapieco, F. A., \& Ho, A. W. (2001). Potential {Associations Between
-Chronic Respiratory Disease} and {Periodontal Disease}: {Analysis} of
-{National Health} and {Nutrition Examination Survey III}. \emph{Journal
-of Periodontology}, \emph{72}(1), 50--56.
-\url{https://doi.org/10.1902/jop.2001.72.1.50}
+Scannapieco FA, Ho AW (2001) Potential {Associations Between Chronic
+Respiratory Disease} and {Periodontal Disease}: {Analysis} of {National
+Health} and {Nutrition Examination Survey III}. \emph{Journal of
+Periodontology}, \textbf{72}, 50--56.
+https://doi.org/\href{https://doi.org/10.1902/jop.2001.72.1.50}{10.1902/jop.2001.72.1.50}
 
 \bibitem[\citeproctext]{ref-scheltemaOpiumTrade1907}
-Scheltema, J. F. (1907). The {Opium Trade} in the {Dutch East Indies}.
-{I}. \emph{American Journal of Sociology}, \emph{13}(1), 79--112.
+Scheltema JF (1907) The {Opium Trade} in the {Dutch East Indies}. {I}.
+\emph{American Journal of Sociology}, \textbf{13}, 79--112.
 
 \bibitem[\citeproctext]{ref-schuijtemakerTeTheegasten2011}
-Schuijtemaker, D. (2011). Te Theegasten. \emph{De Nieuwe Schouwschuit},
-\emph{9}, 16--17.
+Schuijtemaker D (2011) Te Theegasten. \emph{De Nieuwe Schouwschuit},
+\textbf{9}, 16--17.
 
 \bibitem[\citeproctext]{ref-slavinDiagnosisManagement2005}
-Slavin, R. G., Spector, S. L., Bernstein, I. L., Slavin, R. G., Kaliner,
-M. A., Kennedy, D. W., Virant, F. S., Wald, E. R., Khan, D. A.,
-Blessing-Moore, J., Lang, D. M., Nicklas, R. A., Oppenheimer, J. J.,
-Portnoy, J. M., Schuller, D. E., Tilles, S. A., Borish, L., Nathan, R.
-A., Smart, B. A., \& Vandewalker, M. L. (2005). The diagnosis and
-management of sinusitis: {A} practice parameter update. \emph{Journal of
-Allergy and Clinical Immunology}, \emph{116}, S13--S47.
-\url{https://doi.org/10.1016/j.jaci.2005.09.048}
+Slavin RG, Spector SL, Bernstein IL, Slavin RG, Kaliner MA, Kennedy DW,
+Virant FS, Wald ER, Khan DA, Blessing-Moore J, Lang DM, Nicklas RA,
+Oppenheimer JJ, Portnoy JM, Schuller DE, Tilles SA, Borish L, Nathan RA,
+Smart BA, Vandewalker ML (2005) The diagnosis and management of
+sinusitis: {A} practice parameter update. \emph{Journal of Allergy and
+Clinical Immunology}, \textbf{116}, S13--S47.
+https://doi.org/\href{https://doi.org/10.1016/j.jaci.2005.09.048}{10.1016/j.jaci.2005.09.048}
 
 \bibitem[\citeproctext]{ref-smithDetectionOpium2018}
-Smith, R. K., Stacey, R. J., Bergström, E., \& Thomas-Oates, J. (2018).
-Detection of opium alkaloids in a {Cypriot} base-ring juglet.
-\emph{Analyst}, \emph{143}(21), 5127--5136.
-\url{https://doi.org/10.1039/C8AN01040D}
+Smith RK, Stacey RJ, Bergström E, Thomas-Oates J (2018) Detection of
+opium alkaloids in a {Cypriot} base-ring juglet. \emph{Analyst},
+\textbf{143}, 5127--5136.
+https://doi.org/\href{https://doi.org/10.1039/C8AN01040D}{10.1039/C8AN01040D}
 
 \bibitem[\citeproctext]{ref-sorensenEffectAntioxidants2018}
-Sørensen, L. K., \& Hasselstrøm, J. B. (2018). The effect of
-antioxidants on the long-term stability of {THC} and related
-cannabinoids in sampled whole blood. \emph{Drug Testing and Analysis},
-\emph{10}(2), 301--309. \url{https://doi.org/10.1002/dta.2221}
+Sørensen LK, Hasselstrøm JB (2018) The effect of antioxidants on the
+long-term stability of {THC} and related cannabinoids in sampled whole
+blood. \emph{Drug Testing and Analysis}, \textbf{10}, 301--309.
+https://doi.org/\href{https://doi.org/10.1002/dta.2221}{10.1002/dta.2221}
 
 \bibitem[\citeproctext]{ref-sorensenDrugsCalculus2021}
-Sørensen, L. K., Hasselstrøm, J. B., Larsen, L. S., \& Bindslev, D. A.
-(2021). Entrapment of drugs in dental calculus -- {Detection} validation
-based on test results from post-mortem investigations. \emph{Forensic
-Science International}, \emph{319}, 110647.
-\url{https://doi.org/10.1016/j.forsciint.2020.110647}
+Sørensen LK, Hasselstrøm JB, Larsen LS, Bindslev DA (2021) Entrapment of
+drugs in dental calculus -- {Detection} validation based on test results
+from post-mortem investigations. \emph{Forensic Science International},
+\textbf{319}, 110647.
+https://doi.org/\href{https://doi.org/10.1016/j.forsciint.2020.110647}{10.1016/j.forsciint.2020.110647}
 
 \bibitem[\citeproctext]{ref-srdjenovicSimultaneousHPLC2008}
-Srdjenovic, B., Djordjevic-Milic, V., Grujic, N., Injac, R., \&
-Lepojevic, Z. (2008). Simultaneous {HPLC Determination} of {Caffeine},
-{Theobromine}, and {Theophylline} in {Food}, {Drinks}, and {Herbal
-Products}. \emph{Journal of Chromatographic Science}, \emph{46}(2),
-144--149. \url{https://doi.org/10.1093/chromsci/46.2.144}
+Srdjenovic B, Djordjevic-Milic V, Grujic N, Injac R, Lepojevic Z (2008)
+Simultaneous {HPLC Determination} of {Caffeine}, {Theobromine}, and
+{Theophylline} in {Food}, {Drinks}, and {Herbal Products}. \emph{Journal
+of Chromatographic Science}, \textbf{46}, 144--149.
+https://doi.org/\href{https://doi.org/10.1093/chromsci/46.2.144}{10.1093/chromsci/46.2.144}
 
 \bibitem[\citeproctext]{ref-stavricVariabilityCaffeine1988}
-Stavric, B., Klassen, R., Watkinson, B., Karpinski, K., Stapley, R., \&
-Fried, P. (1988). Variability in caffeine consumption from coffee and
-tea: {Possible} significance for epidemiological studies. \emph{Food and
-Chemical Toxicology}, \emph{26}(2), 111--118.
-\url{https://doi.org/10.1016/0278-6915(88)90107-X}
+Stavric B, Klassen R, Watkinson B, Karpinski K, Stapley R, Fried P
+(1988) Variability in caffeine consumption from coffee and tea:
+{Possible} significance for epidemiological studies. \emph{Food and
+Chemical Toxicology}, \textbf{26}, 111--118.
+https://doi.org/\href{https://doi.org/10.1016/0278-6915(88)90107-X}{10.1016/0278-6915(88)90107-X}
 
 \bibitem[\citeproctext]{ref-sunMetabolomicsEvaluation2016}
-Sun, J., Jin, J., Beger, R. D., Cerniglia, C. E., Yang, M., \& Chen, H.
-(2016). Metabolomics evaluation of the impact of smokeless tobacco
-exposure on the oral bacterium {Capnocytophaga} sputigena.
-\emph{Toxicology in Vitro}, \emph{36}, 133--141.
-\url{https://doi.org/10.1016/j.tiv.2016.07.020}
+Sun J, Jin J, Beger RD, Cerniglia CE, Yang M, Chen H (2016) Metabolomics
+evaluation of the impact of smokeless tobacco exposure on the oral
+bacterium {Capnocytophaga} sputigena. \emph{Toxicology in Vitro},
+\textbf{36}, 133--141.
+https://doi.org/\href{https://doi.org/10.1016/j.tiv.2016.07.020}{10.1016/j.tiv.2016.07.020}
 
 \bibitem[\citeproctext]{ref-takahashiOralMicrobiome2015}
-Takahashi, N. (2015). Oral {Microbiome Metabolism}: {From} {``{Who Are
+Takahashi N (2015) Oral {Microbiome Metabolism}: {From} {``{Who Are
 They}?''} To {``{What Are They Doing}?''} \emph{Journal of Dental
-Research}, \emph{94}(12), 1628--1637.
-\url{https://doi.org/10.1177/0022034515606045}
+Research}, \textbf{94}, 1628--1637.
+https://doi.org/\href{https://doi.org/10.1177/0022034515606045}{10.1177/0022034515606045}
 
 \bibitem[\citeproctext]{ref-tushinghamHuntergathererTobacco2013}
-Tushingham, S., Ardura, D., Eerkens, J. W., Palazoglu, M., Shahbaz, S.,
-\& Fiehn, O. (2013). Hunter-gatherer tobacco smoking: Earliest evidence
-from the {Pacific Northwest Coast} of {North America}. \emph{Journal of
-Archaeological Science}, \emph{40}(2), 1397--1407.
-\url{https://doi.org/10.1016/j.jas.2012.09.019}
+Tushingham S, Ardura D, Eerkens JW, Palazoglu M, Shahbaz S, Fiehn O
+(2013) Hunter-gatherer tobacco smoking: Earliest evidence from the
+{Pacific Northwest Coast} of {North America}. \emph{Journal of
+Archaeological Science}, \textbf{40}, 1397--1407.
+https://doi.org/\href{https://doi.org/10.1016/j.jas.2012.09.019}{10.1016/j.jas.2012.09.019}
 
 \bibitem[\citeproctext]{ref-unoSexAgedependent2017}
-Uno, Y., Takata, R., Kito, G., Yamazaki, H., Nakagawa, K., Nakamura, Y.,
-Kamataki, T., \& Katagiri, T. (2017). Sex- and age-dependent gene
-expression in human liver: {An} implication for drug-metabolizing
-enzymes. \emph{Drug Metabolism and Pharmacokinetics}, \emph{32}(1),
-100--107. \url{https://doi.org/10.1016/j.dmpk.2016.10.409}
+Uno Y, Takata R, Kito G, Yamazaki H, Nakagawa K, Nakamura Y, Kamataki T,
+Katagiri T (2017) Sex- and age-dependent gene expression in human liver:
+{An} implication for drug-metabolizing enzymes. \emph{Drug Metabolism
+and Pharmacokinetics}, \textbf{32}, 100--107.
+https://doi.org/\href{https://doi.org/10.1016/j.dmpk.2016.10.409}{10.1016/j.dmpk.2016.10.409}
 
 \bibitem[\citeproctext]{ref-valenDetermination212017}
-Valen, A., Leere Øiestad, Å. M., Strand, D. H., Skari, R., \& Berg, T.
-(2017). Determination of 21 drugs in oral fluid using fully automated
-supported liquid extraction and {UHPLC-MS}/{MS}. \emph{Drug Testing and
-Analysis}, \emph{9}(5), 808--823. \url{https://doi.org/10.1002/dta.2045}
+Valen A, Leere Øiestad ÅM, Strand DH, Skari R, Berg T (2017)
+Determination of 21 drugs in oral fluid using fully automated supported
+liquid extraction and {UHPLC-MS}/{MS}. \emph{Drug Testing and Analysis},
+\textbf{9}, 808--823.
+https://doi.org/\href{https://doi.org/10.1002/dta.2045}{10.1002/dta.2045}
 
 \bibitem[\citeproctext]{ref-velskoMicrobialDifferences2019}
-Velsko, I. M., Fellows Yates, J. A., Aron, F., Hagan, R. W., Frantz, L.
-A. F., Loe, L., Martinez, J. B. R., Chaves, E., Gosden, C., Larson, G.,
-\& Warinner, C. (2019). Microbial differences between dental plaque and
-historic dental calculus are related to oral biofilm maturation stage.
-\emph{Microbiome}, \emph{7}(1), 102.
-\url{https://doi.org/10.1186/s40168-019-0717-3}
+Velsko IM, Fellows Yates JA, Aron F, Hagan RW, Frantz LAF, Loe L,
+Martinez JBR, Chaves E, Gosden C, Larson G, Warinner C (2019) Microbial
+differences between dental plaque and historic dental calculus are
+related to oral biofilm maturation stage. \emph{Microbiome}, \textbf{7},
+102.
+https://doi.org/\href{https://doi.org/10.1186/s40168-019-0717-3}{10.1186/s40168-019-0717-3}
 
 \bibitem[\citeproctext]{ref-velskoDentalCalculus2017}
-Velsko, I. M., Overmyer, K. A., Speller, C., Klaus, L., Collins, M. J.,
-Loe, L., Frantz, L. A. F., Sankaranarayanan, K., Lewis, C. M., Martinez,
-J. B. R., Chaves, E., Coon, J. J., Larson, G., \& Warinner, C. (2017).
-The dental calculus metabolome in modern and historic samples.
-\emph{Metabolomics}, \emph{13}(11), 134.
-\url{https://doi.org/10.1007/s11306-017-1270-3}
+Velsko IM, Overmyer KA, Speller C, Klaus L, Collins MJ, Loe L, Frantz
+LAF, Sankaranarayanan K, Lewis CM, Martinez JBR, Chaves E, Coon JJ,
+Larson G, Warinner C (2017) The dental calculus metabolome in modern and
+historic samples. \emph{Metabolomics}, \textbf{13}, 134.
+https://doi.org/\href{https://doi.org/10.1007/s11306-017-1270-3}{10.1007/s11306-017-1270-3}
 
 \bibitem[\citeproctext]{ref-warinnerEvidenceMilk2014}
-Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R.,
-Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O. E., Swallow, D. M.,
-Fotakis, A., Christensen, R. J., Olsen, J. V., Liebert, A., Montalva,
-N., Fiddyment, S., Charlton, S., Mackie, M., Canci, A., \ldots{}
-Collins, M. J. (2014). Direct evidence of milk consumption from ancient
-human dental calculus. \emph{Scientific Reports}, \emph{4}, 7104.
-\url{https://doi.org/10.1038/srep07104}
+Warinner C, Hendy J, Speller C, Cappellini E, Fischer R, Trachsel C,
+Arneborg J, Lynnerup N, Craig OE, Swallow DM, Fotakis A, Christensen RJ,
+Olsen JV, Liebert A, Montalva N, Fiddyment S, Charlton S, Mackie M,
+Canci A, Bouwman A, Ruhli F, Gilbert MT, Collins MJ (2014) Direct
+evidence of milk consumption from ancient human dental calculus.
+\emph{Scientific Reports}, \textbf{4}, 7104.
+https://doi.org/\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}
 
 \bibitem[\citeproctext]{ref-whiteDentalCalculus1997}
-White, D. J. (1997). Dental calculus: Recent insights into occurrence,
+White DJ (1997) Dental calculus: Recent insights into occurrence,
 formation, prevention, removal and oral health effects of supragingival
 and subgingival deposits. \emph{European Journal of Oral Sciences},
-\emph{105}(5), 508--522.
-\url{https://doi.org/10.1111/j.1600-0722.1997.tb00238.x}
+\textbf{105}, 508--522.
+https://doi.org/\href{https://doi.org/10.1111/j.1600-0722.1997.tb00238.x}{10.1111/j.1600-0722.1997.tb00238.x}
 
 \bibitem[\citeproctext]{ref-ggplot2}
-Wickham, H. (2016). \emph{Ggplot2: {Elegant Graphics} for {Data
-Analysis}}. {Springer-Verlag}. \url{https://ggplot2.tidyverse.org}
+Wickham H (2016) \emph{\href{https://ggplot2.tidyverse.org}{Ggplot2:
+{Elegant Graphics} for {Data Analysis}}}. {Springer-Verlag}, {New York}.
 
 \bibitem[\citeproctext]{ref-tidyverse2019}
-Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D.,
-François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M.,
-Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J.,
-Robinson, D., Seidel, D. P., Spinu, V., \ldots{} Yutani, H. (2019).
-Welcome to the {tidyverse}. \emph{Journal of Open Source Software},
-\emph{4}(43), 1686. \url{https://doi.org/10.21105/joss.01686}
+Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
+Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
+Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K,
+Vaughan D, Wilke C, Woo K, Yutani H (2019) Welcome to the {tidyverse}.
+\emph{Journal of Open Source Software}, \textbf{4}, 1686.
+https://doi.org/\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}
 
 \bibitem[\citeproctext]{ref-willeRelationshipOral2009}
-Wille, S. M. R., Raes, E., Lillsunde, P., Gunnar, T., Laloup, M., Samyn,
-N., Christophersen, A. S., Moeller, M. R., Hammer, K. P., \& Verstraete,
-A. G. (2009). Relationship {Between Oral Fluid} and {Blood
-Concentrations} of {Drugs} of {Abuse} in {Drivers Suspected} of {Driving
-Under} the {Influence} of {Drugs}. \emph{Therapeutic Drug Monitoring},
-\emph{31}(4), 511. \url{https://doi.org/10.1097/FTD.0b013e3181ae46ea}
+Wille SMR, Raes E, Lillsunde P, Gunnar T, Laloup M, Samyn N,
+Christophersen AS, Moeller MR, Hammer KP, Verstraete AG (2009)
+Relationship {Between Oral Fluid} and {Blood Concentrations} of {Drugs}
+of {Abuse} in {Drivers Suspected} of {Driving Under} the {Influence} of
+{Drugs}. \emph{Therapeutic Drug Monitoring}, \textbf{31}, 511.
+https://doi.org/\href{https://doi.org/10.1097/FTD.0b013e3181ae46ea}{10.1097/FTD.0b013e3181ae46ea}
 
 \bibitem[\citeproctext]{ref-yaussyCalculusSurvivorship2019}
-Yaussy, S. L., \& DeWitte, S. N. (2019). Calculus and survivorship in
-medieval {London}: {The} association between dental disease and a
-demographic measure of general health. \emph{American Journal of
-Physical Anthropology}, \emph{168}(3), 552--565.
-\url{https://doi.org/10.1002/ajpa.23772}
+Yaussy SL, DeWitte SN (2019) Calculus and survivorship in medieval
+{London}: {The} association between dental disease and a demographic
+measure of general health. \emph{American Journal of Physical
+Anthropology}, \textbf{168}, 552--565.
+https://doi.org/\href{https://doi.org/10.1002/ajpa.23772}{10.1002/ajpa.23772}
 
 \bibitem[\citeproctext]{ref-ziesemer16SChallenges2015}
-Ziesemer, K. A., Mann, A. E., Sankaranarayanan, K., Schroeder, H., Ozga,
-A. T., Brandt, B. W., Zaura, E., Waters-Rist, A., Hoogland, M.,
-Salazar-Garcia, D. C., Aldenderfer, M., Speller, C., Hendy, J., Weston,
-D. A., MacDonald, S. J., Thomas, G. H., Collins, M. J., Lewis, C. M.,
-Hofman, C., \& Warinner, C. (2015). Intrinsic challenges in ancient
+Ziesemer KA, Mann AE, Sankaranarayanan K, Schroeder H, Ozga AT, Brandt
+BW, Zaura E, Waters-Rist A, Hoogland M, Salazar-Garcia DC, Aldenderfer
+M, Speller C, Hendy J, Weston DA, MacDonald SJ, Thomas GH, Collins MJ,
+Lewis CM, Hofman C, Warinner C (2015) Intrinsic challenges in ancient
 microbiome reconstruction using {16S rRNA} gene amplification. \emph{Sci
-Rep}, \emph{5}, 16498. \url{https://doi.org/10.1038/srep16498}
+Rep}, \textbf{5}, 16498.
+https://doi.org/\href{https://doi.org/10.1038/srep16498}{10.1038/srep16498}
 
 \bibitem[\citeproctext]{ref-ziesemerGenomeCalculus2018}
-Ziesemer, K. A., Ramos‐Madrigal, J., Mann, A. E., Brandt, B. W.,
-Sankaranarayanan, K., Ozga, A. T., Hoogland, M., Hofman, C. A.,
-Salazar‐García, D. C., Frohlich, B., Milner, G. R., Stone, A. C.,
-Aldenderfer, M., Lewis, C. M., Hofman, C. L., Warinner, C., \&
-Schroeder, H. (2018). The efficacy of whole human genome capture on
-ancient dental calculus and dentin. \emph{American Journal of Physical
-Anthropology}. \url{https://doi.org/10.1002/ajpa.23763}
+Ziesemer KA, Ramos‐Madrigal J, Mann AE, Brandt BW, Sankaranarayanan K,
+Ozga AT, Hoogland M, Hofman CA, Salazar‐García DC, Frohlich B, Milner
+GR, Stone AC, Aldenderfer M, Lewis CM, Hofman CL, Warinner C, Schroeder
+H (2018) The efficacy of whole human genome capture on ancient dental
+calculus and dentin. \emph{American Journal of Physical Anthropology}.
+https://doi.org/\href{https://doi.org/10.1002/ajpa.23763}{10.1002/ajpa.23763}
 
 \bibitem[\citeproctext]{ref-zijngeBiofilmArchitecture2010}
-Zijnge, V., van Leeuwen, M. B. M., Degener, J. E., Abbas, F., Thurnheer,
-T., Gmür, R., \& M. Harmsen, H. J. (2010). Oral {Biofilm Architecture}
-on {Natural Teeth}. \emph{PLoS ONE}, \emph{5}(2), e9321.
-\url{https://doi.org/10.1371/journal.pone.0009321}
+Zijnge V, van Leeuwen MBM, Degener JE, Abbas F, Thurnheer T, Gmür R, M.
+Harmsen HJ (2010) Oral {Biofilm Architecture} on {Natural Teeth} (S
+Bereswill, Ed,). \emph{PLoS ONE}, \textbf{5}, e9321.
+https://doi.org/\href{https://doi.org/10.1371/journal.pone.0009321}{10.1371/journal.pone.0009321}
 

### 2024-04-22 analysis/paper/_results.qmd
reduce number of text lines
add pcj extension
minor formatting changes

--- 2023-10-27 analysis/paper/_results.qmd
@@
 but not in females.
-In batch 2, the method was able to detect some form
-of tobacco in
+In batch 2, the method was able to detect some form of tobacco in
 `r nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0))`
@@
 (`r scales::percent(nrow(filter(tobacco_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(tobacco_accuracy, pipe_notch == "Y")), accuracy = 0.1)`).
-When also considering correct the absence of a tobacco alkaloid together with the absence
-of a pipe notch, the accuracy of the method is
+When also considering correct the absence of a tobacco alkaloid together with the
+absence of a pipe notch, the accuracy of the method is
 `r scales::percent(mean(tobacco_accuracy$correct, na.rm = T), accuracy = 0.1)`.
@@
 OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid; PN = pipe notches."
+#| tbl-colwidths: [15,10,10,10,10,10,15,10,10]
 

### 2024-04-24 analysis/paper/_methods.qmd
incoporate changes from pcj branch

@@
 and divided into the following categories:
-early young adult (18-24 years), late young adult (25-34 years), middle adult (
-35-49 years), old adult (50+ years). Preservation was visually scored on a
+early young adult (18-24 years), late young adult (25-34 years), middle adult
+(35-49 years), old adult (50+ years). Preservation was visually scored on a
 four-stage scale (excellent, good, fair, poor) based on the surface condition of

### 2024-04-24 analysis/paper/paper.qmd
incoporate changes from pcj branch

--- 2023-11-03 analysis/paper/paper.qmd
@@
 de Groot from the Middenbeemster Historical Society for their input on early
-draft manuscripts and a lovely guided tour of Middenbeemster. We also thank
-Louise Le Meillour for taking on our preprint and providing helpful comments,
-and Mario Zimmerman and two other anonymous reviewers for their time and invaluable
-insights.
+draft manuscripts and a lovely guided tour of Middenbeemster.
+
+We also thank Louise Le Meillour for taking on our preprint and providing helpful
+comments, and Mario Zimmerman and two other anonymous reviewers for their time
+and invaluable insights. This paper was evaluated during a CODECHECK event at TU
+Delft by Daniel Nüst and Stephen Eglen, who provided helpful feedback on the
+reproducibility of all code used in this paper.
+
+Preprint version 5 of this article has been peer-reviewed and recommended by PCI
+Archaeology <https://doi.org/10.24072/pci.archaeo.100389> [@meillourDetectionPlantderived2023].
 
@@
 
-All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.8061483>).
+All raw data is available on Zenodo [<https://doi.org/10.5281/zenodo.8061483>,
+@bartholdyDentalPathological2023].
 Analysis scripts, and the source code for the manuscript and
 supplementary materials are available as a research compendium
-(<https://doi.org/10.5281/zenodo.7649824>)
+[<https://doi.org/10.5281/zenodo.7649824>, @bartholdyMB11CalculusPilot]
 using the structure recommended by the **rrtools** R package [@Rrrtools].

### 2024-04-24 analysis/paper/paper.tex
add extra files from post-rendering

--- 2023-11-20 analysis/paper/paper.tex
@@
 \usepackage{xcolor}
-\usepackage[top=30mm,bottom=30mm,left=30mm,heightrounded]{geometry}
+\usepackage[top=10mm,bottom=10mm,left=30mm,heightrounded]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \usepackage{xcolor}
-\usepackage{lineno}
+\usepackage{fancyhdr, graphicx}
+\usepackage[hidelinks]{hyperref}
+
+\fancyhead{\vspace{70pt}}
+\renewcommand{\headrulewidth}{1.8pt}
+
 \makeatletter
-\makeatother
-\makeatletter
+\def\@maketitle{%
+  \newpage
+  \hfill\href{https://doi.org/10.24072/pci.archaeo.100389}{\includegraphics[width=3.5cm]{badge_PCI_Archaeology.png}}\hfill
+  \null
+ \begin{center}
+{\LARGE
+\textbf\newline{\@title}
+}
+%\newline
+
+\end{center}
+  \par}
 \makeatother
@@
 \ifdefined\figurename
-  \renewcommand*\figurename{Fig}
+  \renewcommand*\figurename{Figure}
 \else
-  \newcommand\figurename{Fig}
+  \newcommand\figurename{Figure}
 \fi
@@
 
-%\maketitle
+\maketitle
 
@@
 
-%\thispagestyle{fancy}
-
-\LARGE{Multiproxy analysis exploring patterns of diet and disease in
-dental calculus and skeletal remains from a 19th century Dutch
-population}
+\thispagestyle{fancy}
 
@@
 \normalsize{Bjørn Peare
-Bartholdy}\textsuperscript{1,2*}, \normalsize{Jørgen B.
-Hasselstrøm}\textsuperscript{3}, \normalsize{Lambert K.
-Sørensen}\textsuperscript{3}, \normalsize{Maia
+Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
+Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
+Sørensen}\textsuperscript{2}, \normalsize{Maia
 Casna}\textsuperscript{1}, \normalsize{Menno
 Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
-Beemster}\textsuperscript{4}, \normalsize{Amanda G.
+Beemster}\textsuperscript{3}, \normalsize{Amanda G.
 Henry}\textsuperscript{1}\vspace{12pt}
 
-%\rule{\textwidth}{1.8pt}\vspace{8pt}
+\rule{\textwidth}{1.8pt}\vspace{8pt}
 
@@
 Archaeological Sciences, Leiden
-University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Faculty
-of Mechanical, Maritime and Materials Engineering, Delft University of
-Technology}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Department
+University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Department
 of Forensic Medicine, Aarhus
-University}}\\ \textsuperscript{\footnotesize{4}}\textit{\small{Historisch
+University}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Historisch
 Genootschap Beemster}}
@@
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* \small{b.p.bartholdy@tudelft.nl}
+* \small{b.p.bartholdy@arch.leidenuniv.nl}
 
 \vspace{12pt}
-
-\linenumbers
 
@@
 
-%\vfill
-
-%\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\vfill
+
+\newgeometry{top=30mm,left=30mm}
+\section{Introduction}\label{introduction}
 
@@
 
-\hypertarget{materials}{%
-\section{Materials}\label{materials}}
+\section{Materials}\label{materials}
 
@@
 White, 1997). The sample consists of 27 males, 11 probable males, 2
-probable females, and 1 female (\textbf{?@fig-sample-demography}). We
+probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
 selected males due to a higher occurrence of pipe notches and dental
@@
 
-\emph{Fig 1 (fig-sample-demography): Overview of sample demography. Left
-plot is the first batch and right plot is the replication batch with 29
-of the individuals from the first batch. eya = early young adult (18-24
-years); lya = late young adult (25-34 years); ma = middle adult (35-49
-years); old = old adult (50+ years). Male? = probable male; Female? =
-probable female.}
-
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
-
-\hypertarget{skeletal-analysis}{%
-\subsection{Skeletal analysis}\label{skeletal-analysis}}
+\begin{figure}
+
+\centering{
+
+\includegraphics{../figures/fig-sample-demography-1.pdf}
+
+}
+
+\caption{\label{fig-sample-demography}Overview of sample demography.
+Left plot is the first batch and right plot is the replication batch
+with 29 of the individuals from the first batch. eya = early young adult
+(18-24 years); lya = late young adult (25-34 years); ma = middle adult
+(35-49 years); old = old adult (50+ years). Male? = probable male;
+Female? = probable female.}
+
+\end{figure}%
+
+\section{Methods}\label{methods}
+
+\subsection{Skeletal analysis}\label{skeletal-analysis}
 
@@
 divided into the following categories: early young adult (18-24 years),
-late young adult (25-34 years), middle adult ( 35-49 years), old adult
+late young adult (25-34 years), middle adult (35-49 years), old adult
 (50+ years). Preservation was visually scored on a four-stage scale
@@
 
-\hypertarget{dental-pathology}{%
-\subsubsection{Dental pathology}\label{dental-pathology}}
+\subsubsection{Dental pathology}\label{dental-pathology}
 
@@
 
-\hypertarget{calculus-sampling}{%
-\subsection{Calculus sampling}\label{calculus-sampling}}
+\subsection{Calculus sampling}\label{calculus-sampling}
 
@@
 
-\hypertarget{statistical-analysis}{%
-\subsection{Statistical analysis}\label{statistical-analysis}}
+\subsection{Statistical analysis}\label{statistical-analysis}
 
@@
 
-All statistical analysis was conducted in R version 4.3.2 (2023-10-31),
-Eye Holes, (R Core Team, 2020). Data wrangling was conducted with the
-\textbf{tidyverse} (Wickham et al., 2019) and visualisations were
+All statistical analysis was conducted in R version 4.3.3 (2024-02-29),
+Angel Food Cake, (R Core Team, 2020). Data wrangling was conducted with
+the \textbf{tidyverse} (Wickham et al., 2019) and visualisations were
 created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\section{Results}\label{results}
 
@@
 nicotine also had the highest extracted quantity of cotinine
-(\textbf{?@fig-auth-plot-batch2}).
-
-\emph{Fig 2 (fig-auth-plot-batch2): (A) Number of samples in which each
+(Figure~\ref{fig-auth-plot-batch2}).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{../figures/fig-auth-plot-batch2-1.pdf}
+
+}
+
+\caption{\label{fig-auth-plot-batch2}(A) Number of samples in which each
 compound was detected in the first and second batch. (B) Quantity (ng)
@@
 
+\end{figure}%
+
 To see if preservation of the skeletal remains had any effect on the
@@
 of the skeleton, with better preservation leading to increased
-extraction quantity (\textbf{?@fig-detection-preservation}A). We also
+extraction quantity (Figure~\ref{fig-detection-preservation}A). We also
 find a weak positive correlation between the weight of the calculus
 sample and the quantity of compound extracted from the calculus
-(\textbf{?@fig-detection-preservation}B).
-
-\emph{Fig 3 (fig-detection-preservation): (A) Violin plot with overlaid
+(Figure~\ref{fig-detection-preservation}B).
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{../figures/fig-detection-preservation-1.pdf}
+
+}
+
+\caption{\label{fig-detection-preservation}(A) Violin plot with overlaid
 box plots depicting the distribution of extracted quantities of each
@@
 coefficient.}
+
+\end{figure}%
 
@@
 CMS (0.43). Remaining correlations were weak or absent
-(\textbf{?@fig-polycorr}). Correlations with age will be depressed
+(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
 because age was largely controlled for in the sample selection.
 
-\emph{Fig 4 (fig-polycorr): Plot of the polychoric correlations
-(}rho\emph{). Larger circles and increased opacity indicates a stronger
+\begin{figure}
+
+\centering{
+
+\includegraphics{../figures/fig-polycorr-1.pdf}
+
+}
+
+\caption{\label{fig-polycorr}Plot of the polychoric correlations
+(\emph{rho}). Larger circles and increased opacity indicates a stronger
 correlation coefficient. OA = osteoarthritis; VOP = vertebral
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\end{figure}%
+
+\section{Discussion}\label{discussion}
 
@@
 positively correlated, both in concentration and presence/absence in
-individuals (Table~\ref{tbl-pearson} and \textbf{?@fig-polycorr}). The
+individuals (Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). The
 detection of nicotine and cotinine is not surprising, as pipe-smoking in
@@
 to determine if they originated from the same source
-(Table~\ref{tbl-pearson} and \textbf{?@fig-polycorr}). Caffeine and
+(Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). Caffeine and
 theophylline have very similar chemical structures, so we expect they
@@
 to see if the salicylic acid was incorporated during life
-(\textbf{?@fig-auth-plot-batch2}). It is important to note that,
+(Figure~\ref{fig-auth-plot-batch2}). It is important to note that,
 especially with salicylic acid, there is a possibility for the compound
@@
 their input on early draft manuscripts and a lovely guided tour of
-Middenbeemster. We also thank Louise Le Meillour for taking on our
-preprint and providing helpful comments, and Mario Zimmerman and two
-other anonymous reviewers for their time and invaluable insights.
-
-\hypertarget{funding}{%
-\section*{Funding}\label{funding}}
+Middenbeemster.
+
+We also thank Louise Le Meillour for taking on our preprint and
+providing helpful comments, and Mario Zimmerman and two other anonymous
+reviewers for their time and invaluable insights. This paper was
+evaluated during a CODECHECK event at TU Delft by Daniel Nüst and
+Stephen Eglen, who provided helpful feedback on the reproducibility of
+all code used in this paper.
+
+Preprint version 5 of this article has been peer-reviewed and
+recommended by PCI Archaeology
+\url{https://doi.org/10.24072/pci.archaeo.100389} (Meillour, 2023).
+
+\section*{Funding}\label{funding}
 \addcontentsline{toc}{section}{Funding}
@@
 All raw data is available on Zenodo
-(\url{https://doi.org/10.5281/zenodo.8061483}). Analysis scripts, and
-the source code for the manuscript and supplementary materials are
-available as a research compendium
-(\url{https://doi.org/10.5281/zenodo.7649824}) using the structure
-recommended by the \textbf{rrtools} R package (Marwick, 2019).
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
+(\url{https://doi.org/10.5281/zenodo.8061483}, Bartholdy, 2023).
+Analysis scripts, and the source code for the manuscript and
+supplementary materials are available as a research compendium
+(\url{https://doi.org/10.5281/zenodo.7649824}, Bartholdy, 2024) using
+the structure recommended by the \textbf{rrtools} R package (Marwick,
+2019).
+
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-abucaCocaTrade2019}{}}%
+\bibitem[\citeproctext]{ref-abucaCocaTrade2019}
 Abduca, R. (2019). Coca leaf transfers to {Europe}. {Effects} on the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bispoSimultaneousDetermination2002}{}}%
+\bibitem[\citeproctext]{ref-bartholdyDentalPathological2023}
+Bartholdy, B. P. (2023). \emph{Dental, pathological, and {UHPLC} data
+from {Middenbeemster} archaeological site} (Version 1.0.0). {Zenodo}.
+\url{https://doi.org/10.5281/zenodo.8061483}
+
+\bibitem[\citeproctext]{ref-bartholdyMB11CalculusPilot}
+Bartholdy, B. P. (2024). \emph{Bbartholdy/{mb11CalculusPilot}: {Version}
+of record} (Version 1.0.0). {Zenodo}.
+\url{https://doi.org/10.5281/zenodo.11040640}
+
+\bibitem[\citeproctext]{ref-bispoSimultaneousDetermination2002}
 Bispo, M. S., Veloso, M. C. C., Pinheiro, H. L. C., De Oliveira, R. F.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bruinsmaBijdragenTot1872}{}}%
+\bibitem[\citeproctext]{ref-bruinsmaBijdragenTot1872}
 Bruinsma, J. J. (1872). \emph{Bijdragen tot de {Geneeskundige
 Plaatsbeschrijving} van {Nederland}}. {Van Weelden en Mingelen}.
-\url{https://dlcs.io/pdf/wellcome/pdf-item/b24874140/0}
-
-\leavevmode\vadjust pre{\hypertarget{ref-buckberryAuricular2002}{}}%
+\url{https://wellcomecollection.org/works/sttrpc2v}
+
+\bibitem[\citeproctext]{ref-buckberryAuricular2002}
 Buckberry, J. L., \& Chamberlain, A. T. (2002). Age estimation from the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-meindlSutureClosure1985}{}}%
+\bibitem[\citeproctext]{ref-meillourDetectionPlantderived2023}
+Meillour, L. L. (2023). Detection of plant-derived compounds in {XIXth}
+c. {Dutch} dental calculus. \emph{Peer Community in Archaeology},
+\emph{1}, 100389. \url{https://doi.org/10.24072/pci.archaeo.100389}
+
+\bibitem[\citeproctext]{ref-meindlSutureClosure1985}
 Meindl, R. S., \& Lovejoy, C. O. (1985). Ectocranial suture closure: {A}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
+\bibitem[\citeproctext]{ref-Rbase}
 R Core Team. (2020). \emph{R: {A} language and environment for
 statistical computing} {[}Manual{]}. {R Foundation for Statistical
-Computing}; {R Foundation for Statistical Computing}.
-\url{https://www.R-project.org/}
-
-\leavevmode\vadjust pre{\hypertarget{ref-raffertyCurrentResearch2012}{}}%
+Computing}. \url{https://www.R-project.org/}
+
+\bibitem[\citeproctext]{ref-raffertyCurrentResearch2012}
 Rafferty, S. M., Lednev, I., Virkler, K., \& Chovanec, Z. (2012).
@@
 science.} (1st ed, pp. 163--182). {Cambridge University Press}.
-\url{https://login.ezproxy.leidenuniv.nl:2443/login?URL=https://search.ebscohost.com/login.aspx?direct=true\&db=e000xww\&AN=40641\&site=ehost-live}
-
-\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoRoleOral1999}{}}%
+
+\bibitem[\citeproctext]{ref-scannapiecoRoleOral1999}
 Scannapieco, F. A. (1999). Role of {Oral Bacteria} in {Respiratory

### 2025-02-13 analysis/paper/paper.tex
update renv and other minor modifications

--- 2024-04-16 analysis/paper/paper.tex
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 \usepackage[a4paper,headheight=13pt,footskip=20pt,margin=2.54cm]{geometry}
-\usepackage{indentfirst}
 
 % Modify section headings
-\usepackage{titlesec}
-\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{}
-\titleformat{\subsection}{\normalfont\bfseries}{}{.5em}{}
-\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{\addperiod}
-\newcommand{\addperiod}[1]{#1.} %https://tex.stackexchange.com/questions/137193/adding-a-period-after-section
+\usepackage[indentafter,explicit]{titlesec}
+\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{#1}
+%\titleformat{\section}{\normalfont\large\bfseries\centering}{\cdr@ifeq{\sectionname}{}{}{\sectionname~}\thesection.}{.5em}{#1}
+\titleformat{\subsection}{\normalfont\bfseries}{\thesubsection.}{.5em}{#1}
+\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{#1. }
 
@@
     belowskip=12pt,
-    font=footnotesize,
+    font={footnotesize,stretch=1},
     justification=justified]{caption}
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+
 \usepackage{bookmark}
@@
 
-\setstretch{1.5}
+\pagenumbering{gobble}
+
+\setstretch{1.1}
 \section{Introduction}\label{introduction}
@@
 
-\includegraphics{../figures/fig-sample-demography-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{../figures/pcj-fig-sample-demography-1.pdf}}
 
@@
 
-All statistical analysis was conducted in R version 4.3.3 (2024-02-29),
-Angel Food Cake, (\citeproc{ref-Rbase}{R Core Team, 2020}). Data
+All statistical analysis was conducted in R version 4.4.2 (2024-10-31),
+Pile of Leaves, (\citeproc{ref-Rbase}{R Core Team, 2020}). Data
 wrangling was conducted with the \textbf{tidyverse}
@@
 
-\includegraphics{../figures/fig-auth-plot-batch2-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{../figures/pcj-fig-auth-plot-batch2-1.pdf}}
 
@@
 
-\includegraphics{../figures/fig-detection-preservation-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{../figures/pcj-fig-detection-preservation-1.pdf}}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1413}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0761}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0978}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0652}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0978}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.0652}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 16\tabcolsep) * \real{0.1522}}@{}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1500}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1000}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1000}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1000}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1000}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1000}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1500}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1000}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1000}}@{}}
 
@@
 
-\includegraphics{../figures/fig-polycorr-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{../figures/pcj-fig-polycorr-1.pdf}}
 
@@
 their input on early draft manuscripts and a lovely guided tour of
-Middenbeemster. We also thank Louise Le Meillour for taking on our
-preprint and providing helpful comments, and Mario Zimmerman and two
-other anonymous reviewers for their time and invaluable insights.
+Middenbeemster.
+
+We also thank Louise Le Meillour for taking on our preprint and
+providing helpful comments, and Mario Zimmerman and two other anonymous
+reviewers for their time and invaluable insights. This paper was
+evaluated during a CODECHECK event at TU Delft by Daniel Nüst and
+Stephen Eglen, who provided helpful feedback on the reproducibility of
+all code used in this paper.
+
+Preprint version 5 of this article has been peer-reviewed and
+recommended by PCI Archaeology
+\url{https://doi.org/10.24072/pci.archaeo.100389}
+(\citeproc{ref-meillourDetectionPlantderived2023}{\textbf{meillourDetectionPlantderived2023?}}).
 
@@
 All raw data is available on Zenodo
-(\url{https://doi.org/10.5281/zenodo.8061483}). Analysis scripts, and
-the source code for the manuscript and supplementary materials are
-available as a research compendium
-(\url{https://doi.org/10.5281/zenodo.7649824}) using the structure
-recommended by the \textbf{rrtools} R package
+(\url{https://doi.org/10.5281/zenodo.8061483},
+\citeproc{ref-bartholdyDentalPathological2023}{\textbf{bartholdyDentalPathological2023?}}).
+Analysis scripts, and the source code for the manuscript and
+supplementary materials are available as a research compendium
+(\url{https://doi.org/10.5281/zenodo.7649824},
+\citeproc{ref-bartholdyMB11CalculusPilot}{\textbf{bartholdyMB11CalculusPilot?}})
+using the structure recommended by the \textbf{rrtools} R package
 (\citeproc{ref-Rrrtools}{Marwick, 2019}).

### 2025-02-13 README.Rmd
fix: update renv version
update linux dependencies
update README

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/mb11CalculusPilot/main?urlpath=rstudio) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.7649825.svg)](https://zenodo.org/badge/DOI/10.5281/zenodo.7649825)
+[![R-CMD-check](https://github.com/bbartholdy/mb11CalculusPilot/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bbartholdy/mb11CalculusPilot/actions/workflows/R-CMD-check.yaml) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.7649825.svg)](https://zenodo.org/badge/DOI/10.5281/zenodo.7649825)
 
@@
 
-<!--
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
-
-Our pre-print is online here:
--->
-
-> Bartholdy, B.P., Hasselstrøm, J.B., Sørensen, L.K., Casna, M., Hoogland, M., Historisch Genootschap Beemster, & Henry, A.G.
-(2023).
-_`r Title`_.
-Zenodo, Accessed `r format(Sys.Date(), "%d %b %Y")`.
-Online at <https://doi.org/10.5281/zenodo.7649151>
-
-### How to cite
-
-Use the above paper to cite this compendium.
-
-<!--
-Please cite this compendium as:
-
-> Bartholdy, B.P., Hasselstrøm, J.B., Sørensen, L.K., Casna, M., Hoogland, M., Historisch Genootschap Beemster, & Henry, A.G., (`r format(Sys.Date(), "%Y")`).
-_Compendium of R code and data for `r Title`_.
-Accessed `r format(Sys.Date(), "%d %b %Y")`.
-Online at <https://doi.org/10.5281/zenodo.7649825>
--->
+> Bartholdy, Bjørn Peare; Hasselstrøm, Jørgen B.; Sørensen, Lambert K.; Casna, Maia; Hoogland, Menno; Historisch Genootschap Beemster; Henry, Amanda G.
+Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population.
+Peer Community Journal, Volume 4 (2024), article no.
+e46.
+doi : 10.24072/pcjournal.414.
+https://peercommunityjournal.org/articles/10.24072/pcjournal.414/
 
@@
 
-## How to run in your browser or download and run locally
+## How to reproduce the paper
 
-**If you are running Linux, you will need to install `gfortran`.**
+**Software requirements:**
 
-e.g. `sudo apt install gfortran` for Ubuntu.
+- [R](https://cran.r-project.org/)
+- [Quarto](https://quarto.org/)
+- [Git](https://git-scm.com/downloads) (optional)
 
-This research compendium has been developed using the statistical programming
-language R. To work with the compendium, you will need
-installed on your computer the [R software](https://cloud.r-project.org/)
-itself, optionally (but recommended) [RStudio Desktop](https://rstudio.com/products/rstudio/download/), and [Quarto](https://quarto.org/docs/get-started/).
+**Linux dependencies:**
+
+`libx11-dev libicu-dev make zlib1g-dev libcurl4-openssl-dev libssl-dev libfontconfig1-dev libfreetype6-dev libfribidi-dev libharfbuzz-dev libjpeg-dev libpng-dev libtiff-dev pandoc libxml2-dev libgit2-dev`
 
@@
 
-`git clone git@github.com:bbartholdy/mb11CalculusPilot.git`
+`git clone https://github.com/bbartholdy/mb11CalculusPilot.git`
 
@@
 
-- open the `.Rproj` file in RStudio
-- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
-[DESCRIPTION](/DESCRIPTION) file).
-- finally, open `analysis/paper/paper.qmd` and **render** to produce the PDF manuscript (`_output/`),
-  or run `quarto::quarto_render("analysis/paper/paper.qmd")` in the R console.
+- Open the folder in your preferred IDE
+- run `renv::restore()` to install the required R packages
+- run `quarto::quarto_render("analysis/paper/paper.qmd", output_format = "pdf")` in the R console (requires installing **quarto** R package) or `quarto render analysis/paper/paper.qmd --to pdf` to produce a PDF
+- find the pdf in `./analysis/paper/_output`
 

### 2025-02-13 README.Rmd
update usage instructions

@@
 
-# Research compendium for Bartholdy et al. 2023
+# Research compendium for Bartholdy et al. 2024
 
@@
 - [Quarto](https://quarto.org/)
+- [TeX](https://www.latex-project.org/get/)
 - [Git](https://git-scm.com/downloads) (optional)
@@
 
-`libx11-dev libicu-dev make zlib1g-dev libcurl4-openssl-dev libssl-dev libfontconfig1-dev libfreetype6-dev libfribidi-dev libharfbuzz-dev libjpeg-dev libpng-dev libtiff-dev pandoc libxml2-dev libgit2-dev`
+`libx11-dev libicu-dev make zlib1g-dev libcurl4-openssl-dev libssl-dev libfontconfig1-dev libfreetype6-dev libfribidi-dev libharfbuzz-dev libjpeg-dev libpng-dev libtiff-dev pandoc libxml2-dev libgit2-dev gfortran`
 
@@
 
-- Open the folder in your preferred IDE
-- run `renv::restore()` to install the required R packages
-- run `quarto::quarto_render("analysis/paper/paper.qmd", output_format = "pdf")` in the R console (requires installing **quarto** R package) or `quarto render analysis/paper/paper.qmd --to pdf` to produce a PDF
-- find the pdf in `./analysis/paper/_output`
+1. Open the folder in your preferred IDE
+2. run `renv::restore()` to install the required R packages
+3.
+run `quarto::quarto_render("analysis/paper/paper.qmd", output_format = "pdf")` in the R console (requires installing **quarto** R package) or `quarto render analysis/paper/paper.qmd --to pdf` to produce a PDF
+4. find the pdf in `./analysis/paper/_output`
 
-### Licenses
+## Licenses
 

### 2025-02-14 README.Rmd
update README

@@
 
-[main.zip](/archive/refs/heads/main.zip).
+[main.zip](https://github.com/bbartholdy/mb11CalculusPilot/archive/refs/heads/main.zip)
 
@@
 
-1. Open the folder in your preferred IDE
+1. open the folder in your preferred IDE
 2. run `renv::restore()` to install the required R packages
 3.
-run `quarto::quarto_render("analysis/paper/paper.qmd", output_format = "pdf")` in the R console (requires installing **quarto** R package) or `quarto render analysis/paper/paper.qmd --to pdf` to produce a PDF
-4. find the pdf in `./analysis/paper/_output`
+run `quarto::quarto_render("analysis/paper/paper.qmd", output_format = "html")` in the R console (requires installing **quarto** R package) or `quarto render analysis/paper/paper.qmd --to html` to produce an HTML
+4. open the output `./analysis/paper/_output/version-of-record.html` in a browser
 

### 2025-02-14 analysis/scripts/generate-bib.R
fix: rearrange bibliography function to fix R-CMD-Checks

@@ -1,7 +1,4 @@
-#' Generate the bibliography
-#'
-#' @importFrom rbbt bbt_detect_citations bbt_write_bib
-#' @importFrom here here
+# Generate the bibliography
 
 generate_bib <- function(){
   cite_keys <- bbt_detect_citations(list.files(here("analysis/"), "*.qmd", full.names = T, recursive = T))
